---
title: "Phishing URL Detection — Hypothesis & EDA"
author: "K-o-m-A"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---
```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(corrplot)
library(patchwork)
library(scales)
```

---

# 1. Problem Statement

## 1.1 Real-world scenario

A user clicks a link in an email, chat, or search result. The browser or a
corporate proxy has to decide **before the page loads** whether the link is
phishing. The cheapest signal at that moment is the URL string itself — no
DNS lookup, no HTML fetch, no rendering. If a URL-only check is good enough,
we can block the link for free. If it isn't, we have to fall back to a
slower page-level check.

The question we want to answer is simple:
**how much phishing signal is already in the URL, and how few features do
we really need?** If a URL-only model works well, we can run it as a fast
first-line filter. Where it fails, the slower trust- and behavior-level
checks take over.

This lines up with our two chosen scenarios:

- **Scenario 2** — parametric vs. non-parametric models on the URL-only
  tier. Tells us whether a linear model is enough, or whether the URL hides
  non-linear patterns that only RF/SVM/GBM can pick up.
- **Scenario 3** — algorithmic vs. embedded feature selection on the full
  feature matrix. Tells us how few features we really need, and which ones
  survive penalisation when features are correlated.

## 1.2 Dataset

The PhiUSIIL dataset has ~235 k rows and 50 predictor features. We split
them into three families by how expensive they are to compute:

| Family | What it captures | Example features |
|--------|------------------|-----------------|
| **Lexical** | Properties of the URL string | `URLLength`, `NoOfSubDomain`, `NoOfObfuscatedChar`, `CharContinuationRate` |
| **Trust** | Domain / title consistency | `IsHTTPS`, `HasTitle`, `HasFavicon`, `Bank`, `Pay`, `Crypto` |
| **Behavior** | Page-level content signals | `NoOfURLRedirect`, `NoOfiFrame`, `HasExternalFormSubmit`, `HasHiddenFields` |

Lexical is the URL-only first-line case. Trust and Behavior need more
network / page work, and serve as a strong baseline to compare against.

---

# 2. Hypotheses

## H1 (Scenario 2) — On the URL-only tier, non-parametric models pick up meaningfully more signal than parametric ones

Phishing URLs usually leave lexical traces: odd subdomain depth, paths full
of digits, obfuscated characters. None of this needs the page to be
downloaded. We expect a model trained only on Lexical features (URL string,
no DNS, no HTML) to still reach a reasonable AUC — but with a **bigger**
parametric vs. non-parametric gap than any other tier. The reason: lexical
patterns interact in non-additive ways (lengths × digit counts × special
chars), and a linear decision boundary can't fully use those interactions.
We compare three parametric methods (LR, LDA, Naive Bayes) against three
non-parametric ones (RF, SVM-RBF, KNN).

**Concrete success criteria.** To keep the claim precise, H1 is tested
against the 5-fold CV protocol from `scenario_2.rmd`:

| # | Criterion | Threshold |
|---|-----------|-----------|
| C1 | Mean CV AUC gap between the best non-parametric and best parametric model on the **Lexical (URLOnly) tier** | **ΔAUC ≥ 0.02** |
| C2 | Paired Wilcoxon signed-rank test on the 5 per-fold AUCs of those two models | **p < 0.05** |
| C3 | Gradient across tiers: the Lexical gap is **strictly larger** than the Full-tier gap (Full saturates at AUC ≈ 0.99 and the gap should drop to ≤ 0.005). | gap(Lexical) > gap(Full) |

H1 is **supported** only if C1 **and** C2 hold. C3 turns the claim from
"non-parametric wins on one tier" into the stronger statement *"the
non-parametric advantage shrinks as the feature tier gets stronger."* H1 is
**rejected** if either C1 or C2 fails on Lexical.

## H2 (Scenario 3) — Embedded feature selection keeps clearly fewer features than algorithmic selection, because it collapses correlated clusters

The full feature matrix has severe multicollinearity (a URL-length cluster
with VIF > 1000, see Section 4.3) and several pairs with |r| > 0.7. We
expect **embedded methods** (LASSO, ElasticNet, Random Forest permutation
importance) to keep fewer features than **algorithmic methods**
(forward/backward stepwise, RFE). The reason: embedded penalisation pulls
each correlated cluster down to one or two survivors, while stepwise search
treats correlated features as independent candidates. We will report (a)
how many features each method keeps, (b) how much the kept sets overlap,
and (c) which features stop being statistically significant when their
correlated partners are added or removed.

---

# 3. Data Loading & Cleaning

```{r load-data}
raw <- read_csv(
  "PhiUSIIL_Phishing_URL_Dataset.csv",
  show_col_types = FALSE
)

# Strip UTF-8 BOM that appears on the first column name
names(raw) <- map_chr(names(raw), ~ str_replace(.x, "^\ufeff", ""))

# glimpse(raw)
```

```{r define-roles}
# Columns excluded from modelling (identifiers / free text)
ID_COLS <- c("FILENAME", "URL", "Domain", "TLD", "Title")

# Binary target as factor
df <- raw %>%
  mutate(label_bin = factor(label, levels = c(0, 1),
                            labels = c("Legitimate", "Phishing"))) %>%
  select(-all_of(ID_COLS))

# Feature matrix (numeric predictors only)
features <- df %>% select(-label, -label_bin)

cat("Observations:", nrow(df), "\n")
cat("Predictors:  ", ncol(features), "\n")
```

```{r family-map}
# ── Feature-family mapping (reused throughout EDA) ──────────────────────────
# The three families (Lexical / Trust / Behavior) are our own taxonomy, not
# a property of the dataset. They reflect the cost of obtaining each signal
# at decision time:
#   Lexical  — derivable from the raw URL string alone (zero network cost).
#   Trust    — requires light page/domain inspection or an external lookup
#              (DNS, TLD database, HTTPS handshake, title parsing).
#   Behavior — requires fetching, parsing and partially rendering the page.
#
# Note on `IsHTTPS`: although the "https://" scheme is part of the URL string,
# we place it in the **Trust** family. Modern browsers refuse plain HTTP for
# any sensitive interaction and warn the user prominently, so the scheme is
# already enforced by the trust layer of the browser rather than acting as a
# free lexical clue. Treating it as Lexical would let the URLOnly tier cheat
# off a binary that is, in practice, a deployment/trust signal.
family_map <- tribble(
  ~feature,                    ~family,
  "URLLength",                 "Lexical",
  "DomainLength",              "Lexical",
  "IsDomainIP",                "Lexical",
  "URLSimilarityIndex",        "Trust",
  "CharContinuationRate",      "Lexical",
  "TLDLegitimateProb",         "Trust",
  "URLCharProb",               "Trust",
  "TLDLength",                 "Lexical",
  "NoOfSubDomain",             "Lexical",
  "HasObfuscation",            "Lexical",
  "NoOfObfuscatedChar",        "Lexical",
  "ObfuscationRatio",          "Lexical",
  "NoOfLettersInURL",          "Lexical",
  "LetterRatioInURL",          "Lexical",
  "NoOfDegitsInURL",           "Lexical",
  "DegitRatioInURL",           "Lexical",
  "NoOfEqualsInURL",           "Lexical",
  "NoOfQMarkInURL",            "Lexical",
  "NoOfAmpersandInURL",        "Lexical",
  "NoOfOtherSpecialCharsInURL","Lexical",
  "SpacialCharRatioInURL",     "Lexical",
  "IsHTTPS",                   "Trust",
  "LineOfCode",                "Behavior",
  "LargestLineLength",         "Behavior",
  "HasTitle",                  "Trust",
  "DomainTitleMatchScore",     "Trust",
  "URLTitleMatchScore",        "Trust",
  "HasFavicon",                "Trust",
  "Robots",                    "Trust",
  "IsResponsive",              "Behavior",
  "NoOfURLRedirect",           "Behavior",
  "NoOfSelfRedirect",          "Behavior",
  "HasDescription",            "Behavior",
  "NoOfPopup",                 "Behavior",
  "NoOfiFrame",                "Behavior",
  "HasExternalFormSubmit",     "Behavior",
  "HasSocialNet",              "Behavior",
  "HasSubmitButton",           "Behavior",
  "HasHiddenFields",           "Behavior",
  "HasPasswordField",          "Behavior",
  "Bank",                      "Trust",
  "Pay",                       "Trust",
  "Crypto",                    "Trust",
  "HasCopyrightInfo",          "Behavior",
  "NoOfImage",                 "Behavior",
  "NoOfCSS",                   "Behavior",
  "NoOfJS",                    "Behavior",
  "NoOfSelfRef",               "Behavior",
  "NoOfEmptyRef",              "Behavior",
  "NoOfExternalRef",           "Behavior"
)
```

## 3.3 Feature Exclusion

Before the EDA we drop three groups of features that would bias the
analysis or trivialise the modelling. All three decisions are based on
what the feature actually means.

### Category 1 — Computed probability / similarity scores

These are not raw URL measurements. They are outputs of other models or
expert-written scoring functions that already encode phishing-detection
logic. Keeping them would mean we are classifying another classifier's
output, not the URL itself.

```{r exclude-scores}
SCORES <- c(
  "URLSimilarityIndex",   # similarity to known-phishing URL patterns
  "TLDLegitimateProb",    # P(TLD is legitimate) from an external lookup
  "URLCharProb",          # P(character sequence) from a language model
  "DomainTitleMatchScore",# computed domain↔title consistency score
  "URLTitleMatchScore"    # computed URL↔title consistency score
)
cat("Computed scores to remove:", paste(SCORES, collapse = ", "), "\n")
```

### Category 2 — Redundant binary indicator

`HasObfuscation` is literally `(NoOfObfuscatedChar > 0)`. It holds no
information that the count feature doesn't already carry.

```{r exclude-binary}
REDUNDANT_BINARY <- "HasObfuscation"
cat("Redundant binary to remove:", REDUNDANT_BINARY, "\n")
```

### Category 3 — Redundant ratio features

Each ratio below equals `count / URLLength`, and both the count and
`URLLength` are already in the dataset. Keeping the ratio gives us a
perfect linear dependence — and no extra signal.

```{r exclude-ratios}
RATIOS <- c(
  "LetterRatioInURL",      # = NoOfLettersInURL   / URLLength
  "DegitRatioInURL",       # = NoOfDegitsInURL    / URLLength
  "ObfuscationRatio",      # = NoOfObfuscatedChar / URLLength
  "SpacialCharRatioInURL"  # = NoOfOtherSpecialCharsInURL / URLLength
)
cat("Linearly dependent ratios to remove:", paste(RATIOS, collapse = ", "), "\n")
```

```{r apply-exclusions}
EXCLUDE_COLS <- c(SCORES, REDUNDANT_BINARY, RATIOS)

df       <- df %>% select(-all_of(EXCLUDE_COLS))
features <- df %>% select(-label, -label_bin)

cat("\nFeatures after exclusion:", ncol(features),
    "(removed", length(EXCLUDE_COLS), "of", ncol(features) + length(EXCLUDE_COLS), ")\n")
cat("Removed:", paste(EXCLUDE_COLS, collapse = ", "), "\n")
```

Update the family map to reflect the reduced feature set:

```{r family-map-update}
family_map <- family_map %>%
  filter(!feature %in% EXCLUDE_COLS)

family_map %>% count(family) %>% print()
```

```{r feature-types, include=FALSE}
# Silent type split used by downstream VIF/skewness chunks
binary_features <- names(features)[map_lgl(features, ~ all(na.omit(.x) %in% c(0, 1)))]
continuous_features <- setdiff(names(features), binary_features)
```

## 3.4 Feature Tiers Used in Modelling

To test H1 we define three nested feature tiers. The modelling notebook
reads these tier definitions directly from here, so both notebooks stay
in sync.

```{r feature-tiers}
tier_lexical  <- family_map %>% filter(family == "Lexical")  %>% pull(feature)
tier_trust    <- family_map %>% filter(family == "Trust")    %>% pull(feature)
tier_behavior <- family_map %>% filter(family == "Behavior") %>% pull(feature)
tier_full     <- c(tier_lexical, tier_trust, tier_behavior)

cat("URLOnly tier (Lexical, ",  length(tier_lexical),  " features)\n", sep = "")
cat("Trust tier              (",  length(tier_trust),    " features)\n", sep = "")
cat("Behavior tier           (",  length(tier_behavior), " features)\n", sep = "")
cat("Full tier               (",  length(tier_full),     " features)\n", sep = "")
```

---

# 4. Exploratory Data Analysis


## 4.1 Data Overview

### 4.1.1 Class Balance

```{r class-balance, fig.height=3.5}
df %>%
  count(label_bin) %>%
  mutate(pct = n / sum(n)) %>%
  ggplot(aes(label_bin, n, fill = label_bin)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = scales::percent(pct, 0.1)), vjust = -0.4, size = 4) +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, .1))) +
  labs(title = "Class distribution",
       x = NULL, y = "Count") +
  theme_minimal(base_size = 12)
```

**Takeaway:** The dataset is roughly balanced, so we don't need resampling
or class-weighting. Accuracy works as a secondary metric next to AUC.

### 4.1.2 Missing Values

No missing values are present in any predictor (verified with
`colSums(is.na(features))`).


## 4.2 H1 — Lexical features overlap between classes, so we'll need non-linear models

H1 claims that a URL-only detector (Lexical features) will show a bigger
parametric vs. non-parametric gap than any other tier. We don't test
non-linearity directly in the EDA. Instead we show that **per-feature
discriminative power is lowest in the Lexical family**. When no single
feature tells the classes apart, the model has to combine several features
at once — and that's where non-linear models have an edge.

### 4.2.1 Discriminative Power by Family

We measure discriminative power separately for continuous and binary
features. The standardised mean difference (SMD) only makes sense for
continuous features. For binary features the "pooled SD" collapses to a
function of the class proportion, so SMD stops being comparable across
features. For those we use **Cramér's V** instead — the standard
effect-size measure for a 2×2 contingency table, on the same 0–1 scale
as |correlation|.

```{r discriminative-power, fig.height=7, fig.width=8}
# --- Continuous features: standardised mean difference ----------------------
smd_df <- df %>%
  select(label, all_of(continuous_features)) %>%
  pivot_longer(-label, names_to = "feature", values_to = "value") %>%
  group_by(feature, label) %>%
  summarise(m = mean(value, na.rm = TRUE), s = sd(value, na.rm = TRUE),
            .groups = "drop") %>%
  pivot_wider(names_from = label, values_from = c(m, s),
              names_glue = "{.value}_{label}") %>%
  mutate(
    pooled_sd = sqrt((s_0^2 + s_1^2) / 2),
    effect    = abs(m_1 - m_0) / (pooled_sd + 1e-9),
    metric    = "SMD (continuous)"
  ) %>%
  select(feature, effect, metric)

# --- Binary features: Cramer's V on a 2x2 contingency table -----------------
cramers_v <- function(x, y) {
  tab <- table(x, y)
  if (any(dim(tab) < 2)) return(NA_real_)
  chi2 <- suppressWarnings(chisq.test(tab, correct = FALSE)$statistic)
  n <- sum(tab)
  as.numeric(sqrt(chi2 / (n * (min(dim(tab)) - 1))))
}

cv_df <- tibble(
  feature = binary_features,
  effect  = map_dbl(binary_features, ~ cramers_v(features[[.x]], df$label)),
  metric  = "Cramer's V (binary)"
)

discrim_df <- bind_rows(smd_df, cv_df) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  arrange(desc(effect))

ggplot(discrim_df, aes(x = reorder(feature, effect), y = effect, fill = family)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ metric, scales = "free", ncol = 2) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Per-feature discriminative power",
       subtitle = "Continuous: |SMD|. Binary: Cramer's V. Both >= 0.",
       x = NULL, y = "Effect size", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")
```

Effect-size distribution per family (continuous features only, so the SMD
values are directly comparable):

```{r smd-by-family, fig.height=4, fig.width=6}
smd_family <- discrim_df %>%
  filter(metric == "SMD (continuous)", family != "Other")

ggplot(smd_family, aes(family, effect, fill = family)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  geom_jitter(width = 0.15, alpha = 0.5, size = 2) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "SMD distribution by family (continuous features only)",
       subtitle = "H1: Lexical SMD lowest -> largest parametric gap expected at URLOnly tier",
       x = NULL, y = "| SMD |") +
  theme_minimal(base_size = 12)
```

**Takeaway (H1):** Among the continuous features, Lexical has the lowest
median |SMD| and the smallest spread. Trust and Behavior carry more
per-feature signal. Weak per-feature evidence plus visible non-linear
structure is exactly the setting where parametric models struggle most —
and where the parametric vs. non-parametric gap predicted by H1 should
show up.

## 4.3 H2 (Scenario 3) — Multicollinearity is what makes algorithmic vs. embedded selection differ

H2 says that algorithmic and embedded feature selection methods will keep
different feature sets on the Full tier. The EDA evidence is the
correlation structure: where are the correlated clusters, and how bad is
the inflation? Embedded methods (LASSO, ElasticNet, RF importance) collapse
each cluster down to one survivor. Algorithmic methods (stepwise, RFE)
treat correlated features as independent candidates and tend to keep more
of them.

### 4.3.1 Lexical Correlation Matrix

The collinearity that drives H2 lives inside the Lexical family. Trust and
Behavior have `mean_abs_r` around 0.18 and 0.14 and no dense cluster. The
plot below makes the URL-length count cluster visible.

```{r cor-lexical, fig.height=6, fig.width=7}
lexical_cor <- features %>%
  select(all_of(tier_lexical)) %>%
  cor(use = "pairwise.complete.obs")

  corrplot(
  lexical_cor,
  method  = "color",
  order   = "hclust",
  tl.cex  = 0.7,
  tl.col  = "black",
  col     = colorRampPalette(c("firebrick", "white", "steelblue"))(200),
  title   = "Correlation - Lexical features",
  mar     = c(0, 0, 1.5, 0),
  addCoef.col = "black",
  number.cex  = 0.55
)
```

Across the full feature matrix, 10 pairs pass `|r| > 0.7` — and all of
them sit inside this Lexical count cluster (URLLength ↔ NoOfLettersInURL
↔ NoOfDegitsInURL and friends). These are the correlated pairs we expect
algorithmic and embedded selection to handle differently.

### 4.3.2 Variance Inflation Factors

```{r vif}
library(knitr)

# Target-free VIF from the design matrix (avoids non-converged GLM).
X_vif <- features %>%
  select(all_of(continuous_features)) %>%
  as.data.frame()

vif_values <- vapply(seq_along(X_vif), function(i) {
  r2 <- summary(lm(X_vif[[i]] ~ ., data = X_vif[, -i, drop = FALSE]))$r.squared
  1 / (1 - r2)
}, numeric(1))
names(vif_values) <- names(X_vif)

vif_df <- tibble(
  feature = names(vif_values),
  vif     = as.numeric(vif_values)
) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  mutate(severity = case_when(
    vif > 10            ~ "severe (> 10)",
    vif >= 5 & vif <= 10 ~ "moderate (5-10)",
    TRUE                ~ "ok (< 5)"
  )) %>%
  arrange(desc(vif))

vif_df %>%
  transmute(Feature = feature,
            Family  = family,
            VIF     = round(vif, 2),
            Severity = severity) %>%
  kable(caption = "Variance Inflation Factors - continuous features (lm-based, target-free).")
```

**Takeaway (H2):** Dropping the four ratio features in Section 3.3 was
necessary but not enough. `URLLength`, `NoOfLettersInURL` and
`NoOfDegitsInURL` stay near-collinear because they measure overlapping
properties of the same URL string. Stepwise/RFE will likely keep most of
them; LASSO and RF importance should keep only one or two. Measuring that
difference is the H2 task. Binary features are left out of the VIF
calculation by design — regressing a 0/1 indicator on continuous
predictors is not comparable on the same scale.


## 4.4 Skewness Assessment *(Scenario 2 preparation)*

```{r outlier-skewness, fig.height=6, fig.width=8}
# Skewness for all continuous features
skewness_fn <- function(x) {
  x <- x[!is.na(x)]
  n <- length(x)
  m <- mean(x)
  s <- sd(x)
  (n / ((n-1)*(n-2))) * sum(((x - m) / s)^3)
}

skew_df <- tibble(
  feature  = continuous_features,
  skewness = map_dbl(continuous_features, ~ skewness_fn(features[[.x]]))
) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  mutate(abs_skew = abs(skewness)) %>%
  arrange(desc(abs_skew))

ggplot(skew_df, aes(x = reorder(feature, abs_skew), y = skewness, fill = family)) +
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = c(-2, 2), linetype = "dashed", colour = "grey40") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Skewness of continuous features",
       subtitle = "Dashed lines at +/- 2 mark heavy skew; LR/LDA may struggle here",
       x = NULL, y = "Skewness", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")

# Count features with |skewness| > 2
n_heavy_skew <- sum(skew_df$abs_skew > 2)
cat("\nFeatures with |skewness| > 2:", n_heavy_skew, "of",
    length(continuous_features), "\n")
cat("These are:\n")
skew_df %>% filter(abs_skew > 2) %>% select(feature, skewness, family) %>% print()
```

**Takeaway (Scenario 2):** Several count features are heavily right-skewed
with long outlier tails. This shapes our modelling choices:

- **Parametric models** (LR, LDA): need scaling and a log-transform on
  skewed features to make the distributional assumption hold.
- **Non-parametric models** (trees, KNN, SVM): robust to skew, but KNN and
  SVM still need standardisation because they use distances.



---

# 5. EDA Summary & Modelling Implications

**A note on separability:** After the Section 3.3 exclusions, 40 features
remain (13 Lexical + 7 Trust + 20 Behavior). Even on this reduced set,
every model type already reaches AUC ≈ 0.99 on the Full tier — the task
is near-perfectly separable once all three families are present. So we
test H1 mainly on the **URLOnly tier** (13 Lexical features), where the
parametric vs. non-parametric contrast has room to show. H2 is tested on
the **Full tier**, where the multicollinearity that drives the
algorithmic-vs-embedded gap actually lives.

## 5.1 Implications for modelling scenarios

| Finding | Scenario 2 implication (H1) | Scenario 3 implication (H2) |
|---------|----------------------------|------------------------------|
| Near-perfect separability on Full tier | Comparison focused on URLOnly tier where models actually differ | Use Full tier; selection methods need a non-trivial pool of redundant features |
| Lexical SMD lowest + features must be combined | Largest parametric/non-parametric AUC gap expected on URLOnly | Algorithmic methods may keep all weak Lexical features; embedded penalisation should drop them |
| Heavy skewness in 19 of 22 continuous features | Log-transform + standardisation for LR/LDA; trees/RF unaffected | Apply same transform before LASSO/ElasticNet (they are scale-sensitive) |
| URL-length cluster: VIF > 500 and abs(r) > 0.84 | Use stable solvers / regularised LR for fairness | Strongest test case for H2 — embedded should collapse cluster, stepwise should not |
| Trust binary features: low intra-correlation | No special treatment | All three Trust HTTPS-style flags likely retained by both selection families — informative-feature overlap |

