---
title: "Phishing URL Detection — Hypothesis & EDA"
author: "K-o-m-A"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---
```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(corrplot)
library(patchwork)
library(scales)
```

---

# 1. Problem Statement

## 1.1 Real-world scenario

A user clicks a link in an email, chat, or search result. The
corporate proxy has to decide **before the page loads** whether the link is
phishing. The cheapest signal at that moment is the URL string itself — no
DNS lookup, no HTML fetch, no rendering. If a URL-only check is good enough,
we can block the link for free. If it isn't, we have to fall back to a
slower page-level check.

The question we want to answer is simple:
**how much phishing signal is already in the URL?** If a URL-only model works well, we can run it as a fast
first-line filter. Where it fails, the slower trust- and behavior-level checks have to take over.

Explored scenarios for hypothesis driven modeling: 2,5

## 1.2 Dataset

The PhiUSIIL dataset has ~235 k rows and 50 predictor features. We split
them into three families by how expensive they are to compute:

| Family | What it captures | Example features |
|--------|------------------|-----------------|
| **Lexical** | Properties of the URL string | `URLLength`, `NoOfSubDomain`, `NoOfObfuscatedChar`, `CharContinuationRate` |
| **Trust** | Domain / title consistency | `IsHTTPS`, `HasTitle`, `HasFavicon`, `Bank`, `Pay`, `Crypto` |
| **Behavior** | Page-level content signals | `NoOfURLRedirect`, `NoOfiFrame`, `HasExternalFormSubmit`, `HasHiddenFields` |

---

# 2. Hypotheses

## H1 (Scenario 2) — The gap between parametric and non-parametric model families is tier-dependent

Phishing URLs usually leave lexical traces: odd subdomain depth, paths full
of digits, obfuscated characters. None of this needs the page to be
downloaded. We expect a model trained only on Lexical features to still reach a reasonable AUC, but the **size of the
gap** between parametric and non-parametric model families should change
with the family.

We compare three parametric methods (LR, LDA, Naive Bayes)
against three non-parametric ones (RF, SVM-RBF, KNN).

| # | Criterion | Threshold |
|---|-----------|-----------|
| C1 | Mean CV AUC gap between the best non-parametric and best parametric model on the **Lexical (URLOnly) tier** | **ΔAUC ≥ 0.02** |
| C2 | Paired Wilcoxon signed-rank test on the 5 per-fold AUCs of those two models | **p < 0.05** |
| C3 | Gradient across tiers: the Lexical gap is **strictly larger** than the FullLite-tier gap . | gap(Lexical) > gap(FullLite) |

H1 is **supported** only if C1 **and** C2 hold. C3 turns the claim from
"the two families differ on one tier" into the stronger statement *"the
difference between families shrinks as the feature tier gets stronger."*
H1 is **rejected** if either C1 or C2 fails on Lexical.

Note:
The FullLite tier is the scenario_2 replacement for a naive "Full" tier: six Behavior
near-leakers (`LineOfCode`, `NoOfJS`, `NoOfCSS`, `NoOfImage`,
`NoOfSelfRef`, `NoOfExternalRef`, each with univariate AUC > 0.95) are
stripped because they push Full to AUC ≈ 1.0 and hide every contrast H1
is trying to measure.

## H2 (Scenario 5) — The cost of dimensionality reduction is not the same on every family

Every feature family can be compressed into a few new axes via PCA
(unsupervised) or LDA projection (supervised). We claim the cost of that
compression is not uniform across families.

On the URL-only (Lexical) tier it is high, because the phishing signal lives in interactions between features and a linear projection loses those interactions; on Behavior and FullLite it approaches zero, because the signal is additive enough that any sufficiently-large PC basis carries it.

| # | Criterion | Threshold |
|---|-----------|-----------|
| C1 | AUC drop from raw Lexical to Lexical compressed into 3 PCs, best downstream model | **ΔAUC ≥ 0.05** |
| C2 | AUC drop from raw FullLite to FullLite compressed into 10 PCs, same model | **ΔAUC ≤ 0.01** |
| C3 | Gradient: drop(Lexical) > drop(FullLite) | **yes** |

H2 is **supported** only if C1 **and** C2 hold on the same downstream
classifier family. C3 turns the claim into a gradient parallel to H1's —
the compression cost rises as the tier gets weaker.

---

# 3. Data Loading & Cleaning

```{r load-data}
raw <- read_csv(
  "PhiUSIIL_Phishing_URL_Dataset.csv",
  show_col_types = FALSE
)

# Strip UTF-8 BOM that appears on the first column name
names(raw) <- map_chr(names(raw), ~ str_replace(.x, "^\ufeff", ""))

# glimpse(raw)
```

```{r define-roles}
# Columns excluded from modeling (identifiers / free text)
ID_COLS <- c("FILENAME", "URL", "Domain", "TLD", "Title")

# Binary target as factor
df <- raw %>%
  mutate(label_bin = factor(label, levels = c(0, 1),
                            labels = c("Legitimate", "Phishing"))) %>%
  select(-all_of(ID_COLS))

# Feature matrix (numeric predictors only)
features <- df %>% select(-label, -label_bin)

cat("Observations:", nrow(df), "\n")
cat("Predictors:  ", ncol(features), "\n")
```


## Feature-family mapping
The three families (Lexical / Trust / Behavior) are our own taxonomy, not
a property of the dataset. They reflect the cost of obtaining each signal
at decision time:
   Lexical  — derivable from the raw URL string alone (zero network cost).
   Trust    — requires light page/domain inspection or an external lookup
              (DNS, TLD database, HTTPS handshake, title parsing).
   Behavior — requires fetching, parsing and partially rendering the page.

Note on `IsHTTPS`: although the "https://" scheme is part of the URL string,
we place it in the **Trust** family. Modern browsers refuse plain HTTP for
any sensitive interaction and warn the user prominently, so the scheme is
already enforced by the trust layer of the browser rather than acting as a
free lexical clue. Treating it as Lexical would let the URLOnly tier cheat
off a binary that is, in practice, a deployment/trust signal.

```{r family-map}

family_map <- tribble(
  ~feature,                    ~family,
  "URLLength",                 "Lexical",
  "DomainLength",              "Lexical",
  "IsDomainIP",                "Lexical",
  "URLSimilarityIndex",        "Trust",
  "CharContinuationRate",      "Lexical",
  "TLDLegitimateProb",         "Trust",
  "URLCharProb",               "Trust",
  "TLDLength",                 "Lexical",
  "NoOfSubDomain",             "Lexical",
  "HasObfuscation",            "Lexical",
  "NoOfObfuscatedChar",        "Lexical",
  "ObfuscationRatio",          "Lexical",
  "NoOfLettersInURL",          "Lexical",
  "LetterRatioInURL",          "Lexical",
  "NoOfDegitsInURL",           "Lexical",
  "DegitRatioInURL",           "Lexical",
  "NoOfEqualsInURL",           "Lexical",
  "NoOfQMarkInURL",            "Lexical",
  "NoOfAmpersandInURL",        "Lexical",
  "NoOfOtherSpecialCharsInURL","Lexical",
  "SpacialCharRatioInURL",     "Lexical",
  "IsHTTPS",                   "Trust",
  "LineOfCode",                "Behavior",
  "LargestLineLength",         "Behavior",
  "HasTitle",                  "Trust",
  "DomainTitleMatchScore",     "Trust",
  "URLTitleMatchScore",        "Trust",
  "HasFavicon",                "Trust",
  "Robots",                    "Trust",
  "IsResponsive",              "Behavior",
  "NoOfURLRedirect",           "Behavior",
  "NoOfSelfRedirect",          "Behavior",
  "HasDescription",            "Behavior",
  "NoOfPopup",                 "Behavior",
  "NoOfiFrame",                "Behavior",
  "HasExternalFormSubmit",     "Behavior",
  "HasSocialNet",              "Behavior",
  "HasSubmitButton",           "Behavior",
  "HasHiddenFields",           "Behavior",
  "HasPasswordField",          "Behavior",
  "Bank",                      "Trust",
  "Pay",                       "Trust",
  "Crypto",                    "Trust",
  "HasCopyrightInfo",          "Behavior",
  "NoOfImage",                 "Behavior",
  "NoOfCSS",                   "Behavior",
  "NoOfJS",                    "Behavior",
  "NoOfSelfRef",               "Behavior",
  "NoOfEmptyRef",              "Behavior",
  "NoOfExternalRef",           "Behavior"
)
```

## 3.3 Feature Exclusion

Before the EDA we drop three groups of features that would bias the
analysis or trivialise the modeling. All three decisions are based on
what the feature actually means.

### Category 1 — Computed probability / similarity scores

These are not raw URL measurements. They are outputs of other models or
expert-written scoring functions that already encode phishing-detection
logic. Keeping them would mean we are classifying another classifier's
output, not the URL itself.

```{r exclude-scores}
SCORES <- c(
  "URLSimilarityIndex",   # similarity to known-phishing URL patterns
  "TLDLegitimateProb",    # P(TLD is legitimate) from an external lookup
  "URLCharProb",          # P(character sequence) from a language model
  "DomainTitleMatchScore",# computed domain↔title consistency score
  "URLTitleMatchScore"    # computed URL↔title consistency score
)
cat("Computed scores to remove:", paste(SCORES, collapse = ", "), "\n")
```

### Category 2 — Redundant binary indicator

`HasObfuscation` is literally `(NoOfObfuscatedChar > 0)`. It holds no
information that the count feature doesn't already carry.

```{r exclude-binary}
REDUNDANT_BINARY <- "HasObfuscation"
cat("Redundant binary to remove:", REDUNDANT_BINARY, "\n")
```

### Category 3 — Redundant ratio features

Each ratio below equals `count / URLLength`, and both the count and
`URLLength` are already in the dataset. Keeping the ratio gives us a
perfect linear dependence — and no extra signal.

```{r exclude-ratios}
RATIOS <- c(
  "LetterRatioInURL",      # = NoOfLettersInURL   / URLLength
  "DegitRatioInURL",       # = NoOfDegitsInURL    / URLLength
  "ObfuscationRatio",      # = NoOfObfuscatedChar / URLLength
  "SpacialCharRatioInURL"  # = NoOfOtherSpecialCharsInURL / URLLength
)
cat("Linearly dependent ratios to remove:", paste(RATIOS, collapse = ", "), "\n")
```

```{r apply-exclusions}
EXCLUDE_COLS <- c(SCORES, REDUNDANT_BINARY, RATIOS)

df       <- df %>% select(-all_of(EXCLUDE_COLS))
features <- df %>% select(-label, -label_bin)

cat("\nFeatures after exclusion:", ncol(features), "\n")
cat("Removed:", paste(EXCLUDE_COLS, collapse = ", "), "\n")
```

Update the family map to reflect the reduced feature set:

```{r family-map-update}
family_map <- family_map %>%
  filter(!feature %in% EXCLUDE_COLS)

family_map %>% count(family) %>% print()
```

```{r feature-types, include=FALSE}
# Type split used by downstream VIF/skewness chunks
binary_features <- names(features)[map_lgl(features, ~ all(na.omit(.x) %in% c(0, 1)))]
continuous_features <- setdiff(names(features), binary_features)
```

## 3.4 Feature Tiers Used in Modeling

To test H1 we define three nested feature tiers. The modeling notebook
reads these tier definitions directly from here, so both notebooks stay
in sync. The `tier_full` built here (40 features) is the EDA baseline
used by §4.5's PCA diagnostics; the scenario_2 modeling notebook
replaces it with **FullLite** (34 features) by stripping six Behavior
near-leakers — see §5 for the reasoning.

```{r feature-tiers}
tier_lexical  <- family_map %>% filter(family == "Lexical")  %>% pull(feature)
tier_trust    <- family_map %>% filter(family == "Trust")    %>% pull(feature)
tier_behavior <- family_map %>% filter(family == "Behavior") %>% pull(feature)
tier_full     <- c(tier_lexical, tier_trust, tier_behavior)

cat("URLOnly tier (Lexical, ",  length(tier_lexical),  " features)\n", sep = "")
cat("Trust tier              (",  length(tier_trust),    " features)\n", sep = "")
cat("Behavior tier           (",  length(tier_behavior), " features)\n", sep = "")
cat("Full tier               (",  length(tier_full),     " features)\n", sep = "")
```

---

# 4. Exploratory Data Analysis


## 4.1 Data Overview

### 4.1.1 Class Balance

```{r class-balance, fig.height=3.5}
df %>%
  count(label_bin) %>%
  mutate(pct = n / sum(n)) %>%
  ggplot(aes(label_bin, n, fill = label_bin)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = scales::percent(pct, 0.1)), vjust = -0.4, size = 4) +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, .1))) +
  labs(title = "Class distribution",
       x = NULL, y = "Count") +
  theme_minimal(base_size = 12)
```

**Takeaway:** The dataset is roughly balanced, so we don't need resampling
or class-weighting. Accuracy works as a secondary metric next to AUC.

### 4.1.2 Missing Values

No missing values are present in any predictor (verified with `colSums(is.na(features))`).

## 4.2 H1 preparation

H1 claims that a URL-only detector (Lexical features) will show a bigger
parametric vs. non-parametric gap than any other tier. We don't test
non-linearity directly in the EDA. Instead we show that **per-feature
discriminative power is lowest in the Lexical family**.

### 4.2.1 Discriminative Power by Family

We measure discriminative power separately for continuous and binary
features. The standardised mean difference (SMD) only makes sense for
continuous features. For binary features the "pooled SD" collapses to a
function of the class proportion, so SMD stops being comparable across
features. For those we use **Cramér's V** instead — the standard
effect-size measure for a 2×2 contingency table, on the same 0–1 scale
as |correlation|.

```{r discriminative-power, fig.height=7, fig.width=8}
# --- Continuous features: standardised mean difference ----------------------
smd_df <- df %>%
  select(label, all_of(continuous_features)) %>%
  pivot_longer(-label, names_to = "feature", values_to = "value") %>%
  group_by(feature, label) %>%
  summarise(m = mean(value, na.rm = TRUE), s = sd(value, na.rm = TRUE),
            .groups = "drop") %>%
  pivot_wider(names_from = label, values_from = c(m, s),
              names_glue = "{.value}_{label}") %>%
  mutate(
    pooled_sd = sqrt((s_0^2 + s_1^2) / 2),
    effect    = abs(m_1 - m_0) / (pooled_sd + 1e-9),
    metric    = "SMD (continuous)"
  ) %>%
  select(feature, effect, metric)

# --- Binary features: Cramer's V on a 2x2 contingency table -----------------
cramers_v <- function(x, y) {
  tab <- table(x, y)
  if (any(dim(tab) < 2)) return(NA_real_)
  chi2 <- suppressWarnings(chisq.test(tab, correct = FALSE)$statistic)
  n <- sum(tab)
  as.numeric(sqrt(chi2 / (n * (min(dim(tab)) - 1))))
}

cv_df <- tibble(
  feature = binary_features,
  effect  = map_dbl(binary_features, ~ cramers_v(features[[.x]], df$label)),
  metric  = "Cramer's V (binary)"
)

discrim_df <- bind_rows(smd_df, cv_df) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  arrange(desc(effect))

ggplot(discrim_df, aes(x = reorder(feature, effect), y = effect, fill = family)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ metric, scales = "free", ncol = 2) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Per-feature discriminative power",
       subtitle = "Continuous: |SMD|. Binary: Cramer's V. Both >= 0.",
       x = NULL, y = "Effect size", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")
```

Effect-size distribution per family (continuous features only):

```{r smd-by-family, fig.height=4, fig.width=6}
smd_family <- discrim_df %>%
  filter(metric == "SMD (continuous)", family != "Other")

ggplot(smd_family, aes(family, effect, fill = family)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  geom_jitter(width = 0.15, alpha = 0.5, size = 2) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "SMD distribution by family (continuous features only)",
       x = NULL, y = "| SMD |") +
  theme_minimal(base_size = 12)
```

**Takeaway (H1):** Among the continuous features, Lexical has the lowest
median |SMD| and the smallest spread. Trust and Behavior carry more
per-feature signal. This supports H1's claim that the URL-only tier has less signal in individual features and relies more on feature combinations.

## 4.3 H2 preparation

H2 says the cost of linear dimensionality reduction differs sharply across
families. A quicklook at multicollinearity tells us how much redundant dimensionality each family has to start with. If a family is already low-rank in variance, aggressive compression is cheap. If it has many independent directions of variance, compression is more expensive. The question is whether the class signal lives in those dominant variance directions or orthogonal to them.

### 4.3.1 Lexical Correlation Matrix

```{r cor-lexical, fig.height=6, fig.width=7}
lexical_cor <- features %>%
  select(all_of(tier_lexical)) %>%
  cor(use = "pairwise.complete.obs")

  corrplot(
  lexical_cor,
  method  = "color",
  order   = "hclust",
  tl.cex  = 0.7,
  tl.col  = "black",
  col     = colorRampPalette(c("firebrick", "white", "steelblue"))(200),
  title   = "Correlation - Lexical features",
  mar     = c(0, 0, 1.5, 0),
  addCoef.col = "black",
  number.cex  = 0.55
)
```

Across the full feature matrix, 10 pairs pass `|r| > 0.7` — and all of
them sit inside this Lexical count cluster (URLLength ↔ NoOfLettersInURL
↔ NoOfDegitsInURL and friends).

### 4.3.2 Variance Inflation Factors

```{r vif}
library(knitr)

# Target-free VIF from the design matrix (avoids non-converged GLM).
X_vif <- features %>%
  select(all_of(continuous_features)) %>%
  as.data.frame()

vif_values <- vapply(seq_along(X_vif), function(i) {
  r2 <- summary(lm(X_vif[[i]] ~ ., data = X_vif[, -i, drop = FALSE]))$r.squared
  1 / (1 - r2)
}, numeric(1))
names(vif_values) <- names(X_vif)

vif_df <- tibble(
  feature = names(vif_values),
  vif     = as.numeric(vif_values)
) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  mutate(severity = case_when(
    vif > 10            ~ "severe (> 10)",
    vif >= 5 & vif <= 10 ~ "moderate (5-10)",
    TRUE                ~ "ok (< 5)"
  )) %>%
  arrange(desc(vif))

vif_df %>%
  transmute(Feature = feature,
            Family  = family,
            VIF     = round(vif, 2),
            Severity = severity) %>%
  kable(caption = "Variance Inflation Factors - continuous features (lm-based, target-free).")
```

**Takeaway (H2):** `URLLength`, `NoOfLettersInURL` and `NoOfDegitsInURL`
stay near-collinear because they measure overlapping properties of the
same URL string. For H2 this means Lexical has far fewer independent
directions of variance than it has features — PCA will compress it
aggressively. The open question is whether the class signal lives in
those dominant variance directions or orthogonal to them.


## 4.4 Skewness Assessment *(Scenario 2 preparation)*

```{r outlier-skewness, fig.height=6, fig.width=8}
# Skewness for all continuous features
skewness_fn <- function(x) {
  x <- x[!is.na(x)]
  n <- length(x)
  m <- mean(x)
  s <- sd(x)
  (n / ((n-1)*(n-2))) * sum(((x - m) / s)^3)
}

skew_df <- tibble(
  feature  = continuous_features,
  skewness = map_dbl(continuous_features, ~ skewness_fn(features[[.x]]))
) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  mutate(abs_skew = abs(skewness)) %>%
  arrange(desc(abs_skew))

ggplot(skew_df, aes(x = reorder(feature, abs_skew), y = skewness, fill = family)) +
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = c(-2, 2), linetype = "dashed", colour = "grey40") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Skewness of continuous features",
       subtitle = "Dashed lines at +/- 2 mark heavy skew; LR/LDA may struggle here",
       x = NULL, y = "Skewness", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")

# Count features with |skewness| > 2
n_heavy_skew <- sum(skew_df$abs_skew > 2)
cat("\nFeatures with |skewness| > 2:", n_heavy_skew, "of",
    length(continuous_features), "\n")
cat("These are:\n")
skew_df %>% filter(abs_skew > 2) %>% select(feature, skewness, family) %>% print()
```

**Takeaway (Scenario 2):** Several count features are heavily right-skewed
with long outlier tails. This shapes our modeling choices:

- **Parametric models** (LR, LDA): need scaling and a log-transform on
  skewed features to make the distributional assumption hold.
- **Non-parametric models** (trees, KNN, SVM): robust to skew, but KNN and
  SVM still need standardisation because they use distances.


## 4.5 H2 preparation — where the class signal lives across principal components

§4.3 showed that Lexical has fewer independent directions of variance
than it has features. That alone only tells us about variance — not
whether any of those directions carry the *class* signal. For H2 the
crucial question is whether the top few PCs of each family line up with
phishing vs. legitimate separation. If they don't, any low-dimensional
projection will drop detection accuracy. Two PCA-based diagnostics answer
that: a scree plot (how many PCs each tier needs to capture most of its
variance) and the univariate AUC of each leading PC (how much class
signal sits along that direction).

### 4.5.1 Scree plot per tier

```{r pca-scree, fig.height=4.5, fig.width=8}
prep_for_pca <- function(cols) {
  X <- features[, cols, drop = FALSE]
  cont <- setdiff(names(X), binary_features)
  if (length(cont) > 0) {
    X[, cont] <- lapply(X[, cont], function(x) log1p(pmax(x, 0)))
  }
  X_mat <- as.matrix(X)
  sds <- apply(X_mat, 2, sd)
  X_mat <- X_mat[, sds > 0, drop = FALSE]
  scale(X_mat)
}

pca_tiers <- list(
  Lexical  = tier_lexical,
  Trust    = tier_trust,
  Behavior = tier_behavior,
  Full     = tier_full
)

scree_df <- map_dfr(names(pca_tiers), function(tier) {
  Xp <- prep_for_pca(pca_tiers[[tier]])
  pc <- prcomp(Xp, center = FALSE, scale. = FALSE)
  var_exp <- pc$sdev^2 / sum(pc$sdev^2)
  tibble(tier = tier, k = seq_along(var_exp), cum_var = cumsum(var_exp))
})

scree_df %>%
  mutate(tier = factor(tier, levels = c("Lexical", "Trust",
                                         "Behavior", "Full"))) %>%
  ggplot(aes(k, cum_var, colour = tier)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0.9, linetype = "dashed", colour = "grey40") +
  scale_x_continuous(breaks = seq(0, 40, 5)) +
  scale_y_continuous(labels = scales::percent) +
  scale_colour_brewer(palette = "Set2") +
  labs(title = "Cumulative variance explained by leading PCs",
       subtitle = "Dashed line at 90% variance - fewer PCs = tier is more low-rank",
       x = "Number of PCs retained", y = "Cumulative variance",
       colour = "Tier") +
  theme_minimal(base_size = 11)
```

Lexical reaches 90 % of its variance with 3–4 PCs — the URL-length
cluster is doing most of the work. Trust spends all 7 directions because
its features are binary and near-uncorrelated. Behavior and Full are
spread out and need 10+ PCs to reach 90 %. That is the *variance*
picture; it predicts that PCA compression on Lexical will look cheap.
The class picture below tells a different story.

### 4.5.2 Class signal carried by each leading PC

```{r pca-per-pc-auc, fig.height=4, fig.width=9}
pca_auc_df <- map_dfr(names(pca_tiers), function(tier) {
  Xp <- prep_for_pca(pca_tiers[[tier]])
  pc <- prcomp(Xp, center = FALSE, scale. = FALSE)
  k_max <- min(10, ncol(pc$x))
  map_dfr(seq_len(k_max), function(i) {
    a <- suppressMessages(as.numeric(
      pROC::roc(df$label_bin, pc$x[, i],
                levels = c("Legitimate", "Phishing"),
                quiet = TRUE)$auc))
    tibble(tier = tier, PC = i, auc = max(a, 1 - a))
  })
})

pca_auc_df %>%
  mutate(tier = factor(tier, levels = c("Lexical", "Trust",
                                         "Behavior", "Full"))) %>%
  ggplot(aes(PC, auc, fill = tier)) +
  geom_col() +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey40") +
  facet_wrap(~ tier, nrow = 1) +
  scale_x_continuous(breaks = 1:10) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Phishing signal carried by each leading PC",
       subtitle = "Orientation-free univariate AUC; PC1 = direction of largest variance",
       x = "Principal component index", y = "Univariate AUC") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")
```

**Takeaway (H2):** On Lexical, PC1 captures almost all the variance but
the class signal is smeared across several PCs with low individual AUC —
variance directions and class directions do not line up. Trust has so few
features that each PC carries a comparable share of the (already weak)
signal. On Behavior and Full the top 2–3 PCs each reach AUC above 0.8, so
a handful of components recover most of the task. This is the direct EDA
prediction H2 tests quantitatively in `scenario_5.rmd`: compression is
cheap on Behavior and Full, expensive on Lexical.

## 4.6 Labelled-data heatmap (Task 4 companion)

A heatmap colours labelled data by feature value; if the two classes
live in different regions of feature space, they should emerge as two
stripes on the plot without any supervised learning. Task 4 asks us to
contrast exactly this "what does the raw data look like" view against
the model-based decision rule. A heatmap is **not** a model — it is an
unsupervised rendering — so the comparison is:

- the heatmap shows **what you can see by eye before fitting anything**,
- the surrogate tree in `scenario_2.rmd §7` shows **what a trained
  tree learns when it tries to mimic the RF**,
- and the RF / SVM numbers in `scenario_2.rmd §6` show **what a
  supervised model actually extracts**.

The gap between the heatmap's cluster purity and the RF's AUC is a
direct measure of *how much supervision buys us* on this dataset.

**Sample and feature choice — the knobs we tuned.** The raw dataset
has 235 k rows and 40 predictors after §3.3. A heatmap over that is
visual noise, so we:

1. Stratified-sample **500 rows** (250 per class). Small enough to
   stay readable, large enough that cluster purity is not dominated by
   individual outliers.
2. Take the **top-10 features** by the discriminative-power ranking
   already computed in §4.2.1 (`discrim_df`, union of |SMD| for
   continuous and Cramér's V for binary). Using all 40 features would
   bury the signal under weak predictors; using only the top 2–3 would
   bake the conclusion in. Ten is the natural breakpoint where
   effect-size drops off the shoulder of the ranking curve.
3. Log-transform + z-score each continuous feature (same recipe the
   parametric / distance models use in `scenario_2.rmd §3.3`).
   Binaries stay on 0/1.

**Clustering knobs.** Hierarchical clustering on rows has two knobs:

- **Distance**: euclidean vs. correlation-based (`1 − |cor|`).
- **Linkage**: `ward.D2` vs. `complete` vs. `average`.

We evaluate all six combinations on two-class cluster recovery and
keep the one with the highest adjusted Rand index (ARI) against the
ground-truth label. Cluster count is fixed at 2 (we know the task is
binary; asking the heatmap to "discover" a different number would be
a different question).

```{r heatmap-task4, fig.height=8, fig.width=10}
set.seed(2026)

# 1) Stratified 500-row sample from the cleaned feature matrix + label.
label_fac <- df$label_bin
heatmap_idx <- c(
  sample(which(label_fac == "Legitimate"), 250),
  sample(which(label_fac == "Phishing"),   250)
)
heatmap_idx <- sample(heatmap_idx)   # shuffle so hclust drives order

# 2) Top-10 features by the §4.2.1 discriminative-power ranking.
top_feats <- discrim_df %>%
  filter(feature %in% names(features)) %>%
  arrange(desc(effect)) %>%
  slice_head(n = 10) %>%
  pull(feature)

X_sample <- features[heatmap_idx, top_feats, drop = FALSE]
y_sample <- label_fac[heatmap_idx]

# 3) Apply the scenario_2 parametric recipe: log1p + z-score on
#    continuous features; binaries left alone.
cont_in_top <- intersect(top_feats, continuous_features)
bin_in_top  <- setdiff(top_feats,   cont_in_top)

X_scaled <- X_sample
X_scaled[cont_in_top] <- lapply(X_scaled[cont_in_top], function(x) {
  v <- log1p(pmax(x, 0))
  as.numeric(scale(v))
})
X_mat <- as.matrix(X_scaled)
stopifnot(!anyNA(X_mat))

# 4) Grid over (distance, linkage); pick best by ARI against label.
adjusted_rand <- function(a, b) {
  tab <- table(a, b)
  n   <- sum(tab)
  sum_comb <- function(x) sum(choose(x, 2))
  index   <- sum_comb(tab)
  sum_a   <- sum_comb(rowSums(tab))
  sum_b   <- sum_comb(colSums(tab))
  expected <- sum_a * sum_b / choose(n, 2)
  max_idx  <- (sum_a + sum_b) / 2
  if (max_idx == expected) return(0)
  (index - expected) / (max_idx - expected)
}

majority_purity <- function(clusters, truth) {
  tab <- table(clusters, truth)
  sum(apply(tab, 1, max)) / sum(tab)
}

grid <- expand.grid(
  distance = c("euclidean", "correlation"),
  linkage  = c("ward.D2", "complete", "average"),
  stringsAsFactors = FALSE
)
grid_res <- grid %>%
  mutate(
    result = pmap(list(distance, linkage), function(d, l) {
      D <- if (d == "euclidean") dist(X_mat) else
           as.dist(1 - abs(cor(t(X_mat))))
      hc <- hclust(D, method = l)
      cl <- cutree(hc, k = 2)
      list(hc = hc, cl = cl,
           ari    = adjusted_rand(cl, y_sample),
           purity = majority_purity(cl, y_sample))
    })
  ) %>%
  mutate(ari    = map_dbl(result, "ari"),
         purity = map_dbl(result, "purity"))

grid_res %>%
  select(distance, linkage, ari, purity) %>%
  mutate(across(c(ari, purity), ~ round(.x, 3))) %>%
  kable(caption = "Unsupervised cluster recovery (k=2) across (distance, linkage) combinations.")

best     <- grid_res %>% slice_max(ari, n = 1, with_ties = FALSE)
best_hc  <- best$result[[1]]$hc
best_cl  <- best$result[[1]]$cl

# 5) Build the heatmap data in hclust row order.
row_order <- best_hc$order
long_df <- as_tibble(X_mat, rownames = "row_id") %>%
  mutate(row_idx = seq_len(n()),
         label   = y_sample) %>%
  pivot_longer(-c(row_id, row_idx, label),
               names_to = "feature", values_to = "value") %>%
  mutate(feature = factor(feature, levels = top_feats),
         row_rank = match(row_idx, row_order))

cap_lo <- quantile(long_df$value, 0.02, na.rm = TRUE)
cap_hi <- quantile(long_df$value, 0.98, na.rm = TRUE)
long_df$value_capped <- pmax(pmin(long_df$value, cap_hi), cap_lo)

ann_df <- long_df %>%
  distinct(row_rank, label)

p_heat <- ggplot(long_df, aes(x = feature, y = row_rank,
                              fill = value_capped)) +
  geom_tile() +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                       midpoint = 0, name = "z-score / 0-1") +
  labs(title = "Top-10 discriminative features, 500 rows clustered",
       subtitle = sprintf(
         "Best combo: distance=%s, linkage=%s -> ARI %.2f, purity %.1f%%",
         best$distance, best$linkage, best$ari, 100 * best$purity),
       x = NULL, y = "Row (ordered by hclust)") +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        axis.text.y = element_blank(),
        panel.grid = element_blank())

p_ann <- ggplot(ann_df, aes(x = 1, y = row_rank, fill = label)) +
  geom_tile() +
  scale_fill_manual(values = c(Legitimate = "#1A9850",
                               Phishing   = "#D73027")) +
  labs(x = NULL, y = NULL, fill = "True label") +
  theme_void() +
  theme(legend.position = "left",
        plot.margin = unit(c(0, 0, 0, 0), "pt"))

p_ann + p_heat + plot_layout(widths = c(0.05, 1))
```

**What the heatmap shows.** Even with only 10 features and no
supervision, the best (distance, linkage) combination recovers the
class boundary with ARI in the 0.3–0.6 range and purity typically ≥
0.8. The phishing block tends to score high on URL length, digit
count, and the `HasObfuscation`-adjacent features; legitimate rows
are darker on those same columns. That's the "you can see it by eye"
baseline.

**What the heatmap does not show — and why the surrogate tree does.**
Two structural limits:

1. The heatmap doesn't know there is a label. It orders rows by
   whatever geometry `(distance, linkage)` imposes — the label strip
   on the left is annotation, not input. Tuning `(distance, linkage)`
   against ARI is the closest thing to "making the visualisation
   align with class structure", but that alignment is already capped
   by how separable the raw features are. Our supervised RF on
   Lexical reaches AUC ≈ 0.97 — substantially above anything this
   heatmap achieves — because it uses the label during training.
2. The heatmap is a 500 × 10 view; any interaction between three or
   more features (URL length × digit count × subdomain depth) is
   invisible. The surrogate tree in `scenario_2.rmd §7`, by contrast,
   is trained on the full 24 k rows and all features in the tier,
   and its tuning target is **the RF itself**, not the truth. Those
   two differences — supervision and model-faithfulness — are why a
   tree diagram can answer Task 4's "capture model-based
   classification" question while a heatmap can only preview the
   geometry.

The heatmap is therefore the **lower bound** of Task 4: it shows how
much of the label is already visible in the features before any
learning. The surrogate tree is the **model-faithful counterpart**
that makes the trained decision rule human-readable.

---

# 5. EDA Summary & Modeling Implications

**A note on separability.** After the Section 3.3 exclusions, 40 features
remain (13 Lexical + 7 Trust + 20 Behavior). On this reduced set the task
is near-perfectly separable once Behavior is present — six Behavior
counts have univariate AUC > 0.95 on their own, so a naive Full tier
saturates at AUC ≈ 1.0 and the parametric vs. non-parametric contrast
disappears.

For H1 we therefore use the scenario_2 family layout: Lexical
(13), Trust (7), Behavior with the six near-leakers stripped (14), and
**FullLite** (34) = Lexical + Trust + reduced Behavior. H1 is tested
primarily on the **URLOnly tier**, where the family contrast has the most
room to show, and the family gradient (C3) is measured against **FullLite**,
not Full. H2 is tested across the same families, because the cost of
compression is what differs between them.

## 5.1 Implications for modeling scenarios

| Finding | Scenario 2 implication (H1) | Scenario 5 implication (H2) |
|---------|----------------------------|------------------------------|
| Lexical SMD lowest + features must be combined | Largest parametric/non-parametric AUC gap expected on URLOnly | On Lexical the signal lives in feature combinations, so a linear projection to 3 PCs is expected to drop AUC sharply |
| Heavy skewness in 19 of 22 continuous features | Log-transform + standardisation for LR/LDA; trees/RF unaffected | Apply the same log+standardise transform before PCA/LDA projection (both are scale-sensitive) |
| URL-length cluster: VIF > 1000 and abs(r) > 0.84 | Use stable solvers / regularised LR for fairness | Lexical is low-rank in variance but the class signal is smeared across PCs — the strongest case for "compression is expensive here" |
| Trust binary features: low intra-correlation | No special treatment | Trust spends all 7 PCs on variance; projection cannot shorten it without losing signal |

