---
title: "Phishing URL Detection — Hypothesis & EDA"
author: "K-o-m-A"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---
```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(corrplot)
library(patchwork)
library(scales)
```

---

# 1. Problem Statement

Phishing attacks trick users into disclosing credentials by serving counterfeit websites
at deceptive URLs. Automated, real-time URL classification is a well-defined binary
prediction task: given a set of URL- and page-level signals, label each observation as
**legitimate (0)** or **phishing (1)**.

The dataset (PhiUSIIL) contains ~235 k observations and 50 predictive features drawn
from three signal families:

| Family | Nature | Example features |
|--------|--------|-----------------|
| **Lexical** | URL string properties | `URLLength`, `NoOfSubDomain`, `ObfuscationRatio`, `IsHTTPS` |
| **Trust** | Domain / title consistency | `URLSimilarityIndex`, `TLDLegitimateProb`, `DomainTitleMatchScore` |
| **Behavior** | Page-level content signals | `NoOfURLRedirect`, `NoOfiFrame`, `HasExternalFormSubmit`, `HasHiddenFields` |

---

# 2. Hypotheses

## H1 — Trust signals are the primary discriminators of phishing URLs

Phishing sites mimic legitimate domains but cannot maintain consistency across all
trust signals.
We expect trust features to show the largest standardised mean differences between
classes, consistently across modelling methods (Scenario 2) and feature-selection
procedures (Scenario 3).

## H2 — Lexical obfuscation and URL structure are necessary but not sufficient

Lexical features carry some signal but overlap with legitimate URLs (e-commerce, news).
We expect them to show lower SMD than trust features and to lose marginal value when combined with trust features
(Scenario 3).

---

# 3. Data Loading & Cleaning

```{r load-data}
raw <- read_csv(
  "PhiUSIIL_Phishing_URL_Dataset.csv",
  show_col_types = FALSE
)

# Strip UTF-8 BOM that appears on the first column name
names(raw) <- map_chr(names(raw), ~ str_replace(.x, "^\ufeff", ""))

# glimpse(raw)
```

```{r define-roles}
# Columns excluded from modelling (identifiers / free text)
ID_COLS <- c("FILENAME", "URL", "Domain", "TLD", "Title")

# Binary target as factor
df <- raw %>%
  mutate(label_bin = factor(label, levels = c(0, 1),
                            labels = c("Legitimate", "Phishing"))) %>%
  select(-all_of(ID_COLS))

# Feature matrix (numeric predictors only)
features <- df %>% select(-label, -label_bin)

cat("Observations:", nrow(df), "\n")
cat("Predictors:  ", ncol(features), "\n")
```

```{r family-map}
# ── Feature-family mapping (reused throughout EDA) ──────────────────────────
family_map <- tribble(
  ~feature,                    ~family,
  "URLLength",                 "Lexical",
  "DomainLength",              "Lexical",
  "IsDomainIP",                "Lexical",
  "URLSimilarityIndex",        "Trust",
  "CharContinuationRate",      "Lexical",
  "TLDLegitimateProb",         "Trust",
  "URLCharProb",               "Trust",
  "TLDLength",                 "Lexical",
  "NoOfSubDomain",             "Lexical",
  "HasObfuscation",            "Lexical",
  "NoOfObfuscatedChar",        "Lexical",
  "ObfuscationRatio",          "Lexical",
  "NoOfLettersInURL",          "Lexical",
  "LetterRatioInURL",          "Lexical",
  "NoOfDegitsInURL",           "Lexical",
  "DegitRatioInURL",           "Lexical",
  "NoOfEqualsInURL",           "Lexical",
  "NoOfQMarkInURL",            "Lexical",
  "NoOfAmpersandInURL",        "Lexical",
  "NoOfOtherSpecialCharsInURL","Lexical",
  "SpacialCharRatioInURL",     "Lexical",
  "IsHTTPS",                   "Trust",
  "LineOfCode",                "Behavior",
  "LargestLineLength",         "Behavior",
  "HasTitle",                  "Trust",
  "DomainTitleMatchScore",     "Trust",
  "URLTitleMatchScore",        "Trust",
  "HasFavicon",                "Trust",
  "Robots",                    "Trust",
  "IsResponsive",              "Behavior",
  "NoOfURLRedirect",           "Behavior",
  "NoOfSelfRedirect",          "Behavior",
  "HasDescription",            "Behavior",
  "NoOfPopup",                 "Behavior",
  "NoOfiFrame",                "Behavior",
  "HasExternalFormSubmit",     "Behavior",
  "HasSocialNet",              "Behavior",
  "HasSubmitButton",           "Behavior",
  "HasHiddenFields",           "Behavior",
  "HasPasswordField",          "Behavior",
  "Bank",                      "Trust",
  "Pay",                       "Trust",
  "Crypto",                    "Trust",
  "HasCopyrightInfo",          "Behavior",
  "NoOfImage",                 "Behavior",
  "NoOfCSS",                   "Behavior",
  "NoOfJS",                    "Behavior",
  "NoOfSelfRef",               "Behavior",
  "NoOfEmptyRef",              "Behavior",
  "NoOfExternalRef",           "Behavior"
)
```

---

# 4. Exploratory Data Analysis


## 4.1 Data Overview

### 4.1.1 Class Balance

```{r class-balance, fig.height=3.5}
df %>%
  count(label_bin) %>%
  mutate(pct = n / sum(n)) %>%
  ggplot(aes(label_bin, n, fill = label_bin)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = scales::percent(pct, 0.1)), vjust = -0.4, size = 4) +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, .1))) +
  labs(title = "Class distribution",
       x = NULL, y = "Count") +
  theme_minimal(base_size = 12)
```

**Takeaway:** The dataset is roughly balanced, so no resampling or class-weight
adjustment is needed before modelling. Accuracy is a valid metric alongside AUC.

### 4.1.2 Missing Values

```{r missing-values}
miss_summary <- features %>%
  summarise(across(everything(), ~ mean(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "miss_rate") %>%
  filter(miss_rate > 0)

if (nrow(miss_summary) == 0) {
  cat("✓ No missing values found in any predictor.\n")
  cat("  All", ncol(features), "features can enter the modelling pipeline",
      "without imputation.\n")
} else {
  print(miss_summary)
}
```

### 4.1.3 Feature Types Overview

```{r feature-types}
binary_features <- features %>%
  summarise(across(everything(), ~ all(.x %in% c(0, 1), na.rm = TRUE))) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "is_binary") %>%
  filter(is_binary) %>%
  pull(feature)

continuous_features <- setdiff(names(features), binary_features)

cat("Binary features (0/1):", length(binary_features), "\n")
cat(str_wrap(paste(binary_features, collapse = ", "), width = 80), "\n\n")
cat("Continuous / count features:", length(continuous_features), "\n")
cat(str_wrap(paste(continuous_features, collapse = ", "), width = 80), "\n")
```


## 4.2 H1 — Trust Signals are Primary Discriminators

### 4.2.1 Continuous Feature Distributions

Two features per family — density plots show class-conditional separation.

```{r density-continuous, fig.height=9, fig.width=8}
cont_selected <- c(
  # Trust
  "URLSimilarityIndex", "DomainTitleMatchScore",
  # Lexical
  "ObfuscationRatio", "DegitRatioInURL",
  # Behavior (count-based)
  "NoOfExternalRef", "NoOfJS"
)

df %>%
  select(label_bin, all_of(cont_selected)) %>%
  pivot_longer(-label_bin, names_to = "feature", values_to = "value") %>%
  mutate(feature = factor(feature, levels = cont_selected)) %>%
  # Apply log1p to count features to handle extreme skew
  group_by(feature) %>%
  mutate(value = ifelse(feature %in% c("NoOfExternalRef", "NoOfJS"),
                        log1p(value), value)) %>%
  ungroup() %>%
  mutate(feature = recode(feature,
    "NoOfExternalRef" = "NoOfExternalRef (log scale)",
    "NoOfJS" = "NoOfJS (log scale)")) %>%
  ggplot(aes(value, fill = label_bin, colour = label_bin)) +
  geom_density(alpha = 0.35, linewidth = 0.4) +
  facet_wrap(~ feature, scales = "free", ncol = 2) +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  scale_colour_manual(values = c("steelblue", "firebrick")) +
  labs(title = "Class-conditional density — continuous features (2 per family)",
       subtitle = "Count features shown on log(1+x) scale to reveal class differences",
       x = NULL, fill = "Class", colour = "Class") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

**Takeaway (H1):** Trust features (`URLSimilarityIndex`, `DomainTitleMatchScore`)
show the clearest class separation — visible confirmation of H1.  Notably,
`URLSimilarityIndex` is *higher* for phishing: phishing URLs deliberately mimic
legitimate domains.  Lexical ratio features show substantial class overlap, previewing
H2.  Behavioral count features require log-transformation to reveal separation, with
implications for parametric modelling (Scenario 2).

### 4.2.2 Discriminative Power by Family (SMD)

```{r smd, fig.height=7, fig.width=8}
smd_df <- df %>%
  select(label, all_of(names(features %>% select(where(is.numeric))))) %>%
  pivot_longer(-label, names_to = "feature", values_to = "value") %>%
  group_by(feature, label) %>%
  summarise(m = mean(value, na.rm = TRUE), s = sd(value, na.rm = TRUE),
            .groups = "drop") %>%
  pivot_wider(names_from = label, values_from = c(m, s),
              names_glue = "{.value}_{label}") %>%
  mutate(
    pooled_sd = sqrt((s_0^2 + s_1^2) / 2),
    smd       = abs(m_1 - m_0) / (pooled_sd + 1e-9)
  ) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  arrange(desc(smd))

ggplot(smd_df, aes(x = reorder(feature, smd), y = smd, fill = family)) +
  geom_col() +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Standardised mean difference by feature and family",
       subtitle = "Higher = more discriminative between phishing and legitimate",
       x = NULL, y = "| SMD |", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")
```

SMD distribution per family with Kruskal-Wallis test:

```{r smd-by-family, fig.height=4, fig.width=6}
smd_family <- smd_df %>% filter(family != "Other")

ggplot(smd_family, aes(family, smd, fill = family)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  geom_jitter(width = 0.15, alpha = 0.5, size = 2) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "SMD distribution by feature family",
       subtitle = "H1 predicts Trust > Lexical; H2 predicts Lexical is secondary",
       x = NULL, y = "| SMD |") +
  theme_minimal(base_size = 12)

# Kruskal–Wallis test: do families differ in median SMD?
kw_test <- kruskal.test(smd ~ family, data = smd_family)
cat("\nKruskal–Wallis test (SMD across families):\n")
cat("  χ² =", round(kw_test$statistic, 2),
    ", df =", kw_test$parameter,
    ", p =", formatC(kw_test$p.value, format = "e", digits = 2), "\n")

# Median SMD per family
smd_family %>%
  group_by(family) %>%
  summarise(
    n_features = n(),
    median_smd = round(median(smd), 3),
    mean_smd   = round(mean(smd), 3),
    max_smd    = round(max(smd), 3)
  ) %>%
  arrange(desc(median_smd))
```

**Takeaway (H1, H2):** Family-level median SMD: Trust (1.005) > Behavior (0.595) > Lexical (0.336), confirmed by Kruskal-Wallis (p=0.012). However, individually `URLSimilarityIndex` (Trust) is #1 but positions #2-5 are Behavior features. Trust is the most *consistent* family; Lexical is weakest (H2 confirmed).

## 4.3 H2 — Lexical Features are Necessary but Not Sufficient

### 4.3.1 Correlation Structure

```{r correlation, fig.height=8, fig.width=9}
num_features <- features %>% select(where(is.numeric))

cor_mat <- cor(num_features, use = "pairwise.complete.obs")

corrplot(
  cor_mat,
  method  = "color",
  order   = "hclust",
  addrect = 4,
  tl.cex  = 0.55,
  tl.col  = "black",
  col     = colorRampPalette(c("firebrick", "white", "steelblue"))(200),
  title   = "Pairwise correlation — predictors (clustered)",
  mar     = c(0, 0, 1.5, 0)
)
```

```{r high-corr-pairs}
# Report feature pairs with |r| > 0.7
cor_long <- cor_mat %>%
  as.data.frame() %>%
  rownames_to_column("feat_a") %>%
  pivot_longer(-feat_a, names_to = "feat_b", values_to = "r") %>%
  filter(feat_a < feat_b, abs(r) > 0.7) %>%
  arrange(desc(abs(r)))

cat("Feature pairs with |r| > 0.7:\n")
print(cor_long, n = Inf)
```

**Takeaway (H2):** 14 feature pairs exceed |r|=0.7. Lexical features dominate (`NoOfLettersInURL`-`URLLength` r=0.956). `DomainTitleMatchScore`-`URLTitleMatchScore` (r=0.961) is the highest pair overall. LASSO and RFE will likely retain different subsets from these correlated groups.

### 4.3.2 Variance Inflation Factors

```{r vif, fig.height=6, fig.width=7}
library(car)

# Fit a logistic regression on all continuous features to compute VIF
glm_vif <- glm(label ~ ., data = df %>% select(label, all_of(continuous_features)),
               family = binomial)

vif_values <- vif(glm_vif)

vif_df <- tibble(
  feature = names(vif_values),
  vif     = as.numeric(vif_values)
) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  arrange(desc(vif))

ggplot(vif_df, aes(x = reorder(feature, vif), y = vif, fill = family)) +
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = 5,  linetype = "dashed", colour = "orange") +
  geom_hline(yintercept = 10, linetype = "dashed", colour = "red") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Variance Inflation Factors — continuous features",
       subtitle = "Orange = 5 (moderate), Red = 10 (severe multicollinearity)",
       x = NULL, y = "VIF", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")

cat("Features with VIF > 10 (severe):\n")
vif_df %>% filter(vif > 10) %>% select(feature, vif, family) %>% print()
cat("\nFeatures with VIF 5–10 (moderate):\n")
vif_df %>% filter(vif >= 5, vif <= 10) %>% select(feature, vif, family) %>% print()
```

**Takeaway (H2):** Lexical features dominate extreme VIF: `URLLength` (2410), `NoOfLettersInURL` (1198), `NoOfDigitsInURL` (203). Two Trust features also severe: `DomainTitleMatchScore` (16.9), `URLSimilarityIndex` (15.5). LR/LDA need regularisation; LASSO will zero out redundant features while tree-based RFE tolerates collinearity.


## 4.4 Skewness Assessment *(Scenario 2 preparation)*

```{r outlier-skewness, fig.height=6, fig.width=8}
# Skewness for all continuous features
skewness_fn <- function(x) {
  x <- x[!is.na(x)]
  n <- length(x)
  m <- mean(x)
  s <- sd(x)
  (n / ((n-1)*(n-2))) * sum(((x - m) / s)^3)
}

skew_df <- tibble(
  feature  = continuous_features,
  skewness = map_dbl(continuous_features, ~ skewness_fn(features[[.x]]))
) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  mutate(abs_skew = abs(skewness)) %>%
  arrange(desc(abs_skew))

ggplot(skew_df, aes(x = reorder(feature, abs_skew), y = skewness, fill = family)) +
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = c(-2, 2), linetype = "dashed", colour = "grey40") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Skewness of continuous features",
       subtitle = "Dashed lines at ±2 mark heavy skew; LR/LDA may struggle here",
       x = NULL, y = "Skewness", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")

# Count features with |skewness| > 2
n_heavy_skew <- sum(skew_df$abs_skew > 2)
cat("\nFeatures with |skewness| > 2:", n_heavy_skew, "of",
    length(continuous_features), "\n")
cat("These are:\n")
skew_df %>% filter(abs_skew > 2) %>% select(feature, skewness, family) %>% print()
```

**Takeaway (Scenario 2):** Several count-based features are heavily right-skewed with
extreme outliers. This has direct implications for model selection:

- **Parametric models** (LR, LDA): will require feature scaling and log-transforms
  of skewed features to satisfy distributional assumptions.
- **Non-parametric models** (trees, KNN, SVM): are robust to skewness but KNN and SVM
  still require standardisation for distance-based computations.



---

# 5. EDA Summary & Modelling Implications

## 5.1 Hypothesis support from EDA

| Hypothesis | EDA evidence | Supported? |
|-----------|-------------|-----------|
| **H1** — Trust signals dominate | Trust features have the highest median SMD (Trust 1.005 > Behavior 0.595 > Lexical 0.336); `URLSimilarityIndex` is the single strongest discriminator. Trust wins the median but the top individual features are a mix of Trust and Behavior. | ✓ Supported (with nuance) |
| **H2** — Lexical features are secondary | Lexical features have the lowest median SMD, the highest VIF values, and high within-group correlations (redundancy). Several have negligible practical discriminative power despite statistical significance. | ✓ Strongly supported |

## 5.2 Implications for modelling scenarios

| Finding | Scenario 2 implication | Scenario 3 implication |
|---------|------------------------|------------------------|
| Class balance ≈ equal | Accuracy valid alongside AUC; no resampling needed | LASSO penalty not confounded by imbalance |
| Severe VIF in Lexical features | LR/LDA coefficients unstable without regularisation | LASSO expected to zero out redundant Lexical features |
| Trust features highest median SMD | Non-parametric methods can exploit non-linear structure | Trust + top Behavior flags expected to dominate selection |
| Heavy skewness in count features | LR/LDA need scaling/transforms; trees handle natively | Log-transforms may change selection rankings |
| Behavioral flags are sparse binary | KNN/SVM require scaling; trees handle natively | Most carry non-redundant signal → likely survive selection |

