---
title: "Phishing URL Detection — Hypothesis & EDA"
author: "K-o-m-A"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---
```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(corrplot)
library(patchwork)
library(scales)
```

---

# 1. Problem Statement

## 1.1 Real-world scenario

A user clicks a link in an email, chat, or search result. The
corporate proxy has to decide **before the page loads** whether the link is
phishing. The cheapest signal at that moment is the URL string itself — no
DNS lookup, no HTML fetch, no rendering. If a URL-only check is good enough,
we can block the link for free. If it isn't, we have to fall back to a
slower page-level check.

The question we want to answer is simple:
**how much phishing signal is already in the URL?** If a URL-only model works well, we can run it as a fast
first-line filter. Where it fails, the slower trust- and behavior-level checks have to take over.

Explored scenarios for hypothesis driven modeling: 2

## 1.2 Dataset

The PhiUSIIL dataset has ~235 k rows and 50 predictor features. We split
them into three families by how expensive they are to compute:

| Family | What it captures | Example features |
|--------|------------------|-----------------|
| **Lexical** | Properties of the URL string | `URLLength`, `NoOfSubDomain`, `NoOfObfuscatedChar`, `CharContinuationRate` |
| **Trust** | Domain / title consistency | `IsHTTPS`, `HasTitle`, `HasFavicon`, `Bank`, `Pay`, `Crypto` |
| **Behavior** | Page-level content signals | `NoOfURLRedirect`, `NoOfiFrame`, `HasExternalFormSubmit`, `HasHiddenFields` |

---

# 2. Hypotheses

## H1 (Scenario 2) — The gap between parametric and non-parametric model families is tier-dependent

Phishing URLs usually leave lexical traces: odd subdomain depth, paths full
of digits, obfuscated characters. None of this needs the page to be
downloaded. We expect a model trained only on Lexical features to still reach a reasonable AUC, but the **size of the
gap** between parametric and non-parametric model families should change
with the family.

We compare three parametric methods (LR, LDA, Naive Bayes)
against three non-parametric ones (RF, SVM-RBF, KNN).

Because the project frame is a **corporate-proxy URL phishing detector**
(EDA §1.1), the deployment cost is not a ranking — it is a single
block/allow verdict at threshold 0.5 per click. A model with high AUC but
low Specificity floods the helpdesk with false blocks; one with low
Sensitivity quietly lets phish through. The H1 gating criteria therefore
target the **weakest-link operating-point metric**
$\mathrm{minSS} = \min(\mathrm{Sensitivity}, \mathrm{Specificity})$ at the
default 0.5 cut, with AUC kept as a complementary ranking criterion.

| # | Criterion | Threshold |
|---|-----------|-----------|
| C1 | Test-set $\mathrm{minSS}$ gap between the best non-parametric and best parametric model on the **Lexical (URLOnly) tier** | **Δ minSS ≥ 0.10** |
| C2 | Paired Wilcoxon signed-rank test on the 10 per-fold AUCs of those two model families (one-sided, non-param > param) | **p < 0.05** |
| C3 | Gradient across tiers in deployment-relevant terms: $\mathrm{minSS}$ gap is **strictly larger** on Lexical than on FullLite | gap(Lexical) > gap(FullLite) |
| C4 | Sanity check on ranking discrimination: mean test AUC gap on Lexical | **ΔAUC ≥ 0.02** |

H1 is **supported** only if C1 **and** C2 hold. C3 turns the claim from
"the two families differ on one tier" into the stronger statement *"the
difference between families shrinks as the feature tier gets stronger."*
C4 is informative — it confirms the families also differ on the
threshold-free metric — but is not gating on its own, since a
non-deployable model with a high AUC (e.g. NB Lexical: AUC 0.85, Spec
0.38) cannot satisfy the project goal regardless of how it ranks. H1 is
**rejected** if either C1 or C2 fails on Lexical.

Note:
The FullLite tier is the scenario_2 replacement for a naive "Full" tier: six Behavior
near-leakers (`LineOfCode`, `NoOfJS`, `NoOfCSS`, `NoOfImage`,
`NoOfSelfRef`, `NoOfExternalRef`, each with univariate AUC > 0.95) are
stripped because they push Full to AUC ≈ 1.0 and hide every contrast H1
is trying to measure.

## H2 (Scenario 3) — Predictor redundancy increases with tier richness

H1 measures *whether* the parametric–non-parametric gap shrinks as the
feature tier gets stronger; it does not say *why*. Section §4 of this
EDA flags two correlated blocks in the predictor space — the URL-length
cluster (`URLLength`, `DomainLength`, `URLCharProb`, with pairwise
$|r| > 0.7$ and VIF > 1000 on the FullLite design matrix) and the
page-count block on Behavior (`NoOfImage`, `NoOfSelfRef`,
`NoOfExternalRef`, etc., even after the six near-leakers are stripped).
On the **Lexical** tier (13 features, no near-saturating signal) every
predictor carries non-trivial unique information; on **FullLite**
(34 features) the same correlated blocks make multiple equally-good
subsets exist. H2 turns this observation into a measurable claim about
feature selection itself.

We compare **one algorithmic** method (Forward stepwise on AIC) against
**two embedded** methods (Lasso, Ridge — all three required by the
assignment) on the same Lexical and FullLite tiers used in S2.

| # | Criterion | Threshold |
|---|-----------|-----------|
| C1 | Difference in retained-feature count between Lasso and Forward | $\lvert k_\text{Lasso} - k_\text{Forward}\rvert \le 1$ on Lexical, $\ge 3$ on FullLite |
| C2 | Jaccard overlap of retained feature sets between Lasso and Forward | $J \ge 0.85$ on Lexical, $J \le 0.70$ on FullLite |
| C3 | Wald p-value flips along the Forward path (significant ↔ non-significant for the same predictor as other predictors enter) | $\ge 1$ flip on FullLite, $0$ flips on Lexical |

H2 is **supported** only if at least two of {C1, C2, C3} hold in the
predicted direction *on both tiers*. If Lexical and FullLite behave
the same way (e.g. Jaccard collapses on both, or no flips on either),
H2 is **rejected** — the redundancy story would not match the data.

The bridge to H1 is left as a **narrative observation** in S3's verdict
rather than a gating criterion: if regularised parametric models on
FullLite reach AUC within striking distance of the S2 non-parametric
champions, that is a useful corroboration but not what H2 tests.

---

# 3. Data Loading & Cleaning

```{r load-data}
raw <- read_csv(
  "PhiUSIIL_Phishing_URL_Dataset.csv",
  show_col_types = FALSE
)

# Strip UTF-8 BOM that appears on the first column name
names(raw) <- map_chr(names(raw), ~ str_replace(.x, "^\ufeff", ""))

# glimpse(raw)
```

```{r define-roles}
# Columns excluded from modeling (identifiers / free text)
ID_COLS <- c("FILENAME", "URL", "Domain", "TLD", "Title")

# Binary target as factor
df <- raw %>%
  mutate(label_bin = factor(label, levels = c(0, 1),
                            labels = c("Legitimate", "Phishing"))) %>%
  select(-all_of(ID_COLS))

# Feature matrix (numeric predictors only)
features <- df %>% select(-label, -label_bin)

cat("Observations:", nrow(df), "\n")
cat("Predictors:  ", ncol(features), "\n")
```


## Feature-family mapping
The three families (Lexical / Trust / Behavior) are our own taxonomy, not
a property of the dataset. They reflect the cost of obtaining each signal
at decision time:
   Lexical  — derivable from the raw URL string alone (zero network cost).
   Trust    — requires light page/domain inspection or an external lookup
              (DNS, TLD database, HTTPS handshake, title parsing).
   Behavior — requires fetching, parsing and partially rendering the page.

Note on `IsHTTPS`: although the "https://" scheme is part of the URL string,
we place it in the **Trust** family. Modern browsers refuse plain HTTP for
any sensitive interaction and warn the user prominently, so the scheme is
already enforced by the trust layer of the browser rather than acting as a
free lexical clue. Treating it as Lexical would let the URLOnly tier cheat
off a binary that is, in practice, a deployment/trust signal.

```{r family-map}

family_map <- tribble(
  ~feature,                    ~family,
  "URLLength",                 "Lexical",
  "DomainLength",              "Lexical",
  "IsDomainIP",                "Lexical",
  "URLSimilarityIndex",        "Trust",
  "CharContinuationRate",      "Lexical",
  "TLDLegitimateProb",         "Trust",
  "URLCharProb",               "Trust",
  "TLDLength",                 "Lexical",
  "NoOfSubDomain",             "Lexical",
  "HasObfuscation",            "Lexical",
  "NoOfObfuscatedChar",        "Lexical",
  "ObfuscationRatio",          "Lexical",
  "NoOfLettersInURL",          "Lexical",
  "LetterRatioInURL",          "Lexical",
  "NoOfDegitsInURL",           "Lexical",
  "DegitRatioInURL",           "Lexical",
  "NoOfEqualsInURL",           "Lexical",
  "NoOfQMarkInURL",            "Lexical",
  "NoOfAmpersandInURL",        "Lexical",
  "NoOfOtherSpecialCharsInURL","Lexical",
  "SpacialCharRatioInURL",     "Lexical",
  "IsHTTPS",                   "Trust",
  "LineOfCode",                "Behavior",
  "LargestLineLength",         "Behavior",
  "HasTitle",                  "Trust",
  "DomainTitleMatchScore",     "Trust",
  "URLTitleMatchScore",        "Trust",
  "HasFavicon",                "Trust",
  "Robots",                    "Trust",
  "IsResponsive",              "Behavior",
  "NoOfURLRedirect",           "Behavior",
  "NoOfSelfRedirect",          "Behavior",
  "HasDescription",            "Behavior",
  "NoOfPopup",                 "Behavior",
  "NoOfiFrame",                "Behavior",
  "HasExternalFormSubmit",     "Behavior",
  "HasSocialNet",              "Behavior",
  "HasSubmitButton",           "Behavior",
  "HasHiddenFields",           "Behavior",
  "HasPasswordField",          "Behavior",
  "Bank",                      "Trust",
  "Pay",                       "Trust",
  "Crypto",                    "Trust",
  "HasCopyrightInfo",          "Behavior",
  "NoOfImage",                 "Behavior",
  "NoOfCSS",                   "Behavior",
  "NoOfJS",                    "Behavior",
  "NoOfSelfRef",               "Behavior",
  "NoOfEmptyRef",              "Behavior",
  "NoOfExternalRef",           "Behavior"
)
```

## 3.3 Feature Exclusion

Before the EDA we drop three groups of features that would bias the
analysis or trivialise the modeling. All three decisions are based on
what the feature actually means.

### Category 1 — Computed probability / similarity scores

These are not raw URL measurements. They are outputs of other models or
expert-written scoring functions that already encode phishing-detection
logic. Keeping them would mean we are classifying another classifier's
output, not the URL itself.

```{r exclude-scores}
SCORES <- c(
  "URLSimilarityIndex",   # similarity to known-phishing URL patterns
  "TLDLegitimateProb",    # P(TLD is legitimate) from an external lookup
  "URLCharProb",          # P(character sequence) from a language model
  "DomainTitleMatchScore",# computed domain↔title consistency score
  "URLTitleMatchScore"    # computed URL↔title consistency score
)
cat("Computed scores to remove:", paste(SCORES, collapse = ", "), "\n")
```

### Category 2 — Redundant binary indicator

`HasObfuscation` is literally `(NoOfObfuscatedChar > 0)`. It holds no
information that the count feature doesn't already carry.

```{r exclude-binary}
REDUNDANT_BINARY <- "HasObfuscation"
cat("Redundant binary to remove:", REDUNDANT_BINARY, "\n")
```

### Category 3 — Redundant ratio features

Each ratio below equals `count / URLLength`, and both the count and
`URLLength` are already in the dataset. Keeping the ratio gives us a
perfect linear dependence — and no extra signal.

```{r exclude-ratios}
RATIOS <- c(
  "LetterRatioInURL",      # = NoOfLettersInURL   / URLLength
  "DegitRatioInURL",       # = NoOfDegitsInURL    / URLLength
  "ObfuscationRatio",      # = NoOfObfuscatedChar / URLLength
  "SpacialCharRatioInURL"  # = NoOfOtherSpecialCharsInURL / URLLength
)
cat("Linearly dependent ratios to remove:", paste(RATIOS, collapse = ", "), "\n")
```

```{r apply-exclusions}
EXCLUDE_COLS <- c(SCORES, REDUNDANT_BINARY, RATIOS)

df       <- df %>% select(-all_of(EXCLUDE_COLS))
features <- df %>% select(-label, -label_bin)

cat("\nFeatures after exclusion:", ncol(features), "\n")
cat("Removed:", paste(EXCLUDE_COLS, collapse = ", "), "\n")
```

Update the family map to reflect the reduced feature set:

```{r family-map-update}
family_map <- family_map %>%
  filter(!feature %in% EXCLUDE_COLS)

family_map %>% count(family) %>% print()
```

```{r feature-types, include=FALSE}
# Type split used by downstream VIF/skewness chunks
binary_features <- names(features)[map_lgl(features, ~ all(na.omit(.x) %in% c(0, 1)))]
continuous_features <- setdiff(names(features), binary_features)
```

## 3.4 Feature Tiers Used in Modeling

To test H1 we define three nested feature tiers. The modeling notebook
reads these tier definitions directly from here, so both notebooks stay
in sync. The `tier_full` built here (40 features) is the EDA baseline
used by §4.5's PCA diagnostics; the scenario_2 modeling notebook
replaces it with **FullLite** (34 features) by stripping six Behavior
near-leakers — see §5 for the reasoning.

```{r feature-tiers}
tier_lexical  <- family_map %>% filter(family == "Lexical")  %>% pull(feature)
tier_trust    <- family_map %>% filter(family == "Trust")    %>% pull(feature)
tier_behavior <- family_map %>% filter(family == "Behavior") %>% pull(feature)
tier_full     <- c(tier_lexical, tier_trust, tier_behavior)

cat("URLOnly tier (Lexical, ",  length(tier_lexical),  " features)\n", sep = "")
cat("Trust tier              (",  length(tier_trust),    " features)\n", sep = "")
cat("Behavior tier           (",  length(tier_behavior), " features)\n", sep = "")
cat("Full tier               (",  length(tier_full),     " features)\n", sep = "")
```

---

# 4. Exploratory Data Analysis


## 4.1 Data Overview

### 4.1.1 Class Balance

```{r class-balance, fig.height=3.5}
df %>%
  count(label_bin) %>%
  mutate(pct = n / sum(n)) %>%
  ggplot(aes(label_bin, n, fill = label_bin)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = scales::percent(pct, 0.1)), vjust = -0.4, size = 4) +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, .1))) +
  labs(title = "Class distribution",
       x = NULL, y = "Count") +
  theme_minimal(base_size = 12)
```

**Takeaway:** The dataset is roughly balanced, so we don't need resampling
or class-weighting. Accuracy works as a secondary metric next to AUC.

### 4.1.2 Missing Values

No missing values are present in any predictor (verified with `colSums(is.na(features))`).

## 4.2 H1 preparation

H1 claims that a URL-only detector (Lexical features) will show a bigger
parametric vs. non-parametric gap than any other tier. We don't test
non-linearity directly in the EDA. Instead we show that **per-feature
discriminative power is lowest in the Lexical family**.

### 4.2.1 Discriminative Power by Family

We measure discriminative power separately for continuous and binary
features. The standardised mean difference (SMD) only makes sense for
continuous features. For binary features the "pooled SD" collapses to a
function of the class proportion, so SMD stops being comparable across
features. For those we use **Cramér's V** instead — the standard
effect-size measure for a 2×2 contingency table, on the same 0–1 scale
as |correlation|.

```{r discriminative-power, fig.height=7, fig.width=8}
# --- Continuous features: standardised mean difference ----------------------
smd_df <- df %>%
  select(label, all_of(continuous_features)) %>%
  pivot_longer(-label, names_to = "feature", values_to = "value") %>%
  group_by(feature, label) %>%
  summarise(m = mean(value, na.rm = TRUE), s = sd(value, na.rm = TRUE),
            .groups = "drop") %>%
  pivot_wider(names_from = label, values_from = c(m, s),
              names_glue = "{.value}_{label}") %>%
  mutate(
    pooled_sd = sqrt((s_0^2 + s_1^2) / 2),
    effect    = abs(m_1 - m_0) / (pooled_sd + 1e-9),
    metric    = "SMD (continuous)"
  ) %>%
  select(feature, effect, metric)

# --- Binary features: Cramer's V on a 2x2 contingency table -----------------
cramers_v <- function(x, y) {
  tab <- table(x, y)
  if (any(dim(tab) < 2)) return(NA_real_)
  chi2 <- suppressWarnings(chisq.test(tab, correct = FALSE)$statistic)
  n <- sum(tab)
  as.numeric(sqrt(chi2 / (n * (min(dim(tab)) - 1))))
}

cv_df <- tibble(
  feature = binary_features,
  effect  = map_dbl(binary_features, ~ cramers_v(features[[.x]], df$label)),
  metric  = "Cramer's V (binary)"
)

discrim_df <- bind_rows(smd_df, cv_df) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  arrange(desc(effect))

ggplot(discrim_df, aes(x = reorder(feature, effect), y = effect, fill = family)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ metric, scales = "free", ncol = 2) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Per-feature discriminative power",
       subtitle = "Continuous: |SMD|. Binary: Cramer's V. Both >= 0.",
       x = NULL, y = "Effect size", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")
```

Effect-size distribution per family (continuous features only):

```{r smd-by-family, fig.height=4, fig.width=6}
smd_family <- discrim_df %>%
  filter(metric == "SMD (continuous)", family != "Other")

ggplot(smd_family, aes(family, effect, fill = family)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  geom_jitter(width = 0.15, alpha = 0.5, size = 2) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "SMD distribution by family (continuous features only)",
       x = NULL, y = "| SMD |") +
  theme_minimal(base_size = 12)
```

**Takeaway (H1):** Among the continuous features, Lexical has the lowest
median |SMD| and the smallest spread. Trust and Behavior carry more
per-feature signal. This supports H1's claim that the URL-only tier has less signal in individual features and relies more on feature combinations.

## 4.3 Collinearity inside Lexical

Before dimensionality diagnostics, we
check how much redundant dimensionality Lexical has to start with.
Heavy collinearity means a few linear combinations already carry most
of the variance — so PCA's first few axes will dominate the variance
budget, and the open question shifts to whether those dominant
variance directions also carry the *class* signal (answered in §4.5).
The URL-length count cluster is the main source of collinearity; we
keep Trust and Behavior out of this panel because their features are
largely binary and near-uncorrelated.

### 4.3.1 Lexical Correlation Matrix

```{r cor-lexical, fig.height=6, fig.width=7}
lexical_cor <- features %>%
  select(all_of(tier_lexical)) %>%
  cor(use = "pairwise.complete.obs")

  corrplot(
  lexical_cor,
  method  = "color",
  order   = "hclust",
  tl.cex  = 0.7,
  tl.col  = "black",
  col     = colorRampPalette(c("firebrick", "white", "steelblue"))(200),
  title   = "Correlation - Lexical features",
  mar     = c(0, 0, 1.5, 0),
  addCoef.col = "black",
  number.cex  = 0.55
)
```

Across the full feature matrix, 10 pairs pass `|r| > 0.7` — and all of
them sit inside this Lexical count cluster (URLLength ↔ NoOfLettersInURL
↔ NoOfDegitsInURL and friends).

### 4.3.2 Variance Inflation Factors

```{r vif}
library(knitr)

# Target-free VIF from the design matrix (avoids non-converged GLM).
X_vif <- features %>%
  select(all_of(continuous_features)) %>%
  as.data.frame()

vif_values <- vapply(seq_along(X_vif), function(i) {
  r2 <- summary(lm(X_vif[[i]] ~ ., data = X_vif[, -i, drop = FALSE]))$r.squared
  1 / (1 - r2)
}, numeric(1))
names(vif_values) <- names(X_vif)

vif_df <- tibble(
  feature = names(vif_values),
  vif     = as.numeric(vif_values)
) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  mutate(severity = case_when(
    vif > 10            ~ "severe (> 10)",
    vif >= 5 & vif <= 10 ~ "moderate (5-10)",
    TRUE                ~ "ok (< 5)"
  )) %>%
  arrange(desc(vif))

vif_df %>%
  transmute(Feature = feature,
            Family  = family,
            VIF     = round(vif, 2),
            Severity = severity) %>%
  kable(caption = "Variance Inflation Factors - continuous features (lm-based, target-free).")
```

**Takeaway:** `URLLength`, `NoOfLettersInURL` and
`NoOfDegitsInURL` stay near-collinear because they measure overlapping
properties of the same URL string. Lexical therefore has fewer
independent directions of variance than it has features — PCA will
concentrate 95 % of its variance into a small *k*. That alone does
not yet say anything about *class separation* in those directions;
the §4.5 scree and per-PC AUC answer whether variance axes and
class-separation axes coincide or diverge.


## 4.4 Skewness Assessment *(Scenario 2 preparation)*

```{r outlier-skewness, fig.height=6, fig.width=8}
# Skewness for all continuous features
skewness_fn <- function(x) {
  x <- x[!is.na(x)]
  n <- length(x)
  m <- mean(x)
  s <- sd(x)
  (n / ((n-1)*(n-2))) * sum(((x - m) / s)^3)
}

skew_df <- tibble(
  feature  = continuous_features,
  skewness = map_dbl(continuous_features, ~ skewness_fn(features[[.x]]))
) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  mutate(abs_skew = abs(skewness)) %>%
  arrange(desc(abs_skew))

ggplot(skew_df, aes(x = reorder(feature, abs_skew), y = skewness, fill = family)) +
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = c(-2, 2), linetype = "dashed", colour = "grey40") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Skewness of continuous features",
       subtitle = "Dashed lines at +/- 2 mark heavy skew; LR/LDA may struggle here",
       x = NULL, y = "Skewness", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")

# Count features with |skewness| > 2
n_heavy_skew <- sum(skew_df$abs_skew > 2)
cat("\nFeatures with |skewness| > 2:", n_heavy_skew, "of",
    length(continuous_features), "\n")
cat("These are:\n")
skew_df %>% filter(abs_skew > 2) %>% select(feature, skewness, family) %>% print()
```

**Takeaway (Scenario 2):** Several count features are heavily right-skewed
with long outlier tails. This shapes our modeling choices:

- **Parametric models** (LR, LDA): need scaling and a log-transform on
  skewed features to make the distributional assumption hold.
- **Non-parametric models** (trees, KNN, SVM): robust to skew, but KNN and
  SVM still need standardisation because they use distances.

---

# 5. EDA Summary & Modeling Implications

**A note on separability.** After the Section 3.3 exclusions, 40 features
remain (13 Lexical + 7 Trust + 20 Behavior). On this reduced set the task
is near-perfectly separable once Behavior is present — six Behavior
counts have univariate AUC > 0.95 on their own, so a naive Full tier
saturates at AUC ≈ 1.0 and the parametric vs. non-parametric contrast
disappears.

For H1 we therefore use the scenario_2 family layout: Lexical
(13), Trust (7), Behavior with the six near-leakers stripped (14), and
**FullLite** (34) = Lexical + Trust + reduced Behavior. H1 is tested
primarily on the **URLOnly tier**, where the family contrast has the
most room to show, and the family gradient (C3) is measured against
**FullLite**, not Full. The cross-tier §4.5 scree / per-PC AUC is
retained as context for where variance and class-separation directions
diverge most clearly.

## 5.1 Implications for modeling scenarios

| Finding | Scenario 2 implication (H1) | Additional EDA implication |
|---------|----------------------------|----------------------------|
| Lexical SMD lowest + features must be combined | Largest parametric/non-parametric AUC gap expected on URLOnly | Feature interactions matter more than single-feature thresholds in URL-only detection |
| Heavy skewness in 19 of 22 continuous features | Log-transform + standardisation for LR/LDA; trees/RF unaffected | Keep one consistent preprocessing recipe across model families for fair comparisons |
| URL-length cluster: VIF > 1000 and abs(r) > 0.84 | Use stable solvers / regularised LR for fairness | Lexical is low-rank in variance — strong compression is possible without preserving all class signal |
| Lexical panel §4.5.2: PC1 carries most variance but low per-PC AUC across PC1–PC5 | — | Variance axes can be orthogonal to class-separation axes, so variance retention alone is not enough |
| Scenario 2: widest linear/non-linear AUC gap on Lexical (LR 0.861 vs SVM 0.997) | — | Lexical remains the best tier for studying where linear assumptions break |
| FullLite top 2–3 PCs each carry AUC > 0.8 (§4.5.2) | — | On stronger tiers, many methods saturate; contrast is easier to observe on weaker tiers |

