---
title: "Phishing URL Detection — Hypothesis & EDA"
author: "K-o-m-A"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---
```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(corrplot)
library(patchwork)
library(scales)
```

---

# 1. Problem Statement
The dataset (PhiUSIIL) contains ~235 k observations and 50 predictive features drawn
from three signal families:

| Family | Nature | Example features |
|--------|--------|-----------------|
| **Lexical** | URL string properties | `URLLength`, `NoOfSubDomain`, `ObfuscationRatio`, `IsHTTPS` |
| **Trust** | Domain / title consistency | `URLSimilarityIndex`, `TLDLegitimateProb`, `DomainTitleMatchScore` |
| **Behavior** | Page-level content signals | `NoOfURLRedirect`, `NoOfiFrame`, `HasExternalFormSubmit`, `HasHiddenFields` |

---

# 2. Hypotheses

## H1 — A URL-string-only detector provides useful first-line defence, but non-parametric models extract significantly more signal from it than parametric models

Phishing URLs leave lexical traces in the URL string itself — unusual subdomain
depth, digit-heavy paths, obfuscated characters — that do not require downloading
the target page. We expect a model trained exclusively on Lexical features (URL
string, no DNS lookup, no HTML parse) to achieve meaningful AUC, but with a
larger parametric vs. non-parametric gap than any other feature tier. This gap
arises because lexical patterns are non-linear (combinations of character counts,
lengths, special chars) and linear models cannot fully exploit them.

## H2 — Lexical + Trust achieves higher AUC than Lexical + Behavior, despite Trust having fewer features

When augmenting the Lexical baseline with a second feature family, we
hypothesise that Trust features are the stronger complement. Trust features
are fewer in number but individually more discriminative (higher per-feature
SMD) and mutually less redundant (lower intra-family correlation). Behavior
features are numerous but internally correlated, so much of their signal is
shared. We expect Lexical + Trust to outperform Lexical + Behavior in AUC,
demonstrating that feature quality and independence matter more than feature
count.

---

# 3. Data Loading & Cleaning

```{r load-data}
raw <- read_csv(
  "PhiUSIIL_Phishing_URL_Dataset.csv",
  show_col_types = FALSE
)

# Strip UTF-8 BOM that appears on the first column name
names(raw) <- map_chr(names(raw), ~ str_replace(.x, "^\ufeff", ""))

# glimpse(raw)
```

```{r define-roles}
# Columns excluded from modelling (identifiers / free text)
ID_COLS <- c("FILENAME", "URL", "Domain", "TLD", "Title")

# Binary target as factor
df <- raw %>%
  mutate(label_bin = factor(label, levels = c(0, 1),
                            labels = c("Legitimate", "Phishing"))) %>%
  select(-all_of(ID_COLS))

# Feature matrix (numeric predictors only)
features <- df %>% select(-label, -label_bin)

cat("Observations:", nrow(df), "\n")
cat("Predictors:  ", ncol(features), "\n")
```

```{r family-map}
# ── Feature-family mapping (reused throughout EDA) ──────────────────────────
family_map <- tribble(
  ~feature,                    ~family,
  "URLLength",                 "Lexical",
  "DomainLength",              "Lexical",
  "IsDomainIP",                "Lexical",
  "URLSimilarityIndex",        "Trust",
  "CharContinuationRate",      "Lexical",
  "TLDLegitimateProb",         "Trust",
  "URLCharProb",               "Trust",
  "TLDLength",                 "Lexical",
  "NoOfSubDomain",             "Lexical",
  "HasObfuscation",            "Lexical",
  "NoOfObfuscatedChar",        "Lexical",
  "ObfuscationRatio",          "Lexical",
  "NoOfLettersInURL",          "Lexical",
  "LetterRatioInURL",          "Lexical",
  "NoOfDegitsInURL",           "Lexical",
  "DegitRatioInURL",           "Lexical",
  "NoOfEqualsInURL",           "Lexical",
  "NoOfQMarkInURL",            "Lexical",
  "NoOfAmpersandInURL",        "Lexical",
  "NoOfOtherSpecialCharsInURL","Lexical",
  "SpacialCharRatioInURL",     "Lexical",
  "IsHTTPS",                   "Trust",
  "LineOfCode",                "Behavior",
  "LargestLineLength",         "Behavior",
  "HasTitle",                  "Trust",
  "DomainTitleMatchScore",     "Trust",
  "URLTitleMatchScore",        "Trust",
  "HasFavicon",                "Trust",
  "Robots",                    "Trust",
  "IsResponsive",              "Behavior",
  "NoOfURLRedirect",           "Behavior",
  "NoOfSelfRedirect",          "Behavior",
  "HasDescription",            "Behavior",
  "NoOfPopup",                 "Behavior",
  "NoOfiFrame",                "Behavior",
  "HasExternalFormSubmit",     "Behavior",
  "HasSocialNet",              "Behavior",
  "HasSubmitButton",           "Behavior",
  "HasHiddenFields",           "Behavior",
  "HasPasswordField",          "Behavior",
  "Bank",                      "Trust",
  "Pay",                       "Trust",
  "Crypto",                    "Trust",
  "HasCopyrightInfo",          "Behavior",
  "NoOfImage",                 "Behavior",
  "NoOfCSS",                   "Behavior",
  "NoOfJS",                    "Behavior",
  "NoOfSelfRef",               "Behavior",
  "NoOfEmptyRef",              "Behavior",
  "NoOfExternalRef",           "Behavior"
)
```

## 3.3 Feature Exclusion

Before EDA we remove three categories of features that would bias or
trivialise downstream modeling. All decisions are based on feature
semantics.

### Category 1 — Computed probability / similarity scores

These features are not raw URL measurements. They are outputs of
prior models or expert-designed scoring functions that already encode
phishing-detection logic. Including them would mean classifying the
outputs of another classifier, not the URL itself.

```{r exclude-scores}
SCORES <- c(
  "URLSimilarityIndex",   # similarity to known-phishing URL patterns
  "TLDLegitimateProb",    # P(TLD is legitimate) from an external lookup
  "URLCharProb",          # P(character sequence) from a language model
  "DomainTitleMatchScore",# computed domain↔title consistency score
  "URLTitleMatchScore"    # computed URL↔title consistency score
)
cat("Computed scores to remove:", paste(SCORES, collapse = ", "), "\n")
```

### Category 2 — Redundant binary indicator

`HasObfuscation` is defined as `(NoOfObfuscatedChar > 0)`. It carries
no information beyond the count feature already in the dataset.

```{r exclude-binary}
REDUNDANT_BINARY <- "HasObfuscation"
cat("Redundant binary to remove:", REDUNDANT_BINARY, "\n")
```

### Category 3 — Linearly dependent ratio features

Each ratio below equals `count / URLLength`, where both the count and
`URLLength` are already present. Keeping both introduces perfect linear
dependence (VIF → ∞) without adding signal.

```{r exclude-ratios}
RATIOS <- c(
  "LetterRatioInURL",      # = NoOfLettersInURL   / URLLength
  "DegitRatioInURL",       # = NoOfDegitsInURL    / URLLength
  "ObfuscationRatio",      # = NoOfObfuscatedChar / URLLength
  "SpacialCharRatioInURL"  # = NoOfOtherSpecialCharsInURL / URLLength
)
cat("Linearly dependent ratios to remove:", paste(RATIOS, collapse = ", "), "\n")
```

```{r apply-exclusions}
EXCLUDE_COLS <- c(SCORES, REDUNDANT_BINARY, RATIOS)

df       <- df %>% select(-all_of(EXCLUDE_COLS))
features <- df %>% select(-label, -label_bin)

cat("\nFeatures after exclusion:", ncol(features),
    "(removed", length(EXCLUDE_COLS), "of", ncol(features) + length(EXCLUDE_COLS), ")\n")
cat("Removed:", paste(EXCLUDE_COLS, collapse = ", "), "\n")
```

Update the family map to reflect the reduced feature set:

```{r family-map-update}
family_map <- family_map %>%
  filter(!feature %in% EXCLUDE_COLS)

family_map %>% count(family) %>% print()
```

---

# 4. Exploratory Data Analysis


## 4.1 Data Overview

### 4.1.1 Class Balance

```{r class-balance, fig.height=3.5}
df %>%
  count(label_bin) %>%
  mutate(pct = n / sum(n)) %>%
  ggplot(aes(label_bin, n, fill = label_bin)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = scales::percent(pct, 0.1)), vjust = -0.4, size = 4) +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, .1))) +
  labs(title = "Class distribution",
       x = NULL, y = "Count") +
  theme_minimal(base_size = 12)
```

**Takeaway:** The dataset is roughly balanced, so no resampling or class-weight
adjustment is needed before modelling. Accuracy is a valid metric alongside AUC.

### 4.1.2 Missing Values

```{r missing-values}
miss_summary <- features %>%
  summarise(across(everything(), ~ mean(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "miss_rate") %>%
  filter(miss_rate > 0)

if (nrow(miss_summary) == 0) {
  cat("✓ No missing values found in any predictor.\n")
  cat("  All", ncol(features), "features can enter the modelling pipeline",
      "without imputation.\n")
} else {
  print(miss_summary)
}
```


## 4.2 H1 — Lexical Features Show Class Overlap That Requires Non-Linear Models

H1 predicts that a URL-string-only detector (Lexical features) will show a
larger parametric vs. non-parametric gap than other tiers, because Lexical
patterns are non-linear combinations of character counts and URL structure
that linear decision boundaries cannot fully capture.

### 4.2.1 Class-Conditional Distributions — Lexical URLOnly Tier

Six Lexical features — these are all that a URL-string-only detector has access
to without any network call or page download.

```{r density-continuous, fig.height=9, fig.width=8}
cont_selected <- c(
  "URLLength", "NoOfSubDomain",
  "NoOfObfuscatedChar", "NoOfDegitsInURL",
  "NoOfEqualsInURL", "NoOfAmpersandInURL"
)

df %>%
  select(label_bin, all_of(cont_selected)) %>%
  pivot_longer(-label_bin, names_to = "feature", values_to = "value") %>%
  mutate(feature = factor(feature, levels = cont_selected)) %>%
  group_by(feature) %>%
  mutate(value = log1p(value)) %>%
  ungroup() %>%
  mutate(feature = paste0(feature, " (log1p)")) %>%
  ggplot(aes(value, fill = label_bin, colour = label_bin)) +
  geom_density(alpha = 0.35, linewidth = 0.4) +
  facet_wrap(~ feature, scales = "free", ncol = 2) +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  scale_colour_manual(values = c("steelblue", "firebrick")) +
  labs(title = "Class-conditional density — Lexical features (URLOnly tier)",
       subtitle = "Substantial class overlap predicts parametric models will struggle without non-linear boundaries",
       x = NULL, fill = "Class", colour = "Class") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

**Takeaway (H1):** Lexical features show substantial class overlap — phishing
and legitimate URLs share similar URL lengths, digit counts, and special
character patterns. This overlap is the evidence that linear models will underperform non-parametric
ones on the URLOnly tier: parametric models draw a single linear boundary
through overlapping distributions, while RF/SVM find local non-linear
decision rules in the character-count space.

### 4.2.2 Discriminative Power by Family (SMD)

```{r smd, fig.height=7, fig.width=8}
smd_df <- df %>%
  select(label, all_of(names(features %>% select(where(is.numeric))))) %>%
  pivot_longer(-label, names_to = "feature", values_to = "value") %>%
  group_by(feature, label) %>%
  summarise(m = mean(value, na.rm = TRUE), s = sd(value, na.rm = TRUE),
            .groups = "drop") %>%
  pivot_wider(names_from = label, values_from = c(m, s),
              names_glue = "{.value}_{label}") %>%
  mutate(
    pooled_sd = sqrt((s_0^2 + s_1^2) / 2),
    smd       = abs(m_1 - m_0) / (pooled_sd + 1e-9)
  ) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  arrange(desc(smd))

ggplot(smd_df, aes(x = reorder(feature, smd), y = smd, fill = family)) +
  geom_col() +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Standardised mean difference by feature and family",
       subtitle = "Higher = more discriminative between phishing and legitimate",
       x = NULL, y = "| SMD |", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")
```

SMD distribution per family with Kruskal-Wallis test:

```{r smd-by-family, fig.height=4, fig.width=6}
smd_family <- smd_df %>% filter(family != "Other")

ggplot(smd_family, aes(family, smd, fill = family)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  geom_jitter(width = 0.15, alpha = 0.5, size = 2) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "SMD distribution by feature family (after removing computed scores)",
       subtitle = "H1: Lexical SMD lowest → largest parametric gap expected at URLOnly tier",
       x = NULL, y = "| SMD |") +
  theme_minimal(base_size = 12)

# Kruskal–Wallis test: do families differ in median SMD?
kw_test <- kruskal.test(smd ~ family, data = smd_family)
cat("\nKruskal–Wallis test (SMD across families):\n")
cat("  χ² =", round(kw_test$statistic, 2),
    ", df =", kw_test$parameter,
    ", p =", formatC(kw_test$p.value, format = "e", digits = 2), "\n")

# Median SMD per family
smd_family %>%
  group_by(family) %>%
  summarise(
    n_features = n(),
    median_smd = round(median(smd), 3),
    mean_smd   = round(mean(smd), 3),
    max_smd    = round(max(smd), 3)
  ) %>%
  arrange(desc(median_smd))
```

**Takeaway (H1):** The key question for H1 is whether Lexical shows a wider parametric vs. non-parametric gap than Trust and Behavior tiers, which lower SMD and non-linear structure would predict.

## 4.3 H2 — Lexical + Trust Outperforms Lexical + Behavior Despite Fewer Features

H2 compares two ways of augmenting the Lexical baseline: Lexical + Trust
vs. Lexical + Behavior. If Trust features are individually more discriminative
and mutually less redundant, the smaller Trust family should yield a stronger
combined model. EDA evidence comes from per-feature SMD and intra-family
correlation.

### 4.3.1 Per-Feature SMD: Trust vs. Behavior

```{r h2-smd-comparison, fig.height=5, fig.width=9}
trust_binary <- family_map %>% filter(family == "Trust") %>% pull(feature)
behavior_all <- family_map %>% filter(family == "Behavior") %>% pull(feature)

smd_h2 <- smd_df %>%
  filter(family %in% c("Trust", "Behavior"))

ggplot(smd_h2, aes(x = reorder(feature, smd), y = smd, fill = family)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("Trust" = "steelblue",
                               "Behavior" = "firebrick")) +
  labs(title = "Per-feature SMD: Trust vs. Behavior",
       subtitle = "H2: Trust features should show higher per-feature discriminative power",
       x = NULL, y = "| SMD |", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")

smd_h2 %>%
  group_by(family) %>%
  summarise(
    n_features  = n(),
    median_smd  = round(median(smd), 3),
    mean_smd    = round(mean(smd), 3),
    total_signal = round(sum(smd), 3),
    .groups = "drop"
  ) %>%
  print()
```

**Takeaway (H2):** Trust features have higher per-feature SMD than Behavior
features. Despite being fewer in number, their `total_signal` (sum of SMD)
indicates how much raw discriminative power each family brings. Combined with
the redundancy analysis below, this predicts that Lexical + Trust will
outperform Lexical + Behavior in AUC.

### 4.3.2 Intra-Family Correlation Matrices

H2 relies on a redundancy argument: if Trust features are internally
independent while Behavior features are highly correlated, then 7 Trust
features carry more *effective* signal than their count suggests, while 20
Behavior features carry less. We show correlation matrices for all three
families to make this comparison transparent.

#### Lexical family

```{r cor-lexical, fig.height=6, fig.width=7}
lexical_all <- family_map %>% filter(family == "Lexical") %>% pull(feature)

lexical_cor <- features %>%
  select(all_of(lexical_all)) %>%
  cor(use = "pairwise.complete.obs")

corrplot(
  lexical_cor,
  method  = "color",
  order   = "hclust",
  tl.cex  = 0.7,
  tl.col  = "black",
  col     = colorRampPalette(c("firebrick", "white", "steelblue"))(200),
  title   = "Correlation — Lexical features",
  mar     = c(0, 0, 1.5, 0),
  addCoef.col = "black",
  number.cex  = 0.55
)
```

#### Trust family

```{r cor-trust, fig.height=5, fig.width=6}
trust_cor <- features %>%
  select(all_of(trust_binary)) %>%
  cor(use = "pairwise.complete.obs")

corrplot(
  trust_cor,
  method  = "color",
  tl.cex  = 0.8,
  tl.col  = "black",
  col     = colorRampPalette(c("firebrick", "white", "steelblue"))(200),
  title   = "Correlation — Trust features",
  mar     = c(0, 0, 1.5, 0),
  addCoef.col = "black",
  number.cex  = 0.7
)
```

#### Behavior family

```{r cor-behavior, fig.height=7, fig.width=8}
behavior_cor <- features %>%
  select(all_of(behavior_all)) %>%
  cor(use = "pairwise.complete.obs")

corrplot(
  behavior_cor,
  method  = "color",
  order   = "hclust",
  tl.cex  = 0.7,
  tl.col  = "black",
  col     = colorRampPalette(c("firebrick", "white", "steelblue"))(200),
  title   = "Correlation — Behavior features",
  mar     = c(0, 0, 1.5, 0),
  addCoef.col = "black",
  number.cex  = 0.55
)
```

#### Cross-family redundancy summary

```{r family-redundancy-summary}
# Helper: mean and max absolute correlation from upper triangle
cor_summary <- function(cor_mat) {
  vals <- cor_mat[upper.tri(cor_mat)] %>% abs()
  tibble(
    n_features = nrow(cor_mat),
    n_pairs    = length(vals),
    mean_abs_r = round(mean(vals), 3),
    median_abs_r = round(median(vals), 3),
    max_abs_r  = round(max(vals), 3),
    pct_above_05 = round(mean(vals > 0.5) * 100, 1)
  )
}

bind_rows(
  cor_summary(lexical_cor)  %>% mutate(family = "Lexical"),
  cor_summary(trust_cor)    %>% mutate(family = "Trust"),
  cor_summary(behavior_cor) %>% mutate(family = "Behavior")
) %>%
  select(family, everything()) %>%
  print()
```

**Takeaway (H2):** Trust features show the lowest internal correlation — each
carries independent signal. Behavior features are more numerous but also more
redundant internally, meaning their effective dimensionality is lower than the
raw feature count suggests. This supports H2: Lexical + Trust should
outperform Lexical + Behavior, because Trust adds more independent
discriminative signal per feature than Behavior does.

### 4.3.3 Correlation Structure — Full Feature Set

```{r correlation, fig.height=8, fig.width=9}
num_features <- features %>% select(where(is.numeric))
cor_mat      <- cor(num_features, use = "pairwise.complete.obs")

corrplot(
  cor_mat,
  method  = "color",
  order   = "hclust",
  addrect = 4,
  tl.cex  = 0.55,
  tl.col  = "black",
  col     = colorRampPalette(c("firebrick", "white", "steelblue"))(200),
  title   = "Pairwise correlation — all predictors (clustered)",
  mar     = c(0, 0, 1.5, 0)
)
```

```{r high-corr-pairs}
cor_long <- cor_mat %>%
  as.data.frame() %>%
  rownames_to_column("feat_a") %>%
  pivot_longer(-feat_a, names_to = "feat_b", values_to = "r") %>%
  filter(feat_a < feat_b, abs(r) > 0.7) %>%
  arrange(desc(abs(r)))

cat("Feature pairs with |r| > 0.7 after cleanup:\n")
print(cor_long, n = Inf)
```

### 4.3.4 Variance Inflation Factors

```{r vif, fig.height=6, fig.width=7}
library(car)

glm_vif <- glm(label ~ .,
               data   = df %>% select(label, all_of(continuous_features)),
               family = binomial)
vif_values <- vif(glm_vif)

vif_df <- tibble(
  feature = names(vif_values),
  vif     = as.numeric(vif_values)
) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  arrange(desc(vif))

ggplot(vif_df, aes(x = reorder(feature, vif), y = vif, fill = family)) +
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = 5,  linetype = "dashed", colour = "orange") +
  geom_hline(yintercept = 10, linetype = "dashed", colour = "red") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Variance Inflation Factors — continuous features",
       subtitle = "Lexical VIF reduced after removing ratios; Trust binary features not shown (binary → excluded from VIF)",
       x = NULL, y = "VIF", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")

cat("Features with VIF > 10 (severe):\n")
vif_df %>% filter(vif > 10) %>% select(feature, vif, family) %>% print()
cat("\nFeatures with VIF 5–10 (moderate):\n")
vif_df %>% filter(vif >= 5, vif <= 10) %>% select(feature, vif, family) %>% print()
```

**Takeaway (H2):** Trust binary features are excluded from VIF (binary
variables are not collinear in the classical sense). Remaining Lexical VIF
is lower after removing ratio features. Any severe Lexical VIF will be
handled by iterative VIF filtering in Scenario 2.


## 4.4 Skewness Assessment *(Scenario 2 preparation)*

```{r outlier-skewness, fig.height=6, fig.width=8}
# Skewness for all continuous features
skewness_fn <- function(x) {
  x <- x[!is.na(x)]
  n <- length(x)
  m <- mean(x)
  s <- sd(x)
  (n / ((n-1)*(n-2))) * sum(((x - m) / s)^3)
}

skew_df <- tibble(
  feature  = continuous_features,
  skewness = map_dbl(continuous_features, ~ skewness_fn(features[[.x]]))
) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  mutate(abs_skew = abs(skewness)) %>%
  arrange(desc(abs_skew))

ggplot(skew_df, aes(x = reorder(feature, abs_skew), y = skewness, fill = family)) +
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = c(-2, 2), linetype = "dashed", colour = "grey40") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Skewness of continuous features",
       subtitle = "Dashed lines at ±2 mark heavy skew; LR/LDA may struggle here",
       x = NULL, y = "Skewness", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")

# Count features with |skewness| > 2
n_heavy_skew <- sum(skew_df$abs_skew > 2)
cat("\nFeatures with |skewness| > 2:", n_heavy_skew, "of",
    length(continuous_features), "\n")
cat("These are:\n")
skew_df %>% filter(abs_skew > 2) %>% select(feature, skewness, family) %>% print()
```

**Takeaway (Scenario 2):** Several count-based features are heavily right-skewed with
extreme outliers. This has direct implications for model selection:

- **Parametric models** (LR, LDA): will require feature scaling and log-transforms
  of skewed features to satisfy distributional assumptions.
- **Non-parametric models** (trees, KNN, SVM): are robust to skewness but KNN and SVM
  still require standardisation for distance-based computations.



---

# 5. EDA Summary & Modelling Implications

## 5.1 Hypothesis support from EDA

**Note on dataset separability:** After removing computed scores and redundant
features (Section 3.3), 40 raw features remain. Initial modelling shows that
even this reduced set yields AUC ≈ 0.99 for all model types — the dataset is
intrinsically near-perfectly separable. This is a finding in itself (addressed
in Scenario 5 via PCA), not a modelling failure. The hypotheses below are
therefore tested primarily on the **URLOnly tier** (13 Lexical features) where
meaningful parametric vs. non-parametric differences emerge.

| Hypothesis | EDA evidence | Testable in modelling? |
|-----------|-------------|----------------------|
| **H1** — URLOnly useful, large parametric gap | Lexical features have the lowest median SMD after removing computed scores. Low SMD + non-linear structure → parametric models expected to underperform RF/SVM on URL string alone. | Yes — compare AUC gap across tiers in Scenario 2 |
| **H2** — Lexical + Trust > Lexical + Behavior | Trust features have higher per-feature SMD and lower intra-family correlation than Behavior. Fewer but more independent signals should yield a stronger complement to Lexical. | Yes — compare AUC of Lexical+Trust vs. Lexical+Behavior in Scenario 2 |

## 5.2 Implications for modelling scenarios

| Finding | Scenario 2 implication | Scenario 5 implication |
|---------|------------------------|------------------------|
| Near-perfect separability on Full tier | Focus comparison on URLOnly tier; Full tier used to satisfy 30+ feature requirement | PCA will show few components needed → confirms redundancy |
| Lexical SMD lowest, non-linear structure | Parametric/non-parametric gap largest at URLOnly tier | PC1–2 likely dominated by Behavior/Trust, not Lexical |
| Heavy skewness in count features | Log-transform before scaling; affects parametric models more | Skewed features compress onto few PCA dimensions |
| Trust features: high SMD, low internal correlation | Lexical+Trust expected to outperform Lexical+Behavior | Trust features contribute distinct PCA components |
| Behavior features: numerous but redundant | Lexical+Behavior adds volume but less independent signal | Many Behavior PCs needed to explain variance, but low discriminative power per PC |

