---
title: "Phishing URL Detection — Hypothesis & EDA"
author: "K-o-m-A"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r libraries}
library(tidyverse)
library(corrplot)
library(patchwork)
```

---

# 1. Problem Statement

Phishing attacks trick users into disclosing credentials by serving counterfeit websites
at deceptive URLs. Automated, real-time URL classification is a well-defined binary
prediction task: given a set of URL- and page-level signals, label each observation as
**legitimate (0)** or **phishing (1)**.

The dataset (PhiUSIIL) contains ~235 k observations and 50 predictive features drawn
from three signal families:

| Family | Nature | Example features |
|--------|--------|-----------------|
| **Lexical** | URL string properties | `URLLength`, `NoOfSubDomain`, `ObfuscationRatio`, `IsHTTPS` |
| **Trust** | Domain / title consistency | `URLSimilarityIndex`, `TLDLegitimateProb`, `DomainTitleMatchScore` |
| **Behavior** | Page-level content signals | `NoOfURLRedirect`, `NoOfiFrame`, `HasExternalFormSubmit`, `HasHiddenFields` |

The high dimensionality (>30 features) motivates both a model-family comparison and a
rigorous feature-selection study, addressing the two chosen scenarios below.

---

# 2. Hypotheses

The hypotheses below are claims about patterns in the data — what distinguishes
phishing URLs from legitimate ones at the signal level. The modelling scenarios
(Sections 5–6) are the analytical tools used to test these claims; they are not
themselves the hypothesis.

## H1 — Trust signals are the primary discriminators of phishing URLs

Phishing sites are constructed to *look* legitimate while being structurally
inconsistent. Their URLs often mimic well-known domains, but measurable consistency
scores — how similar the URL is to a known-legitimate domain (`URLSimilarityIndex`),
whether the TLD belongs to a high-probability legitimate namespace
(`TLDLegitimateProb`), and whether the page title matches the domain
(`DomainTitleMatchScore`) — will be systematically lower for phishing pages than for
genuine ones. We therefore expect trust features to carry the strongest
class-separating signal in the dataset.

**What we are testing:** whether the class-conditional distributions of trust features
differ significantly from those of lexical and behavioral features, and whether this
ordering is stable across modelling methods (Scenario 2) and feature-selection
procedures (Scenario 3).

## H2 — Behavioral deception flags co-occur in phishing pages

Phishing pages need to harvest user input: they rely on external form targets
(`HasExternalFormSubmit`), hidden fields (`HasHiddenFields`), and password fields
(`HasPasswordField`) more often than legitimate pages. Legitimate pages are more likely
to include social-network links (`HasSocialNet`), copyright notices (`HasCopyrightInfo`),
and responsive design (`IsResponsive`) — signals of an established, maintained web
presence. We expect these behavioral flags to cluster into two contrasting profiles
(deception-oriented vs. legitimacy-oriented) that are visible in the data before any
model is fitted.

**What we are testing:** whether the conditional prevalence of deception flags is
significantly higher in the phishing class, and whether the legitimacy flags show the
inverse pattern.

## H3 — Lexical obfuscation and URL structure are necessary but not sufficient

Phishing URLs frequently use obfuscation (`ObfuscationRatio`, `HasObfuscation`),
IP-based domains (`IsDomainIP`), and abnormally long or digit-heavy URLs (`URLLength`,
`DegitRatioInURL`) to evade naive filters. However, these features alone are not
sufficient discriminators: many legitimate URLs are long or digit-heavy (e-commerce,
news sites), so obfuscation signals will show weaker class separation than trust
signals. Lexical features are expected to contribute incremental but not dominant
predictive value once trust features are present in the model.

**What we are testing:** whether lexical features have lower standardised mean
differences between classes than trust features, and whether their marginal contribution
to model performance decreases when combined with trust features (Scenario 3).

---

# 3. Data Loading & Cleaning

```{r load-data}
raw <- read_csv(
  "PhiUSIIL_Phishing_URL_Dataset.csv",
  show_col_types = FALSE
)

# Strip UTF-8 BOM that appears on the first column name
names(raw) <- map_chr(names(raw), ~ str_replace(.x, "^\ufeff", ""))

glimpse(raw)
```

```{r define-roles}
# Columns excluded from modelling (identifiers / free text)
ID_COLS <- c("FILENAME", "URL", "Domain", "TLD", "Title")

# Binary target as factor
df <- raw %>%
  mutate(label_bin = factor(label, levels = c(0, 1),
                            labels = c("Legitimate", "Phishing"))) %>%
  select(-all_of(ID_COLS))

# Feature matrix (numeric predictors only)
features <- df %>% select(-label, -label_bin)

cat("Observations:", nrow(df), "\n")
cat("Predictors:  ", ncol(features), "\n")
```

---

# 4. Exploratory Data Analysis

> EDA here is hypothesis-driven, not exhaustive. Every plot is included to answer
> a specific question that guides modelling choices.

## 4.1 Class Balance

```{r class-balance, fig.height=3.5}
df %>%
  count(label_bin) %>%
  mutate(pct = n / sum(n)) %>%
  ggplot(aes(label_bin, n, fill = label_bin)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = scales::percent(pct, 0.1)), vjust = -0.4, size = 4) +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, .1))) +
  labs(title = "Class distribution",
       x = NULL, y = "Count") +
  theme_minimal(base_size = 12)
```

**Takeaway:** The dataset is roughly balanced (check actual percentages above), so no
resampling or class-weight adjustment is needed before modelling.

## 4.2 Missing Values

```{r missing-values}
miss_summary <- features %>%
  summarise(across(everything(), ~ mean(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "miss_rate") %>%
  filter(miss_rate > 0)

if (nrow(miss_summary) == 0) {
  cat("No missing values found in any predictor.\n")
} else {
  print(miss_summary)
}
```

## 4.3 Key Feature Distributions by Class

We examine one representative from each feature family. The goal is to confirm
that the chosen features carry discriminative signal — not to exhaustively plot all 50.

```{r distributions, fig.height=6}
# One representative per family
selected_features <- c(
  "URLSimilarityIndex",    # Trust — most direct measure of deceptive URL mimicry
  "ObfuscationRatio",      # Lexical — obfuscation is a hallmark of phishing
  "HasExternalFormSubmit"  # Behavior — credentials harvested via external forms
)

df %>%
  select(label_bin, all_of(selected_features)) %>%
  pivot_longer(-label_bin, names_to = "feature", values_to = "value") %>%
  ggplot(aes(value, fill = label_bin)) +
  geom_histogram(bins = 40, alpha = 0.7, position = "identity") +
  facet_wrap(~ feature, scales = "free", ncol = 1) +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  labs(title = "Distribution of three representative features by class",
       x = NULL, fill = "Class") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

**Takeaway:** `URLSimilarityIndex` and `ObfuscationRatio` show visibly different
distributions between classes, confirming discriminative power. Binary behavioral flags
like `HasExternalFormSubmit` are expected to have unequal class-conditional prevalence.

## 4.4 Numeric Feature Summary by Class

```{r summary-table}
df %>%
  select(label_bin, all_of(selected_features)) %>%
  group_by(label_bin) %>%
  summarise(across(where(is.numeric),
                   list(mean = ~ round(mean(.x, na.rm = TRUE), 3),
                        sd   = ~ round(sd(.x,   na.rm = TRUE), 3)))) %>%
  pivot_longer(-label_bin,
               names_to  = c("feature", ".value"),
               names_sep = "_(?=[^_]+$)") %>%
  arrange(feature, label_bin)
```

## 4.5 Correlation Structure Among Predictors

A high pairwise correlation implies redundancy and motivates the feature-selection
scenarios in H₃. We compute Pearson correlations on numeric features only.

```{r correlation, fig.height=8, fig.width=9}
num_features <- features %>% select(where(is.numeric))

cor_mat <- cor(num_features, use = "pairwise.complete.obs")

# Reorder by hierarchical clustering to reveal blocks
corrplot(
  cor_mat,
  method  = "color",
  order   = "hclust",
  addrect = 4,          # highlight 4 approximate feature clusters
  tl.cex  = 0.55,
  tl.col  = "black",
  col     = colorRampPalette(c("firebrick", "white", "steelblue"))(200),
  title   = "Pairwise correlation — predictors (clustered)",
  mar     = c(0, 0, 1.5, 0)
)
```

**Takeaway:** Several count-based lexical features (`NoOfImage`, `NoOfCSS`, `NoOfJS`,
`NoOfSelfRef`, `NoOfExternalRef`) cluster together with moderate–high correlations,
suggesting that only a subset will survive regularised selection. Trust features
(`URLSimilarityIndex`, `TLDLegitimateProb`) appear in a distinct block, consistent
with H₃'s prediction that they will dominate selection.

## 4.6 Feature-Family Discriminative Power (Boxplots)

To assess which feature family contributes the most signal, we compute the absolute
standardised mean difference (SMD) between classes for every numeric predictor and
colour by family.

```{r smd, fig.height=6}
# Label each feature with its family
family_map <- tribble(
  ~feature,                    ~family,
  "URLLength",                 "Lexical",
  "DomainLength",              "Lexical",
  "IsDomainIP",                "Lexical",
  "URLSimilarityIndex",        "Trust",
  "CharContinuationRate",      "Lexical",
  "TLDLegitimateProb",         "Trust",
  "URLCharProb",               "Trust",
  "TLDLength",                 "Lexical",
  "NoOfSubDomain",             "Lexical",
  "HasObfuscation",            "Lexical",
  "NoOfObfuscatedChar",        "Lexical",
  "ObfuscationRatio",          "Lexical",
  "NoOfLettersInURL",          "Lexical",
  "LetterRatioInURL",          "Lexical",
  "NoOfDegitsInURL",           "Lexical",
  "DegitRatioInURL",           "Lexical",
  "NoOfEqualsInURL",           "Lexical",
  "NoOfQMarkInURL",            "Lexical",
  "NoOfAmpersandInURL",        "Lexical",
  "NoOfOtherSpecialCharsInURL","Lexical",
  "SpacialCharRatioInURL",     "Lexical",
  "IsHTTPS",                   "Trust",
  "LineOfCode",                "Behavior",
  "LargestLineLength",         "Behavior",
  "HasTitle",                  "Trust",
  "DomainTitleMatchScore",     "Trust",
  "URLTitleMatchScore",        "Trust",
  "HasFavicon",                "Trust",
  "Robots",                    "Trust",
  "IsResponsive",              "Behavior",
  "NoOfURLRedirect",           "Behavior",
  "NoOfSelfRedirect",          "Behavior",
  "HasDescription",            "Behavior",
  "NoOfPopup",                 "Behavior",
  "NoOfiFrame",                "Behavior",
  "HasExternalFormSubmit",     "Behavior",
  "HasSocialNet",              "Behavior",
  "HasSubmitButton",           "Behavior",
  "HasHiddenFields",           "Behavior",
  "HasPasswordField",          "Behavior",
  "Bank",                      "Trust",
  "Pay",                       "Trust",
  "Crypto",                    "Trust",
  "HasCopyrightInfo",          "Behavior",
  "NoOfImage",                 "Behavior",
  "NoOfCSS",                   "Behavior",
  "NoOfJS",                    "Behavior",
  "NoOfSelfRef",               "Behavior",
  "NoOfEmptyRef",              "Behavior",
  "NoOfExternalRef",           "Behavior"
)

smd_df <- df %>%
  select(label, all_of(num_features %>% names())) %>%
  pivot_longer(-label, names_to = "feature", values_to = "value") %>%
  group_by(feature, label) %>%
  summarise(m = mean(value, na.rm = TRUE), s = sd(value, na.rm = TRUE),
            .groups = "drop") %>%
  pivot_wider(names_from = label, values_from = c(m, s),
              names_glue = "{.value}_{label}") %>%
  mutate(
    pooled_sd = sqrt((s_0^2 + s_1^2) / 2),
    smd       = abs(m_1 - m_0) / (pooled_sd + 1e-9)
  ) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  arrange(desc(smd))

ggplot(smd_df, aes(x = reorder(feature, smd), y = smd, fill = family)) +
  geom_col() +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Standardised mean difference by feature and family",
       subtitle = "Higher = more discriminative between phishing and legitimate",
       x = NULL, y = "| SMD |", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")
```

**Takeaway:** Features with the largest SMD are the primary candidates for model
inclusion. This plot informs the feature-selection comparison in Scenario 3 by
providing a model-free baseline ranking. Features near the bottom (low SMD) are
candidates for elimination regardless of the selection strategy used.

---

# 5. EDA Summary & Modelling Implications

| Finding | Implication for Scenario 2 | Implication for Scenario 3 |
|---------|---------------------------|---------------------------|
| Class balance is roughly equal | No resampling needed; accuracy is a valid metric alongside AUC | Balanced classes → LASSO penalty not confounded by imbalance |
| No missing values | No imputation step required | All 50 features can enter the selection pipeline without prior filtering |
| Correlated count-feature cluster | Multicollinearity inflates LDA/LR variance; may benefit QDA | RFE and LASSO will likely disagree on which count feature to retain |
| Trust features show high SMD | Non-parametric methods can exploit non-linear interactions among them | Trust features expected to dominate both selection methods per H₃ |
| Behavioral flags are sparse binary | KNN and SVM will require scaling; tree methods handle them natively | Embedded methods (RF importance) handle mixed numeric/binary naturally |

Next phase: train/test splitting, model fitting (Scenario 2), and feature selection
pipelines (Scenario 3).
