---
title: "Phishing URL Detection — Hypothesis & EDA"
author: "K-o-m-A"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---
```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(corrplot)
library(scales)
```

---

# 1. Problem Statement

## 1.1 Real-world scenario

A user clicks a link in an email, chat, or search result. The
corporate proxy has to decide **before the page loads** whether the link is
phishing. The cheapest signal at that moment is the URL string itself — no
DNS lookup, no HTML fetch, no rendering. If a URL-only check is good enough,
we can block the link for free. If it isn't, we have to fall back to a
slower page-level check.

The question we want to answer is simple:
**how much phishing signal is already in the URL?** If a URL-only model works well, we can run it as a fast
first-line filter. Where it fails, the slower trust- and behavior-level checks have to take over.

Explored scenarios for hypothesis driven modeling: 3

## 1.2 Dataset

The PhiUSIIL dataset has ~235 k rows and 50 predictor features. We split
them into three families by how expensive they are to compute:

| Family | What it captures | Example features |
|--------|------------------|-----------------|
| **Lexical** | Properties of the URL string | `URLLength`, `NoOfSubDomain`, `NoOfObfuscatedChar`, `CharContinuationRate` |
| **Trust** | Domain / title consistency | `IsHTTPS`, `HasTitle`, `HasFavicon`, `Bank`, `Pay`, `Crypto` |
| **Behavior** | Page-level content signals | `NoOfURLRedirect`, `NoOfiFrame`, `HasExternalFormSubmit`, `HasHiddenFields` |

---

# 2. Hypotheses

## H1 (Scenario 2) — The gap between parametric and non-parametric model families is tier-dependent

Phishing URLs usually leave lexical traces: odd subdomain depth, paths full
of digits, obfuscated characters. None of this needs the page to be
downloaded. We expect a model trained only on Lexical features to still reach a reasonable AUC, but the **size of the
gap** between parametric and non-parametric model families should change
with the family.

We compare three parametric methods (LR, LDA, Naive Bayes)
against three non-parametric ones (RF, SVM-RBF, KNN).

| # | Criterion | Threshold |
|---|-----------|-----------|
| C1 | Mean CV AUC gap between the best non-parametric and best parametric model on the **Lexical (URLOnly) tier** | **ΔAUC ≥ 0.02** |
| C2 | Paired Wilcoxon signed-rank test on the 5 per-fold AUCs of those two models | **p < 0.05** |
| C3 | Gradient across tiers: the Lexical gap is **strictly larger** than the FullLite-tier gap . | gap(Lexical) > gap(FullLite) |

H1 is **supported** only if C1 **and** C2 hold. C3 turns the claim from
"the two families differ on one tier" into the stronger statement *"the
difference between families shrinks as the feature tier gets stronger."*
H1 is **rejected** if either C1 or C2 fails on Lexical.

Note:
The FullLite tier is the scenario_2 replacement for a naive "Full" tier: six Behavior
near-leakers (`LineOfCode`, `NoOfJS`, `NoOfCSS`, `NoOfImage`,
`NoOfSelfRef`, `NoOfExternalRef`, each with univariate AUC > 0.95) are
stripped because they push Full to AUC ≈ 1.0 and hide every contrast H1
is trying to measure.

## H1 (Scenario 3) — A smaller URL-only filter can preserve operating quality

Scenario 3 keeps the classifier fixed as logistic regression and compares
feature-selection strategies instead of model families: one algorithmic method
(bidirectional `stepAIC`) and two embedded methods (`lasso`, `elastic-net`).

The primary pool is now **Lexical URL-only**, because that is the deployable
first-line filter the proxy can evaluate before loading the page. FullLite stays
as a fallback benchmark: if URL-only is not enough and the proxy must inspect
Trust/Behavior signals too, can we shrink the combined filter to a much smaller
set without losing AUC, sensitivity, or specificity?

The EDA motivation is the strong redundancy in the URL-length cluster
(`URLLength`, `NoOfLettersInURL`, `NoOfDegitsInURL`, and related counts)
plus the near-saturated FullLite signal. That setting should make feature
selection meaningful: several predictors explain overlapping variance, so
different selection rules may keep different active sets.

**Scenario 3 H1:** on the Lexical URL-only pool, feature selection can reduce
the deployable filter below the full 13-feature lexical set without losing
held-out operating quality.

| # | Criterion | Threshold |
|---|-----------|-----------|
| C1 | Selected-predictor count on the lexical final fit | k_selected < 13 for at least one FS method |
| C2 | Held-out operating quality on the same 20% test set | AUC, Sensitivity, and Specificity remain operationally usable relative to the full lexical GLM |

This hypothesis **cannot be answered from EDA alone**. EDA only motivates
why feature selection should matter (correlation, VIF, redundant signal);
the actual claim is about the fitted active sets and their held-out
AUC/Sensitivity/Specificity, so it is tested in `scenario_3.rmd`.

---

# 3. Data Loading & Cleaning

```{r load-data}
raw <- read_csv(
  "PhiUSIIL_Phishing_URL_Dataset.csv",
  show_col_types = FALSE
)

# Strip UTF-8 BOM that appears on the first column name
names(raw) <- map_chr(names(raw), ~ str_replace(.x, "^\ufeff", ""))

# glimpse(raw)
```

```{r define-roles}
# Columns excluded from modeling (identifiers / free text)
ID_COLS <- c("FILENAME", "URL", "Domain", "TLD", "Title")

# Binary target as factor
df <- raw %>%
  mutate(label_bin = factor(label, levels = c(0, 1),
                            labels = c("Legitimate", "Phishing"))) %>%
  select(-all_of(ID_COLS))

# Feature matrix (numeric predictors only)
features <- df %>% select(-label, -label_bin)

cat("Observations:", nrow(df), "\n")
cat("Predictors:  ", ncol(features), "\n")
```


## Feature-family mapping
The three families (Lexical / Trust / Behavior) are our own taxonomy, not
a property of the dataset. They reflect the cost of obtaining each signal
at decision time:
   Lexical  — derivable from the raw URL string alone (zero network cost).
   Trust    — requires light page/domain inspection or an external lookup
              (DNS, TLD database, HTTPS handshake, title parsing).
   Behavior — requires fetching, parsing and partially rendering the page.

Note on `IsHTTPS`: although the "https://" scheme is part of the URL string,
we place it in the **Trust** family. Modern browsers refuse plain HTTP for
any sensitive interaction and warn the user prominently, so the scheme is
already enforced by the trust layer of the browser rather than acting as a
free lexical clue. Treating it as Lexical would let the URLOnly tier cheat
off a binary that is, in practice, a deployment/trust signal.

```{r family-map}

family_map <- tribble(
  ~feature,                    ~family,
  "URLLength",                 "Lexical",
  "DomainLength",              "Lexical",
  "IsDomainIP",                "Lexical",
  "URLSimilarityIndex",        "Trust",
  "CharContinuationRate",      "Lexical",
  "TLDLegitimateProb",         "Trust",
  "URLCharProb",               "Trust",
  "TLDLength",                 "Lexical",
  "NoOfSubDomain",             "Lexical",
  "HasObfuscation",            "Lexical",
  "NoOfObfuscatedChar",        "Lexical",
  "ObfuscationRatio",          "Lexical",
  "NoOfLettersInURL",          "Lexical",
  "LetterRatioInURL",          "Lexical",
  "NoOfDegitsInURL",           "Lexical",
  "DegitRatioInURL",           "Lexical",
  "NoOfEqualsInURL",           "Lexical",
  "NoOfQMarkInURL",            "Lexical",
  "NoOfAmpersandInURL",        "Lexical",
  "NoOfOtherSpecialCharsInURL","Lexical",
  "SpacialCharRatioInURL",     "Lexical",
  "IsHTTPS",                   "Trust",
  "LineOfCode",                "Behavior",
  "LargestLineLength",         "Behavior",
  "HasTitle",                  "Trust",
  "DomainTitleMatchScore",     "Trust",
  "URLTitleMatchScore",        "Trust",
  "HasFavicon",                "Trust",
  "Robots",                    "Trust",
  "IsResponsive",              "Behavior",
  "NoOfURLRedirect",           "Behavior",
  "NoOfSelfRedirect",          "Behavior",
  "HasDescription",            "Behavior",
  "NoOfPopup",                 "Behavior",
  "NoOfiFrame",                "Behavior",
  "HasExternalFormSubmit",     "Behavior",
  "HasSocialNet",              "Behavior",
  "HasSubmitButton",           "Behavior",
  "HasHiddenFields",           "Behavior",
  "HasPasswordField",          "Behavior",
  "Bank",                      "Trust",
  "Pay",                       "Trust",
  "Crypto",                    "Trust",
  "HasCopyrightInfo",          "Behavior",
  "NoOfImage",                 "Behavior",
  "NoOfCSS",                   "Behavior",
  "NoOfJS",                    "Behavior",
  "NoOfSelfRef",               "Behavior",
  "NoOfEmptyRef",              "Behavior",
  "NoOfExternalRef",           "Behavior"
)
```

## 3.3 Feature Exclusion

Before the EDA we drop three groups of features that would bias the
analysis or trivialise the modeling. All three decisions are based on
what the feature actually means.

### Category 1 — Computed probability / similarity scores

These are not raw URL measurements. They are outputs of other models or
expert-written scoring functions that already encode phishing-detection
logic. Keeping them would mean we are classifying another classifier's
output, not the URL itself.

```{r exclude-scores}
SCORES <- c(
  "URLSimilarityIndex",   # similarity to known-phishing URL patterns
  "TLDLegitimateProb",    # P(TLD is legitimate) from an external lookup
  "URLCharProb",          # P(character sequence) from a language model
  "DomainTitleMatchScore",# computed domain↔title consistency score
  "URLTitleMatchScore"    # computed URL↔title consistency score
)
cat("Computed scores to remove:", paste(SCORES, collapse = ", "), "\n")
```

### Category 2 — Redundant binary indicator

`HasObfuscation` is literally `(NoOfObfuscatedChar > 0)`. It holds no
information that the count feature doesn't already carry.

```{r exclude-binary}
REDUNDANT_BINARY <- "HasObfuscation"
cat("Redundant binary to remove:", REDUNDANT_BINARY, "\n")
```

### Category 3 — Redundant ratio features

Each ratio below equals `count / URLLength`, and both the count and
`URLLength` are already in the dataset. Keeping the ratio gives us a
perfect linear dependence — and no extra signal.

```{r exclude-ratios}
RATIOS <- c(
  "LetterRatioInURL",      # = NoOfLettersInURL   / URLLength
  "DegitRatioInURL",       # = NoOfDegitsInURL    / URLLength
  "ObfuscationRatio",      # = NoOfObfuscatedChar / URLLength
  "SpacialCharRatioInURL"  # = NoOfOtherSpecialCharsInURL / URLLength
)
cat("Linearly dependent ratios to remove:", paste(RATIOS, collapse = ", "), "\n")
```

```{r apply-exclusions}
EXCLUDE_COLS <- c(SCORES, REDUNDANT_BINARY, RATIOS)

df       <- df %>% select(-all_of(EXCLUDE_COLS))
features <- df %>% select(-label, -label_bin)

cat("\nFeatures after exclusion:", ncol(features), "\n")
cat("Removed:", paste(EXCLUDE_COLS, collapse = ", "), "\n")
```

Update the family map to reflect the reduced feature set:

```{r family-map-update}
family_map <- family_map %>%
  filter(!feature %in% EXCLUDE_COLS)

family_map %>% count(family) %>% print()
```

```{r feature-types, include=FALSE}
# Type split used by downstream VIF/skewness chunks
binary_features <- names(features)[map_lgl(features, ~ all(na.omit(.x) %in% c(0, 1)))]
continuous_features <- setdiff(names(features), binary_features)
```

## 3.4 Feature Tiers Used in Modeling

To test H1 we define three nested feature tiers. The modeling notebook
reads these tier definitions directly from here, so both notebooks stay
in sync. The `tier_full` built here (40 features) is the EDA baseline;
the scenario_2 modeling notebook replaces it with **FullLite**
(34 features) by stripping six Behavior near-leakers — see §5 for
the reasoning.

```{r feature-tiers}
tier_lexical  <- family_map %>% filter(family == "Lexical")  %>% pull(feature)
tier_trust    <- family_map %>% filter(family == "Trust")    %>% pull(feature)
tier_behavior <- family_map %>% filter(family == "Behavior") %>% pull(feature)
tier_full     <- c(tier_lexical, tier_trust, tier_behavior)

cat("URLOnly tier (Lexical, ",  length(tier_lexical),  " features)\n", sep = "")
cat("Trust tier              (",  length(tier_trust),    " features)\n", sep = "")
cat("Behavior tier           (",  length(tier_behavior), " features)\n", sep = "")
cat("Full tier               (",  length(tier_full),     " features)\n", sep = "")
```

---

# 4. Exploratory Data Analysis


## 4.1 Data Overview

### 4.1.1 Class Balance

```{r class-balance, fig.height=3.5}
df %>%
  count(label_bin) %>%
  mutate(pct = n / sum(n)) %>%
  ggplot(aes(label_bin, n, fill = label_bin)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = scales::percent(pct, 0.1)), vjust = -0.4, size = 4) +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, .1))) +
  labs(title = "Class distribution",
       x = NULL, y = "Count") +
  theme_minimal(base_size = 12)
```

**Takeaway:** The dataset is roughly balanced, so we don't need resampling
or class-weighting. Accuracy works as a secondary metric next to AUC.

### 4.1.2 Missing Values

No missing values are present in any predictor (verified with `colSums(is.na(features))`).

## 4.2 H1 preparation

H1 claims that a URL-only detector (Lexical features) will show a bigger
parametric vs. non-parametric gap than any other tier. We don't test
non-linearity directly in the EDA. Instead we show that **per-feature
discriminative power is lowest in the Lexical family**.

### 4.2.1 Discriminative Power by Family

We measure discriminative power separately for continuous and binary
features. The standardised mean difference (SMD) only makes sense for
continuous features. For binary features the "pooled SD" collapses to a
function of the class proportion, so SMD stops being comparable across
features. For those we use **Cramér's V** instead — the standard
effect-size measure for a 2×2 contingency table, on the same 0–1 scale
as |correlation|.

```{r discriminative-power, fig.height=7, fig.width=8}
# --- Continuous features: standardised mean difference ----------------------
smd_df <- df %>%
  select(label, all_of(continuous_features)) %>%
  pivot_longer(-label, names_to = "feature", values_to = "value") %>%
  group_by(feature, label) %>%
  summarise(m = mean(value, na.rm = TRUE), s = sd(value, na.rm = TRUE),
            .groups = "drop") %>%
  pivot_wider(names_from = label, values_from = c(m, s),
              names_glue = "{.value}_{label}") %>%
  mutate(
    pooled_sd = sqrt((s_0^2 + s_1^2) / 2),
    effect    = abs(m_1 - m_0) / (pooled_sd + 1e-9),
    metric    = "SMD (continuous)"
  ) %>%
  select(feature, effect, metric)

# --- Binary features: Cramer's V on a 2x2 contingency table -----------------
cramers_v <- function(x, y) {
  tab <- table(x, y)
  if (any(dim(tab) < 2)) return(NA_real_)
  chi2 <- suppressWarnings(chisq.test(tab, correct = FALSE)$statistic)
  n <- sum(tab)
  as.numeric(sqrt(chi2 / (n * (min(dim(tab)) - 1))))
}

cv_df <- tibble(
  feature = binary_features,
  effect  = map_dbl(binary_features, ~ cramers_v(features[[.x]], df$label)),
  metric  = "Cramer's V (binary)"
)

discrim_df <- bind_rows(smd_df, cv_df) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  arrange(desc(effect))

ggplot(discrim_df, aes(x = reorder(feature, effect), y = effect, fill = family)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ metric, scales = "free", ncol = 2) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Per-feature discriminative power",
       subtitle = "Continuous: |SMD|. Binary: Cramer's V. Both >= 0.",
       x = NULL, y = "Effect size", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")
```

Effect-size distribution per family (continuous features only):

```{r smd-by-family, fig.height=4, fig.width=6}
smd_family <- discrim_df %>%
  filter(metric == "SMD (continuous)", family != "Other")

ggplot(smd_family, aes(family, effect, fill = family)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  geom_jitter(width = 0.15, alpha = 0.5, size = 2) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "SMD distribution by family (continuous features only)",
       x = NULL, y = "| SMD |") +
  theme_minimal(base_size = 12)
```

**Takeaway (H1):** Among the continuous features, Lexical has the lowest
median |SMD| and the smallest spread. Trust and Behavior carry more
per-feature signal. This supports H1's claim that the URL-only tier has less signal in individual features and relies more on feature combinations.

## 4.3 Collinearity

A quicklook at multicollinearity tells us how much redundant dimensionality each family has to start with. If a family is already low-rank in variance, multiple features carry overlapping information; if it has many independent directions of variance, each predictor adds something new.

### 4.3.1 Lexical Correlation Matrix

```{r cor-lexical, fig.height=6, fig.width=7}
lexical_cor <- features %>%
  select(all_of(tier_lexical)) %>%
  cor(use = "pairwise.complete.obs")

  corrplot(
  lexical_cor,
  method  = "color",
  order   = "hclust",
  tl.cex  = 0.7,
  tl.col  = "black",
  col     = colorRampPalette(c("firebrick", "white", "steelblue"))(200),
  title   = "Correlation - Lexical features",
  mar     = c(0, 0, 1.5, 0),
  addCoef.col = "black",
  number.cex  = 0.55
)
```

Across the full feature matrix, 10 pairs pass `|r| > 0.7` — and all of
them sit inside this Lexical count cluster (URLLength ↔ NoOfLettersInURL
↔ NoOfDegitsInURL and friends).

### 4.3.2 Variance Inflation Factors

```{r vif}
library(knitr)

# Target-free VIF from the design matrix (avoids non-converged GLM).
X_vif <- features %>%
  select(all_of(continuous_features)) %>%
  as.data.frame()

vif_values <- vapply(seq_along(X_vif), function(i) {
  r2 <- summary(lm(X_vif[[i]] ~ ., data = X_vif[, -i, drop = FALSE]))$r.squared
  1 / (1 - r2)
}, numeric(1))
names(vif_values) <- names(X_vif)

vif_df <- tibble(
  feature = names(vif_values),
  vif     = as.numeric(vif_values)
) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  mutate(severity = case_when(
    vif > 10            ~ "severe (> 10)",
    vif >= 5 & vif <= 10 ~ "moderate (5-10)",
    TRUE                ~ "ok (< 5)"
  )) %>%
  arrange(desc(vif))

vif_df %>%
  transmute(Feature = feature,
            Family  = family,
            VIF     = round(vif, 2),
            Severity = severity) %>%
  kable(caption = "Variance Inflation Factors - continuous features (lm-based, target-free).")
```

**Takeaway:** `URLLength`, `NoOfLettersInURL` and `NoOfDegitsInURL`
stay near-collinear because they measure overlapping properties of the
same URL string. Lexical has far fewer independent directions of variance
than it has features.


## 4.4 Skewness Assessment *(Scenario 2 preparation)*

```{r outlier-skewness, fig.height=6, fig.width=8}
# Skewness for all continuous features
skewness_fn <- function(x) {
  x <- x[!is.na(x)]
  n <- length(x)
  m <- mean(x)
  s <- sd(x)
  (n / ((n-1)*(n-2))) * sum(((x - m) / s)^3)
}

skew_df <- tibble(
  feature  = continuous_features,
  skewness = map_dbl(continuous_features, ~ skewness_fn(features[[.x]]))
) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  mutate(abs_skew = abs(skewness)) %>%
  arrange(desc(abs_skew))

ggplot(skew_df, aes(x = reorder(feature, abs_skew), y = skewness, fill = family)) +
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = c(-2, 2), linetype = "dashed", colour = "grey40") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Skewness of continuous features",
       subtitle = "Dashed lines at +/- 2 mark heavy skew; LR/LDA may struggle here",
       x = NULL, y = "Skewness", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")

# Count features with |skewness| > 2
n_heavy_skew <- sum(skew_df$abs_skew > 2)
cat("\nFeatures with |skewness| > 2:", n_heavy_skew, "of",
    length(continuous_features), "\n")
cat("These are:\n")
skew_df %>% filter(abs_skew > 2) %>% select(feature, skewness, family) %>% print()
```

**Takeaway (Scenario 2):** Several count features are heavily right-skewed
with long outlier tails. This shapes our modeling choices:

- **Parametric models** (LR, LDA): need scaling and a log-transform on
  skewed features to make the distributional assumption hold.
- **Non-parametric models** (trees, KNN, SVM): robust to skew, but KNN and
  SVM still need standardisation because they use distances.


---

# 5. EDA Summary & Modeling Implications

**A note on separability.** After the Section 3.3 exclusions, 40 features
remain (13 Lexical + 7 Trust + 20 Behavior). On this reduced set the task
is near-perfectly separable once Behavior is present — six Behavior
counts have univariate AUC > 0.95 on their own, so a naive Full tier
saturates at AUC ≈ 1.0 and the parametric vs. non-parametric contrast
disappears.

For H1 we therefore use the scenario_2 family layout: Lexical
(13), Trust (7), Behavior with the six near-leakers stripped (14), and
**FullLite** (34) = Lexical + Trust + reduced Behavior. H1 is tested
primarily on the **URLOnly tier**, where the family contrast has the most
room to show, and the family gradient (C3) is measured against **FullLite**,
not Full.

This near-perfect separability is a dataset-internal finding, not a production
claim. A random split of PhiUSIIL preserves the same collection process,
website templates and campaign mix on both train and test; a real deployment
would face temporal drift, new phishing kits and domain/campaign shift. Any
0.99+ FullLite result should therefore be read as an upper-bound benchmark for
this curated dataset. A production-grade claim would require a temporal or
external holdout.

## 5.1 Implications for modeling scenarios

| Finding | Scenario 2 implication (H1) | Scenario 3 implication (H1) |
|---------|----------------------------|----------------------------|
| Lexical SMD lowest + features must be combined | Largest parametric/non-parametric AUC gap expected on URLOnly | Use Lexical as the primary first-line FS pool |
| Heavy skewness in 19 of 22 continuous features | Log-transform + standardisation for LR/LDA; trees/RF unaffected | Use the same transformed/scaled matrix for lexical FS and FullLite fallback |
| URL-length cluster: VIF > 1000 and abs(r) > 0.84 | Use stable solvers / regularised LR for fairness | Redundant URL predictors make lasso/EN/stepwise active sets worth comparing |
| Trust binary features: low intra-correlation | No special treatment | Trust/Behavior features belong to the fallback FullLite benchmark |

EDA therefore supports the **motivation** for Scenario 3, but not its
result. The selected-predictor counts and the AUC/Sensitivity/Specificity
checks belong in `scenario_3.rmd`, because they depend on the fitted
feature-selection paths rather than on marginal EDA summaries.

