---
title: "Phishing URL Detection — Hypothesis & EDA"
author: "K-o-m-A"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(corrplot)
library(patchwork)
library(scales)
```

---

# 1. Problem Statement

Phishing attacks trick users into disclosing credentials by serving counterfeit websites
at deceptive URLs. Automated, real-time URL classification is a well-defined binary
prediction task: given a set of URL- and page-level signals, label each observation as
**legitimate (0)** or **phishing (1)**.

The dataset (PhiUSIIL) contains ~235 k observations and 50 predictive features drawn
from three signal families:

| Family | Nature | Example features |
|--------|--------|-----------------|
| **Lexical** | URL string properties | `URLLength`, `NoOfSubDomain`, `ObfuscationRatio`, `IsHTTPS` |
| **Trust** | Domain / title consistency | `URLSimilarityIndex`, `TLDLegitimateProb`, `DomainTitleMatchScore` |
| **Behavior** | Page-level content signals | `NoOfURLRedirect`, `NoOfiFrame`, `HasExternalFormSubmit`, `HasHiddenFields` |

The high dimensionality (50 features, well above the ≥30 threshold required by the
assignment) motivates both a model-family comparison and a rigorous feature-selection
study, addressing the two chosen scenarios below.

---

# 2. Hypotheses

The hypotheses below are claims about patterns in the data — what distinguishes
phishing URLs from legitimate ones at the signal level. The modelling scenarios
(Sections 5–6) are the analytical tools used to test these claims; they are not
themselves the hypothesis.

## H1 — Trust signals are the primary discriminators of phishing URLs

Phishing sites are constructed to *look* legitimate while being structurally
inconsistent. Trust-related features capture this tension: phishing URLs often achieve
a high `URLSimilarityIndex` because they deliberately mimic well-known domains, yet
other trust signals — `TLDLegitimateProb`, `DomainTitleMatchScore`, `HasFavicon` — tend
to be weaker for phishing pages because perfect imitation across all dimensions is
difficult to maintain. We therefore expect trust features to exhibit the largest
absolute class-conditional differences (regardless of direction) and to carry the
strongest class-separating signal in the dataset as a family.

**What we are testing:** whether trust features show systematically higher standardised
mean differences between classes than lexical and behavioral features, and whether this
ordering is stable across modelling methods (Scenario 2) and feature-selection
procedures (Scenario 3).

## H2 — Behavioral flags form two contrasting profiles

Phishing pages need to harvest user input: they rely on external form targets
(`HasExternalFormSubmit`), hidden fields (`HasHiddenFields`), and password fields
(`HasPasswordField`) more often than legitimate pages. Conversely, legitimate pages are
more likely to include social-network links (`HasSocialNet`), copyright notices
(`HasCopyrightInfo`), and responsive design (`IsResponsive`) — signals of an
established, maintained web presence. We expect these behavioral flags to form two
contrasting prevalence profiles (deception-oriented vs. legitimacy-oriented) that are
visible in the data before any model is fitted. Notably, the dominant signal may come
from the *absence* of legitimacy features in phishing pages rather than the presence of
deception features.

**What we are testing:** whether the conditional prevalence of deception flags is
higher in the phishing class, whether legitimacy flags show the inverse pattern, and
which direction carries the stronger discriminative signal.

## H3 — Lexical obfuscation and URL structure are necessary but not sufficient

Phishing URLs frequently use obfuscation (`ObfuscationRatio`, `HasObfuscation`),
IP-based domains (`IsDomainIP`), and abnormally long or digit-heavy URLs (`URLLength`,
`DegitRatioInURL`) to evade naive filters. However, these features alone are not
sufficient discriminators: many legitimate URLs are long or digit-heavy (e-commerce,
news sites), so obfuscation signals will show weaker class separation than trust
signals. Lexical features are expected to contribute incremental but not dominant
predictive value once trust features are present in the model.

**What we are testing:** whether lexical features have lower standardised mean
differences between classes than trust features, and whether their marginal contribution
to model performance decreases when combined with trust features (Scenario 3).

---

# 3. Data Loading & Cleaning

```{r load-data}
raw <- read_csv(
  "PhiUSIIL_Phishing_URL_Dataset.csv",
  show_col_types = FALSE
)

# Strip UTF-8 BOM that appears on the first column name
names(raw) <- map_chr(names(raw), ~ str_replace(.x, "^\ufeff", ""))

glimpse(raw)
```

```{r define-roles}
# Columns excluded from modelling (identifiers / free text)
ID_COLS <- c("FILENAME", "URL", "Domain", "TLD", "Title")

# Binary target as factor
df <- raw %>%
  mutate(label_bin = factor(label, levels = c(0, 1),
                            labels = c("Legitimate", "Phishing"))) %>%
  select(-all_of(ID_COLS))

# Feature matrix (numeric predictors only)
features <- df %>% select(-label, -label_bin)

cat("Observations:", nrow(df), "\n")
cat("Predictors:  ", ncol(features), "\n")
```

```{r family-map}
# ── Feature-family mapping (reused throughout EDA) ──────────────────────────
family_map <- tribble(
  ~feature,                    ~family,
  "URLLength",                 "Lexical",
  "DomainLength",              "Lexical",
  "IsDomainIP",                "Lexical",
  "URLSimilarityIndex",        "Trust",
  "CharContinuationRate",      "Lexical",
  "TLDLegitimateProb",         "Trust",
  "URLCharProb",               "Trust",
  "TLDLength",                 "Lexical",
  "NoOfSubDomain",             "Lexical",
  "HasObfuscation",            "Lexical",
  "NoOfObfuscatedChar",        "Lexical",
  "ObfuscationRatio",          "Lexical",
  "NoOfLettersInURL",          "Lexical",
  "LetterRatioInURL",          "Lexical",
  "NoOfDegitsInURL",           "Lexical",
  "DegitRatioInURL",           "Lexical",
  "NoOfEqualsInURL",           "Lexical",
  "NoOfQMarkInURL",            "Lexical",
  "NoOfAmpersandInURL",        "Lexical",
  "NoOfOtherSpecialCharsInURL","Lexical",
  "SpacialCharRatioInURL",     "Lexical",
  "IsHTTPS",                   "Trust",
  "LineOfCode",                "Behavior",
  "LargestLineLength",         "Behavior",
  "HasTitle",                  "Trust",
  "DomainTitleMatchScore",     "Trust",
  "URLTitleMatchScore",        "Trust",
  "HasFavicon",                "Trust",
  "Robots",                    "Trust",
  "IsResponsive",              "Behavior",
  "NoOfURLRedirect",           "Behavior",
  "NoOfSelfRedirect",          "Behavior",
  "HasDescription",            "Behavior",
  "NoOfPopup",                 "Behavior",
  "NoOfiFrame",                "Behavior",
  "HasExternalFormSubmit",     "Behavior",
  "HasSocialNet",              "Behavior",
  "HasSubmitButton",           "Behavior",
  "HasHiddenFields",           "Behavior",
  "HasPasswordField",          "Behavior",
  "Bank",                      "Trust",
  "Pay",                       "Trust",
  "Crypto",                    "Trust",
  "HasCopyrightInfo",          "Behavior",
  "NoOfImage",                 "Behavior",
  "NoOfCSS",                   "Behavior",
  "NoOfJS",                    "Behavior",
  "NoOfSelfRef",               "Behavior",
  "NoOfEmptyRef",              "Behavior",
  "NoOfExternalRef",           "Behavior"
)
```

---

# 4. Exploratory Data Analysis

> EDA here is **hypothesis-driven**, not exhaustive. Every plot and test is included to
> answer a specific question that guides modelling choices. Section references (H1, H2,
> H3) link each analysis back to the hypothesis it supports.

## 4.1 Class Balance

```{r class-balance, fig.height=3.5}
df %>%
  count(label_bin) %>%
  mutate(pct = n / sum(n)) %>%
  ggplot(aes(label_bin, n, fill = label_bin)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = scales::percent(pct, 0.1)), vjust = -0.4, size = 4) +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, .1))) +
  labs(title = "Class distribution",
       x = NULL, y = "Count") +
  theme_minimal(base_size = 12)
```

**Takeaway:** The dataset is roughly balanced, so no resampling or class-weight
adjustment is needed before modelling. Accuracy is a valid metric alongside AUC.

## 4.2 Missing Values

```{r missing-values}
miss_summary <- features %>%
  summarise(across(everything(), ~ mean(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "miss_rate") %>%
  filter(miss_rate > 0)

if (nrow(miss_summary) == 0) {
  cat("✓ No missing values found in any predictor.\n")
  cat("  All", ncol(features), "features can enter the modelling pipeline",
      "without imputation.\n")
} else {
  print(miss_summary)
}
```

## 4.3 Feature Types Overview

Before diving into distributions, we classify features by their actual data type
(continuous vs. binary) to choose appropriate visualisation and test strategies.

```{r feature-types}
binary_features <- features %>%
  summarise(across(everything(), ~ all(.x %in% c(0, 1), na.rm = TRUE))) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "is_binary") %>%
  filter(is_binary) %>%
  pull(feature)

continuous_features <- setdiff(names(features), binary_features)

cat("Binary features (0/1):", length(binary_features), "\n")
cat(str_wrap(paste(binary_features, collapse = ", "), width = 80), "\n\n")
cat("Continuous / count features:", length(continuous_features), "\n")
cat(str_wrap(paste(continuous_features, collapse = ", "), width = 80), "\n")
```

## 4.4 Key Feature Distributions by Class *(H1, H3)*

### 4.4.1 Continuous features — density plots

Density plots (rather than histograms) allow a cleaner comparison of class-conditional
distributions. We show two features per family.

```{r density-continuous, fig.height=9, fig.width=8}
cont_selected <- c(
  # Trust
  "URLSimilarityIndex", "DomainTitleMatchScore",
  # Lexical
  "ObfuscationRatio", "DegitRatioInURL",
  # Behavior (count-based)
  "NoOfExternalRef", "NoOfJS"
)

df %>%
  select(label_bin, all_of(cont_selected)) %>%
  pivot_longer(-label_bin, names_to = "feature", values_to = "value") %>%
  mutate(feature = factor(feature, levels = cont_selected)) %>%
  # Apply log1p to count features to handle extreme skew
  group_by(feature) %>%
  mutate(value = ifelse(feature %in% c("NoOfExternalRef", "NoOfJS"),
                        log1p(value), value)) %>%
  ungroup() %>%
  mutate(feature = recode(feature,
    "NoOfExternalRef" = "NoOfExternalRef (log scale)",
    "NoOfJS" = "NoOfJS (log scale)")) %>%
  ggplot(aes(value, fill = label_bin, colour = label_bin)) +
  geom_density(alpha = 0.35, linewidth = 0.4) +
  facet_wrap(~ feature, scales = "free", ncol = 2) +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  scale_colour_manual(values = c("steelblue", "firebrick")) +
  labs(title = "Class-conditional density — continuous features (2 per family)",
       subtitle = "Count features shown on log(1+x) scale to reveal class differences",
       x = NULL, fill = "Class", colour = "Class") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

**Takeaway (H1, H3):** Trust features (`URLSimilarityIndex`, `DomainTitleMatchScore`)
show the clearest class separation. Notably, `URLSimilarityIndex` is *higher* for
phishing — phishing URLs deliberately mimic legitimate domains, pushing this score up.
Lexical ratio features (`ObfuscationRatio`, `DegitRatioInURL`) show substantial
overlap between classes — confirming H3's claim that they are informative but not
sufficient alone. Behavioral count features require log-transformation to reveal any
separation, which has implications for parametric modelling (Scenario 2).

### 4.4.2 Binary features — class-conditional prevalence *(H2)*

For binary (0/1) features, prevalence (proportion of 1s) within each class is the
natural summary. This directly tests H2: are deception flags more prevalent in phishing?

```{r binary-prevalence, fig.height=6, fig.width=8}
prevalence_df <- df %>%
  select(label_bin, all_of(binary_features)) %>%
  pivot_longer(-label_bin, names_to = "feature", values_to = "value") %>%
  group_by(label_bin, feature) %>%
  summarise(prevalence = mean(value, na.rm = TRUE), .groups = "drop") %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other"))

# Compute the difference (phishing − legitimate) for ordering
prev_diff <- prevalence_df %>%
  pivot_wider(names_from = label_bin, values_from = prevalence) %>%
  mutate(diff = Phishing - Legitimate) %>%
  arrange(diff)

prevalence_df %>%
  mutate(feature = factor(feature, levels = prev_diff$feature)) %>%
  ggplot(aes(feature, prevalence, fill = label_bin)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.65) +
  coord_flip() +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "Binary feature prevalence by class",
       subtitle = "Ordered by (Phishing − Legitimate) difference",
       x = NULL, y = "Prevalence (% of class = 1)", fill = "Class") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")
```

**Takeaway (H2):** The strongest discriminative signal among binary features comes from
the **absence of legitimacy markers** in phishing pages: `HasSocialNet` (~85% legit
vs. ~10% phishing), `HasCopyrightInfo`, `HasDescription`, and `IsResponsive` all drop
dramatically for phishing. Deception flags (`HasExternalFormSubmit`, `HasHiddenFields`,
`HasPasswordField`) are more prevalent in phishing but their absolute prevalence is
low (<10%), producing a smaller absolute difference. This refines H2: the two-profile
pattern is real, but the legitimacy-absence signal dominates over the deception-presence
signal. Surprisingly, `IsHTTPS` and `HasTitle` are more prevalent in phishing, reflecting
modern phishing tactics (free TLS certificates, deliberate title mimicry).

## 4.5 Numeric Feature Summary by Class

```{r summary-table}
cont_selected_all <- c(
  "URLSimilarityIndex", "DomainTitleMatchScore", "TLDLegitimateProb",
  "ObfuscationRatio", "URLLength", "DegitRatioInURL",
  "NoOfURLRedirect", "LineOfCode", "NoOfExternalRef"
)

df %>%
  select(label_bin, all_of(cont_selected_all)) %>%
  group_by(label_bin) %>%
  summarise(across(where(is.numeric),
                   list(mean = ~ round(mean(.x, na.rm = TRUE), 3),
                        sd   = ~ round(sd(.x,   na.rm = TRUE), 3)))) %>%
  pivot_longer(-label_bin,
               names_to  = c("feature", ".value"),
               names_sep = "_(?=[^_]+$)") %>%
  arrange(feature, label_bin)
```

## 4.6 Formal Statistical Tests *(H1, H2, H3)*

Visual impressions must be confirmed with formal tests. We use:

- **Wilcoxon rank-sum test** for continuous features (non-parametric, robust to
  non-normality and outliers)
- **Chi-squared test** for binary features (tests independence from class)

```{r statistical-tests}
# ── Wilcoxon for continuous features ─────────────────────────────────────────
wilcox_results <- map_dfr(continuous_features, function(feat) {
  x0 <- df %>% filter(label == 0) %>% pull(!!sym(feat))
  x1 <- df %>% filter(label == 1) %>% pull(!!sym(feat))
  tryCatch({
    test <- wilcox.test(x0, x1, conf.int = FALSE, exact = FALSE)
    tibble(feature = feat, test = "Wilcoxon", p_value = test$p.value)
  }, error = function(e) {
    tibble(feature = feat, test = "Wilcoxon", p_value = NA_real_)
  })
})

# ── Chi-squared for binary features ─────────────────────────────────────────
chisq_results <- map_dfr(binary_features, function(feat) {
  tbl <- table(df$label_bin, df[[feat]])
  tryCatch({
    test <- chisq.test(tbl)
    tibble(feature = feat, test = "Chi-squared", p_value = test$p.value)
  }, error = function(e) {
    tibble(feature = feat, test = "Chi-squared", p_value = NA_real_)
  })
})

test_results <- bind_rows(wilcox_results, chisq_results) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  mutate(
    significant = ifelse(p_value < 0.001, "***",
                  ifelse(p_value < 0.01,  "**",
                  ifelse(p_value < 0.05,  "*", "ns"))),
    p_display   = ifelse(p_value < 2.2e-16, "< 2.2e-16",
                         formatC(p_value, format = "e", digits = 2))
  ) %>%
  arrange(p_value)

# Show all results
test_results %>%
  select(feature, family, test, p_display, significant) %>%
  print(n = Inf)

# Count significant features per family
test_results %>%
  filter(significant != "ns") %>%
  count(family, name = "n_significant") %>%
  arrange(desc(n_significant))
```

**Takeaway (H1, H2, H3):** Virtually all features show highly significant differences
between classes (p < 0.001), confirming that the dataset carries strong discriminative
signal across all three families. This motivates the need for formal feature selection
(Scenario 3) rather than relying on simple filtering.

## 4.7 Correlation Structure Among Predictors *(H3)*

A high pairwise correlation implies redundancy and motivates the feature-selection
comparison in Scenario 3. We compute Pearson correlations on numeric features only.

```{r correlation, fig.height=8, fig.width=9}
num_features <- features %>% select(where(is.numeric))

cor_mat <- cor(num_features, use = "pairwise.complete.obs")

corrplot(
  cor_mat,
  method  = "color",
  order   = "hclust",
  addrect = 4,
  tl.cex  = 0.55,
  tl.col  = "black",
  col     = colorRampPalette(c("firebrick", "white", "steelblue"))(200),
  title   = "Pairwise correlation — predictors (clustered)",
  mar     = c(0, 0, 1.5, 0)
)
```

```{r high-corr-pairs}
# Report feature pairs with |r| > 0.7
cor_long <- cor_mat %>%
  as.data.frame() %>%
  rownames_to_column("feat_a") %>%
  pivot_longer(-feat_a, names_to = "feat_b", values_to = "r") %>%
  filter(feat_a < feat_b, abs(r) > 0.7) %>%
  arrange(desc(abs(r)))

cat("Feature pairs with |r| > 0.7:\n")
print(cor_long, n = Inf)
```

**Takeaway (H3):** Several count-based features are highly correlated (e.g.,
`NoOfExternalRef`–`NoOfJS`, `URLLength`–`NoOfLettersInURL`). This multicollinearity
inflates variance for parametric models (LR, LDA) in Scenario 2 and means that
LASSO and RFE will likely retain different subsets from these correlated groups
in Scenario 3.

## 4.8 Feature-Family Discriminative Power *(H1, H3)*

### 4.8.1 SMD per feature

The absolute standardised mean difference (SMD) between classes provides a model-free
measure of discriminative power. We colour by family to test H1 (trust features
dominate) and H3 (lexical features are secondary).

```{r smd, fig.height=7, fig.width=8}
smd_df <- df %>%
  select(label, all_of(names(num_features))) %>%
  pivot_longer(-label, names_to = "feature", values_to = "value") %>%
  group_by(feature, label) %>%
  summarise(m = mean(value, na.rm = TRUE), s = sd(value, na.rm = TRUE),
            .groups = "drop") %>%
  pivot_wider(names_from = label, values_from = c(m, s),
              names_glue = "{.value}_{label}") %>%
  mutate(
    pooled_sd = sqrt((s_0^2 + s_1^2) / 2),
    smd       = abs(m_1 - m_0) / (pooled_sd + 1e-9)
  ) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  arrange(desc(smd))

ggplot(smd_df, aes(x = reorder(feature, smd), y = smd, fill = family)) +
  geom_col() +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Standardised mean difference by feature and family",
       subtitle = "Higher = more discriminative between phishing and legitimate",
       x = NULL, y = "| SMD |", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")
```

### 4.8.2 SMD distribution per family — formal test of H1 vs H3

To formally compare discriminative power across families, we compare the distribution
of SMD values per family using boxplots and a Kruskal–Wallis test.

```{r smd-by-family, fig.height=4, fig.width=6}
smd_family <- smd_df %>% filter(family != "Other")

ggplot(smd_family, aes(family, smd, fill = family)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  geom_jitter(width = 0.15, alpha = 0.5, size = 2) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "SMD distribution by feature family",
       subtitle = "H1 predicts Trust > Lexical; H3 predicts Lexical is secondary",
       x = NULL, y = "| SMD |") +
  theme_minimal(base_size = 12)

# Kruskal–Wallis test: do families differ in median SMD?
kw_test <- kruskal.test(smd ~ family, data = smd_family)
cat("\nKruskal–Wallis test (SMD across families):\n")
cat("  χ² =", round(kw_test$statistic, 2),
    ", df =", kw_test$parameter,
    ", p =", formatC(kw_test$p.value, format = "e", digits = 2), "\n")

# Median SMD per family
smd_family %>%
  group_by(family) %>%
  summarise(
    n_features = n(),
    median_smd = round(median(smd), 3),
    mean_smd   = round(mean(smd), 3),
    max_smd    = round(max(smd), 3)
  ) %>%
  arrange(desc(median_smd))
```

**Takeaway (H1, H3):** Trust features show the highest median SMD, confirming H1 — as a
family they carry the strongest discriminative signal. Lexical features have the lowest
median SMD, confirming H3 — they are the weakest family overall. Behavior features fall
in between but with high variance: some behavioral flags (e.g., `HasSocialNet`,
`HasCopyrightInfo`) rival top trust features, while others (e.g., `NoOfPopup`) are
weak. The Kruskal–Wallis test indicates whether these family-level differences are
statistically significant.

## 4.9 Behavioral Flag Co-occurrence *(H2)*

H2 predicts that deception-oriented and legitimacy-oriented flags form two contrasting
profiles. We examine whether these flags also co-occur within phishing pages by
computing the correlation matrix among behavioral binary features, conditioned on the
phishing class.

```{r h2-cooccurrence, fig.height=5.5, fig.width=6}
# Deception-oriented flags
deception_flags <- c("HasExternalFormSubmit", "HasHiddenFields",
                     "HasPasswordField", "NoOfPopup", "NoOfiFrame")

# Legitimacy-oriented flags (HasTitle excluded — it is more prevalent in phishing)
legit_flags <- c("HasSocialNet", "HasCopyrightInfo", "IsResponsive",
                 "HasFavicon", "HasDescription")

h2_features <- c(deception_flags, legit_flags)
h2_available <- intersect(h2_features, names(features))

# Correlation among H2 features within phishing class
cor_phish <- df %>%
  filter(label == 1) %>%
  select(all_of(h2_available)) %>%
  cor(use = "pairwise.complete.obs")

corrplot(
  cor_phish,
  method  = "color",
  type    = "upper",
  addCoef.col = "black",
  number.cex  = 0.65,
  tl.cex  = 0.7,
  tl.col  = "black",
  col     = colorRampPalette(c("firebrick", "white", "steelblue"))(200),
  title   = "Correlation among behavioral flags (phishing pages only)",
  mar     = c(0, 0, 1.5, 0)
)
```

```{r h2-profile-comparison, fig.height=4.5, fig.width=7}
# Mean prevalence of each H2 flag by class
h2_prevalence <- df %>%
  select(label_bin, all_of(h2_available)) %>%
  pivot_longer(-label_bin, names_to = "feature", values_to = "value") %>%
  group_by(label_bin, feature) %>%
  summarise(prevalence = mean(value, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    profile = case_when(
      feature %in% deception_flags ~ "Deception",
      feature %in% legit_flags     ~ "Legitimacy",
      TRUE                         ~ "Other"
    )
  )

ggplot(h2_prevalence, aes(feature, prevalence, fill = label_bin)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  facet_wrap(~ profile, scales = "free_x") +
  coord_flip() +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "H2: Deception vs. legitimacy flag profiles by class",
       x = NULL, y = "Prevalence", fill = "Class") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

**Takeaway (H2):** The correlation matrix reveals that behavioral flags have **low
pairwise correlations** even within the phishing class (most r < 0.15, max ≈ 0.28 for
`HasPasswordField`–`HasHiddenFields`). This means deception flags do **not** tightly
co-occur — a phishing page typically uses one or two deception tactics, not all at once.
The prevalence comparison (right panel) confirms the two-profile pattern: deception
flags are consistently more prevalent in phishing, legitimacy flags in legitimate pages.
However, the discrimination comes from **each flag independently**, not from a
co-occurring cluster. For feature selection (Scenario 3), this low within-group
correlation means most behavioral flags carry non-redundant information and may
each survive selection.

## 4.10 Outlier & Skewness Assessment *(Scenario 2 preparation)*

Parametric methods (LR, LDA) assume approximately normal features. Heavy skewness or
extreme outliers would require transformation or argue for non-parametric alternatives.

```{r outlier-skewness, fig.height=6, fig.width=8}
# Skewness for all continuous features
skewness_fn <- function(x) {
  x <- x[!is.na(x)]
  n <- length(x)
  m <- mean(x)
  s <- sd(x)
  (n / ((n-1)*(n-2))) * sum(((x - m) / s)^3)
}

skew_df <- tibble(
  feature  = continuous_features,
  skewness = map_dbl(continuous_features, ~ skewness_fn(features[[.x]]))
) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  mutate(abs_skew = abs(skewness)) %>%
  arrange(desc(abs_skew))

ggplot(skew_df, aes(x = reorder(feature, abs_skew), y = skewness, fill = family)) +
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = c(-2, 2), linetype = "dashed", colour = "grey40") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Skewness of continuous features",
       subtitle = "Dashed lines at ±2 mark heavy skew; LR/LDA may struggle here",
       x = NULL, y = "Skewness", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")

# Count features with |skewness| > 2
n_heavy_skew <- sum(skew_df$abs_skew > 2)
cat("\nFeatures with |skewness| > 2:", n_heavy_skew, "of",
    length(continuous_features), "\n")
cat("These are:\n")
skew_df %>% filter(abs_skew > 2) %>% select(feature, skewness, family) %>% print()
```

```{r boxplots-skewed, fig.height=5, fig.width=8}
# Boxplots for the most skewed features — log(1+x) scale for readability
top_skewed <- skew_df %>% slice_head(n = 6) %>% pull(feature)

df %>%
  select(label_bin, all_of(top_skewed)) %>%
  pivot_longer(-label_bin, names_to = "feature", values_to = "value") %>%
  mutate(value = log1p(value)) %>%
  ggplot(aes(label_bin, value, fill = label_bin)) +
  geom_boxplot(outlier.alpha = 0.1, outlier.size = 0.5) +
  facet_wrap(~ feature, scales = "free_y", ncol = 3) +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  labs(title = "Boxplots of most skewed features — log(1+x) scale",
       subtitle = "Log transform compresses outliers; class differences now visible",
       x = NULL, y = "log(1 + value)") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "none")
```

**Takeaway (Scenario 2):** Several count-based features are heavily right-skewed with
extreme outliers. This has direct implications for model selection:

- **Parametric models** (LR, LDA): will require feature scaling and potentially
  log-transforms of skewed features to satisfy distributional assumptions.
- **Non-parametric models** (trees, KNN, SVM): are robust to skewness but KNN and SVM
  still require standardisation for distance-based computations.
- **QDA**: may benefit from the non-equal variance structure visible in the boxplots.

## 4.11 Effect Sizes *(H1, H2, H3)*

With ~235 k observations every feature returns p < 2.2e-16.  P-values alone cannot
distinguish strong from trivial effects at this sample size.  We therefore compute:

- **Rank-biserial correlation** for continuous features (non-parametric effect size
  aligned with the Wilcoxon test in 4.6).
- **Cramér's V** for binary features (effect size aligned with the Chi-squared test).

```{r effect-sizes, fig.height=8, fig.width=8}
# ── Rank-biserial for continuous features ────────────────────────────────────
rank_biserial <- map_dfr(continuous_features, function(feat) {
  x0 <- df %>% filter(label == 0) %>% pull(!!sym(feat))
  x1 <- df %>% filter(label == 1) %>% pull(!!sym(feat))
  w  <- wilcox.test(x0, x1, exact = FALSE)$statistic
  n0 <- length(x0); n1 <- length(x1)
  r  <- 1 - (2 * w) / (n0 * n1)
  tibble(feature = feat, effect_size = abs(r), measure = "Rank-biserial |r|")
})

# ── Cramér's V for binary features ──────────────────────────────────────────
cramers_v <- map_dfr(binary_features, function(feat) {
  tbl <- table(df$label_bin, df[[feat]])
  chi <- chisq.test(tbl)$statistic
  n   <- sum(tbl)
  k   <- min(nrow(tbl), ncol(tbl))
  v   <- sqrt(chi / (n * (k - 1)))
  tibble(feature = feat, effect_size = v, measure = "Cramér's V")
})

effect_df <- bind_rows(rank_biserial, cramers_v) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  arrange(desc(effect_size))

ggplot(effect_df, aes(x = reorder(feature, effect_size),
                      y = effect_size, fill = family)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ measure, scales = "free_x") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Effect sizes — rank-biserial (continuous) & Cramér's V (binary)",
       subtitle = "Model-free magnitude of class separation per feature",
       x = NULL, y = "Effect size", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")
```

```{r effect-size-summary}
effect_df %>%
  group_by(family) %>%
  summarise(
    n         = n(),
    median_es = round(median(effect_size), 3),
    mean_es   = round(mean(effect_size), 3),
    max_es    = round(max(effect_size), 3)
  ) %>%
  arrange(desc(median_es))
```

**Takeaway (H1, H3):** Effect sizes make the family ranking concrete.  Trust features
carry the largest median effect size, confirming H1.  Among binary features the
strongest Cramér's V values belong to Behavior legitimacy flags (`HasSocialNet`,
`HasCopyrightInfo`), consistent with H2's finding that *absence* of legitimacy markers
is the dominant signal.  Lexical features cluster at the low end, reinforcing H3.
Unlike the p-value table (4.6), effect sizes reveal real magnitude differences —
several lexical features have negligible effect despite reaching p < 2.2e-16.

## 4.12 Variance Inflation Factors *(Scenario 2, Scenario 3)*

High VIF flags features whose coefficient estimates will be unstable in LR and LDA
due to multicollinearity.  This motivates Scenario 3's feature selection and sets
expectations for which correlated features will be dropped.

```{r vif, fig.height=6, fig.width=7}
library(car)

# Fit a logistic regression on all continuous features to compute VIF
glm_vif <- glm(label ~ ., data = df %>% select(label, all_of(continuous_features)),
               family = binomial)

vif_values <- vif(glm_vif)

vif_df <- tibble(
  feature = names(vif_values),
  vif     = as.numeric(vif_values)
) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  arrange(desc(vif))

ggplot(vif_df, aes(x = reorder(feature, vif), y = vif, fill = family)) +
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = 5,  linetype = "dashed", colour = "orange") +
  geom_hline(yintercept = 10, linetype = "dashed", colour = "red") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Variance Inflation Factors — continuous features",
       subtitle = "Orange = 5 (moderate), Red = 10 (severe multicollinearity)",
       x = NULL, y = "VIF", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")

cat("Features with VIF > 10 (severe):\n")
vif_df %>% filter(vif > 10) %>% select(feature, vif, family) %>% print()
cat("\nFeatures with VIF 5–10 (moderate):\n")
vif_df %>% filter(vif >= 5, vif <= 10) %>% select(feature, vif, family) %>% print()
```

**Takeaway (Scenario 2, 3):** Lexical features dominate the high-VIF region — URL
length, letter count, and digit count are algebraically related and inflate each other's
variance.  This confirms that LR and LDA will produce unstable coefficients without
feature selection or regularisation.  In Scenario 3 we expect LASSO to aggressively
zero out redundant lexical features, while RFE (tree-based) may tolerate them better
because trees are not destabilised by collinearity.

## 4.13 Non-linearity Evidence *(Scenario 2)*

Scenario 2 compares parametric (linear decision boundary) with non-parametric methods.
If the class boundary is non-linear, non-parametric models should win.  We look for
evidence of non-linearity in two ways:

1. **2D hex-bin plots** of top feature pairs coloured by class.
2. **Local class-probability curves** (LOESS) for individual features.

```{r nonlinearity-hex, fig.height=5, fig.width=9}
# 2D hex plots — top trust feature vs top lexical feature
p1 <- df %>%
  ggplot(aes(URLSimilarityIndex, DomainTitleMatchScore, z = label)) +
  stat_summary_hex(fun = mean, bins = 40) +
  scale_fill_gradient2(low = "steelblue", mid = "grey90", high = "firebrick",
                       midpoint = 0.5, limits = c(0, 1),
                       name = "P(phishing)") +
  labs(title = "Trust × Trust",
       x = "URLSimilarityIndex", y = "DomainTitleMatchScore") +
  theme_minimal(base_size = 10)

p2 <- df %>%
  ggplot(aes(URLSimilarityIndex, log1p(NoOfExternalRef), z = label)) +
  stat_summary_hex(fun = mean, bins = 40) +
  scale_fill_gradient2(low = "steelblue", mid = "grey90", high = "firebrick",
                       midpoint = 0.5, limits = c(0, 1),
                       name = "P(phishing)") +
  labs(title = "Trust × Behavior",
       x = "URLSimilarityIndex", y = "log(1 + NoOfExternalRef)") +
  theme_minimal(base_size = 10)

p1 + p2 + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

```{r nonlinearity-loess, fig.height=4.5, fig.width=9}
# LOESS class-probability curves for 3 key features
loess_features <- c("URLSimilarityIndex", "TLDLegitimateProb", "ObfuscationRatio")

df %>%
  select(label, all_of(loess_features)) %>%
  pivot_longer(-label, names_to = "feature", values_to = "value") %>%
  ggplot(aes(value, label)) +
  geom_smooth(method = "loess", span = 0.3, colour = "firebrick",
              fill = "firebrick", alpha = 0.2) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey50") +
  facet_wrap(~ feature, scales = "free_x") +
  labs(title = "LOESS class-probability curves — key features",
       subtitle = "Departure from a straight line = non-linearity a linear model cannot capture",
       x = "Feature value", y = "P(phishing)") +
  theme_minimal(base_size = 11)
```

**Takeaway (Scenario 2):** The hex plots reveal that the phishing/legitimate boundary
is **not a simple linear cut** in 2D feature space — phishing probability varies
non-monotonically across feature combinations.  The LOESS curves confirm this:
`URLSimilarityIndex` shows a sharp threshold effect (not a smooth linear trend), and
`TLDLegitimateProb` has a plateau followed by a steep drop.  These non-linear patterns
mean parametric models (LR, LDA) will under-fit unless features are transformed or
interactions are added, while tree-based and KNN methods can capture these boundaries
natively.

## 4.14 Normality Assessment per Class *(Scenario 2 — LDA/QDA assumptions)*

LDA assumes features are normally distributed within each class with equal covariance
matrices; QDA relaxes the equal-covariance assumption but still requires within-class
normality.  We assess this with Q-Q plots for selected features — one from each family.

```{r qq-plots, fig.height=8, fig.width=9}
qq_features <- c("URLSimilarityIndex", "ObfuscationRatio",
                 "NoOfExternalRef", "LineOfCode")

qq_data <- df %>%
  select(label_bin, all_of(qq_features)) %>%
  pivot_longer(-label_bin, names_to = "feature", values_to = "value")

ggplot(qq_data, aes(sample = value, colour = label_bin)) +
  stat_qq(alpha = 0.15, size = 0.5) +
  stat_qq_line(linewidth = 0.7) +
  facet_wrap(~ feature, scales = "free", ncol = 2) +
  scale_colour_manual(values = c("steelblue", "firebrick")) +
  labs(title = "Q-Q plots — within-class normality check",
       subtitle = "Points deviating from the line indicate non-normality",
       colour = "Class") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

```{r normality-test}
# Shapiro-Wilk on a random subsample (test limited to n ≤ 5000)
set.seed(42)
subsample_idx <- sample(nrow(df), 5000)

shapiro_results <- map_dfr(qq_features, function(feat) {
  map_dfr(c("Legitimate", "Phishing"), function(cls) {
    vals <- df %>%
      slice(subsample_idx) %>%
      filter(label_bin == cls) %>%
      pull(!!sym(feat))
    sw <- shapiro.test(vals)
    tibble(feature = feat, class = cls,
           W = round(sw$statistic, 4), p_value = sw$p.value)
  })
})

shapiro_results %>%
  mutate(normal = ifelse(p_value > 0.05, "Yes", "No"),
         p_display = formatC(p_value, format = "e", digits = 2)) %>%
  select(feature, class, W, p_display, normal)
```

**Takeaway (Scenario 2):** Every tested feature strongly deviates from normality within
both classes — Q-Q plots show heavy tails and discrete spikes, and all Shapiro-Wilk
p-values are essentially zero.  This means:

- **LDA's** within-class Gaussian assumption is violated across all feature families.
  LDA may still perform reasonably as a linear classifier, but its probabilistic
  outputs and theoretical optimality guarantees do not hold.
- **QDA** is equally affected by non-normality, though its flexibility with
  class-specific covariances may partially compensate.
- This provides a concrete, data-driven motivation for including non-parametric methods
  in Scenario 2: they make no distributional assumptions at all.

## 4.15 Within-Class Covariance Comparison *(LDA vs QDA)*

LDA assumes equal covariance matrices across classes; QDA does not.  If the matrices
differ substantially, QDA has a structural advantage.  We compare per-feature within-class
variances and formally test homogeneity with Box's M test (approximated for this
dataset size).

```{r covariance-comparison, fig.height=5, fig.width=8}
# Per-feature variance ratio: Var(phishing) / Var(legitimate)
var_ratio <- map_dfr(continuous_features, function(feat) {
  v0 <- var(df %>% filter(label == 0) %>% pull(!!sym(feat)), na.rm = TRUE)
  v1 <- var(df %>% filter(label == 1) %>% pull(!!sym(feat)), na.rm = TRUE)
  tibble(feature = feat, var_legit = v0, var_phish = v1,
         ratio = v1 / (v0 + 1e-12))
}) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  mutate(log_ratio = log2(ratio)) %>%
  arrange(desc(abs(log_ratio)))

ggplot(var_ratio, aes(x = reorder(feature, abs(log_ratio)),
                      y = log_ratio, fill = family)) +
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = c(-1, 1), linetype = "dashed", colour = "grey40") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Within-class variance ratio — log₂(Var_phishing / Var_legitimate)",
       subtitle = "Values beyond ±1 indicate >2× variance difference between classes",
       x = NULL, y = "log₂(variance ratio)", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")

cat("Features with variance ratio > 2× or < 0.5× (|log₂ ratio| > 1):\n")
var_ratio %>%
  filter(abs(log_ratio) > 1) %>%
  select(feature, family, var_legit, var_phish, ratio) %>%
  mutate(across(c(var_legit, var_phish), ~ round(.x, 2)),
         ratio = round(ratio, 2)) %>%
  print(n = Inf)
```

**Takeaway (LDA vs QDA):** Many features show variance ratios well beyond the 2× threshold
(|log₂ ratio| > 1), particularly among Behavior count features (`NoOfExternalRef`,
`NoOfJS`, `LineOfCode`) where phishing pages have dramatically different spread than
legitimate ones.  The covariance matrices are clearly unequal, violating LDA's
homoscedasticity assumption.  This gives QDA a principled advantage: it can fit
class-specific covariances to capture these differences.  However, QDA estimates
more parameters (one full covariance matrix per class), so with 50 features it
risks overfitting — another reason to run feature selection (Scenario 3) first.

## 4.16 Univariate Predictive Power Baseline *(Scenario 3)*

Before multivariate feature selection, we establish a **univariate baseline**: the
AUC of a single-feature logistic regression for each predictor.  This ranking serves
as a reference point — in Scenario 3 we can see which features gain importance
(due to complementary information) or lose importance (due to redundancy) when
considered jointly by LASSO and RFE.

```{r univariate-auc, fig.height=8, fig.width=8}
library(pROC)

univariate_auc <- map_dfr(names(features), function(feat) {
  tryCatch({
    fit <- glm(label ~ ., data = df %>% select(label, all_of(feat)),
               family = binomial)
    pred <- predict(fit, type = "response")
    auc_val <- as.numeric(auc(roc(df$label, pred, quiet = TRUE)))
    tibble(feature = feat, auc = auc_val)
  }, error = function(e) {
    tibble(feature = feat, auc = 0.5)
  })
}) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  arrange(desc(auc))

ggplot(univariate_auc, aes(x = reorder(feature, auc), y = auc, fill = family)) +
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey50") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Univariate AUC — single-feature logistic regression",
       subtitle = "Baseline ranking before multivariate feature selection (Scenario 3)",
       x = NULL, y = "AUC", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")
```

```{r univariate-auc-table}
# Top 10 and bottom 5 features by univariate AUC
cat("── Top 10 features by univariate AUC ──\n")
univariate_auc %>%
  slice_head(n = 10) %>%
  mutate(auc = round(auc, 4)) %>%
  select(feature, family, auc) %>%
  print()

cat("\n── Bottom 5 features by univariate AUC ──\n")
univariate_auc %>%
  slice_tail(n = 5) %>%
  mutate(auc = round(auc, 4)) %>%
  select(feature, family, auc) %>%
  print()

# Family-level summary
cat("\n── Median AUC by family ──\n")
univariate_auc %>%
  group_by(family) %>%
  summarise(n = n(), median_auc = round(median(auc), 4),
            mean_auc = round(mean(auc), 4)) %>%
  arrange(desc(median_auc))
```

**Takeaway (Scenario 3):** The univariate AUC ranking largely mirrors the SMD ranking
from section 4.8 — Trust and Behavior features dominate the top, Lexical features
cluster in the middle-to-lower range.  This baseline will be compared against
multivariate LASSO and RFE rankings in Scenario 3 to identify:

- **Features that gain importance jointly** — those with modest univariate AUC but
  high complementary value when combined with other predictors.
- **Features that lose importance jointly** — those with decent univariate AUC but
  redundant information (likely the correlated Lexical pairs flagged in 4.7 and 4.12).

---

# 5. EDA Summary & Modelling Implications

## 5.1 Hypothesis support from EDA

| Hypothesis | EDA evidence | Supported? |
|-----------|-------------|-----------|
| **H1** — Trust signals dominate | Trust features have the highest median SMD and effect size (4.8, 4.11); `URLSimilarityIndex` is the single strongest discriminator. Univariate AUC baseline (4.16) confirms the family ranking. Some Behavior features (`HasSocialNet`, `HasCopyrightInfo`) rank 2nd–3rd, so the advantage comes from Trust's consistently high floor, not universal dominance. | ✓ Supported (with nuance) |
| **H2** — Behavioral flag profiles | Two-profile pattern confirmed: deception flags are more prevalent in phishing, legitimacy flags in legitimate. However, flags do **not** co-occur (low within-class correlations, max r ≈ 0.28). The dominant signal is the *absence* of legitimacy features in phishing, not the *presence* of deception features. Cramér's V (4.11) confirms the strongest binary effects belong to legitimacy flags. `HasTitle` and `IsHTTPS` behave counter-intuitively (higher in phishing). | ✓ Partially supported |
| **H3** — Lexical features are secondary | Lexical features have the lowest median SMD and effect sizes across families, the highest VIF values (4.12), and high within-group correlations (redundancy). Univariate AUC (4.16) places most lexical features in the bottom half. Despite p < 2.2e-16 for all features, effect sizes (4.11) reveal several lexical features have negligible practical discriminative power. | ✓ Strongly supported |

## 5.2 Implications for modelling scenarios

| Finding | Implication for Scenario 2 (Parametric vs. Non-parametric) | Implication for Scenario 3 (Feature Selection) |
|---------|---------------------------|---------------------------|
| Class balance ≈ equal | No resampling needed; accuracy is valid alongside AUC | Balanced classes → LASSO penalty not confounded by imbalance |
| No missing values | No imputation step required | All 50 features can enter the selection pipeline |
| Severe VIF in Lexical features (4.12) | Multicollinearity inflates LR/LDA coefficients; unstable parametric estimates | LASSO expected to zero out redundant Lexical features; RFE (tree-based) more tolerant |
| Non-linear class boundaries (4.13) | LOESS and hex plots show threshold effects and non-monotonic boundaries — LR/LDA under-fit without transforms | Non-linearity means univariate ranking (4.16) may not predict multivariate importance |
| Non-normality within classes (4.14) | LDA/QDA Gaussian assumptions violated; probabilistic outputs unreliable | — |
| Unequal within-class covariances (4.15) | LDA's homoscedasticity assumption violated; QDA has structural advantage but risks overfitting with 50 features | Feature selection before QDA reduces parameter count and overfitting risk |
| Trust features show highest effect sizes (4.11) | Non-parametric methods can exploit their non-linear structure | Trust features expected to dominate both selection methods (H1) |
| Heavy skewness in count features | LR/LDA need feature scaling/transforms; trees handle natively | Log-transformed features may change selection rankings |
| Behavioral flags are sparse binary | KNN/SVM require scaling; tree methods handle them natively | Cramér's V (4.11) confirms most carry non-redundant signal → likely survive selection independently (H2) |
| Univariate AUC baseline established (4.16) | — | Reference ranking to detect features that gain or lose importance in multivariate selection |

## 5.3 Next steps

1. **Train/test split** (stratified 80/20)
2. **Scenario 2** — Parametric (Logistic Regression, LDA, QDA) vs. Non-parametric
   (Random Forest, KNN, SVM) model comparison
3. **Scenario 3** — Algorithmic (RFE) vs. Embedded (LASSO, RF importance) feature
   selection comparison
4. **Shiny application** for interactive model exploration
5. **Executive summary** (one-pager)
