---
title: "Phishing URL Detection — Hypothesis & EDA"
author: "K-o-m-A"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(corrplot)
library(patchwork)
library(scales)
```

---

# 1. Problem Statement

Phishing attacks trick users into disclosing credentials by serving counterfeit websites
at deceptive URLs. Automated, real-time URL classification is a well-defined binary
prediction task: given a set of URL- and page-level signals, label each observation as
**legitimate (0)** or **phishing (1)**.

The dataset (PhiUSIIL) contains ~235 k observations and 50 predictive features drawn
from three signal families:

| Family | Nature | Example features |
|--------|--------|-----------------|
| **Lexical** | URL string properties | `URLLength`, `NoOfSubDomain`, `ObfuscationRatio`, `IsHTTPS` |
| **Trust** | Domain / title consistency | `URLSimilarityIndex`, `TLDLegitimateProb`, `DomainTitleMatchScore` |
| **Behavior** | Page-level content signals | `NoOfURLRedirect`, `NoOfiFrame`, `HasExternalFormSubmit`, `HasHiddenFields` |

---

# 2. Hypotheses

## H1 — Trust signals are the primary discriminators of phishing URLs

Phishing sites are constructed to *look* legitimate while being structurally
inconsistent. Trust-related features capture this tension: phishing URLs often achieve
a high `URLSimilarityIndex` because they deliberately mimic well-known domains, yet
other trust signals — `TLDLegitimateProb`, `DomainTitleMatchScore`, `HasFavicon` — tend
to be weaker for phishing pages because perfect imitation across all dimensions is
difficult to maintain. We therefore expect trust features to exhibit the largest
absolute class-conditional differences (regardless of direction) and to carry the
strongest class-separating signal in the dataset as a family.

**What we are testing:** whether trust features show systematically higher standardised
mean differences between classes than lexical and behavioral features, and whether this
ordering is stable across modelling methods (Scenario 2) and feature-selection
procedures (Scenario 3).

## H2 — Phishing pages are over-engineered to appear legitimate

Modern phishing pages are not sparse impostors — they are deliberately constructed to
pass as real sites.  We hypothesise that phishing pages show **higher prevalence across
most binary behavioral signals**, including both deception-oriented flags
(`HasExternalFormSubmit`, `HasHiddenFields`, `HasPasswordField`) and legitimacy-mimicry
signals (`HasSocialNet`, `HasCopyrightInfo`, `HasDescription`, `IsResponsive`, `IsHTTPS`).
The latter group is included precisely because phishing sites inject fake social-network
links, copyright notices, and HTTPS certificates to reduce user suspicion.

We therefore expect behavioral flags to form two groups not by *which class they lean
towards* but by *mechanism*: deception flags (low absolute prevalence, modestly higher
in phishing) vs. legitimacy-mimicry flags (high absolute prevalence, strongly higher in
phishing).  The dominant discriminative signal is predicted to come from the
legitimacy-mimicry group, where the absolute gap between classes is largest.

**What we are testing:** whether phishing pages exhibit higher prevalence for both
deception and legitimacy-mimicry flags, whether legitimacy-mimicry flags produce larger
class differences than deception flags, and whether these flags contribute independently
(low co-occurrence) or as a cluster.

## H3 — Lexical obfuscation and URL structure are necessary but not sufficient

Phishing URLs frequently use obfuscation (`ObfuscationRatio`, `HasObfuscation`),
IP-based domains (`IsDomainIP`), and abnormally long or digit-heavy URLs (`URLLength`,
`DegitRatioInURL`) to evade naive filters. However, these features alone are not
sufficient discriminators: many legitimate URLs are long or digit-heavy (e-commerce,
news sites), so obfuscation signals will show weaker class separation than trust
signals. Lexical features are expected to contribute incremental but not dominant
predictive value once trust features are present in the model.

**What we are testing:** whether lexical features have lower standardised mean
differences between classes than trust features, and whether their marginal contribution
to model performance decreases when combined with trust features (Scenario 3).

---

# 3. Data Loading & Cleaning

```{r load-data}
raw <- read_csv(
  "PhiUSIIL_Phishing_URL_Dataset.csv",
  show_col_types = FALSE
)

# Strip UTF-8 BOM that appears on the first column name
names(raw) <- map_chr(names(raw), ~ str_replace(.x, "^\ufeff", ""))

# glimpse(raw)
```

```{r define-roles}
# Columns excluded from modelling (identifiers / free text)
ID_COLS <- c("FILENAME", "URL", "Domain", "TLD", "Title")

# Binary target as factor
df <- raw %>%
  mutate(label_bin = factor(label, levels = c(0, 1),
                            labels = c("Legitimate", "Phishing"))) %>%
  select(-all_of(ID_COLS))

# Feature matrix (numeric predictors only)
features <- df %>% select(-label, -label_bin)

cat("Observations:", nrow(df), "\n")
cat("Predictors:  ", ncol(features), "\n")
```

```{r family-map}
# ── Feature-family mapping (reused throughout EDA) ──────────────────────────
family_map <- tribble(
  ~feature,                    ~family,
  "URLLength",                 "Lexical",
  "DomainLength",              "Lexical",
  "IsDomainIP",                "Lexical",
  "URLSimilarityIndex",        "Trust",
  "CharContinuationRate",      "Lexical",
  "TLDLegitimateProb",         "Trust",
  "URLCharProb",               "Trust",
  "TLDLength",                 "Lexical",
  "NoOfSubDomain",             "Lexical",
  "HasObfuscation",            "Lexical",
  "NoOfObfuscatedChar",        "Lexical",
  "ObfuscationRatio",          "Lexical",
  "NoOfLettersInURL",          "Lexical",
  "LetterRatioInURL",          "Lexical",
  "NoOfDegitsInURL",           "Lexical",
  "DegitRatioInURL",           "Lexical",
  "NoOfEqualsInURL",           "Lexical",
  "NoOfQMarkInURL",            "Lexical",
  "NoOfAmpersandInURL",        "Lexical",
  "NoOfOtherSpecialCharsInURL","Lexical",
  "SpacialCharRatioInURL",     "Lexical",
  "IsHTTPS",                   "Trust",
  "LineOfCode",                "Behavior",
  "LargestLineLength",         "Behavior",
  "HasTitle",                  "Trust",
  "DomainTitleMatchScore",     "Trust",
  "URLTitleMatchScore",        "Trust",
  "HasFavicon",                "Trust",
  "Robots",                    "Trust",
  "IsResponsive",              "Behavior",
  "NoOfURLRedirect",           "Behavior",
  "NoOfSelfRedirect",          "Behavior",
  "HasDescription",            "Behavior",
  "NoOfPopup",                 "Behavior",
  "NoOfiFrame",                "Behavior",
  "HasExternalFormSubmit",     "Behavior",
  "HasSocialNet",              "Behavior",
  "HasSubmitButton",           "Behavior",
  "HasHiddenFields",           "Behavior",
  "HasPasswordField",          "Behavior",
  "Bank",                      "Trust",
  "Pay",                       "Trust",
  "Crypto",                    "Trust",
  "HasCopyrightInfo",          "Behavior",
  "NoOfImage",                 "Behavior",
  "NoOfCSS",                   "Behavior",
  "NoOfJS",                    "Behavior",
  "NoOfSelfRef",               "Behavior",
  "NoOfEmptyRef",              "Behavior",
  "NoOfExternalRef",           "Behavior"
)
```

---

# 4. Exploratory Data Analysis


## 4.1 Data Overview

### 4.1.1 Class Balance

```{r class-balance, fig.height=3.5}
df %>%
  count(label_bin) %>%
  mutate(pct = n / sum(n)) %>%
  ggplot(aes(label_bin, n, fill = label_bin)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = scales::percent(pct, 0.1)), vjust = -0.4, size = 4) +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, .1))) +
  labs(title = "Class distribution",
       x = NULL, y = "Count") +
  theme_minimal(base_size = 12)
```

**Takeaway:** The dataset is roughly balanced, so no resampling or class-weight
adjustment is needed before modelling. Accuracy is a valid metric alongside AUC.

### 4.1.2 Missing Values

```{r missing-values}
miss_summary <- features %>%
  summarise(across(everything(), ~ mean(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "miss_rate") %>%
  filter(miss_rate > 0)

if (nrow(miss_summary) == 0) {
  cat("✓ No missing values found in any predictor.\n")
  cat("  All", ncol(features), "features can enter the modelling pipeline",
      "without imputation.\n")
} else {
  print(miss_summary)
}
```

### 4.1.3 Feature Types Overview

Before diving into distributions, we classify features by their actual data type
(continuous vs. binary) to choose appropriate visualisation and test strategies.

```{r feature-types}
binary_features <- features %>%
  summarise(across(everything(), ~ all(.x %in% c(0, 1), na.rm = TRUE))) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "is_binary") %>%
  filter(is_binary) %>%
  pull(feature)

continuous_features <- setdiff(names(features), binary_features)

cat("Binary features (0/1):", length(binary_features), "\n")
cat(str_wrap(paste(binary_features, collapse = ", "), width = 80), "\n\n")
cat("Continuous / count features:", length(continuous_features), "\n")
cat(str_wrap(paste(continuous_features, collapse = ", "), width = 80), "\n")
```


## 4.2 H1 — Trust Signals are Primary Discriminators

H1 predicts that trust-signal features — those capturing URL/domain similarity and TLD
legitimacy — carry the strongest class separation of the three feature families.  We test
this through distributions, standardised mean differences, and effect sizes,
always comparing the Trust family against Lexical and Behavior families.

### 4.2.1 Continuous Feature Distributions

Density plots allow a cleaner comparison of class-conditional distributions.
We show two features per family to place Trust signals in context alongside
Lexical and Behavior features.

```{r density-continuous, fig.height=9, fig.width=8}
cont_selected <- c(
  # Trust
  "URLSimilarityIndex", "DomainTitleMatchScore",
  # Lexical
  "ObfuscationRatio", "DegitRatioInURL",
  # Behavior (count-based)
  "NoOfExternalRef", "NoOfJS"
)

df %>%
  select(label_bin, all_of(cont_selected)) %>%
  pivot_longer(-label_bin, names_to = "feature", values_to = "value") %>%
  mutate(feature = factor(feature, levels = cont_selected)) %>%
  # Apply log1p to count features to handle extreme skew
  group_by(feature) %>%
  mutate(value = ifelse(feature %in% c("NoOfExternalRef", "NoOfJS"),
                        log1p(value), value)) %>%
  ungroup() %>%
  mutate(feature = recode(feature,
    "NoOfExternalRef" = "NoOfExternalRef (log scale)",
    "NoOfJS" = "NoOfJS (log scale)")) %>%
  ggplot(aes(value, fill = label_bin, colour = label_bin)) +
  geom_density(alpha = 0.35, linewidth = 0.4) +
  facet_wrap(~ feature, scales = "free", ncol = 2) +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  scale_colour_manual(values = c("steelblue", "firebrick")) +
  labs(title = "Class-conditional density — continuous features (2 per family)",
       subtitle = "Count features shown on log(1+x) scale to reveal class differences",
       x = NULL, fill = "Class", colour = "Class") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

**Takeaway (H1):** Trust features (`URLSimilarityIndex`, `DomainTitleMatchScore`)
show the clearest class separation — visible confirmation of H1.  Notably,
`URLSimilarityIndex` is *higher* for phishing: phishing URLs deliberately mimic
legitimate domains.  Lexical ratio features show substantial class overlap, previewing
H3.  Behavioral count features require log-transformation to reveal separation, with
implications for parametric modelling (Scenario 2).

### 4.2.2 Discriminative Power by Family

The absolute standardised mean difference (SMD) between classes provides a model-free
measure of discriminative power.  We colour by family to test H1 (trust features
dominate) and contrast with H3 (lexical features are secondary).

#### SMD per feature

```{r smd, fig.height=7, fig.width=8}
smd_df <- df %>%
  select(label, all_of(names(features %>% select(where(is.numeric))))) %>%
  pivot_longer(-label, names_to = "feature", values_to = "value") %>%
  group_by(feature, label) %>%
  summarise(m = mean(value, na.rm = TRUE), s = sd(value, na.rm = TRUE),
            .groups = "drop") %>%
  pivot_wider(names_from = label, values_from = c(m, s),
              names_glue = "{.value}_{label}") %>%
  mutate(
    pooled_sd = sqrt((s_0^2 + s_1^2) / 2),
    smd       = abs(m_1 - m_0) / (pooled_sd + 1e-9)
  ) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  arrange(desc(smd))

ggplot(smd_df, aes(x = reorder(feature, smd), y = smd, fill = family)) +
  geom_col() +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Standardised mean difference by feature and family",
       subtitle = "Higher = more discriminative between phishing and legitimate",
       x = NULL, y = "| SMD |", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")
```

#### SMD distribution per family — formal test of H1 vs H3

To formally compare discriminative power across families, we compare the distribution
of SMD values per family using boxplots and a Kruskal–Wallis test.

```{r smd-by-family, fig.height=4, fig.width=6}
smd_family <- smd_df %>% filter(family != "Other")

ggplot(smd_family, aes(family, smd, fill = family)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  geom_jitter(width = 0.15, alpha = 0.5, size = 2) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "SMD distribution by feature family",
       subtitle = "H1 predicts Trust > Lexical; H3 predicts Lexical is secondary",
       x = NULL, y = "| SMD |") +
  theme_minimal(base_size = 12)

# Kruskal–Wallis test: do families differ in median SMD?
kw_test <- kruskal.test(smd ~ family, data = smd_family)
cat("\nKruskal–Wallis test (SMD across families):\n")
cat("  χ² =", round(kw_test$statistic, 2),
    ", df =", kw_test$parameter,
    ", p =", formatC(kw_test$p.value, format = "e", digits = 2), "\n")

# Median SMD per family
smd_family %>%
  group_by(family) %>%
  summarise(
    n_features = n(),
    median_smd = round(median(smd), 3),
    mean_smd   = round(mean(smd), 3),
    max_smd    = round(max(smd), 3)
  ) %>%
  arrange(desc(median_smd))
```

**Takeaway (H1, H3):** At the **family level** (median SMD), the ranking is Trust (1.005) > Behavior (0.595) > Lexical (0.336), and the Kruskal–Wallis test confirms these differences are statistically significant (χ²=8.82, df=2, p=0.012).  In this narrow sense H1 holds.

However, the **individual feature ranking** tells a more nuanced story.  `URLSimilarityIndex` (Trust) sits at #1, but positions #2–5 are all Behavior features — `HasSocialNet`, `HasCopyrightInfo`, `HasDescription`, `HasSubmitButton` — each outperforming the second-best Trust feature (`DomainTitleMatchScore`).  Trust wins the median comparison because its features cluster tightly around 1.0, whereas Behavior has high within-family variance: a handful of legitimacy flags are very strong while many deception flags are weak, lowering the family median.

**A more precise conclusion:** Trust is the most *consistent* family, but it does not monopolise the top discriminators.  The highest-SMD individual features are a mix of Trust and Behavior.  H1 should be interpreted as "Trust features are reliably useful across the board" rather than "Trust features dominate the top of the ranking."  Lexical features remain the weakest family, confirming H3.

## 4.3 H2 — Behavioral Flags form Contrasting Profiles

H2 predicts that behavioral features fall into two contrasting profiles: deception-oriented
flags elevated in phishing pages, and legitimacy-oriented flags elevated in legitimate pages.
We test this through prevalence, formal tests, co-occurrence structure, and effect sizes.

### 4.3.1 Binary Feature Prevalence

For binary (0/1) features, prevalence (proportion of 1s) within each class is the
natural summary.  This directly tests H2: are deception flags more prevalent in phishing
and legitimacy flags more prevalent in legitimate pages?

```{r binary-prevalence, fig.height=6, fig.width=8}
prevalence_df <- df %>%
  select(label_bin, all_of(binary_features)) %>%
  pivot_longer(-label_bin, names_to = "feature", values_to = "value") %>%
  group_by(label_bin, feature) %>%
  summarise(prevalence = mean(value, na.rm = TRUE), .groups = "drop") %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other"))

# Compute the difference (phishing − legitimate) for ordering
prev_diff <- prevalence_df %>%
  pivot_wider(names_from = label_bin, values_from = prevalence) %>%
  mutate(diff = Phishing - Legitimate) %>%
  arrange(diff)

prevalence_df %>%
  mutate(feature = factor(feature, levels = prev_diff$feature)) %>%
  ggplot(aes(feature, prevalence, fill = label_bin)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.65) +
  coord_flip() +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "Binary feature prevalence by class",
       subtitle = "Ordered by (Phishing − Legitimate) difference",
       x = NULL, y = "Prevalence (% of class = 1)", fill = "Class") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")
```

**Takeaway (H2):** Confirmed.  Phishing pages show higher prevalence for nearly every binary feature — both deception-oriented and legitimacy-mimicry flags lean phishing, as H2 predicts.  The legitimacy-mimicry group drives the largest absolute gaps: `HasSocialNet` (~85% phishing vs. ~5% legitimate), `HasCopyrightInfo`, `HasDescription`, and `IsResponsive` all have far higher prevalence in phishing, consistent with the hypothesis that phishing sites actively inject these signals to appear trustworthy.  `IsHTTPS` (~100% phishing) and `HasTitle` (~100% phishing) confirm that free TLS certificates and deliberate title mimicry are now standard practice.  Deception flags (`HasExternalFormSubmit`, `HasHiddenFields`, `HasPasswordField`) also lean phishing but with low absolute prevalence, producing smaller absolute differences — matching the predicted two-mechanism structure.

### 4.3.2 Effect Sizes — Binary Features

**Cramér's V** provides an effect size aligned with the Chi-squared test, quantifying
the practical magnitude of association between each binary feature and class.

```{r cramers-v, fig.height=5, fig.width=7}
cramers_v <- map_dfr(binary_features, function(feat) {
  tbl <- table(df$label_bin, df[[feat]])
  chi <- chisq.test(tbl)$statistic
  n   <- sum(tbl)
  k   <- min(nrow(tbl), ncol(tbl))
  v   <- sqrt(chi / (n * (k - 1)))
  tibble(feature = feat, effect_size = v)
})

cv_df <- cramers_v %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  arrange(desc(effect_size))

ggplot(cv_df, aes(x = reorder(feature, effect_size), y = effect_size, fill = family)) +
  geom_col() +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Cramér's V — binary features",
       subtitle = "Model-free magnitude of association with class label",
       x = NULL, y = "Cramér's V", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")
```

```{r cramers-v-summary}
cv_df %>%
  group_by(family) %>%
  summarise(
    n         = n(),
    median_es = round(median(effect_size), 3),
    mean_es   = round(mean(effect_size), 3),
    max_es    = round(max(effect_size), 3)
  ) %>%
  arrange(desc(median_es))
```

**Takeaway (H2):** Confirmed — legitimacy-mimicry flags dominate by effect size.
`HasSocialNet` and `HasCopyrightInfo` have the highest Cramér's V among all binary
features, driven by the large prevalence gap between classes.  Deception flags
(`HasExternalFormSubmit`, `HasPasswordField`) have lower Cramér's V because their
absolute prevalence is low in both classes.  This confirms the predicted two-mechanism
structure: legitimacy-mimicry is the dominant discriminative signal, deception flags
provide weaker but independent additional signal.


## 4.4 H3 — Lexical Features are Necessary but Not Sufficient

H3 predicts that lexical and URL-structure features carry discriminative signal but are
weaker than trust features and suffer from inter-feature redundancy.  The density plots
in 4.2.1 and the SMD results in 4.2.2 already show the Trust > Lexical ranking; here we
quantify *why* lexical features are secondary through correlation structure and variance
inflation.

### 4.4.1 Correlation Structure Among Lexical Predictors

A high pairwise correlation implies redundancy and motivates the feature-selection
comparison in Scenario 3.  We compute Pearson correlations on all numeric features;
the lexical cluster is the primary region of concern.

```{r correlation, fig.height=8, fig.width=9}
num_features <- features %>% select(where(is.numeric))

cor_mat <- cor(num_features, use = "pairwise.complete.obs")

corrplot(
  cor_mat,
  method  = "color",
  order   = "hclust",
  addrect = 4,
  tl.cex  = 0.55,
  tl.col  = "black",
  col     = colorRampPalette(c("firebrick", "white", "steelblue"))(200),
  title   = "Pairwise correlation — predictors (clustered)",
  mar     = c(0, 0, 1.5, 0)
)
```

```{r high-corr-pairs}
# Report feature pairs with |r| > 0.7
cor_long <- cor_mat %>%
  as.data.frame() %>%
  rownames_to_column("feat_a") %>%
  pivot_longer(-feat_a, names_to = "feat_b", values_to = "r") %>%
  filter(feat_a < feat_b, abs(r) > 0.7) %>%
  arrange(desc(abs(r)))

cat("Feature pairs with |r| > 0.7:\n")
print(cor_long, n = Inf)
```

**Takeaway (H3):** 14 feature pairs exceed |r|=0.7.  The strongest cluster is lexical:
`NoOfLettersInURL`–`URLLength` (r=0.956), `NoOfDigitsInURL`–`URLLength` (r=0.836), and
several obfuscation pairs.  Notably, two Trust features also correlate strongly:
`DomainTitleMatchScore`–`URLTitleMatchScore` (r=0.961) — the highest pair in the dataset.
And `HasObfuscation`–`ObfuscationRatio` (r=0.799) links a binary and a continuous lexical
feature.  This multicollinearity inflates variance for parametric models (LR, LDA) and
means LASSO and RFE will likely retain different subsets from these correlated groups
in Scenario 3.

### 4.4.2 Variance Inflation Factors

High VIF flags features whose coefficient estimates will be unstable in LR and LDA
due to multicollinearity.  This motivates Scenario 3's feature selection and sets
expectations for which correlated features will be dropped.

```{r vif, fig.height=6, fig.width=7}
library(car)

# Fit a logistic regression on all continuous features to compute VIF
glm_vif <- glm(label ~ ., data = df %>% select(label, all_of(continuous_features)),
               family = binomial)

vif_values <- vif(glm_vif)

vif_df <- tibble(
  feature = names(vif_values),
  vif     = as.numeric(vif_values)
) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  arrange(desc(vif))

ggplot(vif_df, aes(x = reorder(feature, vif), y = vif, fill = family)) +
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = 5,  linetype = "dashed", colour = "orange") +
  geom_hline(yintercept = 10, linetype = "dashed", colour = "red") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Variance Inflation Factors — continuous features",
       subtitle = "Orange = 5 (moderate), Red = 10 (severe multicollinearity)",
       x = NULL, y = "VIF", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")

cat("Features with VIF > 10 (severe):\n")
vif_df %>% filter(vif > 10) %>% select(feature, vif, family) %>% print()
cat("\nFeatures with VIF 5–10 (moderate):\n")
vif_df %>% filter(vif >= 5, vif <= 10) %>% select(feature, vif, family) %>% print()
```

**Takeaway (H3, Scenario 2/3):** Lexical features dominate the most extreme VIF values —
`URLLength` (2410), `NoOfLettersInURL` (1198), and `NoOfDigitsInURL` (203) are algebraically
related and inflate each other's variance severely.  Two Trust features also have severe VIF:
`DomainTitleMatchScore` (16.9) and `URLSimilarityIndex` (15.5), consistent with their high
correlation (r=0.961) found in 4.4.1.  The moderate-VIF features (5–10) include further
lexical and one trust feature.  LR and LDA will produce unstable coefficients for all
severe-VIF features without selection or regularisation.  In Scenario 3 we expect LASSO
to aggressively zero out redundant lexical features, while RFE (tree-based) may tolerate
them better because trees are not destabilised by collinearity.


## 4.5 Model Assumption Checks *(Scenario 2 preparation)*

We assess skewness and outliers in the continuous features, as these directly affect
preprocessing decisions (scaling, log-transforms) for Scenario 2.

### 4.5.1 Outlier & Skewness Assessment

Parametric methods (LR, LDA) assume approximately normal features.  Heavy skewness or
extreme outliers would require transformation or argue for non-parametric alternatives.

```{r outlier-skewness, fig.height=6, fig.width=8}
# Skewness for all continuous features
skewness_fn <- function(x) {
  x <- x[!is.na(x)]
  n <- length(x)
  m <- mean(x)
  s <- sd(x)
  (n / ((n-1)*(n-2))) * sum(((x - m) / s)^3)
}

skew_df <- tibble(
  feature  = continuous_features,
  skewness = map_dbl(continuous_features, ~ skewness_fn(features[[.x]]))
) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  mutate(abs_skew = abs(skewness)) %>%
  arrange(desc(abs_skew))

ggplot(skew_df, aes(x = reorder(feature, abs_skew), y = skewness, fill = family)) +
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = c(-2, 2), linetype = "dashed", colour = "grey40") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Skewness of continuous features",
       subtitle = "Dashed lines at ±2 mark heavy skew; LR/LDA may struggle here",
       x = NULL, y = "Skewness", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")

# Count features with |skewness| > 2
n_heavy_skew <- sum(skew_df$abs_skew > 2)
cat("\nFeatures with |skewness| > 2:", n_heavy_skew, "of",
    length(continuous_features), "\n")
cat("These are:\n")
skew_df %>% filter(abs_skew > 2) %>% select(feature, skewness, family) %>% print()
```

**Takeaway (Scenario 2):** Several count-based features are heavily right-skewed with
extreme outliers. This has direct implications for model selection:

- **Parametric models** (LR, LDA): will require feature scaling and log-transforms
  of skewed features to satisfy distributional assumptions.
- **Non-parametric models** (trees, KNN, SVM): are robust to skewness but KNN and SVM
  still require standardisation for distance-based computations.



---

# 5. EDA Summary & Modelling Implications

## 5.1 Hypothesis support from EDA

| Hypothesis | EDA evidence | Supported? |
|-----------|-------------|-----------|
| **H1** — Trust signals dominate | Trust features have the highest median SMD (Trust 1.005 > Behavior 0.595 > Lexical 0.336); `URLSimilarityIndex` is the single strongest discriminator. Trust wins the median but the top individual features are a mix of Trust and Behavior. | ✓ Supported (with nuance) |
| **H2** — Behavioral flag profiles | Two-profile pattern confirmed: legitimacy-mimicry flags (`HasSocialNet`, `HasCopyrightInfo`) show the largest class gaps and highest Cramér's V. Deception flags lean phishing but with smaller absolute differences. `HasTitle` and `IsHTTPS` behave counter-intuitively (higher in phishing). | ✓ Partially supported |
| **H3** — Lexical features are secondary | Lexical features have the lowest median SMD, the highest VIF values, and high within-group correlations (redundancy). Several have negligible practical discriminative power despite statistical significance. | ✓ Strongly supported |

## 5.2 Implications for modelling scenarios

| Finding | Scenario 2 implication | Scenario 3 implication |
|---------|------------------------|------------------------|
| Class balance ≈ equal | Accuracy valid alongside AUC; no resampling needed | LASSO penalty not confounded by imbalance |
| Severe VIF in Lexical features | LR/LDA coefficients unstable without regularisation | LASSO expected to zero out redundant Lexical features |
| Trust features highest median SMD | Non-parametric methods can exploit non-linear structure | Trust + top Behavior flags expected to dominate selection |
| Heavy skewness in count features | LR/LDA need scaling/transforms; trees handle natively | Log-transforms may change selection rankings |
| Behavioral flags are sparse binary | KNN/SVM require scaling; trees handle natively | Most carry non-redundant signal → likely survive selection |

## 5.3 Next steps

1. **Train/test split** (stratified 80/20)
2. **Scenario 2** — Parametric (Logistic Regression, LDA, QDA) vs. Non-parametric
   (Random Forest, KNN, SVM) model comparison
3. **Scenario 3** — Algorithmic (RFE) vs. Embedded (LASSO, RF importance) feature
   selection comparison
4. **Shiny application** for interactive model exploration
5. **Executive summary** (one-pager)
