---
title: "Phishing URL Detection — Hypothesis & EDA"
author: "K-o-m-A"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(corrplot)
library(scales)
library(knitr)
```

---

# 1. Problem Statement

## 1.1 Real-world scenario

A user clicks a link. The browser or a corporate proxy has to decide
**before the page loads** whether the link is phishing. The cheapest signal
at that moment is the URL string itself — no DNS lookup, no HTML fetch, no
rendering. If a URL-only check is good enough, we block for free; if not, a
slower page-level check takes over.

The question: **how much phishing signal is already in the URL, and which
class of model extracts it best?**

## 1.2 Dataset

PhiUSIIL: ~235 k rows, 54 columns = 5 IDs + `label` + 48 numeric predictors.
After the exclusions in §3.2 we model on **40 predictors** split into three
families by cost at decision time:

| Family | What it captures | Example features |
|--------|------------------|-----------------|
| **Lexical** | Properties of the URL string (zero network cost) | `URLLength`, `NoOfSubDomain`, `NoOfObfuscatedChar`, `CharContinuationRate` |
| **Trust** | Domain / title consistency (light lookup) | `IsHTTPS`, `HasTitle`, `HasFavicon`, `Bank`, `Pay`, `Crypto` |
| **Behavior** | Page-level content signals (fetch + parse) | `NoOfURLRedirect`, `NoOfiFrame`, `HasExternalFormSubmit`, `HasHiddenFields` |

Lexical is the URL-only first-line case. Trust and Behavior require a
network fetch and act as the rich-feature upper bound.

---

# 2. Hypothesis

## H1 — On the URL-only tier, non-parametric models pick up meaningfully more signal than parametric ones

Phishing URLs usually leave lexical traces: odd subdomain depth, digit-heavy
paths, obfuscated characters. Lexical patterns interact in non-additive
ways (length × digit count × special chars), so a linear decision boundary
cannot fully use those interactions. We compare three parametric methods
(LR, LDA, NB) against three non-parametric ones (RF, SVM-RBF, KNN) on the
Lexical tier, and repeat on a richer tier as a saturation check.

**Success criteria** (tested in `scenario_2.rmd` with a 5-fold CV protocol
on shared folds):

| # | Criterion | Threshold |
|---|-----------|-----------|
| C1 | Mean CV AUC gap between the best non-parametric and best parametric model on **Lexical** | **ΔAUC ≥ 0.02** |
| C2 | Paired Wilcoxon signed-rank test on the 5 per-fold AUCs of those two models | **p < 0.05** |
| C3 | Gradient: Lexical gap **strictly larger** than rich-tier gap | gap(Lexical) > gap(FullLite) |

H1 is supported only if C1 **and** C2 hold. C3 turns the claim into *"the
non-parametric advantage shrinks as the feature tier gets stronger."*

---

# 3. Data Loading & Cleaning

```{r load-data}
raw <- read_csv("PhiUSIIL_Phishing_URL_Dataset.csv", show_col_types = FALSE)
names(raw) <- map_chr(names(raw), ~ str_replace(.x, "^\ufeff", ""))
```

```{r define-roles}
ID_COLS <- c("FILENAME", "URL", "Domain", "TLD", "Title")

df <- raw %>%
  mutate(label_bin = factor(label, levels = c(0, 1),
                            labels = c("Legitimate", "Phishing"))) %>%
  select(-all_of(ID_COLS))

features <- df %>% select(-label, -label_bin)

cat("Observations:", nrow(df), "\n")
cat("Predictors (before exclusion):", ncol(features), "\n")
```

## 3.1 Family map

Families reflect the cost of obtaining each signal at decision time.
`IsHTTPS` is placed in **Trust**, not Lexical: modern browsers refuse plain
HTTP for sensitive interactions, so the scheme is enforced at the trust
layer rather than acting as a free lexical clue.

```{r family-map}
family_map <- tribble(
  ~feature,                    ~family,
  "URLLength",                 "Lexical",
  "DomainLength",              "Lexical",
  "IsDomainIP",                "Lexical",
  "URLSimilarityIndex",        "Trust",
  "CharContinuationRate",      "Lexical",
  "TLDLegitimateProb",         "Trust",
  "URLCharProb",               "Trust",
  "TLDLength",                 "Lexical",
  "NoOfSubDomain",             "Lexical",
  "HasObfuscation",            "Lexical",
  "NoOfObfuscatedChar",        "Lexical",
  "ObfuscationRatio",          "Lexical",
  "NoOfLettersInURL",          "Lexical",
  "LetterRatioInURL",          "Lexical",
  "NoOfDegitsInURL",           "Lexical",
  "DegitRatioInURL",           "Lexical",
  "NoOfEqualsInURL",           "Lexical",
  "NoOfQMarkInURL",            "Lexical",
  "NoOfAmpersandInURL",        "Lexical",
  "NoOfOtherSpecialCharsInURL","Lexical",
  "SpacialCharRatioInURL",     "Lexical",
  "IsHTTPS",                   "Trust",
  "LineOfCode",                "Behavior",
  "LargestLineLength",         "Behavior",
  "HasTitle",                  "Trust",
  "DomainTitleMatchScore",     "Trust",
  "URLTitleMatchScore",        "Trust",
  "HasFavicon",                "Trust",
  "Robots",                    "Trust",
  "IsResponsive",              "Behavior",
  "NoOfURLRedirect",           "Behavior",
  "NoOfSelfRedirect",          "Behavior",
  "HasDescription",            "Behavior",
  "NoOfPopup",                 "Behavior",
  "NoOfiFrame",                "Behavior",
  "HasExternalFormSubmit",     "Behavior",
  "HasSocialNet",              "Behavior",
  "HasSubmitButton",           "Behavior",
  "HasHiddenFields",           "Behavior",
  "HasPasswordField",          "Behavior",
  "Bank",                      "Trust",
  "Pay",                       "Trust",
  "Crypto",                    "Trust",
  "HasCopyrightInfo",          "Behavior",
  "NoOfImage",                 "Behavior",
  "NoOfCSS",                   "Behavior",
  "NoOfJS",                    "Behavior",
  "NoOfSelfRef",               "Behavior",
  "NoOfEmptyRef",              "Behavior",
  "NoOfExternalRef",           "Behavior"
)
```

## 3.2 Feature exclusion

Three groups of features are dropped before modelling.

**(a) Computed scores** — outputs of other models or hand-crafted scoring
functions that already encode phishing logic. Keeping them means classifying
a classifier's output (e.g. `URLSimilarityIndex` alone gives AUC ≈ 1.0).

```{r exclude-scores}
SCORES <- c("URLSimilarityIndex", "TLDLegitimateProb", "URLCharProb",
            "DomainTitleMatchScore", "URLTitleMatchScore")
```

**(b) Redundant binary flag.** `HasObfuscation = (NoOfObfuscatedChar > 0)` —
a deterministic function of a count already in the dataset.

```{r exclude-binary}
REDUNDANT_BINARY <- "HasObfuscation"
```

**(c) Redundant ratios.** Each of these equals `count / URLLength` where
both numerator and denominator are already present. We keep the raw counts
(easier units to interpret) and drop the ratio.

```{r exclude-ratios}
RATIOS <- c("LetterRatioInURL", "DegitRatioInURL",
            "ObfuscationRatio", "SpacialCharRatioInURL")
```

```{r apply-exclusions}
EXCLUDE_COLS <- c(SCORES, REDUNDANT_BINARY, RATIOS)

df       <- df       %>% select(-all_of(EXCLUDE_COLS))
features <- df       %>% select(-label, -label_bin)
family_map <- family_map %>% filter(!feature %in% EXCLUDE_COLS)

cat("Features after exclusion:", ncol(features),
    "(dropped", length(EXCLUDE_COLS), "/", ncol(features) + length(EXCLUDE_COLS), ")\n")
family_map %>% count(family) %>% print()
```

```{r feature-types, include=FALSE}
binary_features     <- names(features)[map_lgl(features, ~ all(na.omit(.x) %in% c(0, 1)))]
continuous_features <- setdiff(names(features), binary_features)
```

## 3.3 Feature tiers used in modelling

```{r feature-tiers}
tier_lexical  <- family_map %>% filter(family == "Lexical")  %>% pull(feature)
tier_trust    <- family_map %>% filter(family == "Trust")    %>% pull(feature)
tier_behavior <- family_map %>% filter(family == "Behavior") %>% pull(feature)
tier_full     <- c(tier_lexical, tier_trust, tier_behavior)

cat("Lexical (URLOnly):", length(tier_lexical), "features\n")
cat("Trust:            ", length(tier_trust),   "features\n")
cat("Behavior:         ", length(tier_behavior),"features\n")
cat("Full:             ", length(tier_full),    "features\n")
```

---

# 4. Exploratory Data Analysis

Every plot in this section is tied to a specific modelling decision — we
avoid generic per-feature histograms or a global correlation matrix across
all 40 predictors, following the assignment's guidance to keep EDA
deliberate rather than exhaustive.

## 4.1 Class balance and missingness

```{r class-balance, fig.height=3.5}
df %>%
  count(label_bin) %>%
  mutate(pct = n / sum(n)) %>%
  ggplot(aes(label_bin, n, fill = label_bin)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = scales::percent(pct, 0.1)), vjust = -0.4, size = 4) +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, .1))) +
  labs(title = "Class distribution", x = NULL, y = "Count") +
  theme_minimal(base_size = 12)

cat("Total NAs in feature matrix:", sum(is.na(features)), "\n")
```

Roughly balanced (≈57/43) → AUC and Accuracy are both valid metrics, no
resampling needed. Zero missing values in any predictor. Scenario 2
additionally stratifies the modelling sub-sample to an exact 50/50, so
class weighting is not needed at either stage.

## 4.2 Per-feature discriminative power by family

If no single Lexical feature separates the classes, the model has to combine
several of them — the setting where non-parametric models tend to beat
linear/Gaussian baselines.

We compute discriminative power separately for continuous and binary
features. SMD is only meaningful for continuous features; for binary
features the pooled SD degenerates, so we use **Cramér's V** on a 2×2
contingency table (same 0–1 scale as |correlation|).

```{r discriminative-power, fig.height=7, fig.width=8}
smd_df <- df %>%
  select(label, all_of(continuous_features)) %>%
  pivot_longer(-label, names_to = "feature", values_to = "value") %>%
  group_by(feature, label) %>%
  summarise(m = mean(value, na.rm = TRUE), s = sd(value, na.rm = TRUE),
            .groups = "drop") %>%
  pivot_wider(names_from = label, values_from = c(m, s),
              names_glue = "{.value}_{label}") %>%
  mutate(pooled_sd = sqrt((s_0^2 + s_1^2) / 2),
         effect    = abs(m_1 - m_0) / (pooled_sd + 1e-9),
         metric    = "SMD (continuous)") %>%
  select(feature, effect, metric)

cramers_v <- function(x, y) {
  tab <- table(x, y)
  if (any(dim(tab) < 2)) return(NA_real_)
  chi2 <- suppressWarnings(chisq.test(tab, correct = FALSE)$statistic)
  as.numeric(sqrt(chi2 / (sum(tab) * (min(dim(tab)) - 1))))
}

cv_df <- tibble(
  feature = binary_features,
  effect  = map_dbl(binary_features, ~ cramers_v(features[[.x]], df$label)),
  metric  = "Cramer's V (binary)"
)

discrim_df <- bind_rows(smd_df, cv_df) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other"))

ggplot(discrim_df, aes(reorder(feature, effect), effect, fill = family)) +
  geom_col() + coord_flip() +
  facet_wrap(~ metric, scales = "free", ncol = 2) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Per-feature discriminative power by family",
       subtitle = "Continuous: |SMD|, Binary: Cramer's V. Both >= 0.",
       x = NULL, y = "Effect size", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")
```

**Takeaway.** Lexical features cluster at the bottom of the |SMD| ranking —
no individual URL-length or digit-count feature separates classes on its
own. Trust and Behavior carry stronger per-feature signal. Weak per-feature
evidence means a Lexical classifier must combine many small cues, which is
the setting in which tree ensembles and kernel methods typically beat
linear models — the motivation for H1.

## 4.3 Multicollinearity — preparation for modelling

The Lexical family contains a dense count cluster (`URLLength`,
`NoOfLettersInURL`, `NoOfDegitsInURL`, and the obfuscation / equals /
ampersand triangle). Plain (unpenalised) logistic regression is numerically
unstable on such a cluster, so Scenario 2 uses **ridge-regularised LR** on
every tier. We document the cluster here via a correlation plot and
target-free VIF so the modelling decision is traceable to data.

```{r cor-lexical, fig.height=6, fig.width=7}
lexical_cor <- features %>% select(all_of(tier_lexical)) %>%
  cor(use = "pairwise.complete.obs")

corrplot(lexical_cor, method = "color", order = "hclust",
         tl.cex = 0.7, tl.col = "black",
         col = colorRampPalette(c("firebrick", "white", "steelblue"))(200),
         title = "Correlation — Lexical features",
         mar = c(0, 0, 1.5, 0),
         addCoef.col = "black", number.cex = 0.55)
```

Target-free VIF — regress each continuous feature on the rest of the design
matrix. This avoids the non-converged binomial GLM that `car::vif` produces
on this dataset and is computed directly from `lm`:

```{r vif}
X_vif <- features %>% select(all_of(continuous_features)) %>% as.data.frame()

vif_values <- vapply(seq_along(X_vif), function(i) {
  r2 <- summary(lm(X_vif[[i]] ~ ., data = X_vif[, -i, drop = FALSE]))$r.squared
  1 / (1 - r2)
}, numeric(1))
names(vif_values) <- names(X_vif)

tibble(feature = names(vif_values), vif = as.numeric(vif_values)) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  mutate(severity = case_when(vif > 10 ~ "severe (> 10)",
                              vif >= 5 ~ "moderate (5-10)",
                              TRUE     ~ "ok (< 5)")) %>%
  arrange(desc(vif)) %>%
  transmute(Feature = feature, Family = family,
            VIF = round(vif, 2), Severity = severity) %>%
  kable(caption = "Variance Inflation Factors (lm-based, target-free).")
```

**Takeaway.** `URLLength`, `NoOfLettersInURL` and `NoOfDegitsInURL` remain
near-collinear after the ratio-feature drop because they measure
overlapping properties of the same string. This is what motivates
ridge-regularised LR in Scenario 2; unpenalised LR would swing its
coefficients on this cluster from fold to fold.

## 4.4 Skewness

```{r skewness, fig.height=6, fig.width=8}
skewness_fn <- function(x) {
  x <- x[!is.na(x)]; n <- length(x); m <- mean(x); s <- sd(x)
  (n / ((n-1)*(n-2))) * sum(((x - m) / s)^3)
}

skew_df <- tibble(
  feature  = continuous_features,
  skewness = map_dbl(continuous_features, ~ skewness_fn(features[[.x]]))
) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  mutate(abs_skew = abs(skewness))

ggplot(skew_df, aes(reorder(feature, abs_skew), skewness, fill = family)) +
  geom_col() + coord_flip() +
  geom_hline(yintercept = c(-2, 2), linetype = "dashed", colour = "grey40") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Skewness of continuous features",
       subtitle = "Dashed lines at +/- 2 mark heavy skew",
       x = NULL, y = "Skewness", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")

cat("Features with |skewness| > 2:", sum(skew_df$abs_skew > 2),
    "of", length(continuous_features), "\n")
```

**Takeaway.** Most continuous features are heavily right-skewed (count data
with long tails). Parametric LR/LDA/NB and distance-based KNN/SVM apply
`log1p` + centre/scale in Scenario 2; Random Forest is left untransformed.

---

# 5. EDA summary

After exclusions: **40 features** = 13 Lexical + 7 Trust + 20 Behavior.

| Finding | Modelling implication |
|---------|----------------------|
| Roughly balanced (57/43), zero NAs | AUC + Accuracy both valid; Scenario 2 stratifies sub-sample to exact 50/50 |
| Per-feature SMD lowest on Lexical | Largest parametric/non-parametric gap expected on Lexical — the H1 test |
| Heavy skew on 19/22 continuous features | `log1p` + centre/scale for LR/LDA/NB/SVM/KNN; trees unaffected |
| URL-length cluster with VIF > 500 | Ridge-regularised LR in Scenario 2 |
| Full-tier saturates at AUC ≈ 0.99 | Compare models on Lexical (where the gap has room to show) and on a >30-feature "lite" tier that drops the near-leaky page counts |

Scenario 2 reuses the same exclusions and tier definitions, so the findings
above apply directly to the modelling results there.
