---
title: "Phishing URL Detection — Hypothesis & EDA"
author: "K-o-m-A"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---
```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(corrplot)
library(patchwork)
library(scales)
```

---

# 1. Problem Statement

## 1.1 Real-world scenario

A user clicks a link in an email, chat, or search result. The browser, a mail
gateway, or a corporate proxy needs to decide **before the page loads** whether
the destination is a phishing site. The cheapest signal available at that
moment is the URL string itself — no DNS lookup, no HTML fetch, no rendering.
If this string-only decision is good enough, the system can block or warn the
user without paying any network cost; if it is not, the system has to fall
back to slower page-level analysis.

The question this project tries to answer is therefore concrete:
**how much of the phishing signal is already in the URL, and what is the
minimal feature footprint needed to detect it reliably?** Demonstrating that a
URL-only model is competitive justifies a fast first-line filter; showing
where it breaks justifies the (more expensive) trust- and behavior-level
checks that follow.

This question maps directly onto the two assignment scenarios we picked:

- **Scenario 2** — parametric vs. non-parametric models on the URL-only tier
  tells us whether linear models are sufficient or whether the URL string
  hides non-linear structure that only RF/SVM/GBM can exploit.
- **Scenario 3** — algorithmic vs. embedded feature selection on the full
  feature matrix tells us how few predictors we actually need, and which ones
  survive penalisation under known multicollinearity.

## 1.2 Dataset

The dataset (PhiUSIIL) contains ~235 k observations and 50 predictive features drawn
from three signal families:

| Family | Nature | Example features |
|--------|--------|-----------------|
| **Lexical** | URL string properties | `URLLength`, `NoOfSubDomain`, `NoOfObfuscatedChar`, `CharContinuationRate` |
| **Trust** | Domain / title consistency | `IsHTTPS`, `HasTitle`, `HasFavicon`, `Bank`, `Pay`, `Crypto` |
| **Behavior** | Page-level content signals | `NoOfURLRedirect`, `NoOfiFrame`, `HasExternalFormSubmit`, `HasHiddenFields` |

The Lexical family corresponds to the URL-only first-line scenario above; Trust
and Behavior require progressively more network/page information and serve as
the strong-baseline comparison.

---

# 2. Hypotheses

## H1 (Scenario 2) — On the URL-string-only tier, non-parametric models extract significantly more signal than parametric models

Phishing URLs leave lexical traces in the URL string itself — unusual subdomain
depth, digit-heavy paths, obfuscated characters — that do not require downloading
the target page. We expect a model trained exclusively on Lexical features (URL
string, no DNS lookup, no HTML parse) to achieve meaningful AUC, but with a
larger parametric vs. non-parametric gap than any other feature tier. This gap
arises because lexical patterns interact non-additively (combinations of
character counts, lengths and special chars) and linear decision boundaries
cannot fully exploit them. We test this with three parametric methods (LR,
LDA, Naive Bayes) and three non-parametric methods (RF, SVM-RBF, KNN).

**Quantitative success criteria.** To avoid a vague "significantly more" claim,
H1 is operationalised against the 5-fold CV protocol used in `scenario_2.rmd`:

| # | Criterion | Threshold |
|---|-----------|-----------|
| C1 | Mean CV AUC gap between the best non-parametric and the best parametric model on the **Lexical (URLOnly) tier** | **ΔAUC ≥ 0.02** |
| C2 | Paired Wilcoxon signed-rank test on the 5 per-fold AUCs of those two models | **p < 0.05** |
| C3 | Gradient check across tiers: the gap on Lexical is **strictly larger** than the gap on the Full tier (where AUC ≈ 0.99 saturates and the gap should collapse to ≤ 0.005). | gap(Lexical) > gap(Full) |

H1 is **supported** only if C1 *and* C2 hold; C3 strengthens the claim from a
single-tier statement into the gradient *"the non-parametric advantage shrinks
as the feature tier becomes stronger."* H1 is **rejected** if either C1 or C2
fails on the Lexical tier.

## H2 (Scenario 3) — Embedded feature selection retains substantially fewer features than algorithmic selection by collapsing collinear clusters

The full feature matrix contains severe multicollinearity (URL-length cluster
with VIF > 1000, see Section 4.3) and several pairs with |r| > 0.7. We
hypothesise that **embedded methods** (LASSO, ElasticNet, RandomForest
permutation importance) will retain fewer features than **algorithmic methods**
(forward/backward stepwise, RFE) on the same task, because embedded penalisation
collapses each collinear cluster to one or few survivors while algorithmic
search treats correlated features as independent candidates. We will report
(a) how many features each method ultimately retains, (b) the overlap between
the retained sets, and (c) which features lose statistical significance as
collinear partners are added or removed.

---

# 3. Data Loading & Cleaning

```{r load-data}
raw <- read_csv(
  "PhiUSIIL_Phishing_URL_Dataset.csv",
  show_col_types = FALSE
)

# Strip UTF-8 BOM that appears on the first column name
names(raw) <- map_chr(names(raw), ~ str_replace(.x, "^\ufeff", ""))

# glimpse(raw)
```

```{r define-roles}
# Columns excluded from modelling (identifiers / free text)
ID_COLS <- c("FILENAME", "URL", "Domain", "TLD", "Title")

# Binary target as factor
df <- raw %>%
  mutate(label_bin = factor(label, levels = c(0, 1),
                            labels = c("Legitimate", "Phishing"))) %>%
  select(-all_of(ID_COLS))

# Feature matrix (numeric predictors only)
features <- df %>% select(-label, -label_bin)

cat("Observations:", nrow(df), "\n")
cat("Predictors:  ", ncol(features), "\n")
```

```{r family-map}
# ── Feature-family mapping (reused throughout EDA) ──────────────────────────
# The three families (Lexical / Trust / Behavior) are our own taxonomy, not
# a property of the dataset. They reflect the cost of obtaining each signal
# at decision time:
#   Lexical  — derivable from the raw URL string alone (zero network cost).
#   Trust    — requires light page/domain inspection or an external lookup
#              (DNS, TLD database, HTTPS handshake, title parsing).
#   Behavior — requires fetching, parsing and partially rendering the page.
#
# Note on `IsHTTPS`: although the "https://" scheme is part of the URL string,
# we place it in the **Trust** family. Modern browsers refuse plain HTTP for
# any sensitive interaction and warn the user prominently, so the scheme is
# already enforced by the trust layer of the browser rather than acting as a
# free lexical clue. Treating it as Lexical would let the URLOnly tier cheat
# off a binary that is, in practice, a deployment/trust signal.
family_map <- tribble(
  ~feature,                    ~family,
  "URLLength",                 "Lexical",
  "DomainLength",              "Lexical",
  "IsDomainIP",                "Lexical",
  "URLSimilarityIndex",        "Trust",
  "CharContinuationRate",      "Lexical",
  "TLDLegitimateProb",         "Trust",
  "URLCharProb",               "Trust",
  "TLDLength",                 "Lexical",
  "NoOfSubDomain",             "Lexical",
  "HasObfuscation",            "Lexical",
  "NoOfObfuscatedChar",        "Lexical",
  "ObfuscationRatio",          "Lexical",
  "NoOfLettersInURL",          "Lexical",
  "LetterRatioInURL",          "Lexical",
  "NoOfDegitsInURL",           "Lexical",
  "DegitRatioInURL",           "Lexical",
  "NoOfEqualsInURL",           "Lexical",
  "NoOfQMarkInURL",            "Lexical",
  "NoOfAmpersandInURL",        "Lexical",
  "NoOfOtherSpecialCharsInURL","Lexical",
  "SpacialCharRatioInURL",     "Lexical",
  "IsHTTPS",                   "Trust",
  "LineOfCode",                "Behavior",
  "LargestLineLength",         "Behavior",
  "HasTitle",                  "Trust",
  "DomainTitleMatchScore",     "Trust",
  "URLTitleMatchScore",        "Trust",
  "HasFavicon",                "Trust",
  "Robots",                    "Trust",
  "IsResponsive",              "Behavior",
  "NoOfURLRedirect",           "Behavior",
  "NoOfSelfRedirect",          "Behavior",
  "HasDescription",            "Behavior",
  "NoOfPopup",                 "Behavior",
  "NoOfiFrame",                "Behavior",
  "HasExternalFormSubmit",     "Behavior",
  "HasSocialNet",              "Behavior",
  "HasSubmitButton",           "Behavior",
  "HasHiddenFields",           "Behavior",
  "HasPasswordField",          "Behavior",
  "Bank",                      "Trust",
  "Pay",                       "Trust",
  "Crypto",                    "Trust",
  "HasCopyrightInfo",          "Behavior",
  "NoOfImage",                 "Behavior",
  "NoOfCSS",                   "Behavior",
  "NoOfJS",                    "Behavior",
  "NoOfSelfRef",               "Behavior",
  "NoOfEmptyRef",              "Behavior",
  "NoOfExternalRef",           "Behavior"
)
```

## 3.3 Feature Exclusion

Before EDA we remove three categories of features that would bias or
trivialise downstream modeling. All decisions are based on feature
semantics.

### Category 1 — Computed probability / similarity scores

These features are not raw URL measurements. They are outputs of
prior models or expert-designed scoring functions that already encode
phishing-detection logic. Including them would mean classifying the
outputs of another classifier, not the URL itself.

```{r exclude-scores}
SCORES <- c(
  "URLSimilarityIndex",   # similarity to known-phishing URL patterns
  "TLDLegitimateProb",    # P(TLD is legitimate) from an external lookup
  "URLCharProb",          # P(character sequence) from a language model
  "DomainTitleMatchScore",# computed domain↔title consistency score
  "URLTitleMatchScore"    # computed URL↔title consistency score
)
cat("Computed scores to remove:", paste(SCORES, collapse = ", "), "\n")
```

### Category 2 — Redundant binary indicator

`HasObfuscation` is defined as `(NoOfObfuscatedChar > 0)`. It carries
no information beyond the count feature already in the dataset.

```{r exclude-binary}
REDUNDANT_BINARY <- "HasObfuscation"
cat("Redundant binary to remove:", REDUNDANT_BINARY, "\n")
```

### Category 3 — Redundant ratio features

Each ratio below equals `count / URLLength`, where both the count and
`URLLength` are already present. Keeping both introduces perfect linear
dependence without adding signal.

```{r exclude-ratios}
RATIOS <- c(
  "LetterRatioInURL",      # = NoOfLettersInURL   / URLLength
  "DegitRatioInURL",       # = NoOfDegitsInURL    / URLLength
  "ObfuscationRatio",      # = NoOfObfuscatedChar / URLLength
  "SpacialCharRatioInURL"  # = NoOfOtherSpecialCharsInURL / URLLength
)
cat("Linearly dependent ratios to remove:", paste(RATIOS, collapse = ", "), "\n")
```

```{r apply-exclusions}
EXCLUDE_COLS <- c(SCORES, REDUNDANT_BINARY, RATIOS)

df       <- df %>% select(-all_of(EXCLUDE_COLS))
features <- df %>% select(-label, -label_bin)

cat("\nFeatures after exclusion:", ncol(features),
    "(removed", length(EXCLUDE_COLS), "of", ncol(features) + length(EXCLUDE_COLS), ")\n")
cat("Removed:", paste(EXCLUDE_COLS, collapse = ", "), "\n")
```

Update the family map to reflect the reduced feature set:

```{r family-map-update}
family_map <- family_map %>%
  filter(!feature %in% EXCLUDE_COLS)

family_map %>% count(family) %>% print()
```

```{r feature-types, include=FALSE}
# Silent type split used by downstream VIF/skewness chunks
binary_features <- names(features)[map_lgl(features, ~ all(na.omit(.x) %in% c(0, 1)))]
continuous_features <- setdiff(names(features), binary_features)
```

## 3.4 Feature Tiers Used in Modelling

To test H1 we define three nested feature tiers; modelling code consumes them
directly from this notebook so the tier definitions stay consistent.

```{r feature-tiers}
tier_lexical  <- family_map %>% filter(family == "Lexical")  %>% pull(feature)
tier_trust    <- family_map %>% filter(family == "Trust")    %>% pull(feature)
tier_behavior <- family_map %>% filter(family == "Behavior") %>% pull(feature)
tier_full     <- c(tier_lexical, tier_trust, tier_behavior)

cat("URLOnly tier (Lexical, ",  length(tier_lexical),  " features)\n", sep = "")
cat("Trust tier              (",  length(tier_trust),    " features)\n", sep = "")
cat("Behavior tier           (",  length(tier_behavior), " features)\n", sep = "")
cat("Full tier               (",  length(tier_full),     " features)\n", sep = "")
```

---

# 4. Exploratory Data Analysis


## 4.1 Data Overview

### 4.1.1 Class Balance

```{r class-balance, fig.height=3.5}
df %>%
  count(label_bin) %>%
  mutate(pct = n / sum(n)) %>%
  ggplot(aes(label_bin, n, fill = label_bin)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = scales::percent(pct, 0.1)), vjust = -0.4, size = 4) +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, .1))) +
  labs(title = "Class distribution",
       x = NULL, y = "Count") +
  theme_minimal(base_size = 12)
```

**Takeaway:** The dataset is roughly balanced, so no resampling or class-weight
adjustment is needed before modelling. Accuracy is a valid metric alongside AUC.

### 4.1.2 Missing Values

No missing values are present in any predictor (verified with
`colSums(is.na(features))`).


## 4.2 H1 — Lexical Features Show Class Overlap That Requires Non-Linear Models

H1 predicts that a URL-string-only detector (Lexical features) will show a
larger parametric vs. non-parametric gap than other tiers. We do not test
non-linearity directly here; instead we show that **per-feature
discriminative power on Lexical is the lowest among the three families**.
A weak per-feature signal forces a classifier to combine many features at
once.

### 4.2.1 Discriminative Power by Family

We measure per-feature discriminative power separately for the two feature
types, because the standardised mean difference (SMD) is only meaningful
for continuous variables. For binary features the "pooled SD" reduces to a
function of the marginal proportion and SMD becomes hard to compare across
features; we use **Cramer's V** instead, which is the standard effect-size
measure for a 2×2 contingency table and lies on the same 0–1 scale as a
correlation magnitude.

```{r discriminative-power, fig.height=7, fig.width=8}
# --- Continuous features: standardised mean difference ----------------------
smd_df <- df %>%
  select(label, all_of(continuous_features)) %>%
  pivot_longer(-label, names_to = "feature", values_to = "value") %>%
  group_by(feature, label) %>%
  summarise(m = mean(value, na.rm = TRUE), s = sd(value, na.rm = TRUE),
            .groups = "drop") %>%
  pivot_wider(names_from = label, values_from = c(m, s),
              names_glue = "{.value}_{label}") %>%
  mutate(
    pooled_sd = sqrt((s_0^2 + s_1^2) / 2),
    effect    = abs(m_1 - m_0) / (pooled_sd + 1e-9),
    metric    = "SMD (continuous)"
  ) %>%
  select(feature, effect, metric)

# --- Binary features: Cramer's V on a 2x2 contingency table -----------------
cramers_v <- function(x, y) {
  tab <- table(x, y)
  if (any(dim(tab) < 2)) return(NA_real_)
  chi2 <- suppressWarnings(chisq.test(tab, correct = FALSE)$statistic)
  n <- sum(tab)
  as.numeric(sqrt(chi2 / (n * (min(dim(tab)) - 1))))
}

cv_df <- tibble(
  feature = binary_features,
  effect  = map_dbl(binary_features, ~ cramers_v(features[[.x]], df$label)),
  metric  = "Cramer's V (binary)"
)

discrim_df <- bind_rows(smd_df, cv_df) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  arrange(desc(effect))

ggplot(discrim_df, aes(x = reorder(feature, effect), y = effect, fill = family)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ metric, scales = "free", ncol = 2) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Per-feature discriminative power",
       subtitle = "Continuous: |SMD|. Binary: Cramer's V. Both >= 0.",
       x = NULL, y = "Effect size", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")
```

Distribution of effect size per family (combined view, restricted to continuous
features so SMD values are directly comparable):

```{r smd-by-family, fig.height=4, fig.width=6}
smd_family <- discrim_df %>%
  filter(metric == "SMD (continuous)", family != "Other")

ggplot(smd_family, aes(family, effect, fill = family)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  geom_jitter(width = 0.15, alpha = 0.5, size = 2) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "SMD distribution by family (continuous features only)",
       subtitle = "H1: Lexical SMD lowest -> largest parametric gap expected at URLOnly tier",
       x = NULL, y = "| SMD |") +
  theme_minimal(base_size = 12)
```

**Takeaway (H1):** Among the continuous features, Lexical has the lowest
median |SMD| and the smallest spread, while Trust and Behavior carry stronger
per-feature signal. With weak per-feature evidence and visible non-linear
structure, the URLOnly tier is the setting where parametric models will
struggle most — exactly where the parametric vs. non-parametric gap predicted by H1 should appear.

## 4.3 H2 (Scenario 3) — Multicollinearity Motivates Algorithmic vs. Embedded Feature Selection

H2 hypothesises that algorithmic and embedded feature selection methods will
retain different feature sets on the Full tier. The EDA evidence is the
collinearity structure: where do correlated clusters live, and how severe is
the inflation? Embedded methods (LASSO, ElasticNet, RF importance) collapse
each cluster to one survivor; algorithmic methods (forward/backward stepwise,
RFE) treat correlated features as independent candidates and tend to retain
more.

### 4.3.1 Lexical Correlation Matrix

The Lexical family is where the collinearity that drives H2 actually lives
(Trust and Behavior families have `mean_abs_r` around 0.18 and 0.14 respectively
and contain no dense cluster). The plot below makes the URL-length count
cluster visible.

```{r cor-lexical, fig.height=6, fig.width=7}
lexical_cor <- features %>%
  select(all_of(tier_lexical)) %>%
  cor(use = "pairwise.complete.obs")

  corrplot(
  lexical_cor,
  method  = "color",
  order   = "hclust",
  tl.cex  = 0.7,
  tl.col  = "black",
  col     = colorRampPalette(c("firebrick", "white", "steelblue"))(200),
  title   = "Correlation - Lexical features",
  mar     = c(0, 0, 1.5, 0),
  addCoef.col = "black",
  number.cex  = 0.55
)
```

Across the full feature matrix, 10 pairs exceed `|r| > 0.7` and all of them
sit inside this Lexical count cluster (URLLength ↔ NoOfLettersInURL ↔
NoOfDegitsInURL and partners). These are the concrete collinear pairs that
algorithmic and embedded selection are expected to handle differently.

### 4.3.2 Variance Inflation Factors

```{r vif}
library(knitr)

# Target-free VIF from the design matrix (avoids non-converged GLM).
X_vif <- features %>%
  select(all_of(continuous_features)) %>%
  as.data.frame()

vif_values <- vapply(seq_along(X_vif), function(i) {
  r2 <- summary(lm(X_vif[[i]] ~ ., data = X_vif[, -i, drop = FALSE]))$r.squared
  1 / (1 - r2)
}, numeric(1))
names(vif_values) <- names(X_vif)

vif_df <- tibble(
  feature = names(vif_values),
  vif     = as.numeric(vif_values)
) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  mutate(severity = case_when(
    vif > 10            ~ "severe (> 10)",
    vif >= 5 & vif <= 10 ~ "moderate (5-10)",
    TRUE                ~ "ok (< 5)"
  )) %>%
  arrange(desc(vif))

vif_df %>%
  transmute(Feature = feature,
            Family  = family,
            VIF     = round(vif, 2),
            Severity = severity) %>%
  kable(caption = "Variance Inflation Factors - continuous features (lm-based, target-free).")
```

**Takeaway (H2):** Removing the four ratio features in Section 3.3 was
necessary but not sufficient. `URLLength`, `NoOfLettersInURL` and
`NoOfDegitsInURL` remain near-collinear because they measure overlapping
URL-string properties. Stepwise/RFE will likely keep most of them; LASSO and
RF importance are expected to retain only one or two. Quantifying that gap is
the H2 modelling task. Binary features are excluded from the VIF computation
by construction (linear regression of a 0/1 indicator on the rest is not
informative for collinearity diagnostics in the same scale as continuous
predictors).


## 4.4 Skewness Assessment *(Scenario 2 preparation)*

```{r outlier-skewness, fig.height=6, fig.width=8}
# Skewness for all continuous features
skewness_fn <- function(x) {
  x <- x[!is.na(x)]
  n <- length(x)
  m <- mean(x)
  s <- sd(x)
  (n / ((n-1)*(n-2))) * sum(((x - m) / s)^3)
}

skew_df <- tibble(
  feature  = continuous_features,
  skewness = map_dbl(continuous_features, ~ skewness_fn(features[[.x]]))
) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  mutate(abs_skew = abs(skewness)) %>%
  arrange(desc(abs_skew))

ggplot(skew_df, aes(x = reorder(feature, abs_skew), y = skewness, fill = family)) +
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = c(-2, 2), linetype = "dashed", colour = "grey40") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Skewness of continuous features",
       subtitle = "Dashed lines at +/- 2 mark heavy skew; LR/LDA may struggle here",
       x = NULL, y = "Skewness", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")

# Count features with |skewness| > 2
n_heavy_skew <- sum(skew_df$abs_skew > 2)
cat("\nFeatures with |skewness| > 2:", n_heavy_skew, "of",
    length(continuous_features), "\n")
cat("These are:\n")
skew_df %>% filter(abs_skew > 2) %>% select(feature, skewness, family) %>% print()
```

**Takeaway (Scenario 2):** Several count-based features are heavily right-skewed with
extreme outliers. This has direct implications for model selection:

- **Parametric models** (LR, LDA): will require feature scaling and log-transforms
  of skewed features to satisfy distributional assumptions.
- **Non-parametric models** (trees, KNN, SVM): are robust to skewness but KNN and SVM
  still require standardisation for distance-based computations.



---

# 5. EDA Summary & Modelling Implications

**Note on dataset separability:** After removing computed scores and redundant
features (Section 3.3), 40 raw features remain (13 Lexical + 7 Trust +
20 Behavior). Initial modelling shows that this reduced set still yields
AUC ≈ 0.99 for all model types — the dataset is near-perfectly separable on
the Full tier. H1 is therefore tested primarily on the **URLOnly tier**
(13 Lexical features) where meaningful parametric vs. non-parametric
differences are expected to emerge; H2 is tested on the **Full tier**, where
the multicollinearity that drives the algorithmic-vs-embedded gap actually
exists.

## 5.1 Implications for modelling scenarios

| Finding | Scenario 2 implication (H1) | Scenario 3 implication (H2) |
|---------|----------------------------|------------------------------|
| Near-perfect separability on Full tier | Comparison focused on URLOnly tier where models actually differ | Use Full tier; selection methods need a non-trivial pool of redundant features |
| Lexical SMD lowest + features must be combined | Largest parametric/non-parametric AUC gap expected on URLOnly | Algorithmic methods may keep all weak Lexical features; embedded penalisation should drop them |
| Heavy skewness in 19 of 22 continuous features | Log-transform + standardisation for LR/LDA; trees/RF unaffected | Apply same transform before LASSO/ElasticNet (they are scale-sensitive) |
| URL-length cluster: VIF > 500 and abs(r) > 0.84 | Use stable solvers / regularised LR for fairness | Strongest test case for H2 — embedded should collapse cluster, stepwise should not |
| Trust binary features: low intra-correlation | No special treatment | All three Trust HTTPS-style flags likely retained by both selection families — informative-feature overlap |

