---
title: "Phishing URL Detection — Hypothesis & EDA"
author: "K-o-m-A"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---
```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(corrplot)
library(patchwork)
library(scales)
```

---

# 1. Problem Statement

## 1.1 Real-world scenario

A user clicks a link in an email, chat, or search result. The
corporate proxy has to decide **before the page loads** whether the link is
phishing. The cheapest signal at that moment is the URL string itself — no
DNS lookup, no HTML fetch, no rendering. If a URL-only check is good enough,
we can block the link for free. If it isn't, we have to fall back to a
slower page-level check.

The question we want to answer is simple:
**how much phishing signal is already in the URL?** If a URL-only model works well, we can run it as a fast
first-line filter. Where it fails, the slower trust- and behavior-level checks have to take over.

This lines up with our two chosen scenarios:

- **Scenario 2** — parametric vs. non-parametric models on the URL-only
  tier. Tells us whether a linear model is enough, or whether the URL hides
  non-linear patterns that only RF/SVM/GBM can pick up.

## 1.2 Dataset

The PhiUSIIL dataset has ~235 k rows and 50 predictor features. We split
them into three families by how expensive they are to compute:

| Family | What it captures | Example features |
|--------|------------------|-----------------|
| **Lexical** | Properties of the URL string | `URLLength`, `NoOfSubDomain`, `NoOfObfuscatedChar`, `CharContinuationRate` |
| **Trust** | Domain / title consistency | `IsHTTPS`, `HasTitle`, `HasFavicon`, `Bank`, `Pay`, `Crypto` |
| **Behavior** | Page-level content signals | `NoOfURLRedirect`, `NoOfiFrame`, `HasExternalFormSubmit`, `HasHiddenFields` |

---

# 2. Hypotheses

## H1 (Scenario 2) — On the URL-only tier, non-parametric models pick up meaningfully more signal than parametric ones

Phishing URLs usually leave lexical traces: odd subdomain depth, paths full
of digits, obfuscated characters. None of this needs the page to be
downloaded. We expect a model trained only on Lexical features (URL string,
no DNS, no HTML) to still reach a reasonable AUC — but with a **bigger**
parametric vs. non-parametric gap than any other tier. The reason: lexical
patterns interact in non-additive ways (lengths × digit counts × special
chars), and a linear decision boundary can't fully use those interactions.
We compare three parametric methods (LR, LDA, Naive Bayes) against three
non-parametric ones (RF, SVM-RBF, KNN).

**Concrete success criteria.** To keep the claim precise, H1 is tested
against the 5-fold CV protocol from `scenario_2.rmd`:

| # | Criterion | Threshold |
|---|-----------|-----------|
| C1 | Mean CV AUC gap between the best non-parametric and best parametric model on the **Lexical (URLOnly) tier** | **ΔAUC ≥ 0.02** |
| C2 | Paired Wilcoxon signed-rank test on the 5 per-fold AUCs of those two models | **p < 0.05** |
| C3 | Gradient across tiers: the Lexical gap is **strictly larger** than the Full-tier gap (Full saturates at AUC ≈ 0.99 and the gap should drop to ≤ 0.005). | gap(Lexical) > gap(Full) |

H1 is **supported** only if C1 **and** C2 hold. C3 turns the claim from
"non-parametric wins on one tier" into the stronger statement *"the
non-parametric advantage shrinks as the feature tier gets stronger."* H1 is
**rejected** if either C1 or C2 fails on Lexical.

## H2 (Scenario 5) — The cost of dimensionality reduction is not the same on every tier

Every feature tier can be compressed into a few new axes via PCA
(unsupervised) or LDA projection (supervised). We claim the cost of that
compression is not uniform across tiers: on the URL-only (Lexical) tier it
is high, because the phishing signal lives in interactions between
features and a linear projection loses those interactions; on Behavior and
FullLite it approaches zero, because the signal is additive enough that
any sufficiently-large PC basis carries it. This is the same interaction
structure that forced us to reach for non-parametric models in H1 — on
Lexical it blocks linear compression the same way it blocks linear
classification.

**Concrete success criteria** (tested in `scenario_5.rmd`):

| # | Criterion | Threshold |
|---|-----------|-----------|
| C1 | AUC drop from raw Lexical to Lexical compressed into 3 PCs, best downstream model | **ΔAUC ≥ 0.05** |
| C2 | AUC drop from raw FullLite to FullLite compressed into 10 PCs, same model | **ΔAUC ≤ 0.01** |
| C3 | Gradient: drop(Lexical) > drop(FullLite) | **yes** |

H2 is **supported** only if C1 **and** C2 hold on the same downstream
classifier family. C3 turns the claim into a gradient parallel to H1's —
the compression cost rises as the tier gets weaker.

---

# 3. Data Loading & Cleaning

```{r load-data}
raw <- read_csv(
  "PhiUSIIL_Phishing_URL_Dataset.csv",
  show_col_types = FALSE
)

# Strip UTF-8 BOM that appears on the first column name
names(raw) <- map_chr(names(raw), ~ str_replace(.x, "^\ufeff", ""))

# glimpse(raw)
```

```{r define-roles}
# Columns excluded from modelling (identifiers / free text)
ID_COLS <- c("FILENAME", "URL", "Domain", "TLD", "Title")

# Binary target as factor
df <- raw %>%
  mutate(label_bin = factor(label, levels = c(0, 1),
                            labels = c("Legitimate", "Phishing"))) %>%
  select(-all_of(ID_COLS))

# Feature matrix (numeric predictors only)
features <- df %>% select(-label, -label_bin)

cat("Observations:", nrow(df), "\n")
cat("Predictors:  ", ncol(features), "\n")
```


## Feature-family mapping
The three families (Lexical / Trust / Behavior) are our own taxonomy, not
a property of the dataset. They reflect the cost of obtaining each signal
at decision time:
   Lexical  — derivable from the raw URL string alone (zero network cost).
   Trust    — requires light page/domain inspection or an external lookup
              (DNS, TLD database, HTTPS handshake, title parsing).
   Behavior — requires fetching, parsing and partially rendering the page.

Note on `IsHTTPS`: although the "https://" scheme is part of the URL string,
we place it in the **Trust** family. Modern browsers refuse plain HTTP for
any sensitive interaction and warn the user prominently, so the scheme is
already enforced by the trust layer of the browser rather than acting as a
free lexical clue. Treating it as Lexical would let the URLOnly tier cheat
off a binary that is, in practice, a deployment/trust signal.

```{r family-map}

family_map <- tribble(
  ~feature,                    ~family,
  "URLLength",                 "Lexical",
  "DomainLength",              "Lexical",
  "IsDomainIP",                "Lexical",
  "URLSimilarityIndex",        "Trust",
  "CharContinuationRate",      "Lexical",
  "TLDLegitimateProb",         "Trust",
  "URLCharProb",               "Trust",
  "TLDLength",                 "Lexical",
  "NoOfSubDomain",             "Lexical",
  "HasObfuscation",            "Lexical",
  "NoOfObfuscatedChar",        "Lexical",
  "ObfuscationRatio",          "Lexical",
  "NoOfLettersInURL",          "Lexical",
  "LetterRatioInURL",          "Lexical",
  "NoOfDegitsInURL",           "Lexical",
  "DegitRatioInURL",           "Lexical",
  "NoOfEqualsInURL",           "Lexical",
  "NoOfQMarkInURL",            "Lexical",
  "NoOfAmpersandInURL",        "Lexical",
  "NoOfOtherSpecialCharsInURL","Lexical",
  "SpacialCharRatioInURL",     "Lexical",
  "IsHTTPS",                   "Trust",
  "LineOfCode",                "Behavior",
  "LargestLineLength",         "Behavior",
  "HasTitle",                  "Trust",
  "DomainTitleMatchScore",     "Trust",
  "URLTitleMatchScore",        "Trust",
  "HasFavicon",                "Trust",
  "Robots",                    "Trust",
  "IsResponsive",              "Behavior",
  "NoOfURLRedirect",           "Behavior",
  "NoOfSelfRedirect",          "Behavior",
  "HasDescription",            "Behavior",
  "NoOfPopup",                 "Behavior",
  "NoOfiFrame",                "Behavior",
  "HasExternalFormSubmit",     "Behavior",
  "HasSocialNet",              "Behavior",
  "HasSubmitButton",           "Behavior",
  "HasHiddenFields",           "Behavior",
  "HasPasswordField",          "Behavior",
  "Bank",                      "Trust",
  "Pay",                       "Trust",
  "Crypto",                    "Trust",
  "HasCopyrightInfo",          "Behavior",
  "NoOfImage",                 "Behavior",
  "NoOfCSS",                   "Behavior",
  "NoOfJS",                    "Behavior",
  "NoOfSelfRef",               "Behavior",
  "NoOfEmptyRef",              "Behavior",
  "NoOfExternalRef",           "Behavior"
)
```

## 3.3 Feature Exclusion

Before the EDA we drop three groups of features that would bias the
analysis or trivialise the modelling. All three decisions are based on
what the feature actually means.

### Category 1 — Computed probability / similarity scores

These are not raw URL measurements. They are outputs of other models or
expert-written scoring functions that already encode phishing-detection
logic. Keeping them would mean we are classifying another classifier's
output, not the URL itself.

```{r exclude-scores}
SCORES <- c(
  "URLSimilarityIndex",   # similarity to known-phishing URL patterns
  "TLDLegitimateProb",    # P(TLD is legitimate) from an external lookup
  "URLCharProb",          # P(character sequence) from a language model
  "DomainTitleMatchScore",# computed domain↔title consistency score
  "URLTitleMatchScore"    # computed URL↔title consistency score
)
cat("Computed scores to remove:", paste(SCORES, collapse = ", "), "\n")
```

### Category 2 — Redundant binary indicator

`HasObfuscation` is literally `(NoOfObfuscatedChar > 0)`. It holds no
information that the count feature doesn't already carry.

```{r exclude-binary}
REDUNDANT_BINARY <- "HasObfuscation"
cat("Redundant binary to remove:", REDUNDANT_BINARY, "\n")
```

### Category 3 — Redundant ratio features

Each ratio below equals `count / URLLength`, and both the count and
`URLLength` are already in the dataset. Keeping the ratio gives us a
perfect linear dependence — and no extra signal.

```{r exclude-ratios}
RATIOS <- c(
  "LetterRatioInURL",      # = NoOfLettersInURL   / URLLength
  "DegitRatioInURL",       # = NoOfDegitsInURL    / URLLength
  "ObfuscationRatio",      # = NoOfObfuscatedChar / URLLength
  "SpacialCharRatioInURL"  # = NoOfOtherSpecialCharsInURL / URLLength
)
cat("Linearly dependent ratios to remove:", paste(RATIOS, collapse = ", "), "\n")
```

```{r apply-exclusions}
EXCLUDE_COLS <- c(SCORES, REDUNDANT_BINARY, RATIOS)

df       <- df %>% select(-all_of(EXCLUDE_COLS))
features <- df %>% select(-label, -label_bin)

cat("\nFeatures after exclusion:", ncol(features), "\n")
cat("Removed:", paste(EXCLUDE_COLS, collapse = ", "), "\n")
```

Update the family map to reflect the reduced feature set:

```{r family-map-update}
family_map <- family_map %>%
  filter(!feature %in% EXCLUDE_COLS)

family_map %>% count(family) %>% print()
```

```{r feature-types, include=FALSE}
# Silent type split used by downstream VIF/skewness chunks
binary_features <- names(features)[map_lgl(features, ~ all(na.omit(.x) %in% c(0, 1)))]
continuous_features <- setdiff(names(features), binary_features)
```

## 3.4 Feature Tiers Used in Modelling

To test H1 we define three nested feature tiers. The modelling notebook
reads these tier definitions directly from here, so both notebooks stay
in sync.

```{r feature-tiers}
tier_lexical  <- family_map %>% filter(family == "Lexical")  %>% pull(feature)
tier_trust    <- family_map %>% filter(family == "Trust")    %>% pull(feature)
tier_behavior <- family_map %>% filter(family == "Behavior") %>% pull(feature)
tier_full     <- c(tier_lexical, tier_trust, tier_behavior)

cat("URLOnly tier (Lexical, ",  length(tier_lexical),  " features)\n", sep = "")
cat("Trust tier              (",  length(tier_trust),    " features)\n", sep = "")
cat("Behavior tier           (",  length(tier_behavior), " features)\n", sep = "")
cat("Full tier               (",  length(tier_full),     " features)\n", sep = "")
```

---

# 4. Exploratory Data Analysis


## 4.1 Data Overview

### 4.1.1 Class Balance

```{r class-balance, fig.height=3.5}
df %>%
  count(label_bin) %>%
  mutate(pct = n / sum(n)) %>%
  ggplot(aes(label_bin, n, fill = label_bin)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = scales::percent(pct, 0.1)), vjust = -0.4, size = 4) +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, .1))) +
  labs(title = "Class distribution",
       x = NULL, y = "Count") +
  theme_minimal(base_size = 12)
```

**Takeaway:** The dataset is roughly balanced, so we don't need resampling
or class-weighting. Accuracy works as a secondary metric next to AUC.

### 4.1.2 Missing Values

No missing values are present in any predictor (verified with
`colSums(is.na(features))`).


## 4.2 H1 — Lexical features overlap between classes, so we'll need non-linear models

H1 claims that a URL-only detector (Lexical features) will show a bigger
parametric vs. non-parametric gap than any other tier. We don't test
non-linearity directly in the EDA. Instead we show that **per-feature
discriminative power is lowest in the Lexical family**. When no single
feature tells the classes apart, the model has to combine several features
at once — and that's where non-linear models have an edge.

### 4.2.1 Discriminative Power by Family

We measure discriminative power separately for continuous and binary
features. The standardised mean difference (SMD) only makes sense for
continuous features. For binary features the "pooled SD" collapses to a
function of the class proportion, so SMD stops being comparable across
features. For those we use **Cramér's V** instead — the standard
effect-size measure for a 2×2 contingency table, on the same 0–1 scale
as |correlation|.

```{r discriminative-power, fig.height=7, fig.width=8}
# --- Continuous features: standardised mean difference ----------------------
smd_df <- df %>%
  select(label, all_of(continuous_features)) %>%
  pivot_longer(-label, names_to = "feature", values_to = "value") %>%
  group_by(feature, label) %>%
  summarise(m = mean(value, na.rm = TRUE), s = sd(value, na.rm = TRUE),
            .groups = "drop") %>%
  pivot_wider(names_from = label, values_from = c(m, s),
              names_glue = "{.value}_{label}") %>%
  mutate(
    pooled_sd = sqrt((s_0^2 + s_1^2) / 2),
    effect    = abs(m_1 - m_0) / (pooled_sd + 1e-9),
    metric    = "SMD (continuous)"
  ) %>%
  select(feature, effect, metric)

# --- Binary features: Cramer's V on a 2x2 contingency table -----------------
cramers_v <- function(x, y) {
  tab <- table(x, y)
  if (any(dim(tab) < 2)) return(NA_real_)
  chi2 <- suppressWarnings(chisq.test(tab, correct = FALSE)$statistic)
  n <- sum(tab)
  as.numeric(sqrt(chi2 / (n * (min(dim(tab)) - 1))))
}

cv_df <- tibble(
  feature = binary_features,
  effect  = map_dbl(binary_features, ~ cramers_v(features[[.x]], df$label)),
  metric  = "Cramer's V (binary)"
)

discrim_df <- bind_rows(smd_df, cv_df) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  arrange(desc(effect))

ggplot(discrim_df, aes(x = reorder(feature, effect), y = effect, fill = family)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ metric, scales = "free", ncol = 2) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Per-feature discriminative power",
       subtitle = "Continuous: |SMD|. Binary: Cramer's V. Both >= 0.",
       x = NULL, y = "Effect size", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")
```

Effect-size distribution per family (continuous features only, so the SMD
values are directly comparable):

```{r smd-by-family, fig.height=4, fig.width=6}
smd_family <- discrim_df %>%
  filter(metric == "SMD (continuous)", family != "Other")

ggplot(smd_family, aes(family, effect, fill = family)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  geom_jitter(width = 0.15, alpha = 0.5, size = 2) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "SMD distribution by family (continuous features only)",
       subtitle = "H1: Lexical SMD lowest -> largest parametric gap expected at URLOnly tier",
       x = NULL, y = "| SMD |") +
  theme_minimal(base_size = 12)
```

**Takeaway (H1):** Among the continuous features, Lexical has the lowest
median |SMD| and the smallest spread. Trust and Behavior carry more
per-feature signal. Weak per-feature evidence plus visible non-linear
structure is exactly the setting where parametric models struggle most —
and where the parametric vs. non-parametric gap predicted by H1 should
show up.

## 4.3 H2 preparation — How low-rank each tier already is

H2 says the cost of linear dimensionality reduction differs sharply across
tiers. Before we run the actual projections in `scenario_5.rmd`, a quick
look at multicollinearity tells us how much redundant dimensionality each
tier has to start with: if a tier is already low-rank, PCA will find a
compact basis easily — but that only measures variance, not class
separation, so the signal may still be lost. §4.5 checks the class side
next.

### 4.3.1 Lexical Correlation Matrix

The collinearity that drives H2 lives inside the Lexical family. Trust and
Behavior have `mean_abs_r` around 0.18 and 0.14 and no dense cluster. The
plot below makes the URL-length count cluster visible.

```{r cor-lexical, fig.height=6, fig.width=7}
lexical_cor <- features %>%
  select(all_of(tier_lexical)) %>%
  cor(use = "pairwise.complete.obs")

  corrplot(
  lexical_cor,
  method  = "color",
  order   = "hclust",
  tl.cex  = 0.7,
  tl.col  = "black",
  col     = colorRampPalette(c("firebrick", "white", "steelblue"))(200),
  title   = "Correlation - Lexical features",
  mar     = c(0, 0, 1.5, 0),
  addCoef.col = "black",
  number.cex  = 0.55
)
```

Across the full feature matrix, 10 pairs pass `|r| > 0.7` — and all of
them sit inside this Lexical count cluster (URLLength ↔ NoOfLettersInURL
↔ NoOfDegitsInURL and friends). A tight cluster like this means the
Lexical feature matrix has fewer than 13 independent directions of
variance: PCA will collapse the cluster into one dominant component. The
open question is whether the class signal lines up with that dominant
component or sits orthogonal to it — §4.5 tests exactly that.

### 4.3.2 Variance Inflation Factors

```{r vif}
library(knitr)

# Target-free VIF from the design matrix (avoids non-converged GLM).
X_vif <- features %>%
  select(all_of(continuous_features)) %>%
  as.data.frame()

vif_values <- vapply(seq_along(X_vif), function(i) {
  r2 <- summary(lm(X_vif[[i]] ~ ., data = X_vif[, -i, drop = FALSE]))$r.squared
  1 / (1 - r2)
}, numeric(1))
names(vif_values) <- names(X_vif)

vif_df <- tibble(
  feature = names(vif_values),
  vif     = as.numeric(vif_values)
) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  mutate(severity = case_when(
    vif > 10            ~ "severe (> 10)",
    vif >= 5 & vif <= 10 ~ "moderate (5-10)",
    TRUE                ~ "ok (< 5)"
  )) %>%
  arrange(desc(vif))

vif_df %>%
  transmute(Feature = feature,
            Family  = family,
            VIF     = round(vif, 2),
            Severity = severity) %>%
  kable(caption = "Variance Inflation Factors - continuous features (lm-based, target-free).")
```

**Takeaway (H2):** `URLLength`, `NoOfLettersInURL` and `NoOfDegitsInURL`
stay near-collinear because they measure overlapping properties of the
same URL string. For H2 this means Lexical has far fewer independent
directions of variance than it has features — PCA will compress it
aggressively. The open question is whether the class signal lives in
those dominant variance directions or orthogonal to them; §4.5 answers
it. Binary features are left out of the VIF calculation by design —
regressing a 0/1 indicator on continuous predictors is not comparable on
the same scale.


## 4.4 Skewness Assessment *(Scenario 2 preparation)*

```{r outlier-skewness, fig.height=6, fig.width=8}
# Skewness for all continuous features
skewness_fn <- function(x) {
  x <- x[!is.na(x)]
  n <- length(x)
  m <- mean(x)
  s <- sd(x)
  (n / ((n-1)*(n-2))) * sum(((x - m) / s)^3)
}

skew_df <- tibble(
  feature  = continuous_features,
  skewness = map_dbl(continuous_features, ~ skewness_fn(features[[.x]]))
) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  mutate(abs_skew = abs(skewness)) %>%
  arrange(desc(abs_skew))

ggplot(skew_df, aes(x = reorder(feature, abs_skew), y = skewness, fill = family)) +
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = c(-2, 2), linetype = "dashed", colour = "grey40") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Skewness of continuous features",
       subtitle = "Dashed lines at +/- 2 mark heavy skew; LR/LDA may struggle here",
       x = NULL, y = "Skewness", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")

# Count features with |skewness| > 2
n_heavy_skew <- sum(skew_df$abs_skew > 2)
cat("\nFeatures with |skewness| > 2:", n_heavy_skew, "of",
    length(continuous_features), "\n")
cat("These are:\n")
skew_df %>% filter(abs_skew > 2) %>% select(feature, skewness, family) %>% print()
```

**Takeaway (Scenario 2):** Several count features are heavily right-skewed
with long outlier tails. This shapes our modelling choices:

- **Parametric models** (LR, LDA): need scaling and a log-transform on
  skewed features to make the distributional assumption hold.
- **Non-parametric models** (trees, KNN, SVM): robust to skew, but KNN and
  SVM still need standardisation because they use distances.


## 4.5 H2 preparation — where the class signal lives across principal components

§4.3 showed that Lexical has fewer independent directions of variance
than it has features. That alone only tells us about variance — not
whether any of those directions carry the *class* signal. For H2 the
crucial question is whether the top few PCs of each tier line up with
phishing vs. legitimate separation. If they don't, any low-dimensional
projection will drop detection accuracy. Two PCA-based diagnostics answer
that: a scree plot (how many PCs each tier needs to capture most of its
variance) and the univariate AUC of each leading PC (how much class
signal sits along that direction).

`scenario_5.rmd` tests H2 on the scenario_2 tier set (with six Behavior
near-leakers stripped, i.e. FullLite = 34 features). Here we use the
native EDA tiers (Full = 40) to stay consistent with §4.2 and §4.3 — the
direction of the finding is the same either way, and stripping the
near-leakers only makes the Lexical/Full gap smaller, never larger.

### 4.5.1 Scree plot per tier

```{r pca-scree, fig.height=4.5, fig.width=8}
prep_for_pca <- function(cols) {
  X <- features[, cols, drop = FALSE]
  cont <- setdiff(names(X), binary_features)
  if (length(cont) > 0) {
    X[, cont] <- lapply(X[, cont], function(x) log1p(pmax(x, 0)))
  }
  X_mat <- as.matrix(X)
  sds <- apply(X_mat, 2, sd)
  X_mat <- X_mat[, sds > 0, drop = FALSE]
  scale(X_mat)
}

pca_tiers <- list(
  Lexical  = tier_lexical,
  Trust    = tier_trust,
  Behavior = tier_behavior,
  Full     = tier_full
)

scree_df <- map_dfr(names(pca_tiers), function(tier) {
  Xp <- prep_for_pca(pca_tiers[[tier]])
  pc <- prcomp(Xp, center = FALSE, scale. = FALSE)
  var_exp <- pc$sdev^2 / sum(pc$sdev^2)
  tibble(tier = tier, k = seq_along(var_exp), cum_var = cumsum(var_exp))
})

scree_df %>%
  mutate(tier = factor(tier, levels = c("Lexical", "Trust",
                                         "Behavior", "Full"))) %>%
  ggplot(aes(k, cum_var, colour = tier)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0.9, linetype = "dashed", colour = "grey40") +
  scale_x_continuous(breaks = seq(0, 40, 5)) +
  scale_y_continuous(labels = scales::percent) +
  scale_colour_brewer(palette = "Set2") +
  labs(title = "Cumulative variance explained by leading PCs",
       subtitle = "Dashed line at 90% variance — fewer PCs = tier is more low-rank",
       x = "Number of PCs retained", y = "Cumulative variance",
       colour = "Tier") +
  theme_minimal(base_size = 11)
```

Lexical reaches 90 % of its variance with 3–4 PCs — the URL-length
cluster is doing most of the work. Trust spends all 7 directions because
its features are binary and near-uncorrelated. Behavior and Full are
spread out and need 10+ PCs to reach 90 %. That is the *variance*
picture; it predicts that PCA compression on Lexical will look cheap.
The class picture below tells a different story.

### 4.5.2 Class signal carried by each leading PC

```{r pca-per-pc-auc, fig.height=4, fig.width=9}
pca_auc_df <- map_dfr(names(pca_tiers), function(tier) {
  Xp <- prep_for_pca(pca_tiers[[tier]])
  pc <- prcomp(Xp, center = FALSE, scale. = FALSE)
  k_max <- min(10, ncol(pc$x))
  map_dfr(seq_len(k_max), function(i) {
    a <- suppressMessages(as.numeric(
      pROC::roc(df$label_bin, pc$x[, i],
                levels = c("Legitimate", "Phishing"),
                quiet = TRUE)$auc))
    tibble(tier = tier, PC = i, auc = max(a, 1 - a))
  })
})

pca_auc_df %>%
  mutate(tier = factor(tier, levels = c("Lexical", "Trust",
                                         "Behavior", "Full"))) %>%
  ggplot(aes(PC, auc, fill = tier)) +
  geom_col() +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey40") +
  facet_wrap(~ tier, nrow = 1) +
  scale_x_continuous(breaks = 1:10) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Phishing signal carried by each leading PC",
       subtitle = "Orientation-free univariate AUC; PC1 = direction of largest variance",
       x = "Principal component index", y = "Univariate AUC") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")
```

**Takeaway (H2):** On Lexical, PC1 captures almost all the variance but
the class signal is smeared across several PCs with low individual AUC —
variance directions and class directions do not line up. Trust has so few
features that each PC carries a comparable share of the (already weak)
signal. On Behavior and Full the top 2–3 PCs each reach AUC above 0.8, so
a handful of components recover most of the task. This is the direct EDA
prediction H2 tests quantitatively in `scenario_5.rmd`: compression is
cheap on Behavior and Full, expensive on Lexical.

---

# 5. EDA Summary & Modelling Implications

**A note on separability:** After the Section 3.3 exclusions, 40 features
remain (13 Lexical + 7 Trust + 20 Behavior). Even on this reduced set,
every model type already reaches AUC ≈ 0.99 on the Full tier — the task
is near-perfectly separable once all three families are present. So we
test H1 mainly on the **URLOnly tier** (13 Lexical features), where the
parametric vs. non-parametric contrast has room to show. H2 is tested
across tiers, contrasting URL-only against Behavior/FullLite, because the
cost of compression is what differs between them.

## 5.1 Implications for modelling scenarios

| Finding | Scenario 2 implication (H1) | Scenario 5 implication (H2) |
|---------|----------------------------|------------------------------|
| Near-perfect separability on Full tier | Comparison focused on URLOnly tier where models actually differ | Behavior/FullLite has many directions that each carry class signal — compression into a handful of components should be cheap |
| Lexical SMD lowest + features must be combined | Largest parametric/non-parametric AUC gap expected on URLOnly | On Lexical the signal lives in feature combinations, so a linear projection to 3 PCs is expected to drop AUC sharply |
| Heavy skewness in 19 of 22 continuous features | Log-transform + standardisation for LR/LDA; trees/RF unaffected | Apply the same log+standardise transform before PCA/LDA projection (both are scale-sensitive) |
| URL-length cluster: VIF > 1000 and abs(r) > 0.84 | Use stable solvers / regularised LR for fairness | Lexical is low-rank in variance but the class signal is smeared across PCs — the strongest case for "compression is expensive here" |
| Trust binary features: low intra-correlation | No special treatment | Trust spends all 7 PCs on variance; projection cannot shorten it without losing signal |

