---
title: "Phishing URL Detection — Hypothesis & EDA"
author: "K-o-m-A"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---
```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(corrplot)
library(patchwork)
library(scales)
```

---

# 1. Problem Statement

## 1.1 Real-world scenario

A user clicks a link in an email, chat, or search result. The
corporate proxy has to decide **before the page loads** whether the link is
phishing. The cheapest signal at that moment is the URL string itself — no
DNS lookup, no HTML fetch, no rendering. If a URL-only check is good enough,
we can block the link for free. If it isn't, we have to fall back to a
slower page-level check.

The question we want to answer is simple:
**how much phishing signal is already in the URL?** If a URL-only model works well, we can run it as a fast
first-line filter. Where it fails, the slower trust- and behavior-level checks have to take over.

Explored scenarios for hypothesis driven modeling: 2,5

## 1.2 Dataset

The PhiUSIIL dataset has ~235 k rows and 50 predictor features. We split
them into three families by how expensive they are to compute:

| Family | What it captures | Example features |
|--------|------------------|-----------------|
| **Lexical** | Properties of the URL string | `URLLength`, `NoOfSubDomain`, `NoOfObfuscatedChar`, `CharContinuationRate` |
| **Trust** | Domain / title consistency | `IsHTTPS`, `HasTitle`, `HasFavicon`, `Bank`, `Pay`, `Crypto` |
| **Behavior** | Page-level content signals | `NoOfURLRedirect`, `NoOfiFrame`, `HasExternalFormSubmit`, `HasHiddenFields` |

---

# 2. Hypotheses

## H1 (Scenario 2) — The gap between parametric and non-parametric model families is tier-dependent

Phishing URLs usually leave lexical traces: odd subdomain depth, paths full
of digits, obfuscated characters. None of this needs the page to be
downloaded. We expect a model trained only on Lexical features to still reach a reasonable AUC, but the **size of the
gap** between parametric and non-parametric model families should change
with the family.

We compare three parametric methods (LR, LDA, Naive Bayes)
against three non-parametric ones (RF, SVM-RBF, KNN).

| # | Criterion | Threshold |
|---|-----------|-----------|
| C1 | Mean CV AUC gap between the best non-parametric and best parametric model on the **Lexical (URLOnly) tier** | **ΔAUC ≥ 0.02** |
| C2 | Paired Wilcoxon signed-rank test on the 5 per-fold AUCs of those two models | **p < 0.05** |
| C3 | Gradient across tiers: the Lexical gap is **strictly larger** than the FullLite-tier gap . | gap(Lexical) > gap(FullLite) |

H1 is **supported** only if C1 **and** C2 hold. C3 turns the claim from
"the two families differ on one tier" into the stronger statement *"the
difference between families shrinks as the feature tier gets stronger."*
H1 is **rejected** if either C1 or C2 fails on Lexical.

Note:
The FullLite tier is the scenario_2 replacement for a naive "Full" tier: six Behavior
near-leakers (`LineOfCode`, `NoOfJS`, `NoOfCSS`, `NoOfImage`,
`NoOfSelfRef`, `NoOfExternalRef`, each with univariate AUC > 0.95) are
stripped because they push Full to AUC ≈ 1.0 and hide every contrast H1
is trying to measure.

## H2 (Scenario 5) — On FullLite, a linear projection preserves the AUC ceiling, but a supervised axis does it in one dimension while an unsupervised basis needs many

Scenario 5 asks the complementary question to Scenario 2: instead of
varying which columns the model sees (the tier), we keep the tier fixed
and replace the columns themselves with a smaller set of derived
coordinates. FullLite (34 predictors) is the only tier with enough
dimensions for "reduction" to be a meaningful question — Lexical (13),
Trust (7) and Behavior (14) are too small to justify projection, so H2
is tested on FullLite exclusively.

On FullLite the phishing signal is near-linearly separable (AUC ≈ 0.99
in Scenario 2), so we expect a linear projection — PCA (unsupervised)
or LDA (supervised) — to preserve the AUC ceiling. The non-trivial
part of the claim is the asymmetry between the two projections:

- **LDA** optimises class separation directly and, for a binary
  problem, produces a single discriminant axis (LD1). If the two
  classes are near-Gaussian along that axis, one coordinate should
  already carry the full supervised signal.
- **PCA** optimises variance, not separation. It needs strictly more
  than one axis to match LDA-1, because variance directions and class
  directions only coincide by accident.

The cost of either projection is per-feature explainability: after
projection, model coefficients read as per-component weights, not
per-feature weights. A kernel PCA (RBF) at the same *k* as linear PCA
is run as a sanity check — on FullLite we expect it to match, not
beat, linear PCA, because the task is near-linearly separable.

| # | Criterion | Threshold |
|---|-----------|-----------|
| C1 | AUC drop from raw FullLite to LDA-1 projection, best downstream model | **ΔAUC ≤ 0.01** |
| C2 | AUC drop from raw FullLite to PCA-*k* (k = smallest index reaching ≥ 95 % cumulative variance), same model | **ΔAUC ≤ 0.01** |
| C3 | Supervised-vs-unsupervised asymmetry: **k_pca > 1** (PCA needs more than one axis to match LDA-1) | **yes** |

H2 is **supported** only if C1 **and** C2 hold on the same downstream
classifier family. C3 formalises the asymmetry — if PCA reaches the
ceiling at k_pca = 1, there is no meaningful contrast between the
supervised and unsupervised projection and H2's framing collapses.

---

# 3. Data Loading & Cleaning

```{r load-data}
raw <- read_csv(
  "PhiUSIIL_Phishing_URL_Dataset.csv",
  show_col_types = FALSE
)

# Strip UTF-8 BOM that appears on the first column name
names(raw) <- map_chr(names(raw), ~ str_replace(.x, "^\ufeff", ""))

# glimpse(raw)
```

```{r define-roles}
# Columns excluded from modeling (identifiers / free text)
ID_COLS <- c("FILENAME", "URL", "Domain", "TLD", "Title")

# Binary target as factor
df <- raw %>%
  mutate(label_bin = factor(label, levels = c(0, 1),
                            labels = c("Legitimate", "Phishing"))) %>%
  select(-all_of(ID_COLS))

# Feature matrix (numeric predictors only)
features <- df %>% select(-label, -label_bin)

cat("Observations:", nrow(df), "\n")
cat("Predictors:  ", ncol(features), "\n")
```


## Feature-family mapping
The three families (Lexical / Trust / Behavior) are our own taxonomy, not
a property of the dataset. They reflect the cost of obtaining each signal
at decision time:
   Lexical  — derivable from the raw URL string alone (zero network cost).
   Trust    — requires light page/domain inspection or an external lookup
              (DNS, TLD database, HTTPS handshake, title parsing).
   Behavior — requires fetching, parsing and partially rendering the page.

Note on `IsHTTPS`: although the "https://" scheme is part of the URL string,
we place it in the **Trust** family. Modern browsers refuse plain HTTP for
any sensitive interaction and warn the user prominently, so the scheme is
already enforced by the trust layer of the browser rather than acting as a
free lexical clue. Treating it as Lexical would let the URLOnly tier cheat
off a binary that is, in practice, a deployment/trust signal.

```{r family-map}

family_map <- tribble(
  ~feature,                    ~family,
  "URLLength",                 "Lexical",
  "DomainLength",              "Lexical",
  "IsDomainIP",                "Lexical",
  "URLSimilarityIndex",        "Trust",
  "CharContinuationRate",      "Lexical",
  "TLDLegitimateProb",         "Trust",
  "URLCharProb",               "Trust",
  "TLDLength",                 "Lexical",
  "NoOfSubDomain",             "Lexical",
  "HasObfuscation",            "Lexical",
  "NoOfObfuscatedChar",        "Lexical",
  "ObfuscationRatio",          "Lexical",
  "NoOfLettersInURL",          "Lexical",
  "LetterRatioInURL",          "Lexical",
  "NoOfDegitsInURL",           "Lexical",
  "DegitRatioInURL",           "Lexical",
  "NoOfEqualsInURL",           "Lexical",
  "NoOfQMarkInURL",            "Lexical",
  "NoOfAmpersandInURL",        "Lexical",
  "NoOfOtherSpecialCharsInURL","Lexical",
  "SpacialCharRatioInURL",     "Lexical",
  "IsHTTPS",                   "Trust",
  "LineOfCode",                "Behavior",
  "LargestLineLength",         "Behavior",
  "HasTitle",                  "Trust",
  "DomainTitleMatchScore",     "Trust",
  "URLTitleMatchScore",        "Trust",
  "HasFavicon",                "Trust",
  "Robots",                    "Trust",
  "IsResponsive",              "Behavior",
  "NoOfURLRedirect",           "Behavior",
  "NoOfSelfRedirect",          "Behavior",
  "HasDescription",            "Behavior",
  "NoOfPopup",                 "Behavior",
  "NoOfiFrame",                "Behavior",
  "HasExternalFormSubmit",     "Behavior",
  "HasSocialNet",              "Behavior",
  "HasSubmitButton",           "Behavior",
  "HasHiddenFields",           "Behavior",
  "HasPasswordField",          "Behavior",
  "Bank",                      "Trust",
  "Pay",                       "Trust",
  "Crypto",                    "Trust",
  "HasCopyrightInfo",          "Behavior",
  "NoOfImage",                 "Behavior",
  "NoOfCSS",                   "Behavior",
  "NoOfJS",                    "Behavior",
  "NoOfSelfRef",               "Behavior",
  "NoOfEmptyRef",              "Behavior",
  "NoOfExternalRef",           "Behavior"
)
```

## 3.3 Feature Exclusion

Before the EDA we drop three groups of features that would bias the
analysis or trivialise the modeling. All three decisions are based on
what the feature actually means.

### Category 1 — Computed probability / similarity scores

These are not raw URL measurements. They are outputs of other models or
expert-written scoring functions that already encode phishing-detection
logic. Keeping them would mean we are classifying another classifier's
output, not the URL itself.

```{r exclude-scores}
SCORES <- c(
  "URLSimilarityIndex",   # similarity to known-phishing URL patterns
  "TLDLegitimateProb",    # P(TLD is legitimate) from an external lookup
  "URLCharProb",          # P(character sequence) from a language model
  "DomainTitleMatchScore",# computed domain↔title consistency score
  "URLTitleMatchScore"    # computed URL↔title consistency score
)
cat("Computed scores to remove:", paste(SCORES, collapse = ", "), "\n")
```

### Category 2 — Redundant binary indicator

`HasObfuscation` is literally `(NoOfObfuscatedChar > 0)`. It holds no
information that the count feature doesn't already carry.

```{r exclude-binary}
REDUNDANT_BINARY <- "HasObfuscation"
cat("Redundant binary to remove:", REDUNDANT_BINARY, "\n")
```

### Category 3 — Redundant ratio features

Each ratio below equals `count / URLLength`, and both the count and
`URLLength` are already in the dataset. Keeping the ratio gives us a
perfect linear dependence — and no extra signal.

```{r exclude-ratios}
RATIOS <- c(
  "LetterRatioInURL",      # = NoOfLettersInURL   / URLLength
  "DegitRatioInURL",       # = NoOfDegitsInURL    / URLLength
  "ObfuscationRatio",      # = NoOfObfuscatedChar / URLLength
  "SpacialCharRatioInURL"  # = NoOfOtherSpecialCharsInURL / URLLength
)
cat("Linearly dependent ratios to remove:", paste(RATIOS, collapse = ", "), "\n")
```

```{r apply-exclusions}
EXCLUDE_COLS <- c(SCORES, REDUNDANT_BINARY, RATIOS)

df       <- df %>% select(-all_of(EXCLUDE_COLS))
features <- df %>% select(-label, -label_bin)

cat("\nFeatures after exclusion:", ncol(features), "\n")
cat("Removed:", paste(EXCLUDE_COLS, collapse = ", "), "\n")
```

Update the family map to reflect the reduced feature set:

```{r family-map-update}
family_map <- family_map %>%
  filter(!feature %in% EXCLUDE_COLS)

family_map %>% count(family) %>% print()
```

```{r feature-types, include=FALSE}
# Type split used by downstream VIF/skewness chunks
binary_features <- names(features)[map_lgl(features, ~ all(na.omit(.x) %in% c(0, 1)))]
continuous_features <- setdiff(names(features), binary_features)
```

## 3.4 Feature Tiers Used in Modeling

To test H1 we define three nested feature tiers. The modeling notebook
reads these tier definitions directly from here, so both notebooks stay
in sync. The `tier_full` built here (40 features) is the EDA baseline
used by §4.5's PCA diagnostics; the scenario_2 modeling notebook
replaces it with **FullLite** (34 features) by stripping six Behavior
near-leakers — see §5 for the reasoning.

```{r feature-tiers}
tier_lexical  <- family_map %>% filter(family == "Lexical")  %>% pull(feature)
tier_trust    <- family_map %>% filter(family == "Trust")    %>% pull(feature)
tier_behavior <- family_map %>% filter(family == "Behavior") %>% pull(feature)
tier_full     <- c(tier_lexical, tier_trust, tier_behavior)

cat("URLOnly tier (Lexical, ",  length(tier_lexical),  " features)\n", sep = "")
cat("Trust tier              (",  length(tier_trust),    " features)\n", sep = "")
cat("Behavior tier           (",  length(tier_behavior), " features)\n", sep = "")
cat("Full tier               (",  length(tier_full),     " features)\n", sep = "")
```

---

# 4. Exploratory Data Analysis


## 4.1 Data Overview

### 4.1.1 Class Balance

```{r class-balance, fig.height=3.5}
df %>%
  count(label_bin) %>%
  mutate(pct = n / sum(n)) %>%
  ggplot(aes(label_bin, n, fill = label_bin)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = scales::percent(pct, 0.1)), vjust = -0.4, size = 4) +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, .1))) +
  labs(title = "Class distribution",
       x = NULL, y = "Count") +
  theme_minimal(base_size = 12)
```

**Takeaway:** The dataset is roughly balanced, so we don't need resampling
or class-weighting. Accuracy works as a secondary metric next to AUC.

### 4.1.2 Missing Values

No missing values are present in any predictor (verified with `colSums(is.na(features))`).

## 4.2 H1 preparation

H1 claims that a URL-only detector (Lexical features) will show a bigger
parametric vs. non-parametric gap than any other tier. We don't test
non-linearity directly in the EDA. Instead we show that **per-feature
discriminative power is lowest in the Lexical family**.

### 4.2.1 Discriminative Power by Family

We measure discriminative power separately for continuous and binary
features. The standardised mean difference (SMD) only makes sense for
continuous features. For binary features the "pooled SD" collapses to a
function of the class proportion, so SMD stops being comparable across
features. For those we use **Cramér's V** instead — the standard
effect-size measure for a 2×2 contingency table, on the same 0–1 scale
as |correlation|.

```{r discriminative-power, fig.height=7, fig.width=8}
# --- Continuous features: standardised mean difference ----------------------
smd_df <- df %>%
  select(label, all_of(continuous_features)) %>%
  pivot_longer(-label, names_to = "feature", values_to = "value") %>%
  group_by(feature, label) %>%
  summarise(m = mean(value, na.rm = TRUE), s = sd(value, na.rm = TRUE),
            .groups = "drop") %>%
  pivot_wider(names_from = label, values_from = c(m, s),
              names_glue = "{.value}_{label}") %>%
  mutate(
    pooled_sd = sqrt((s_0^2 + s_1^2) / 2),
    effect    = abs(m_1 - m_0) / (pooled_sd + 1e-9),
    metric    = "SMD (continuous)"
  ) %>%
  select(feature, effect, metric)

# --- Binary features: Cramer's V on a 2x2 contingency table -----------------
cramers_v <- function(x, y) {
  tab <- table(x, y)
  if (any(dim(tab) < 2)) return(NA_real_)
  chi2 <- suppressWarnings(chisq.test(tab, correct = FALSE)$statistic)
  n <- sum(tab)
  as.numeric(sqrt(chi2 / (n * (min(dim(tab)) - 1))))
}

cv_df <- tibble(
  feature = binary_features,
  effect  = map_dbl(binary_features, ~ cramers_v(features[[.x]], df$label)),
  metric  = "Cramer's V (binary)"
)

discrim_df <- bind_rows(smd_df, cv_df) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  arrange(desc(effect))

ggplot(discrim_df, aes(x = reorder(feature, effect), y = effect, fill = family)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ metric, scales = "free", ncol = 2) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Per-feature discriminative power",
       subtitle = "Continuous: |SMD|. Binary: Cramer's V. Both >= 0.",
       x = NULL, y = "Effect size", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")
```

Effect-size distribution per family (continuous features only):

```{r smd-by-family, fig.height=4, fig.width=6}
smd_family <- discrim_df %>%
  filter(metric == "SMD (continuous)", family != "Other")

ggplot(smd_family, aes(family, effect, fill = family)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  geom_jitter(width = 0.15, alpha = 0.5, size = 2) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "SMD distribution by family (continuous features only)",
       x = NULL, y = "| SMD |") +
  theme_minimal(base_size = 12)
```

**Takeaway (H1):** Among the continuous features, Lexical has the lowest
median |SMD| and the smallest spread. Trust and Behavior carry more
per-feature signal. This supports H1's claim that the URL-only tier has less signal in individual features and relies more on feature combinations.

## 4.3 H2 preparation

H2 claims that on **FullLite** a linear projection preserves AUC at
small *k*. A quicklook at multicollinearity tells us how much redundant
dimensionality FullLite has to start with — the more collinear the
features, the fewer PCs we need to reach 95 % variance, and the stronger
the case that PCA-*k* can preserve AUC at small *k*. Lexical sits inside
FullLite, so the URL-length cluster below is the main source of
collinearity FullLite inherits; we keep Trust and Behavior out of this
panel because their features are largely binary and near-uncorrelated.

### 4.3.1 Lexical Correlation Matrix

```{r cor-lexical, fig.height=6, fig.width=7}
lexical_cor <- features %>%
  select(all_of(tier_lexical)) %>%
  cor(use = "pairwise.complete.obs")

  corrplot(
  lexical_cor,
  method  = "color",
  order   = "hclust",
  tl.cex  = 0.7,
  tl.col  = "black",
  col     = colorRampPalette(c("firebrick", "white", "steelblue"))(200),
  title   = "Correlation - Lexical features",
  mar     = c(0, 0, 1.5, 0),
  addCoef.col = "black",
  number.cex  = 0.55
)
```

Across the full feature matrix, 10 pairs pass `|r| > 0.7` — and all of
them sit inside this Lexical count cluster (URLLength ↔ NoOfLettersInURL
↔ NoOfDegitsInURL and friends).

### 4.3.2 Variance Inflation Factors

```{r vif}
library(knitr)

# Target-free VIF from the design matrix (avoids non-converged GLM).
X_vif <- features %>%
  select(all_of(continuous_features)) %>%
  as.data.frame()

vif_values <- vapply(seq_along(X_vif), function(i) {
  r2 <- summary(lm(X_vif[[i]] ~ ., data = X_vif[, -i, drop = FALSE]))$r.squared
  1 / (1 - r2)
}, numeric(1))
names(vif_values) <- names(X_vif)

vif_df <- tibble(
  feature = names(vif_values),
  vif     = as.numeric(vif_values)
) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  mutate(severity = case_when(
    vif > 10            ~ "severe (> 10)",
    vif >= 5 & vif <= 10 ~ "moderate (5-10)",
    TRUE                ~ "ok (< 5)"
  )) %>%
  arrange(desc(vif))

vif_df %>%
  transmute(Feature = feature,
            Family  = family,
            VIF     = round(vif, 2),
            Severity = severity) %>%
  kable(caption = "Variance Inflation Factors - continuous features (lm-based, target-free).")
```

**Takeaway (H2):** `URLLength`, `NoOfLettersInURL` and `NoOfDegitsInURL`
stay near-collinear because they measure overlapping properties of the
same URL string. Because Lexical sits inside FullLite, FullLite inherits
this collinearity bloc and therefore has fewer independent directions of
variance than it has features. That is the precondition for C2: PCA can
pack 95 % of FullLite's variance into a *k* well below 34. The open
question the §4.5 scree and per-PC AUC answer is whether the class
signal lives in those dominant variance directions or orthogonal to them
— C2 needs both (enough variance *and* class signal in the top *k* PCs).


## 4.4 Skewness Assessment *(Scenario 2 preparation)*

```{r outlier-skewness, fig.height=6, fig.width=8}
# Skewness for all continuous features
skewness_fn <- function(x) {
  x <- x[!is.na(x)]
  n <- length(x)
  m <- mean(x)
  s <- sd(x)
  (n / ((n-1)*(n-2))) * sum(((x - m) / s)^3)
}

skew_df <- tibble(
  feature  = continuous_features,
  skewness = map_dbl(continuous_features, ~ skewness_fn(features[[.x]]))
) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  mutate(abs_skew = abs(skewness)) %>%
  arrange(desc(abs_skew))

ggplot(skew_df, aes(x = reorder(feature, abs_skew), y = skewness, fill = family)) +
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = c(-2, 2), linetype = "dashed", colour = "grey40") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Skewness of continuous features",
       subtitle = "Dashed lines at +/- 2 mark heavy skew; LR/LDA may struggle here",
       x = NULL, y = "Skewness", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")

# Count features with |skewness| > 2
n_heavy_skew <- sum(skew_df$abs_skew > 2)
cat("\nFeatures with |skewness| > 2:", n_heavy_skew, "of",
    length(continuous_features), "\n")
cat("These are:\n")
skew_df %>% filter(abs_skew > 2) %>% select(feature, skewness, family) %>% print()
```

**Takeaway (Scenario 2):** Several count features are heavily right-skewed
with long outlier tails. This shapes our modeling choices:

- **Parametric models** (LR, LDA): need scaling and a log-transform on
  skewed features to make the distributional assumption hold.
- **Non-parametric models** (trees, KNN, SVM): robust to skew, but KNN and
  SVM still need standardisation because they use distances.


## 4.5 H2 preparation — where the class signal lives across principal components

§4.3 showed that FullLite inherits a heavy collinearity bloc from the
Lexical URL-length cluster, so it is low-rank in variance. That alone
only tells us about variance — not whether the dominant variance
directions carry the *class* signal. For H2 (C2) the crucial question
is whether the top few PCs of **FullLite** line up with phishing vs.
legitimate separation. If they don't, PCA-*k* will drop AUC even though
it captures 95 % of the variance. We compute the same two diagnostics
across all tiers (scree + per-PC AUC) to put FullLite's behaviour in
context — the cross-tier comparison also explains *why* FullLite is the
only meaningful tier to project: Trust and Behavior are too small to
compress, and Lexical's 13 features split their variance and class
signal across different axes.

### 4.5.1 Scree plot per tier

```{r pca-scree, fig.height=4.5, fig.width=8}
prep_for_pca <- function(cols) {
  X <- features[, cols, drop = FALSE]
  cont <- setdiff(names(X), binary_features)
  if (length(cont) > 0) {
    X[, cont] <- lapply(X[, cont], function(x) log1p(pmax(x, 0)))
  }
  X_mat <- as.matrix(X)
  sds <- apply(X_mat, 2, sd)
  X_mat <- X_mat[, sds > 0, drop = FALSE]
  scale(X_mat)
}

pca_tiers <- list(
  Lexical  = tier_lexical,
  Trust    = tier_trust,
  Behavior = tier_behavior,
  Full     = tier_full
)

scree_df <- map_dfr(names(pca_tiers), function(tier) {
  Xp <- prep_for_pca(pca_tiers[[tier]])
  pc <- prcomp(Xp, center = FALSE, scale. = FALSE)
  var_exp <- pc$sdev^2 / sum(pc$sdev^2)
  tibble(tier = tier, k = seq_along(var_exp), cum_var = cumsum(var_exp))
})

scree_df %>%
  mutate(tier = factor(tier, levels = c("Lexical", "Trust",
                                         "Behavior", "Full"))) %>%
  ggplot(aes(k, cum_var, colour = tier)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0.9, linetype = "dashed", colour = "grey40") +
  scale_x_continuous(breaks = seq(0, 40, 5)) +
  scale_y_continuous(labels = scales::percent) +
  scale_colour_brewer(palette = "Set2") +
  labs(title = "Cumulative variance explained by leading PCs",
       subtitle = "Dashed line at 90% variance — fewer PCs = tier is more low-rank",
       x = "Number of PCs retained", y = "Cumulative variance",
       colour = "Tier") +
  theme_minimal(base_size = 11)
```

Lexical reaches 90 % of its variance with 3–4 PCs — the URL-length
cluster is doing most of the work. Trust spends all 7 directions because
its features are binary and near-uncorrelated. Behavior and Full are
spread out and need 10+ PCs to reach 90 %. FullLite sits closer to the
Behavior/Full curve than to Lexical — it has enough independent
variance directions to make "how many PCs do we need" a real question,
and the 95 % variance cutoff that Scenario 5 uses for *k_pca* is
expected to land somewhere in that 10–20 range. The class picture below
tells us whether those axes are the right ones.

### 4.5.2 Class signal carried by each leading PC

```{r pca-per-pc-auc, fig.height=4, fig.width=9}
pca_auc_df <- map_dfr(names(pca_tiers), function(tier) {
  Xp <- prep_for_pca(pca_tiers[[tier]])
  pc <- prcomp(Xp, center = FALSE, scale. = FALSE)
  k_max <- min(10, ncol(pc$x))
  map_dfr(seq_len(k_max), function(i) {
    a <- suppressMessages(as.numeric(
      pROC::roc(df$label_bin, pc$x[, i],
                levels = c("Legitimate", "Phishing"),
                quiet = TRUE)$auc))
    tibble(tier = tier, PC = i, auc = max(a, 1 - a))
  })
})

pca_auc_df %>%
  mutate(tier = factor(tier, levels = c("Lexical", "Trust",
                                         "Behavior", "Full"))) %>%
  ggplot(aes(PC, auc, fill = tier)) +
  geom_col() +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey40") +
  facet_wrap(~ tier, nrow = 1) +
  scale_x_continuous(breaks = 1:10) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Phishing signal carried by each leading PC",
       subtitle = "Orientation-free univariate AUC; PC1 = direction of largest variance",
       x = "Principal component index", y = "Univariate AUC") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")
```

**Takeaway (H2):** On Full (and by inheritance FullLite) the top 2–3
PCs each reach AUC above 0.8, so a handful of components already
recover most of the task and the remaining PCs fill in the residual
signal. That is the direct EDA prediction for C2: PCA-*k* at 95 %
variance should preserve FullLite's AUC within the 0.01 threshold. The
Lexical panel — where PC1 captures almost all the variance but class
signal is smeared across several low-AUC PCs — is shown as context for
why Scenario 5 does not attempt projection on Lexical: variance and
class directions do not line up there, so a 3-PC summary would drop
AUC sharply. It also previews why, in §6 of `scenario_5.rmd`, KPCA is
expected to match rather than beat linear PCA on FullLite — the
non-linearities that KPCA could exploit live on Lexical, not here.

---

# 5. EDA Summary & Modeling Implications

**A note on separability.** After the Section 3.3 exclusions, 40 features
remain (13 Lexical + 7 Trust + 20 Behavior). On this reduced set the task
is near-perfectly separable once Behavior is present — six Behavior
counts have univariate AUC > 0.95 on their own, so a naive Full tier
saturates at AUC ≈ 1.0 and the parametric vs. non-parametric contrast
disappears.

For H1 we therefore use the scenario_2 family layout: Lexical
(13), Trust (7), Behavior with the six near-leakers stripped (14), and
**FullLite** (34) = Lexical + Trust + reduced Behavior. H1 is tested
primarily on the **URLOnly tier**, where the family contrast has the most
room to show, and the family gradient (C3) is measured against **FullLite**,
not Full. H2 is tested on **FullLite only**, because it is the single
tier with enough dimensions for projection to be a meaningful question
— the cross-tier EDA in §4.5 is retained as context for that choice.

## 5.1 Implications for modeling scenarios

| Finding | Scenario 2 implication (H1) | Scenario 5 implication (H2) |
|---------|----------------------------|------------------------------|
| Lexical SMD lowest + features must be combined | Largest parametric/non-parametric AUC gap expected on URLOnly | Explains why Scenario 5 does not project Lexical: a linear 3-PC summary would drop AUC sharply, and KPCA's non-linear advantage also lives here, not on FullLite |
| Heavy skewness in 19 of 22 continuous features | Log-transform + standardisation for LR/LDA; trees/RF unaffected | Apply the same log+standardise recipe before PCA / LDA / KPCA on FullLite (all three projections are scale-sensitive) |
| URL-length cluster: VIF > 1000 and abs(r) > 0.84 | Use stable solvers / regularised LR for fairness | FullLite inherits this cluster, so it is low-rank in variance — PCA's *k_pca* at 95 % variance is expected to land well below 34 (C2 precondition) |
| Top 2–3 PCs of Full each carry AUC > 0.8 (§4.5.2) | — | FullLite's dominant variance directions coincide with class-separation directions, so PCA-*k* should preserve AUC (C2) and LDA-1 should collapse the same signal into one axis (C1) |

