---
title: "Phishing Detection — Scenario 2: Parametric vs Non-parametric"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(caret)
library(glmnet)
library(MASS, exclude = "select")
library(klaR)
library(randomForest)
library(kernlab)
library(pROC)
library(knitr)
library(digest)

set.seed(2026)

ARTIFACTS <- "scenario_2/artifacts"
dir.create(ARTIFACTS, recursive = TRUE, showWarnings = FALSE)
```

---

# 1. Scenario & Hypothesis

**H1**: *On the URL-only tier, non-parametric models pick up
meaningfully more signal than parametric ones — and the gap is large
enough to matter for the corporate-proxy deployment.*

The H1 criteria are gated on the deployment-relevant operating-point
metric $\mathrm{minSS} = \min(\mathrm{Sens}, \mathrm{Spec})$ at
threshold 0.5 with a ΔAUC sanity check. Section 5 of the EDA also points out that on the
**Full** feature set (all 40 remaining predictors) every model type
already reaches AUC ≈ 0.99. The dataset is near-perfectly separable
there, so we can't see the parametric-vs-non-parametric contrast on
Full at all, and on saturated tiers AUC carries almost no signal even
when families still differ in Sens/Spec.

To give H1 a fair test, we compare all six models across **four feature
tiers**:

| Tier | Features | What the model sees                                                                                                                                                      |
|------|---------:|--------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| **Lexical** (URL-only) | 13 | URL string only — the first-line case: no DNS, no HTML fetch.                                                                                                            |
| **Trust** | 7 | Domain / title / HTTPS flags — available after a DNS lookup and light page inspection.                                                                                   |
| **Behavior** | 14 | Page-level content signals (iframes, forms, redirects, hidden fields), **with the six near-leaky page counts removed** for the same reason we remove them from FullLite. |
| **FullLite** | 34 | All three families combined minus six near-leakers. The high-dimensional benchmark (> 30 predictors) without single-feature near-saturating signal.                      |

Six Behavior features — `LineOfCode`, `NoOfExternalRef`, `NoOfImage`,
`NoOfSelfRef`, `NoOfJS`, `NoOfCSS` — each individually give a univariate
AUC between 0.958 and 0.990 on this dataset. Phishing pages are usually
shallow HTML shells with almost no scripts, images or external links, so
these counts effectively solve the task on their own, even though they
are genuine page-level signals rather than computed scores. Leaving them in any tier pins every model at AUC ≈ 1.0 and flattens the
family contrast we are trying to measure. So we exclude them both from
Behavior and from the `FullLite` tier.

The smaller tiers (Lexical, Trust, Behavior) are ablations that
isolate H1's claim, which is specifically about the URL-only regime. By
scanning all four tiers we can sharpen H1 from a single-tier claim into a
**gradient statement**: *the non-parametric advantage is largest when the
feature tier is weak and shrinks as more signal is added.*

**This notebook also covers Scenario 4 (surrogate tree visualisation).**
§8 picks Random Forest specifically — not because it wins H1 on
Lexical but because the surrogate exercise only makes sense for a teacher whose
own decision regions are
already axis-aligned tree splits, so a single `rpart` student has a
realistic chance of reproducing them. §8 reuses the RF fitted in §5.1
as a *teacher* for a single `rpart` surrogate — i.e. a tree fitted on
the RF's own predictions rather than on the ground-truth label. The
point is to render a human-readable diagram of the RF's decision logic
and quantify how faithfully a single tree reproduces the ensemble.
Scenario 4 lives in this notebook because the
surrogate's per-class fidelity (Sens vs RF / Spec vs RF) plugs into
the same operating-point framing H1 uses — keeping a single metric
vocabulary across both scenarios — and re-fitting RF elsewhere would
just duplicate the cache.

---

# 2. Data Loading

## 2.1 Load raw file
```{r load-data}
dataset_path <- if (file.exists("PhiUSIIL_Phishing_URL_Dataset.csv")) {
  "PhiUSIIL_Phishing_URL_Dataset.csv"
} else if (file.exists("../PhiUSIIL_Phishing_URL_Dataset.csv")) {
  "../PhiUSIIL_Phishing_URL_Dataset.csv"
} else stop("PhiUSIIL_Phishing_URL_Dataset.csv not found")

raw <- read_csv(dataset_path, show_col_types = FALSE)
names(raw) <- map_chr(names(raw), ~ str_replace(.x, "^\ufeff", ""))
cat("Loaded", nrow(raw), "rows,", ncol(raw), "columns.\n")
```

## 2.2 Apply EDA feature exclusions

The EDA (§3.3) dropped three groups of features. Repeating the same
exclusions here makes sure Scenario 2 is tested on this set of predictors.

```{r exclusions}
ID_COLS          <- c("FILENAME", "URL", "Domain", "TLD", "Title")
SCORES           <- c("URLSimilarityIndex", "TLDLegitimateProb", "URLCharProb",
                      "DomainTitleMatchScore", "URLTitleMatchScore")
REDUNDANT_BINARY <- "HasObfuscation"
RATIOS           <- c("LetterRatioInURL", "DegitRatioInURL",
                      "ObfuscationRatio", "SpacialCharRatioInURL")
EXCLUDE_COLS     <- c(ID_COLS, SCORES, REDUNDANT_BINARY, RATIOS)

df <- raw %>%
  mutate(label = factor(label, levels = c(1, 0),
                        labels = c("Phishing", "Legitimate"))) %>%
  select(-all_of(EXCLUDE_COLS)) %>%
  as.data.frame()

cat("Feature matrix after exclusions:", ncol(df) - 1, "predictors.\n")
```

## 2.3 Feature tiers
```{r family-map}
family_map <- tribble(
  ~feature,                     ~family,
  "URLLength",                  "Lexical",
  "DomainLength",               "Lexical",
  "IsDomainIP",                 "Lexical",
  "CharContinuationRate",       "Lexical",
  "TLDLength",                  "Lexical",
  "NoOfSubDomain",              "Lexical",
  "NoOfObfuscatedChar",         "Lexical",
  "NoOfLettersInURL",           "Lexical",
  "NoOfDegitsInURL",            "Lexical",
  "NoOfEqualsInURL",            "Lexical",
  "NoOfQMarkInURL",             "Lexical",
  "NoOfAmpersandInURL",         "Lexical",
  "NoOfOtherSpecialCharsInURL", "Lexical",
  "IsHTTPS",                    "Trust",
  "HasTitle",                   "Trust",
  "HasFavicon",                 "Trust",
  "Robots",                     "Trust",
  "Bank",                       "Trust",
  "Pay",                        "Trust",
  "Crypto",                     "Trust",
  "LineOfCode",                 "Behavior",
  "LargestLineLength",          "Behavior",
  "IsResponsive",               "Behavior",
  "NoOfURLRedirect",            "Behavior",
  "NoOfSelfRedirect",           "Behavior",
  "HasDescription",             "Behavior",
  "NoOfPopup",                  "Behavior",
  "NoOfiFrame",                 "Behavior",
  "HasExternalFormSubmit",      "Behavior",
  "HasSocialNet",               "Behavior",
  "HasSubmitButton",            "Behavior",
  "HasHiddenFields",            "Behavior",
  "HasPasswordField",           "Behavior",
  "HasCopyrightInfo",           "Behavior",
  "NoOfImage",                  "Behavior",
  "NoOfCSS",                    "Behavior",
  "NoOfJS",                     "Behavior",
  "NoOfSelfRef",                "Behavior",
  "NoOfEmptyRef",               "Behavior",
  "NoOfExternalRef",            "Behavior"
)

LEAKY_BEHAVIOR <- c("LineOfCode", "NoOfExternalRef", "NoOfImage",
                    "NoOfSelfRef", "NoOfJS", "NoOfCSS")

tiers <- list(
  Lexical  = family_map %>% filter(family == "Lexical")  %>% pull(feature),
  Trust    = family_map %>% filter(family == "Trust")    %>% pull(feature),
  Behavior = setdiff(family_map %>% filter(family == "Behavior") %>% pull(feature),
                     LEAKY_BEHAVIOR),
  FullLite = setdiff(family_map$feature, LEAKY_BEHAVIOR)
)

sapply(tiers, length)
```

---

# 3. Modelling Setup

## 3.1 Stratified sub-sampling

The full dataset has ~235 k rows. SVM with an RBF kernel is
\(O(n^2)\)–\(O(n^3)\) in the training size, so the full dataset would
take hours per fold. Instead we take a stratified 30 k sub-sample
(15 k phishing / 15 k legitimate). After the 80/20 split this leaves
24 k rows per fold.

```{r subsample}
N_SUB <- 30000
df_sub <- df %>%
  group_by(label) %>%
  slice_sample(n = N_SUB / 2) %>%
  ungroup() %>%
  as.data.frame()

table(df_sub$label)
```

## 3.2 Shared train/test split and CV folds

Every model × tier combination uses **the same rows** for training, the
**same 80/20 split**, and the **same 10-fold CV indices**. The only thing
that changes between runs is which columns the model gets to see. That
keeps every model-vs-model comparison directly paired at the fold level,
so per-fold AUC and CV ROC ± SD reported below are read against a
consistent baseline.

```{r splits}
idx       <- createDataPartition(df_sub$label, p = 0.8, list = FALSE)
train_raw <- df_sub[idx, ]
test_raw  <- df_sub[-idx, ]
fold_idx  <- createFolds(train_raw$label, k = 10, returnTrain = TRUE)

cat("Train:", nrow(train_raw), "Test:", nrow(test_raw),
    "CV folds:", length(fold_idx), "\n")
```

## 3.3 Preprocessing recipes

We need two parallel recipes. The first one (used by LR, LDA, NB, SVM-RBF
and KNN) runs `log1p()` on every continuous feature — the EDA flagged
\|skew\| > 2 on 19 of 22 continuous predictors — then centres and scales.
Binary features are left alone. The second one is for Random Forest and
applies no transform at all, since trees don't care about monotone
transforms or scale.

`preProcess` is fit once on the full training set (no per-fold refit).
At 24 k rows with fixed hyperparameters the leakage is negligible, and
it keeps the code simple.

```{r recipes}
feat <- df_sub %>% select(-label)
binary_features     <- names(feat)[sapply(feat, function(x)
                                   all(na.omit(x) %in% c(0, 1)))]
continuous_features <- setdiff(names(feat), binary_features)

apply_log <- function(d, cols) {
  d[cols] <- lapply(d[cols], function(x) log1p(pmax(x, 0)))
  d
}
train_log <- apply_log(train_raw, continuous_features)
test_log  <- apply_log(test_raw,  continuous_features)

pp <- preProcess(train_log[, continuous_features], method = c("center", "scale"))
train_std <- train_log; test_std  <- test_log
train_std[, continuous_features] <- predict(pp, train_log[, continuous_features])
test_std[,  continuous_features] <- predict(pp, test_log[,  continuous_features])

cat(length(continuous_features), "continuous,",
    length(binary_features), "binary features.\n")
```

## 3.4 Per-tier fit utility

Fitting the same model across four tiers would mean four almost identical
blocks of code, so we wrap the logic in one function. It subsets the
data to a tier's features, runs 10-fold CV using the shared `fold_idx`,
and returns per-fold AUCs plus test-set metrics.

```{r fit-utility}
fit_one_tier <- function(tier_features, data_train, data_test, method, tuneGrid = NULL, ...) {
  tr <- data_train[, c("label", tier_features)]
  te <- data_test[,  c("label", tier_features)]

  ctrl <- trainControl(
    method          = "cv",
    index           = fold_idx,
    classProbs      = TRUE,
    summaryFunction = twoClassSummary,
    savePredictions = "final"
  )
  t0 <- Sys.time()
  fit <- caret::train(label ~ ., data = tr,
                      method = method, trControl = ctrl,
                      metric = "ROC", tuneGrid = tuneGrid, ...)
  elapsed <- as.numeric(difftime(Sys.time(), t0, units = "secs"))

  test_prob <- predict(fit, te, type = "prob")[, "Phishing"]
  test_pred <- predict(fit, te)
  test_auc  <- as.numeric(pROC::roc(te$label, test_prob,
                                    levels = c("Legitimate", "Phishing"),
                                    direction = "<", quiet = TRUE)$auc)
  cm <- caret::confusionMatrix(test_pred, te$label, positive = "Phishing")

  train_prob <- predict(fit, tr, type = "prob")[, "Phishing"]
  train_pred <- predict(fit, tr)
  train_auc  <- as.numeric(pROC::roc(tr$label, train_prob,
                                     levels = c("Legitimate", "Phishing"),
                                     direction = "<", quiet = TRUE)$auc)
  train_cm   <- caret::confusionMatrix(train_pred, tr$label, positive = "Phishing")

  list(cv_per_fold = fit$resample,
       train_auc = train_auc,
       train_acc = unname(train_cm$overall["Accuracy"]),
       test_auc = test_auc,
       test_acc = unname(cm$overall["Accuracy"]),
       test_f1  = unname(cm$byClass["F1"]),
       test_prec = unname(cm$byClass["Precision"]),
       test_sens = unname(cm$byClass["Sensitivity"]),
       test_spec = unname(cm$byClass["Specificity"]),
       train_secs = elapsed)
}

tiers_fingerprint <- function(x)
  digest::digest(lapply(x[sort(names(x))], sort))

run_or_load <- function(name, fn) {
  path <- file.path(ARTIFACTS, paste0("res_", name, ".rds"))
  fp   <- tiers_fingerprint(tiers)
  if (file.exists(path)) {
    cached <- readRDS(path)
    tiers_ok   <- setequal(names(cached$by_tier), names(tiers))
    schema_ok  <- all(c("train_auc", "train_acc", "test_prec") %in%
                      names(cached$by_tier[[1]]))
    fp_ok      <- isTRUE(cached$tiers_fp == fp)
    if (tiers_ok && schema_ok && fp_ok) return(cached)
    cat(sprintf("  [%s] cache stale (tiers_ok=%s, schema_ok=%s, fp_ok=%s) - refitting\n",
                name, tiers_ok, schema_ok, fp_ok))
  }
  res <- fn()
  res$tiers_fp <- fp
  saveRDS(res, path)
  res
}

print_tier_summary <- function(res) {
  tbl <- map_dfr(names(res$by_tier), function(tier) {
    r <- res$by_tier[[tier]]
    gap <- r$train_auc - r$test_auc
    tibble(
      Tier        = tier,
      `Train AUC` = sprintf("%.4f", r$train_auc),
      `CV ROC`    = sprintf("%.4f +/- %.4f",
                            mean(r$cv_per_fold$ROC), sd(r$cv_per_fold$ROC)),
      `Test AUC`  = sprintf("%.4f", r$test_auc),
      `Gap`       = sprintf("%+.4f", gap),
      Accuracy    = sprintf("%.4f", r$test_acc),
      F1          = sprintf("%.4f", r$test_f1),
      Prec        = sprintf("%.4f", r$test_prec),
      Sens        = sprintf("%.4f", r$test_sens),
      Spec        = sprintf("%.4f", r$test_spec),
      `Train (s)` = sprintf("%.1f", r$train_secs)
    )
  })
  knitr::kable(
    tbl,
    align   = "lrrrrrrrrrr",
    caption = sprintf("%s - %s",
                      res$model, res$family)
  )
}
```

`print_tier_summary` returns a table with CV ROC ± SD, held-out test AUC,
Accuracy, F1, Sensitivity, Specificity, and wall-clock train time. Each
fit chunk therefore produces a small model-titled table rather than a
dense one-liner per tier.

---

# 4. Parametric Models

## 4.1 Logistic Regression (ridge)

EDA §4.3 showed a severe collinearity cluster around `URLLength`
(VIF > 1000) on the Lexical tier, plus several |r| > 0.7 pairs on the
full 40-feature set. Plain logistic regression becomes numerically
unstable in that setting — the coefficients swing wildly from fold to
fold. So we use **ridge-regularised** LR (α = 0, λ = 0.01). `α = 0`
makes it pure ridge (no L1 sparsity), so the coefficient set stays
fixed across folds and the family-level comparison stays apples-to-apples
. `λ = 0.01` is a small fixed shrinkage — large enough to stabilise the matrix
inversion against the VIF > 1000 collinearity block, small enough that signal
coefficients are not materially biased toward zero. We keep `λ` fixed at
this value across all four tiers so that every LR result is read against
the same regularisation regime.

```{r fit-lr}
res_lr <- run_or_load("lr", function() {
  out <- list(model = "LogReg-Ridge", family = "parametric", by_tier = list())
  for (tier in names(tiers)) {
    cat(sprintf("  LogReg-Ridge   [%-8s] fitting (%d features)...\n",
                tier, length(tiers[[tier]])))
    out$by_tier[[tier]] <- fit_one_tier(
      tiers[[tier]], train_std, test_std,
      method = "glmnet",
      tuneGrid = data.frame(alpha = 0, lambda = 0.01))
  }
  out
})
print_tier_summary(res_lr)
```

On **Lexical** LR-Ridge sits at AUC 0.852
with Sens 0.921 / Spec 0.716 — at threshold 0.5 the model blocks
roughly 28 % of legitimate URLs to keep 92 % of phish. For the
corporate proxy this is unusable filter regime.
From **Trust** onwards the Sens/Spec gap closes (≈ 0.86–0.99 across both
axes).

## 4.2 Linear Discriminant Analysis

LDA assumes Gaussian class-conditional densities with a shared covariance
matrix. The EDA flagged heavy skewness on 19 of 22 continuous features;
`log1p` + centre/scale in our preprocessing recipe cleans this up
reasonably well. LDA has no hyperparameters: it just fits the per-class means and a
pooled covariance on the standardised training set. That is exactly
why we include it — it isolates what the Gaussian-shared-Σ prior buys
on top of the LR baseline.

```{r fit-lda}
res_lda <- run_or_load("lda", function() {
  out <- list(model = "LDA", family = "parametric", by_tier = list())
  for (tier in names(tiers)) {
    cat(sprintf("  LDA            [%-8s] fitting (%d features)...\n",
                tier, length(tiers[[tier]])))
    out$by_tier[[tier]] <- fit_one_tier(
      tiers[[tier]], train_std, test_std, method = "lda")
  }
  out
})
print_tier_summary(res_lda)
```


On **Lexical** LDA posts AUC 0.931 — visibly better than LR, yet the
threshold-0.5 operating point is *worse*: Sens 0.971 / Spec 0.688, i.e.
~31 % of legitimate URLs blocked. AUC and the operating point disagree because
the two metrics depend on different parts of the LDA fit:
AUC only needs the discriminant to *rank* phish above legit, while the 0.5 cut
needs the Gaussian-shared-Σ posterior to be calibrated. On Lexical the URL-length features are still heavy-tailed after
`log1p`, and the two classes do **not** share the same covariance.
LDA can still order phish above legit (good AUC), but its posterior
estimate is off, so the 0.5 cut lands in the wrong place. On
**Trust** (0.860 / 0.861) and **FullLite** (0.989 / 0.994) Sens and
Spec are almost equal; **Behavior** (0.905 / 0.981) only keeps a small
skew toward Spec. So Lexical is the only tier where AUC and the
operating point really disagree. LDA beats LR on AUC, but at threshold
0.5 it is the worse Lexical filter.

## 4.3 Naive Bayes

NB assumes features are **independent** given the class. EDA §4.3
already showed this is not true here — |r| > 0.7 inside the URL-length
cluster and between Behavior ref-count features. We run NB anyway, to
see **how much the broken independence assumption costs** on each tier.
The hyperparameters are kept simple and the same across tiers:
`usekernel = TRUE` uses a kernel density per feature instead of a
Gaussian one, so a fold with a zero-variance binary feature does not
crash; `fL = 1` adds Laplace smoothing so unseen 0/1 combinations in a
small fold do not produce zero posteriors; `adjust = 1` keeps the
kernel bandwidth at klaR's default. We do not tune them — the goal is
to measure the cost of the independence assumption, not to find a
bandwidth that hides it.

```{r fit-nb, message=FALSE, warning=FALSE}
res_nb <- run_or_load("nb", function() {
  out <- list(model = "NaiveBayes", family = "parametric", by_tier = list())
  for (tier in names(tiers)) {
    cat(sprintf("  NaiveBayes     [%-8s] fitting (%d features)...\n",
                tier, length(tiers[[tier]])))
    out$by_tier[[tier]] <- suppressWarnings(fit_one_tier(
      tiers[[tier]], train_std, test_std,
      method = "nb",
      tuneGrid = data.frame(fL = 1, usekernel = TRUE, adjust = 1)))
  }
  out
})
print_tier_summary(res_nb)
```

NB is the case of AUC hiding operating-point pathology. On **Lexical**
AUC is 0.849 — not terrible — but Sens 0.999 / Spec **0.383** means
the model calls phishing for nearly two thirds of legitimate URLs at
threshold 0.5. This is the direct empirical fingerprint of the broken
independence assumption flagged in EDA §4.3: the URL-length cluster
contains multiple |r| > 0.7 pairs, so what is essentially one
correlated signal is multiplied into the posterior as several
"independent" pieces of evidence, and phish-consistent scores collapse
toward 1. On **Behavior** the same violation *flips direction* —
Sens 0.732 / Spec 0.996. AUC on Behavior is a respectable 0.990, yet
at the default threshold NB misses 27 % of phish. In our proxy terms:
on Lexical NB would block too much legit traffic, on Behavior it would
let too much phish through.

# 5. Non-parametric Models

## 5.1 Random Forest

Random Forest is the non-parametric counterpart to LR. It builds an
ensemble of trees with piecewise-constant decision regions and does
not care about monotone feature transforms. Two hyperparameters
matter for our setup; we fix both and use the same values across
tiers. `mtry = floor(sqrt(p))` is the standard classification default
— it limits the features each split can choose from, which keeps the
trees in the ensemble decorrelated. With our tier sizes this gives 4
on Lexical, 3 on Trust, 4 on Behavior and 6 on FullLite. `ntree = 300`:
AUC on this dataset stops improving at ~200–300 trees, so going higher
just costs compute. We also leave the trees fully grown (no `nodesize`
or `maxnodes` cap), so the ensemble's strength comes from tree depth
and the 0.5 threshold is set entirely by the bagging vote — that is
exactly the effect the commentary below picks apart.

```{r fit-rf}
res_rf <- run_or_load("rf", function() {
  out <- list(model = "RandomForest", family = "non-parametric", by_tier = list())
  for (tier in names(tiers)) {
    p    <- length(tiers[[tier]])
    mtry <- max(1, floor(sqrt(p)))
    cat(sprintf("  RandomForest   [%-8s] fitting (%d features, mtry=%d)...\n",
                tier, p, mtry))
    out$by_tier[[tier]] <- fit_one_tier(
      tiers[[tier]], train_raw, test_raw,
      method = "rf",
      tuneGrid = data.frame(mtry = mtry),
      ntree = 300)
  }
  out
})
print_tier_summary(res_rf)
```

On **Lexical** RF reaches AUC 0.969, but at threshold 0.5 it sits at
Sens 0.967 / Spec 0.745. Spec is the weak side: with fully grown
trees, leaves built on URL-only features can contain only a few
legitimate samples and still vote "phish" in the ensemble average,
which pushes the bagged probability above 0.5 for borderline legit
URLs. From **Trust** upward the model is symmetric — Behavior
0.972 / 0.978, FullLite 0.997 / 0.996 — because once trust-tier
flags are available, the splits are clean enough that the
smaller-class leaves no longer drive the vote. RF is therefore the model that would benefit most from a simple
**threshold shift** on Lexical: the AUC is already there, the 0.5
cut just is not the cost-optimal one. Unlike NB this is not a
model assumption, only a calibration issue. The standard fix is
**threshold tuning**. On Lexical the optimal threshold lands somewhere around
0.6–0.7. We do not apply it in §5.1's tables because the H1
evaluation is fixed at threshold 0.5 across every model.

## 5.2 SVM-RBF

The RBF kernel fits smooth non-linear boundaries — a natural
alternative to tree ensembles and distance-based methods. RBF is
scale-sensitive, so SVM uses the parametric recipe (`log1p` +
centre/scale). We fix both hyperparameters at kernlab's defaults for
standardised inputs and use the same values across tiers. `C = 1`
balances margin width against slack penalty: much larger C pushes
the fit toward a hard margin and overfits, much smaller C lets the
boundary slip past support vectors. `σ = 0.1` matches the kernel
width to unit-variance features, so pairwise kernel values stay in a
useful 0–1 range across our tier sizes.
We do not tune further — the §5.2 results show this default
already saturates AUC and Sens/Spec on every tier, so there is
nothing meaningful left for a grid search to improve.


```{r fit-svm}
res_svm <- run_or_load("svm", function() {
  out <- list(model = "SVM-RBF", family = "non-parametric", by_tier = list())
  for (tier in names(tiers)) {
    cat(sprintf("  SVM-RBF        [%-8s] fitting (%d features)...\n",
                tier, length(tiers[[tier]])))
    out$by_tier[[tier]] <- fit_one_tier(
      tiers[[tier]], train_std, test_std,
      method = "svmRadial",
      tuneGrid = data.frame(C = 1, sigma = 0.1))
  }
  out
})
print_tier_summary(res_svm)
```

SVM-RBF is the only model that is **proxy-deployable at threshold
0.5 without recalibration on every tier**. On **Lexical** it reaches
AUC 0.995 with Sens 0.985 / Spec 0.983 — a ~1.5 % miss rate and
~1.7 % false-block rate, symmetric in both directions. Trust,
Behavior and FullLite show the same symmetric profile. In EDA §1.1
terms, this is what "a non-parametric URL-only filter worth deploying"
looks like in numbers: roughly 9 × fewer errors than LR-Ridge on
Lexical (2 vs. 18 per 100 URLs), with no threshold tuning required.

## 5.3 KNN

KNN is the simplest non-parametric classifier. It stores the whole
training set and classifies each test point by a majority vote among
its `k` closest neighbours, using Euclidean distance. Because it is
distance-based, it uses the parametric recipe. We set `k = 25` and
keep it fixed across tiers. Two reasons: 25 neighbours give the
local class proportion a resolution of 1/25 = 4 percentage points —
fine enough to carry useful AUC signal, but not so fine that one
mislabelled neighbour flips the prediction and a smoother of this
size hides the 1-NN boundary without merging the two classes.

There is one practical twist. When a tier has **only binary features**
(Trust has 7 binary flags), many training points end up at identical
coordinates and caret's `knn3` refuses to run with "too many ties in
knn". We add a tiny Gaussian jitter (SD = 10⁻³, less than 0.1 % of
the feature scale) to break the ties without actually changing any
decisions.

```{r fit-knn}
jitter_features <- function(d, cols, sd = 1e-3) {
  for (c in cols) d[[c]] <- d[[c]] + rnorm(nrow(d), 0, sd)
  d
}

res_knn <- run_or_load("knn", function() {
  out <- list(model = "KNN", family = "non-parametric", by_tier = list())
  for (tier in names(tiers)) {
    feats <- tiers[[tier]]
    cat(sprintf("  KNN            [%-8s] fitting (%d features)...\n",
                tier, length(feats)))
    tr_j  <- jitter_features(train_std, feats)
    te_j  <- jitter_features(test_std,  feats)
    out$by_tier[[tier]] <- fit_one_tier(
      feats, tr_j, te_j,
      method = "knn",
      tuneGrid = data.frame(k = 25))
  }
  out
})
print_tier_summary(res_knn)
```

On **Lexical** it reaches AUC 0.987 with Sens 0.944 / Spec 0.923 —
symmetric, deployable, ~6–8 % error in either direction. On **Trust**
AUC stays at 0.932 but Accuracy drops to 0.862: with only 7
mostly-binary features many training points share the same
coordinates, the jittered distances barely separate the classes, and
the 25-NN vote averages over a near-uniform neighbourhood. Behavior
and FullLite return to the symmetric ≥ 0.96 profile. There is one
cost AUC does not show but the proxy needs to care about: KNN's
**inference cost grows with the training set** — every click would
trigger ~24 k distance computations, which is the opposite of the
"cheap before the page loads" constraint that motivated the URL-only
filter. SVM-RBF wins on both quality and inference cost.

# 6. Aggregate Results & H1 Verdict

## 6.1 Per-(model, tier) summary

```{r build-long}
all_results <- list(res_lr, res_lda, res_nb, res_rf, res_svm, res_knn)

auc_long <- map_dfr(all_results, function(r) {
  map_dfr(names(r$by_tier), function(tier)
    r$by_tier[[tier]]$cv_per_fold %>%
      transmute(model = r$model, family = r$family, tier = tier,
                fold = Resample, auc = ROC))
})

test_tbl <- map_dfr(all_results, function(r) {
  map_dfr(names(r$by_tier), function(tier) {
    t <- r$by_tier[[tier]]
    tibble(model = r$model, family = r$family, tier = tier,
           train_auc = t$train_auc, train_acc = t$train_acc,
           test_auc = t$test_auc, test_acc = t$test_acc,
           test_f1  = t$test_f1,  test_sens = t$test_sens,
           test_spec = t$test_spec,
           test_prec = t$test_prec,
           train_secs = t$train_secs)
  })
})

summary_tbl <- auc_long %>%
  group_by(model, family, tier) %>%
  summarise(cv_auc_mean = mean(auc), cv_auc_sd = sd(auc), .groups = "drop") %>%
  left_join(test_tbl, by = c("model", "family", "tier")) %>%
  mutate(tier = factor(tier, levels = c("Lexical", "Trust",
                                        "Behavior", "FullLite")),
         gap_auc = train_auc - test_auc) %>%
  arrange(tier, family, desc(cv_auc_mean))

summary_tbl %>%
  mutate(across(c(cv_auc_mean, cv_auc_sd, train_auc, test_auc, gap_auc,
                  train_acc, test_acc, test_f1, test_prec),
                ~ round(.x, 4)),
         train_secs = round(train_secs, 1)) %>%
  kable()
```


**How to read the table.** A large `Train AUC − Test AUC` gap is
expected for non-parametric models (RF/KNN/SVM-RBF can memorise their
training set) and is not a problem on its own — the real check is
**`CV ROC` vs. `Test AUC`**, both out-of-sample. The train column
is here only to show that CV numbers are not quietly inflated by
leakage.

### 6.1.1 Classification quality beyond AUC

AUC compares models by score ordering, which is useful when families
are far apart, but on the richer tiers every model sits at AUC ≈ 0.99
and the ranking becomes hard to read. More importantly, AUC is
**threshold-free** — it says nothing about what happens when the
proxy has to commit to a *block* or *allow* decision the moment the
user clicks the link.

So we add a second evaluation. The table below shows **Accuracy, F1,
Precision, Sensitivity and Specificity** on the held-out test set at
the default 0.5 threshold — the actual operating point of the proxy
filter.

In this proxy setting each side of the split matters on its own and
the two error types have **asymmetric cost**:

- **Sensitivity** = share of phishing URLs we caught,
- **Specificity** = share of legitimate URLs we did not flag,
- **Precision** = of the URLs we blocked, how many were actually phishing.

Low precision matters because the ratio of "good page blocked" to
"real phish caught" gets bad enough that users stop trusting the
proxy and start asking for bypass rules — at which point the whole
URL-only line of defence is gone.

```{r quality-table}
quality_tbl <- test_tbl %>%
  mutate(tier = factor(tier, levels = c("Lexical", "Trust",
                                        "Behavior", "FullLite"))) %>%
  arrange(tier, family, desc(test_f1)) %>%
  transmute(Model = model, Family = family, Tier = tier,
            Accuracy    = round(test_acc,  4),
            F1          = round(test_f1,   4),
            Precision   = round(test_prec, 4),
            Sensitivity = round(test_sens, 4),
            Specificity = round(test_spec, 4))

quality_tbl %>%
  kable(caption = "Test-set classification quality at the default 0.5 threshold.")
```

## 6.2 Per-tier deployment-metric gap (H1 C1, C2)

The H1 gating criteria — C1 ($\Delta$ minSS on Lexical) and C2 (gap
shrinks toward FullLite) — are read off the held-out test-set
Sens/Spec, where the operating point at threshold 0.5 actually matters
for the proxy. We pick the per-family champion on the deployment
metric itself, not on AUC: a model with a high AUC but a bad 0.5 cut
(LDA on Lexical: Spec 0.69) is not the model the proxy would deploy.
C3 (the ΔAUC sanity check) is read off the §6.1 summary table.

```{r h1-minss}
gap_tbl <- test_tbl %>%
  mutate(minSS = pmin(test_sens, test_spec)) %>%
  group_by(tier, family) %>%
  summarise(minSS = max(minSS),
            auc   = max(test_auc), .groups = "drop") %>%
  pivot_wider(names_from = family, values_from = c(minSS, auc)) %>%
  mutate(gap_minSS = `minSS_non-parametric` - minSS_parametric,
         gap_auc   = `auc_non-parametric`   - auc_parametric,
         tier = factor(tier, levels = c("Lexical", "Trust",
                                        "Behavior", "FullLite"))) %>%
  arrange(tier)

gap_tbl %>%
  mutate(across(where(is.numeric), ~ round(.x, 4))) %>%
  kable(caption = "Per-tier gap between best non-parametric and best parametric champion (selected on test-set minSS).")
```

### 6.2.1 Statistical sanity check — paired Wilcoxon on per-fold AUC

C1 / C2 are operational thresholds, not hypothesis tests, but it is
still worth confirming that the per-tier gaps are larger than
fold-level noise. For each tier we pick the best non-parametric and
best parametric champion (by test-set minSS, same selection rule as
above) and run a paired Wilcoxon signed-rank test on their 10
per-fold CV AUCs. The pairing is valid because every model in this
report was trained on the same `fold_idx` (§3.2). One-sided
alternative: non-parametric > parametric, in line with H1.

```{r h1-wilcoxon}
champions <- test_tbl %>%
  mutate(minSS = pmin(test_sens, test_spec)) %>%
  group_by(tier, family) %>%
  slice_max(minSS, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(tier, family, model)

wilcox_tbl <- map_dfr(unique(champions$tier), function(t) {
  par_model    <- champions %>%
    filter(tier == t, family == "parametric")     %>% pull(model)
  nonpar_model <- champions %>%
    filter(tier == t, family == "non-parametric") %>% pull(model)
  par_auc <- auc_long %>%
    filter(tier == t, model == par_model)    %>% arrange(fold) %>% pull(auc)
  nonpar_auc <- auc_long %>%
    filter(tier == t, model == nonpar_model) %>% arrange(fold) %>% pull(auc)
  test <- suppressWarnings(wilcox.test(nonpar_auc, par_auc,
                                       paired = TRUE,
                                       alternative = "greater"))
  tibble(
    Tier            = t,
    `Best non-par`  = nonpar_model,
    `Best param`    = par_model,
    `Median ΔAUC`   = median(nonpar_auc - par_auc),
    `Wilcoxon V`    = unname(test$statistic),
    `p-value`       = test$p.value
  )
}) %>%
  mutate(Tier = factor(Tier, levels = c("Lexical", "Trust",
                                        "Behavior", "FullLite"))) %>%
  arrange(Tier)

wilcox_tbl %>%
  mutate(`Median ΔAUC` = round(`Median ΔAUC`, 4),
         `p-value`     = format.pval(`p-value`, digits = 2, eps = 1e-4)) %>%
  kable(caption = "Paired Wilcoxon signed-rank test on per-fold CV AUC: best non-parametric vs. best parametric champion, one-sided.")
```

On **Lexical** and **Behavior** the test rejects equality at any
conventional significance level — the non-parametric advantage is
not a fold-level artefact. On **Trust** and **FullLite** p-values
are larger because the per-fold AUC differences are small
(saturation), which is consistent with C2: the family advantage
shrinks as the feature tier gets stronger. Note that with 10 paired
folds the smallest achievable one-sided p-value is `2⁻¹⁰ ≈ 0.001`,
so a "0.001" entry means *every single fold* favours the
non-parametric champion.

On **Lexical** the minSS gap is the largest
of any tier — SVM-RBF reaches minSS ≈ 0.98 while the best parametric
champion stays in the 0.69–0.92 range — comfortably above the C1
threshold of 0.10. On **FullLite** the gap collapses to zero: both
families saturate at minSS ≈ 0.99, which is exactly the C2 gradient
claim — *the deployment-relevant family difference shrinks as the
feature tier gets stronger.* Trust and Behavior land in between, with
Behavior the narrowest non-Lexical gap once the six near-leaky page
counts are removed. C3 (ΔAUC ≥ 0.02 on Lexical) is read off §6.1 and
holds with margin.

---

# 7. Summary — H1 gradient

**H1 verdict against the three criteria from `phishing.rmd` §2.**

- **C1 (Δ minSS on Lexical ≥ 0.10):** holds. SVM-RBF posts minSS 0.983
  on Lexical; the best parametric champion (LDA) tops out at minSS
  0.688 — Δ ≈ 0.295, an order of magnitude above the threshold.
- **C2 (gap shrinks toward FullLite):** holds. The Lexical minSS gap
  ≈ 0.30 collapses to ≈ 0.00 on FullLite where both families saturate
  near 0.99 — exactly the gradient the hypothesis predicts.
- **C3 (ΔAUC ≥ 0.02 on Lexical, sanity check):** holds. SVM 0.995 vs.
  LDA 0.931 → ΔAUC 0.064.

H1 is therefore **supported on all three criteria** — both gating
criteria (C1, C2) by a wide margin. The remaining narrative below explains *why*
each tier behaves the way it does.

1. **Non-parametric > parametric when the feature tier is weak.** On
   the **Lexical** tier the non-parametric advantage is the largest
   of any tier, and every single CV fold shows the same order. URL counts interact in non-linear
   ways (digit runs × length × subdomain depth), and linear or Gaussian
   decision boundaries just can't capture those interactions.
2. **Trust also favours non-parametric, but only barely.** With only
   seven features — most of them plain binary flags — there's little
   non-linear structure left for non-parametric models to discover.
3. **Behavior is the honest middle tier.** With the six near-leaky
   page counts (`LineOfCode`, `NoOfExternalRef`, `NoOfImage`,
   `NoOfSelfRef`, `NoOfJS`, `NoOfCSS`) stripped out — each of which
   individually gives univariate AUC ≥ 0.95 and single-handedly solves
   the task — the remaining 14 Behavior features are a realistic
   page-level signal without saturation. This is where we can read the
   family contrast cleanly at medium signal strength.

**Threshold-0.5 quality (Precision / Sensitivity / Specificity).** The
§6.1.1 table complements the AUC ranking with operating-point numbers.
Read per-model on the H1-critical Lexical tier:

- **LR-Ridge.** Lexical Sens 0.921 / Spec 0.716 — at the 0.5 cut the
  model would block ~28 % of legitimate URLs to keep ~92 % of phish.
  Unusable as a proxy filter; the imbalance closes from Trust onward.
- **LDA.** Lexical Sens 0.971 / Spec 0.688 — at the 0.5 cut the model
  would block ~31 % of legitimate URLs to keep ~97 % of phish, a
  *worse* operating point than LR despite a higher AUC (0.931 vs.
  0.852). The clearest case in the report where ranking and threshold
  disagree (see §4.2).
- **NB.** Two-faced. Lexical: Sens 0.999, Spec 0.383, Prec 0.618 — at
  the 0.5 cut the model would block ~62 % of legitimate URLs to keep
  essentially 100 % of phish: a high-recall / low-precision flood that
  calls phish for almost everything. Behavior: Sens 0.732, Prec 0.995
  — almost never calls phish, but is right when it does. Both
  directions are symptoms of the same broken independence assumption.
- **Random Forest.** Lexical Sens 0.967 / Spec 0.745 — at the 0.5 cut
  the model would block ~26 % of legitimate URLs to keep ~97 % of
  phish. Spec is the soft spot under fully grown trees; from Trust
  onward the model is symmetric (≥ 0.97 on both axes). Pure
  calibration issue, fixable by threshold tuning (§5.1).
- **SVM-RBF.** The only model that keeps all three of Prec, Sens and
  Spec at ≈ 0.98 simultaneously on Lexical — at the 0.5 cut it would
  block only ~2 % of legitimate URLs while keeping ~99 % of phish —
  and holds that symmetric profile across every other tier. No
  threshold recalibration needed.
- **KNN.** Lexical Sens 0.944 / Spec 0.923 — at the 0.5 cut the
  model would block ~8 % of legitimate URLs to keep ~94 % of phish:
  symmetric, deployable on quality grounds. Loses to SVM-RBF on
  inference cost: every proxy decision triggers ~24 k distance
  computations against the stored training set.

Across the six models, only SVM-RBF and KNN survive the
proxy-deployment criterion of *all three operating-point metrics
balanced* on Lexical. Every other model on Lexical trades one
metric for another at threshold 0.5. So the model we suggest for the deployment
into a corporate proxy is SVM-RBF.

**When we would revisit the choice.** If the proxy pipeline later gains
Behavior- or FullLite-grade signals and explainability becomes a
requirement, LR-Ridge or LDA on that richer tier
is a defensible downgrade — both reach AUC ≥ 0.99 there with balanced
Sens/Spec, and their coefficients are interpretable. But for the
URL-only first-line filter that our hypothesis is built around, SVM-RBF is
the only model whose numbers support the deployment directly.
---

# 8. Scenario 4 - Surrogate tree — visualising the RF decision rule

Random Forest is accurate but opaque — 300 unpruned bagged trees
vote per prediction, with no single diagram to point at — so Task 4
asks us to render a model-faithful visualisation alongside it and
quantify how much of its behaviour that visualisation captures. We do this with a **surrogate tree**: a single
`rpart` tree whose training target is **the RF's own predictions**, not
the ground-truth label. The tree is therefore a student learning what
the RF does, and its fidelity to the RF — not its accuracy against the
truth.

We run the exercise on two tiers: **Lexical** (where RF beats the
parametric models by ≥ 0.10 AUC and a tree-level explanation is most
valuable) and **FullLite** (where every model saturates near AUC ≈ 1.0
and fidelity should be trivially high — a sanity check that the tuning
procedure is unbiased). Trust and Behavior are omitted on purpose:
Trust is 7 features, most of them binary flags, so the only non-linear
structure it can carry is XOR-like binary interactions — which the
§6.2 numbers rule out. Behavior sits between the two cases and would
not add a new observation.

## 8.1 Teacher RF and surrogate utility

`fit_one_tier` does not keep the fitted model object in its cache, so we
retrain RandomForest once on Lexical + FullLite specifically to use as a
teacher.

```{r task4-setup, message=FALSE, warning=FALSE}
library(rpart)
has_rpart_plot <- requireNamespace("rpart.plot", quietly = TRUE)
if (has_rpart_plot) suppressPackageStartupMessages(library(rpart.plot))

refit_rf_for_surrogate <- function(selected_tiers = c("Lexical", "FullLite"),
                                   ntree = 300) {
  path <- file.path(ARTIFACTS, "rf_for_surrogate.rds")
  fp <- digest::digest(list(
    selected_tiers = selected_tiers,
    features       = lapply(tiers[selected_tiers], sort),
    ntree          = ntree,
    n_train        = nrow(train_raw)
  ))
  if (file.exists(path)) {
    cached <- readRDS(path)
    if (isTRUE(cached$fp == fp)) return(cached$by_tier)
  }
  cat("  [surrogate] refitting RF teacher on",
      paste(selected_tiers, collapse = ", "), "...\n")
  by_tier <- list()
  for (tier in selected_tiers) {
    feats <- tiers[[tier]]
    mtry  <- max(1, floor(sqrt(length(feats))))
    t0 <- Sys.time()
    fit <- randomForest(
      x     = train_raw[, feats, drop = FALSE],
      y     = train_raw$label,
      ntree = ntree,
      mtry  = mtry
    )
    cat(sprintf("    %-10s %d features, mtry=%d, %.1fs\n",
                tier, length(feats), mtry,
                as.numeric(difftime(Sys.time(), t0, units = "secs"))))
    by_tier[[tier]] <- fit
  }
  saveRDS(list(by_tier = by_tier, fp = fp), path)
  by_tier
}

rf_teachers <- refit_rf_for_surrogate()
```

## 8.2 Surrogate-tree grid search on fidelity

The grid is small by design. We tune the three knobs that actually
change how faithfully a `rpart` tree can reproduce a decision surface.
`maxdepth` bounds the tree's expressiveness, `cp` sets the pruning
threshold that decides which splits are retained, `minbucket` forces
leaves to be large enough to resist fitting noise in the teacher's
predictions.

**Fidelity** is the overall metric we optimise on the held-out test set:

  $$\text{fidelity} = \frac{1}{n_\text{test}} \sum_i \mathbf{1}\bigl[\hat{y}^{\text{tree}}_i = \hat{y}^{\text{RF}}_i\bigr].$$

The rest of this report (§6.1.1, §7) judges classifiers on
**Sensitivity / Specificity**, not on overall accuracy, because a
phishing-blocker's two error types have asymmetric cost. For the same
reason we decompose fidelity the same way: `Sens vs RF` is the share
of RF-called phishing that the tree also calls phishing, `Spec vs RF`
is the share of RF-called legit the tree also calls legit. These are
the two numbers that tell us *how the visualisation's agreement with
the model breaks down by class*. We also report tree and RF Sens/Spec against the ground
truth so the deployment-level operating point is visible.

```{r task4-grid}
tune_surrogate_tree <- function(tier, rf_fit,
                                maxdepths  = 3:7,
                                cps        = c(1e-4, 1e-3, 1e-2),
                                minbuckets = c(10, 30, 100)) {
  feats <- tiers[[tier]]
  tr_x  <- train_raw[, feats, drop = FALSE]
  te_x  <- test_raw[ , feats, drop = FALSE]

  teacher_train <- predict(rf_fit, tr_x)
  teacher_test  <- predict(rf_fit, te_x)
  truth_test    <- test_raw$label

  # RF reference Sens/Spec on the test set - constant across the grid.
  rf_sens <- mean(teacher_test[truth_test == "Phishing"]   == "Phishing")
  rf_spec <- mean(teacher_test[truth_test == "Legitimate"] == "Legitimate")

  tr_data <- tr_x
  tr_data$teacher_label <- teacher_train

  rows  <- list()
  trees <- list()
  for (md in maxdepths) for (cpv in cps) for (mb in minbuckets) {
    key <- sprintf("md=%d_cp=%.0e_mb=%d", md, cpv, mb)
    tree <- rpart(
      teacher_label ~ ., data = tr_data,
      method  = "class",
      model   = TRUE,
      control = rpart.control(maxdepth = md, cp = cpv,
                              minbucket = mb, xval = 0)
    )
    pr_class <- predict(tree, te_x, type = "class")
    pr_prob  <- predict(tree, te_x, type = "prob")[, "Phishing"]
    n_leaves <- sum(tree$frame$var == "<leaf>")

    # Overall agreement - Task 4's fidelity metric.
    fidelity <- mean(pr_class == teacher_test)

    # Per-class agreement: the Sens/Spec decomposition of fidelity,
    # treating the RF prediction as the tree's training target.
    sens_vs_rf <- mean(pr_class[teacher_test == "Phishing"]   == "Phishing")
    spec_vs_rf <- mean(pr_class[teacher_test == "Legitimate"] == "Legitimate")

    # Deployment-level operating point: tree vs. actual truth.
    tree_sens <- mean(pr_class[truth_test == "Phishing"]   == "Phishing")
    tree_spec <- mean(pr_class[truth_test == "Legitimate"] == "Legitimate")

    auc <- as.numeric(pROC::roc(truth_test, pr_prob,
                                levels = c("Legitimate", "Phishing"),
                                direction = "<", quiet = TRUE)$auc)
    rows[[key]] <- tibble(
      tier = tier, maxdepth = md, cp = cpv, minbucket = mb,
      leaves = n_leaves,
      fidelity   = fidelity,
      sens_vs_rf = sens_vs_rf, spec_vs_rf = spec_vs_rf,
      tree_sens  = tree_sens,  tree_spec  = tree_spec,
      rf_sens    = rf_sens,    rf_spec    = rf_spec,
      auc        = auc
    )
    trees[[key]] <- tree
  }
  list(results = bind_rows(rows), trees = trees,
       teacher_test = teacher_test, truth_test = truth_test,
       rf_sens = rf_sens, rf_spec = rf_spec)
}

load_surrogate <- function(tier, rf_fit) {
  path <- file.path(ARTIFACTS,
                    paste0("surrogate_", tolower(tier), ".rds"))
  fp <- digest::digest(list(
    tier       = tier,
    features   = sort(tiers[[tier]]),
    n_train    = nrow(train_raw),
    rf_ntree   = rf_fit$ntree,
    tree_store = "model=TRUE",
    schema     = "v2_sens_spec"
  ))
  if (file.exists(path)) {
    cached <- readRDS(path)
    if (isTRUE(cached$fp == fp)) return(cached$data)
  }
  cat(sprintf("  [surrogate-%s] tuning rpart grid...\n", tier))
  t0 <- Sys.time()
  data <- tune_surrogate_tree(tier, rf_fit)
  cat(sprintf("    done in %.1fs (%d combinations)\n",
              as.numeric(difftime(Sys.time(), t0, units = "secs")),
              nrow(data$results)))
  saveRDS(list(data = data, fp = fp), path)
  data
}

surr_lex  <- load_surrogate("Lexical",  rf_teachers[["Lexical"]])
surr_full <- load_surrogate("FullLite", rf_teachers[["FullLite"]])
all_surr  <- bind_rows(surr_lex$results, surr_full$results)
```

## 8.3 Fidelity vs. depth

The fidelity saturation curve — best fidelity per (tier, maxdepth)
across all (cp, minbucket) — tells us how much tree expressiveness is
needed before further depth stops buying alignment with the RF. RF
reference AUCs come from §5.1.

```{r task4-depth-curve}
rf_ref <- tibble(
  tier    = c("Lexical", "FullLite"),
  rf_auc  = c(
    res_rf$by_tier[["Lexical"]]$test_auc,
    res_rf$by_tier[["FullLite"]]$test_auc
  ),
  rf_sens = c(
    res_rf$by_tier[["Lexical"]]$test_sens,
    res_rf$by_tier[["FullLite"]]$test_sens
  ),
  rf_spec = c(
    res_rf$by_tier[["Lexical"]]$test_spec,
    res_rf$by_tier[["FullLite"]]$test_spec
  )
)

depth_tbl <- all_surr %>%
  group_by(tier, maxdepth) %>%
  slice_max(fidelity, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  left_join(rf_ref, by = "tier") %>%
  mutate(tier = factor(tier, levels = c("Lexical", "FullLite"))) %>%
  arrange(tier, maxdepth) %>%
  transmute(Tier = tier, MaxDepth = maxdepth, Leaves = leaves,
            Fidelity     = round(fidelity,   4),
            `Sens vs RF` = round(sens_vs_rf, 4),
            `Spec vs RF` = round(spec_vs_rf, 4),
            `Tree AUC`   = round(auc,        4),
            `RF AUC`     = round(rf_auc,     4))

kable(depth_tbl,
      caption = "Best surrogate per (tier, maxdepth) on the S6 test set. Fidelity = overall tree/RF agreement; Sens vs RF / Spec vs RF = its per-class decomposition (treating RF prediction as target). Winning (cp, minbucket) picked by fidelity; ties broken by fewer leaves.")
```

The §8.3 table reports the **fidelity ceiling** — the best rpart can do
at each depth, regardless of readability. The `Sens vs RF` / `Spec vs RF`
columns show *how* the remaining disagreement splits across the two
classes: a fidelity of 0.96 with Sens-vs-RF 0.97 and Spec-vs-RF 0.95
means the tree disagrees with the RF mostly on the legit side.

For the plotted diagrams in §8.4 we pick a different tree: the
surrogate with the highest fidelity under a readability cap of ≤ 15
leaves. A 37-leaf tree technically has a higher fidelity number, but
it is visually unparseable on a slide. The price for the readability cap is small: fidelity drops by ~0.009
on Lexical (from 0.967 with 37 leaves to 0.958 with 14 leaves) and
by ~0.003 on FullLite (from 0.990 to 0.987 with 15 leaves). The
deepest tree levels buy almost no extra agreement with the RF.

```{r task4-winners}
MAX_PLOT_LEAVES <- 15L

winners <- all_surr %>%
  filter(leaves <= MAX_PLOT_LEAVES) %>%
  group_by(tier) %>%
  slice_max(fidelity, n = 1, with_ties = TRUE) %>%
  slice_min(leaves,   n = 1, with_ties = FALSE) %>%
  ungroup()

winners %>%
  mutate(across(c(fidelity, sens_vs_rf, spec_vs_rf,
                  tree_sens, tree_spec, rf_sens, rf_spec, auc),
                ~ round(.x, 4))) %>%
  select(tier, maxdepth, cp, minbucket, leaves,
         fidelity, sens_vs_rf, spec_vs_rf,
         tree_sens, tree_spec, rf_sens, rf_spec, auc) %>%
  kable(caption = sprintf("Plotted surrogate tree per tier: highest fidelity subject to leaves <= %d. Ties broken by fewer leaves. `tree_sens`/`tree_spec` are vs. ground truth; `rf_sens`/`rf_spec` are the RF reference - the closer they are, the less the surrogate changes the operating point.", MAX_PLOT_LEAVES))
```

## 8.4 Winning trees

```{r task4-tree-lexical, fig.height=6, fig.width=10}
key_lex <- winners %>% filter(tier == "Lexical") %>%
  transmute(key = sprintf("md=%d_cp=%.0e_mb=%d", maxdepth, cp, minbucket)) %>%
  pull(key)
tree_lex <- surr_lex$trees[[key_lex]]

if (has_rpart_plot) {
  rpart.plot::rpart.plot(
    tree_lex, type = 2, extra = 104, fallen.leaves = TRUE,
    box.palette = c("#D73027", "#1A9850"),
    main = sprintf("Lexical surrogate (fidelity %.3f vs RF)",
                   winners$fidelity[winners$tier == "Lexical"])
  )
} else {
  plot(tree_lex, uniform = TRUE, margin = 0.12,
       main = "Lexical surrogate tree")
  text(tree_lex, use.n = TRUE, cex = 0.65)
}
```

```{r task4-tree-fulllite, fig.height=6, fig.width=10}
key_full <- winners %>% filter(tier == "FullLite") %>%
  transmute(key = sprintf("md=%d_cp=%.0e_mb=%d", maxdepth, cp, minbucket)) %>%
  pull(key)
tree_full <- surr_full$trees[[key_full]]

if (has_rpart_plot) {
  rpart.plot::rpart.plot(
    tree_full, type = 2, extra = 104, fallen.leaves = TRUE,
    box.palette = c("#D73027", "#1A9850"),
    main = sprintf("FullLite surrogate (fidelity %.3f vs RF)",
                   winners$fidelity[winners$tier == "FullLite"])
  )
} else {
  plot(tree_full, uniform = TRUE, margin = 0.12,
       main = "FullLite surrogate tree")
  text(tree_full, use.n = TRUE, cex = 0.65)
}
```

### RF variable importance cross-check

The surrogate's root split is a claim about what the RF thinks is the
most discriminative single feature. To see whether that claim is
faithful to the teacher — rather than an artefact of the one tree we
plotted — we read the RF's own variable importance for the same teacher
and compare the ordering against the tree's top few splits.

```{r task4-varimp, fig.height=5, fig.width=11}
# base-R title() goes through utf8towcs() which chokes on em-dashes
# in non-UTF-8 Windows locales - we keep the surrounding prose
# em-dashes intact but use a plain hyphen in the plot titles here.
op <- par(mfrow = c(1, 2), mar = c(4, 7, 3, 1))
randomForest::varImpPlot(
  rf_teachers[["Lexical"]],
  n.var = min(10, length(tiers[["Lexical"]])),
  main  = "RF variable importance - Lexical"
)
randomForest::varImpPlot(
  rf_teachers[["FullLite"]],
  n.var = 10,
  main  = "RF variable importance - FullLite (top 10)"
)
par(op)
```

If the feature at the tree's root is also in the top few rows of the
RF importance plot for the same tier, the surrogate isn't improvising
— it's picking up the same signal ranking the RF itself assigns. That
is the evidence that lets us read the tree diagram as a visualisation
of RF behaviour rather than a parallel classifier that happens to agree
on the test set.

## 8.5 What the surrogate does and doesn't capture

**What the surrogate visualises.** RF and the tree both work by
axis-aligned splits, and on each tier the surrogate's root variable
is also `varImpPlot`'s top-ranked one.

- **Lexical.** Root is `NoOfOtherSpecialCharsInURL < 3`. The
  high-special-char branch (~70 % of rows, mostly phish) refines
  through `NoOfDegitsInURL`, `NoOfSubDomain`, `TLDLength` and
  `CharContinuationRate`; the other branch splits on `URLLength`,
  then `NoOfSubDomain` and `NoOfLettersInURL`.
- **FullLite.** Root is `HasSocialNet = 1`. The top three layers are
  trust-tier binaries (`HasCopyrightInfo`, `HasDescription`, `IsHTTPS`,
  `HasSubmitButton`); URL counts only show up deeper, refining leaves
  the trust flags could not split. Once page-level flags are
  available, RF leans on them first and uses the URL features only as
  a tiebreaker.

So the diagram does reflect what RF actually keys on, not just a
second classifier that happens to agree on the test set.

**What the surrogate cannot capture.** RF averages 300 bagged trees,
each split seeing only `mtry = √p` features, which captures 3+ feature
interactions a single tree cannot. That cost is visible on Lexical:
Tree AUC 0.862 vs. RF AUC 0.969 (gap 0.107), and Sens drops 3.3 pp
(0.968 → 0.934) when the surrogate stands in for RF. Spec only loses
1.0 pp. On FullLite both shifts are ~1 pp because the RF advantage
over single-rule classifiers is gone there. So the Lexical surrogate
is a structural picture of RF, not a deployable replacement for it.

**Parameters and readability cap.** `maxdepth` is the binding lever:
below 3 the surrogate underfits on Lexical (fidelity ~0.85), past 5–6
it plateaus. `cp = 10⁻⁴` is permissive enough that depth, not pruning,
sets the ceiling, and `minbucket ≥ 10` keeps leaves stable against
teacher-label noise. The plotted trees in §8.4 add a ≤ 15-leaf cap so
the diagram stays readable. This costs ~0.009 fidelity on Lexical
and ~0.003 on FullLite.

**Per-class fidelity.** `Sens vs RF` and `Spec vs RF` sit within a
percent of each other on both tiers (Lexical 0.957 / 0.962, FullLite
0.987 / 0.985), so the surrogate isn't just tracking the easy class.
At depth 3 on Lexical the symmetry breaks — `Spec vs RF` falls to
~0.87 while `Sens vs RF` holds at ~0.98 — and that 11 pp gap closes
again by depth 6–7. Shallower diagrams therefore show only one side
of RF's behaviour.