---
title: "Phishing Detection - Scenario 3: Stepwise vs Lasso vs Elastic-Net"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(caret)
library(glmnet)
library(MASS, exclude = "select")
library(pROC)
library(knitr)
library(digest)

set.seed(2026)

ARTIFACTS <- "scenario_3/artifacts"
dir.create(ARTIFACTS, recursive = TRUE, showWarnings = FALSE)
```

---

# 1. Scenario & Hypothesis

Scenario 2 fixed the predictor pool and asked which **model family** extracts more signal. Scenario 3 holds a single linear classifier (logistic regression) constant and ask which **feature-selection method** keeps the most compact predictor set without losing predictive power.

We pick **bidirectional stepwise** (`MASS::stepAIC`, `direction = "both"`) for the algorithmic side because it reconsiders previously-included terms instead of locking decisions in  pure forward / backward variants. For the embedded pair we pick **Lasso (α = 1)** and **Elastic-Net (α = 0.5)**.

The primary classification task is the same first-line deployment case as the EDA and Scenario 2 motivation, the **Lexical** phishing detection. FullLite remains a secondary fallback benchmark so if URL-only is not enough and we have to inspect trust / behaviour signals as well, then we want the smallest combined-tier filter that preserves the operating quality.

**H2** *On the Lexical predictor pool, at least one feature-selection method can produce a compact deployable URL filter with strong held-out operating quality.*

| # | Criterion | Threshold |
|---|-----------|-----------|
| C1 | Sparsity | at least one FS method keeps **≤ 9 / 13** lexical predictors (≥30% reduction) |
| C2 | Ranking quality | held-out **AUC ≥ 0.95** |
| C3 | Operating point | held-out **Sensitivity ≥ 0.94** and **Specificity ≥ 0.75** at threshold 0.5 |

We evaluate this on the same 80/20 split as Scenario 2 by comparing the three selected feature-selection methods : bidirectional stepwise, lasso and elastic-net.

---

# 2. Data Loading

```{r load-data}
dataset_path <- if (file.exists("PhiUSIIL_Phishing_URL_Dataset.csv")) {
  "PhiUSIIL_Phishing_URL_Dataset.csv"
} else if (file.exists("../PhiUSIIL_Phishing_URL_Dataset.csv")) {
  "../PhiUSIIL_Phishing_URL_Dataset.csv"
} else stop("PhiUSIIL_Phishing_URL_Dataset.csv not found")

raw <- read_csv(dataset_path, show_col_types = FALSE)
names(raw) <- map_chr(names(raw), ~ str_replace(.x, "^﻿", ""))
cat("Loaded", nrow(raw), "rows,", ncol(raw), "columns.\n")
```

## 2.1 EDA exclusions and FullLite tier

Same exclusions as in Scenario 2.

```{r exclusions}
ID_COLS          <- c("FILENAME", "URL", "Domain", "TLD", "Title")
SCORES           <- c("URLSimilarityIndex", "TLDLegitimateProb", "URLCharProb",
                      "DomainTitleMatchScore", "URLTitleMatchScore")
REDUNDANT_BINARY <- "HasObfuscation"
RATIOS           <- c("LetterRatioInURL", "DegitRatioInURL",
                      "ObfuscationRatio", "SpacialCharRatioInURL")
LEAKY_BEHAVIOR   <- c("LineOfCode", "NoOfExternalRef", "NoOfImage",
                      "NoOfSelfRef", "NoOfJS", "NoOfCSS")
EXCLUDE_COLS     <- c(ID_COLS, SCORES, REDUNDANT_BINARY, RATIOS, LEAKY_BEHAVIOR)

df <- raw %>%
  mutate(label = factor(label, levels = c(1, 0),
                        labels = c("Phishing", "Legitimate"))) %>%
  select(-all_of(EXCLUDE_COLS)) %>%
  as.data.frame()

cat("Feature matrix after exclusions:", ncol(df) - 1, "predictors.\n")
```

---

# 3. Modelling Setup

## 3.1 Stratified sub-sampling and split

Identical to Scenario 2 3.1–3.2 for parity. The 30k stratified subsample (15k / 15k) is inherited from Scenario 2's SVM-driven budget. For Scenario 3 we keep it so the reduced-fit AUCs in 6.3 can be compared directly to LogReg-Ridge from Scenario 2 without a sample-size confound.

```{r subsample-split}
N_SUB <- 30000
df_sub <- df %>%
  group_by(label) %>%
  slice_sample(n = N_SUB / 2) %>%
  ungroup() %>%
  as.data.frame()

idx       <- createDataPartition(df_sub$label, p = 0.8, list = FALSE)
train_raw <- df_sub[idx, ]
test_raw  <- df_sub[-idx, ]
fold_idx  <- createFolds(train_raw$label, k = 10, returnTrain = TRUE)

cat("Train:", nrow(train_raw), "Test:", nrow(test_raw),
    "CV folds:", length(fold_idx), "\n")
```

## 3.2 Preprocessing
`log1p` on continuous predictors (the EDA flagged |skew| > 2 on most of them), then centre / scale. Binary indicators are left untouched.

```{r recipes}
feat <- df_sub %>% dplyr::select(-label)
binary_features     <- names(feat)[sapply(feat, function(x)
                                   all(na.omit(x) %in% c(0, 1)))]
continuous_features <- setdiff(names(feat), binary_features)

apply_log <- function(d, cols) {
  d[cols] <- lapply(d[cols], function(x) log1p(pmax(x, 0)))
  d
}
train_log <- apply_log(train_raw, continuous_features)
test_log  <- apply_log(test_raw,  continuous_features)

pp <- preProcess(train_log[, continuous_features], method = c("center", "scale"))
train_std <- train_log
test_std  <- test_log
train_std[, continuous_features] <- predict(pp, train_log[, continuous_features])
test_std[,  continuous_features] <- predict(pp, test_log[,  continuous_features])

predictors <- setdiff(names(train_std), "label")
lexical_predictors <- c("URLLength", "DomainLength", "IsDomainIP",
                        "CharContinuationRate", "TLDLength", "NoOfSubDomain",
                        "NoOfObfuscatedChar", "NoOfLettersInURL",
                        "NoOfDegitsInURL", "NoOfEqualsInURL", "NoOfQMarkInURL",
                        "NoOfAmpersandInURL", "NoOfOtherSpecialCharsInURL")
lexical_predictors <- intersect(lexical_predictors, predictors)
primary_predictors <- lexical_predictors

cat("FullLite pool:", length(predictors), "predictors (",
    length(continuous_features), "continuous,",
    length(binary_features), "binary).\n")
cat("Lexical URL-only pool:", length(lexical_predictors), "predictors.\n")
```

## 3.3 Per-fold cache helper

First knit fits everything and writes `scenario_3/artifacts/*.rds`, later knits just load the files.

```{r cache-helper}
fp_inputs <- digest::digest(list(
  predictors = sort(predictors),
  n_train    = nrow(train_std),
  n_folds    = length(fold_idx),
  fold_sizes = vapply(fold_idx, length, integer(1))
))

run_or_load <- function(name, fn) {
  path <- file.path(ARTIFACTS, paste0(name, ".rds"))
  if (file.exists(path)) {
    cached <- readRDS(path)
    if (isTRUE(cached$fp == fp_inputs)) return(cached)
    cat(sprintf("  [%s] cache stale - refitting\n", name))
  }
  res <- fn()
  res$fp <- fp_inputs
  saveRDS(res, path)
  res
}
```

---

# 4. Feature-selection methods

## 4.1 Bidirectional stepwise

`MASS::stepAIC` at each step considers every possible move, add an unused predictor or drop a current one, and picks the one that improves AIC the most. AIC rewards better fit but charges a small penalty for every extra predictor, so a feature only stays if it earns its keep. We start from the intercept only and grow toward the feature pool we pass in Lexical for the primary Scenario 3 analysis and FullLite for the fallback benchmark.

We use the **bidirectional** variant because the alternatives fail in this setting. Forward-only cannot undo early decisions, so a predictor that gets in first stays in even when later additions make it redundant. Backward-only starts from the full model, which is numerically unstable here because some URL-length features are heavily correlated.

```{r stepwise-helper}
step_keep <- function(model, AIC) {
  s <- tryCatch(summary(model)$coefficients, error = function(e) NULL)
  if (is.null(s) || nrow(s) == 0) {
    return(list(terms = character(0), p = numeric(0), AIC = AIC))
  }
  list(
    terms = rownames(s),
    p     = unname(s[, "Pr(>|z|)"]),
    AIC   = AIC
  )
}

fit_stepwise <- function(train_df, features = predictors) {
  null_mod <- glm(label ~ 1, data = train_df, family = binomial)
  full_fm  <- as.formula(paste("label ~", paste(features, collapse = " + ")))
  out <- MASS::stepAIC(
    null_mod,
    scope     = list(lower = ~ 1, upper = full_fm),
    direction = "both",
    trace     = FALSE,
    keep      = step_keep
  )
  out
}
```

## 4.2 Lasso (embedded, α = 1)

`cv.glmnet` tries a grid of regularisation strengths λ and uses an internal 5-fold cross-validation to score each. We pick `lambda.1se`, the largest λ whose CV error is still within one standard error of the best one. This is the standard parsimony-favouring choice. It gives up a tiny bit of fit in exchange for a smaller and more stable feature set than the strict best-fit `lambda.min` typically returns.

```{r lasso-helper}
fit_glmnet <- function(train_df, alpha) {
  X <- model.matrix(label ~ . - 1, data = train_df)
  y <- train_df$label
  cvfit <- cv.glmnet(X, y, family = "binomial", alpha = alpha,
                     type.measure = "deviance", nfolds = 5)
  cvfit
}

count_nonzero <- function(cvfit, s = "lambda.1se") {
  beta <- as.numeric(coef(cvfit, s = s))[-1]   # drop intercept
  sum(beta != 0)
}
```

## 4.3 Elastic-Net (embedded, α = 0.5)

Same wrapper as lasso, only with `alpha = 0.5`. We add elastic-net because lasso has a known weakness with correlated features. It picks one more or less at random and drops the rest. Elastic-net adds a small ridge penalty that stabilises the choice and tends to keep correlated features together. We fix α = 0.5 instead of tuning it, so the comparison stays about the selection method, not hyperparameters.

[//]: # Zvážiť tú poslednú vetu()

---

# 5. Primary Lexical URL-only H2 check

The primary objective in Scenario 3 is the URL-only version, because it matches what the
proxy can decide before loading the page. We fit the same three
feature-selection methods once on the 80% training split, then score them on the
held-out 20%. Besides AUC we explicitly report
**Sensitivity** and **Specificity** at threshold 0.5, because those are the
operational quantities set in previous scenario.

```{r lexical-final-check}
metrics_from_prob <- function(prob_phishing, truth = test_std$label) {
  pred <- factor(ifelse(prob_phishing >= 0.5, "Phishing", "Legitimate"),
                 levels = levels(truth))
  tp <- sum(pred == "Phishing" & truth == "Phishing")
  tn <- sum(pred == "Legitimate" & truth == "Legitimate")
  fp <- sum(pred == "Phishing" & truth == "Legitimate")
  fn <- sum(pred == "Legitimate" & truth == "Phishing")
  tibble(
    test_auc = as.numeric(pROC::roc(truth, prob_phishing,
                                    levels = c("Legitimate", "Phishing"),
                                    direction = "<", quiet = TRUE)$auc),
    sensitivity = tp / (tp + fn),
    specificity = tn / (tn + fp),
    accuracy = (tp + tn) / length(truth)
  )
}

predict_glm_phishing <- function(fit, test_df) {
  1 - as.numeric(predict(fit, newdata = test_df, type = "response"))
}

predict_glmnet_phishing <- function(cvfit, test_df, s = "lambda.1se") {
  X_test <- model.matrix(label ~ . - 1, data = test_df)
  1 - as.numeric(predict(cvfit, newx = X_test, s = s, type = "response"))
}

lexical_final <- run_or_load("lexical_final", function() {
  trn <- train_std[, c("label", lexical_predictors)]
  sw <- fit_stepwise(trn, lexical_predictors)
  cvl <- fit_glmnet(trn, alpha = 1)
  cve <- fit_glmnet(trn, alpha = 0.5)
  term_names <- colnames(model.matrix(label ~ . - 1, data = trn))
  list(
    stepwise_fit = sw,
    lasso_fit = cvl,
    en_fit = cve,
    sw_terms = setdiff(names(coef(sw)), "(Intercept)"),
    lasso_terms = term_names[as.numeric(coef(cvl, s = "lambda.1se"))[-1] != 0],
    en_terms = term_names[as.numeric(coef(cve, s = "lambda.1se"))[-1] != 0],
    X_names = term_names
  )
})

test_lexical <- test_std[, c("label", lexical_predictors), drop = FALSE]

SPARSITY_MAX_K <- 9
MIN_TEST_AUC <- 0.95
MIN_SENSITIVITY <- 0.94
MIN_SPECIFICITY <- 0.75

lexical_diagnostic <- tibble(
  pool = "Lexical URL-only",
  method = c("Stepwise", "Lasso", "Elastic-Net"),
  k = c(length(lexical_final$sw_terms),
        length(lexical_final$lasso_terms), length(lexical_final$en_terms)),
  prob_phishing = list(
    predict_glm_phishing(lexical_final$stepwise_fit, test_lexical),
    predict_glmnet_phishing(lexical_final$lasso_fit, test_lexical),
    predict_glmnet_phishing(lexical_final$en_fit, test_lexical)
  )
) %>%
  mutate(metrics = map(prob_phishing, metrics_from_prob)) %>%
  unnest(metrics) %>%
  dplyr::select(-prob_phishing) %>%
  mutate(
    k_reduction = length(primary_predictors) - k,
    C1_sparsity = k <= SPARSITY_MAX_K,
    C2_auc = test_auc >= MIN_TEST_AUC,
    C3_operating_point = sensitivity >= MIN_SENSITIVITY &
      specificity >= MIN_SPECIFICITY,
    H2_pass = C1_sparsity & C2_auc & C3_operating_point
  )

lexical_diagnostic %>%
  transmute(
    method,
    k,
    k_reduction,
    test_auc    = sprintf("%.4f", test_auc),
    sensitivity = sprintf("%.4f", sensitivity),
    specificity = sprintf("%.4f", specificity),
    accuracy    = sprintf("%.4f", accuracy),
    C1_sparsity,
    C2_auc,
    C3_operating_point,
    H2_pass
  ) %>%
  kable(caption = "Primary Task 3 result: final selected counts, held-out AUC/Sensitivity/Specificity, and H2 criteria (C1: k <= 9, C2: AUC >= 0.95, C3: Sens >= 0.94 and Spec >= 0.75).")
```

The URL-length features are heavily correlated, and both stepwise and lasso shrink the pool from 13 to **9 predictors** (a 31 % reduction). Elastic-net keeps 11 and so misses the C1 sparsity criterion. The smaller fits clear the operating-point thresholds: stepwise reaches **AUC 0.957, Sensitivity 0.949, Specificity 0.839**, and lasso lands at **AUC 0.951, Sensitivity 0.941, Specificity 0.764** - both pass all three H2 criteria, with stepwise the more comfortable margin on Specificity. Elastic-net (AUC 0.953, Sens 0.943, Spec 0.769) is operationally fine but fails the sparsity bound.

---

# 6. Lexical final fits and diagnostics
### 6.1 Selected predictors per method

A 13-row matrix with one column per method, marking which lexical URL predictors made it. Sorted by total times selected so the "core" set rises to the top.

```{r d1-table}
all_terms <- sort(unique(c(finals$sw_terms, finals$lasso_terms, finals$en_terms,
                           primary_predictors)))
d1 <- tibble(
  predictor = all_terms,
  stepwise    = ifelse(predictor %in% finals$sw_terms,    "x", ""),
  lasso       = ifelse(predictor %in% finals$lasso_terms, "x", ""),
  elastic_net = ifelse(predictor %in% finals$en_terms,    "x", "")
) %>%
  mutate(score = (stepwise == "x") + (lasso == "x") + (elastic_net == "x")) %>%
  arrange(desc(score), predictor)

d1 %>% dplyr::select(-score) %>%
  kable(caption = "Lexical URL-only predictors retained by each FS method.")
```


NEEDS CHECK
## 6.2 Significance loss as predictors are added or removed

### Stepwise: p-value evolution along the AIC path

`stepAIC` recorded a snapshot of `summary(model)$coefficients` at every step via the `keep` callback. We unpack it and watch how each predictor's two-sided z-test p-value moves as terms are added or dropped along the AIC path. Here **GLM** means **Generalized Linear Model**; concretely we use `glm(..., family = binomial)`, i.e. logistic regression for the binary phishing / legitimate label. The summary tables below are therefore not a performance curve - they are a per-feature audit of the selected logistic-regression model along the AIC path. For each predictor we report the first / last step at which it was active, its min / max / final p-value, and whether it ever crossed the descriptive 0.05 threshold.

```{r d2-stepwise}
keep_mat <- finals$stepwise_fit$keep
n_steps <- if (is.matrix(keep_mat)) ncol(keep_mat) else length(keep_mat)

step_path <- map_dfr(seq_len(n_steps), function(i) {
  k_i <- if (is.matrix(keep_mat)) keep_mat[, i] else keep_mat[[i]]
  tibble(step = i,
         predictor = k_i$terms,
         p = k_i$p,
         AIC = k_i$AIC)
}) %>%
  filter(predictor != "(Intercept)")

step_path_summary <- step_path %>%
  group_by(predictor) %>%
  arrange(step, .by_group = TRUE) %>%
  summarise(
    first_step = min(step),
    last_step = max(step),
    min_p = min(p, na.rm = TRUE),
    max_p = max(p, na.rm = TRUE),
    final_p = dplyr::last(p),
    crosses_05 = any(p > 0.05) && any(p < 0.05),
    .groups = "drop"
  ) %>%
  arrange(desc(crosses_05), desc(max_p), predictor)

step_path_summary %>%
  mutate(across(c(min_p, max_p, final_p), ~ signif(.x, 3))) %>%
  kable(caption = "D2 (stepwise): named p-value audit for predictors active along the AIC path. The crosses_05 column flags predictors whose Wald p-value moved across the 0.05 line at any step.")
```

In the rendered Lexical run no predictor has `crosses_05 = TRUE`: nothing moved
from significant to non-significant (or back) across the AIC path. That does not mean
every final AIC-selected term has a small Wald p-value. The late binary terms
`IsDomainIP` and `NoOfQMarkInURL` are useful enough to reduce AIC but have
unstable individual z-tests, so these p-values are reported descriptively rather
than as formal post-selection inference.

What we learn from this plot-and-table pair: the stable lexical signal is the URL
length/count/special-character cluster, which stays extremely far below 0.05.
The weaker late binary indicators are not the main statistical evidence; they
are small AIC improvements kept by the algorithm. This is exactly why we treat
D2 as an interpretability/audit deliverable, not as a new hypothesis test.

### Lasso / Elastic-Net: labelled embedded summaries

Embedded methods do not give per-step frequentist p-values - they give the
*order in which predictors enter or leave the active set as the penalty
changes*. The default `glmnet` coefficient-path plot is useful for debugging,
but it is a weak presentation figure here: 13 coloured lines overlap and the
colours do not identify the feature names. For the report we therefore use two
labelled summaries that carry the same information more directly:

1. the named non-zero coefficients at our final `lambda.1se`, and
2. the largest lambda at which each predictor is active.

If a coefficient is zero at `lambda.1se`, that feature is dropped by the
embedded method. A feature that is active already at a larger lambda is more
robust to regularisation; a feature that appears only at a very small lambda is
weaker or mostly redundant with stronger URL predictors.

```{r d2-embedded-coefs, fig.height=6, fig.width=10}
coef_at_lambda <- function(cvfit, label, s = "lambda.1se") {
  beta <- as.matrix(coef(cvfit, s = s))[-1, , drop = FALSE]
  tibble(
    predictor = rownames(beta),
    method = label,
    coefficient = as.numeric(beta[, 1]),
    retained = coefficient != 0
  )
}

embedded_coefs <- bind_rows(
  coef_at_lambda(finals$lasso_fit, "lasso"),
  coef_at_lambda(finals$en_fit,    "elastic-net")
)

embedded_coefs %>%
  filter(retained) %>%
  mutate(predictor = reorder(predictor, abs(coefficient))) %>%
  ggplot(aes(coefficient, predictor, fill = method)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ method, scales = "free_y") +
  geom_vline(xintercept = 0, colour = "grey50") +
  labs(
    title = "D2 (embedded): named coefficients retained at lambda.1se",
    subtitle = "Only non-zero coefficients are shown; zero means the feature was dropped",
    x = "Coefficient at lambda.1se",
    y = NULL
  ) +
  theme_minimal()
```

```{r d2-zero-lambda}
active_lambda_table <- function(cvfit, label) {
  beta_path <- as.matrix(coef(cvfit$glmnet.fit))[-1, , drop = FALSE]
  lambdas <- cvfit$glmnet.fit$lambda
  active_lambda <- apply(beta_path, 1, function(row) {
    nz <- which(row != 0)
    if (!length(nz)) return(NA_real_)
    max(lambdas[nz])
  })
  tibble(predictor = rownames(beta_path),
         active_lambda = active_lambda,
         method = label) %>%
    arrange(desc(active_lambda))
}

embedded_lambda <- bind_rows(
  active_lambda_table(finals$lasso_fit, "lasso"),
  active_lambda_table(finals$en_fit,    "elastic-net")
)

embedded_lambda %>%
  pivot_wider(names_from = method, values_from = active_lambda) %>%
  mutate(across(c(lasso, `elastic-net`),
                ~ ifelse(is.na(.x), "(never)", signif(.x, 3)))) %>%
  arrange(predictor) %>%
  kable(caption = "D2 (embedded): largest lambda at which each predictor is active. Larger = retained under stronger regularisation.")
```

```{r d2-embedded-lambda-plot, fig.height=6, fig.width=10}
embedded_lambda %>%
  filter(!is.na(active_lambda)) %>%
  mutate(predictor = reorder(predictor, active_lambda)) %>%
  ggplot(aes(active_lambda, predictor, colour = method)) +
  geom_point(size = 2.5, alpha = 0.9,
             position = position_dodge(width = 0.45)) +
  scale_x_log10() +
  labs(
    title = "D2 (embedded): feature robustness along the regularisation path",
    subtitle = "Further right = active even under stronger regularisation; absent = never active",
    x = "Largest lambda where coefficient is non-zero (log scale)",
    y = NULL,
    colour = "Method"
  ) +
  theme_minimal()
```

What we learn from the embedded paths: lasso is the more aggressive selector and
keeps the Lexical model compact at 9 predictors, while elastic-net keeps 11
because its ridge component tends to preserve groups of correlated URL features.
That is a useful robustness check rather than a contradiction: lasso answers
"which single sparse support is enough?", elastic-net answers "which correlated
URL signals remain useful when shrinkage is softened?".


---

# 7. Secondary FullLite fallback benchmark

The primary Lexical analysis is evaluated on a single 80/20 hold-out because H2 is defined operationally (absolute AUC / Sensitivity / Specificity thresholds), the 13-feature pool is small with a dominant URL-length signal, and the selected support is in practice deterministic - a fold distribution would only widen the interval around the result. On Fulllite we are not looking at operational quality (which is near-perfect for all three methods on this pool) but **how many features each method keeps and how stable that count is**. With 34 mixed-tier candidates and several correlated columns, a single fit gives one number per method. The 10-fold loop therefore exists to produce a *distribution of selected-predictor counts*.

For each of the 10 outer CV folds we fit all three methods on the in-fold
training rows and record only the count of selected predictors. This loop is a
selected-predictor-count diagnostic, not a fold-level performance estimate.

Here we intentionally reuse the Scenario 2 train-split preprocessing for parity,
and `glmnet` standardises columns internally again during fitting.

```{r perfold-fit}
perfold <- run_or_load("perfold", function() {
  k_step <- numeric(length(fold_idx))
  k_lasso <- numeric(length(fold_idx))
  k_en   <- numeric(length(fold_idx))
  step_terms_per_fold  <- vector("list", length(fold_idx))
  lasso_terms_per_fold <- vector("list", length(fold_idx))
  en_terms_per_fold    <- vector("list", length(fold_idx))

  t0 <- Sys.time()
  for (k in seq_along(fold_idx)) {
    trn <- train_std[fold_idx[[k]], c("label", predictors), drop = FALSE]

    sw   <- fit_stepwise(trn, predictors)
    sw_terms <- setdiff(names(coef(sw)), "(Intercept)")
    k_step[k] <- length(sw_terms)
    step_terms_per_fold[[k]] <- sw_terms

    cvl <- fit_glmnet(trn, alpha = 1)
    lasso_beta <- as.numeric(coef(cvl, s = "lambda.1se"))[-1]
    lasso_names <- colnames(model.matrix(label ~ . - 1, data = trn))
    lasso_terms <- lasso_names[lasso_beta != 0]
    k_lasso[k] <- length(lasso_terms)
    lasso_terms_per_fold[[k]] <- lasso_terms

    cve <- fit_glmnet(trn, alpha = 0.5)
    en_beta <- as.numeric(coef(cve, s = "lambda.1se"))[-1]
    en_terms <- lasso_names[en_beta != 0]
    k_en[k] <- length(en_terms)
    en_terms_per_fold[[k]] <- en_terms

    cat(sprintf("  fold %2d: step=%2d  lasso=%2d  EN=%2d\n",
                k, k_step[k], k_lasso[k], k_en[k]))
  }
  elapsed <- as.numeric(difftime(Sys.time(), t0, units = "secs"))

  list(k_step = k_step, k_lasso = k_lasso, k_en = k_en,
       step_terms = step_terms_per_fold,
       lasso_terms = lasso_terms_per_fold,
       en_terms = en_terms_per_fold,
       elapsed_secs = elapsed)
})

tibble(
  fold = seq_along(fold_idx),
  k_step  = perfold$k_step,
  k_lasso = perfold$k_lasso,
  k_en    = perfold$k_en
) %>%
  kable(caption = "Secondary FullLite fallback: selected-predictor count per CV fold (out of 34 candidates).")
```

## 7.1 FullLite lasso-vs-stepwise diagnostic

```{r fulllite-diagnostic}
wt <- wilcox.test(perfold$k_lasso, perfold$k_step,
                  paired = TRUE, alternative = "less", exact = FALSE)
wt_opposite <- wilcox.test(perfold$k_lasso, perfold$k_step,
                           paired = TRUE, alternative = "greater", exact = FALSE)
mean_step  <- mean(perfold$k_step)
mean_lasso <- mean(perfold$k_lasso)
mean_en    <- mean(perfold$k_en)
old_fulllite_expectation_supported <- mean_lasso < mean_step && wt$p.value < 0.05

tibble(
  metric = c("mean k (stepwise)", "mean k (lasso)", "mean k (elastic-net)",
              "Wilcoxon V", "p-value (diagnostic: lasso < step)",
             "descriptive p-value (opposite direction)",
              "old FullLite expectation"),
  value  = c(sprintf("%.1f", mean_step),
             sprintf("%.1f", mean_lasso),
             sprintf("%.1f", mean_en),
             sprintf("%.0f", unname(wt$statistic)),
             signif(wt$p.value, 3),
             signif(wt_opposite$p.value, 3),
             ifelse(old_fulllite_expectation_supported, "supported", "not supported"))
) %>% kable()
```

The old FullLite lasso-vs-stepwise expectation is not supported. In this run the observed direction is **stepwise smaller than lasso**: stepwise retains **17.6 predictors** on average across the 10 folds, lasso retains **26.2**, and elastic-net **28.0**. The opposite-direction p-value is shown only as a descriptive diagnostic, not as a replacement hypothesis.

```{r fulllite-perfold-plot, fig.height=4, fig.width=8}
perfold_long <- tibble(
  fold = rep(seq_along(fold_idx), 3),
  method = rep(c("stepwise", "lasso", "elastic-net"), each = length(fold_idx)),
  k = c(perfold$k_step, perfold$k_lasso, perfold$k_en)
) %>%
  mutate(method = factor(method, levels = c("stepwise", "elastic-net", "lasso")))

ggplot(perfold_long, aes(method, k, fill = method)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_point(position = position_jitter(width = 0.1, seed = 1), size = 1.4) +
  geom_hline(yintercept = length(predictors), linetype = "dashed", colour = "grey40") +
  annotate("text", x = 0.6, y = length(predictors) + 0.5,
           label = sprintf("FullLite pool = %d", length(predictors)),
           hjust = 0, size = 3.2, colour = "grey30") +
  labs(x = NULL, y = "selected predictors per fold",
       title = "Secondary FullLite fallback: per-fold predictor count (10 CV folds)") +
  theme_minimal() + theme(legend.position = "none")
```

## 7.2 Selection frequency across folds

The per-fold loop already stored the actual term lists, so we can ask the same question as 6.1 but in a more honest way: instead of reading off a single fit, we count for each FullLite predictor how many of the 10 folds each FS method retained it. Predictors selected in all 10 folds are the robust "core" of the fallback; predictors selected only occasionally are weaker or correlated stand-ins.

```{r fulllite-selection-frequency}
n_folds <- length(fold_idx)

count_in_folds <- function(term_lists, term) {
  sum(vapply(term_lists, function(tl) term %in% tl, logical(1)))
}

all_fulllite_terms <- sort(unique(c(
  unlist(perfold$step_terms),
  unlist(perfold$lasso_terms),
  unlist(perfold$en_terms),
  predictors
)))

d1_full <- tibble(
  predictor   = all_fulllite_terms,
  stepwise    = vapply(all_fulllite_terms, count_in_folds, integer(1),
                       term_lists = perfold$step_terms),
  lasso       = vapply(all_fulllite_terms, count_in_folds, integer(1),
                       term_lists = perfold$lasso_terms),
  elastic_net = vapply(all_fulllite_terms, count_in_folds, integer(1),
                       term_lists = perfold$en_terms)
) %>%
  mutate(score = stepwise + lasso + elastic_net) %>%
  arrange(desc(score), predictor) %>%
  dplyr::select(-score)

d1_full %>%
  kable(caption = sprintf(
    "FullLite predictors: number of CV folds (out of %d) in which each FS method retained each predictor. Higher = more robust to resampling.",
    n_folds))
```

---

# 8. Conclusion

The primary Lexical URL-only result supports the project story. Stepwise and lasso reduce the deployable URL filter from 13 predictors to **9 each** (elastic-net to 11), and the reduced fits meet the explicit AUC ≥ 0.95, Sensitivity ≥ 0.94 and Specificity ≥ 0.75 thresholds: stepwise (AUC 0.957 / Sens 0.949 / Spec 0.839) clears all three by a comfortable margin, lasso (0.951 / 0.941 / 0.764) clears them tightly.

The FullLite fold benchmark still answers a real fallback question with a negative lasso-vs-stepwise result: on this exact FullLite setup, **lasso is not the sparsest combined-tier fallback; stepwise is** (17.6 predictors per fold against 26.2 for lasso and 28.0 for elastic-net, out of a 34-predictor pool). The held-out test metrics on the full-train fits sit at AUC 0.9999 for every method, with stepwise reaching that ceiling on 17 predictors versus 26 for lasso and 29 for elastic-net. The regularised models still matter because they provide a stable embedded-selection baseline and a coefficient path, but the smallest fallback feature set in this experiment comes from the algorithmic AIC path. If a downstream stakeholder wants an audit log of which terms entered and left, stepwise is also the more interpretable deliverable. If the priority is embedded shrinkage with less post-selection instability, elastic-net/lasso remain useful but not more parsimonious here.

