---
title: "Phishing Detection — Scenario 3: Feature Selection"
author: "K-o-m-A"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(caret)
library(glmnet)
library(MASS, exclude = "select")
library(pROC)
library(knitr)

set.seed(2026)

ARTIFACTS <- "scenario_3/artifacts"
CACHE_VERSION <- "v3_phishing_second_level"
dir.create(ARTIFACTS, recursive = TRUE, showWarnings = FALSE)
```

---

# 1. Scenario & Hypothesis
## 1.1 Why Lexical

Scenario 2's §6.1 showed that **Lexical** (13 URL-string features) is
the only tier where classifier family matters — LR-Ridge sits at
0.861 while SVM-RBF reaches 0.997. On the other tiers the linear classifier already sits at
or near the ceiling, so every sensible selection method converges to
"keep 3–5 features, AUC = 1.0, done." That is not a finding — it is
saturation.

Lexical also inherits the EDA §4.3 finding of a **severe
multicollinearity cluster**: `URLLength`, `NoOfLettersInURL` and
`NoOfDegitsInURL` have VIF > 1000 and pairwise `|r| > 0.84`. That is
the textbook regime where ridge and lasso are known to behave
differently — ridge keeps the entire block with shared weight, lasso
sparsifies and re-distributes the bloc onto large
opposing-sign coefficients on the survivors. Any honest comparison
of the two should be anchored where that contrast is visible, not
hidden.

We therefore run S3 on **both Lexical and FullLite** as the two
contrast tiers H2 (§1.3) is built around: Lexical is the
"low-redundancy" pole where methods are expected to converge, and
FullLite is the "high-redundancy" pole where they are expected to
diverge. Trust and Behavior are intentionally omitted — Trust has
only seven features (no room for bloc-level redundancy to manifest)
and Behavior with the six near-leakers stripped is dimensionally
between the two; including either would only blur the contrast that
C1, C2 and C3 measure.

## 1.2 Method choice — one algorithmic + two embedded

The assignment requires **one** algorithmic and **two** embedded
methods. We pick **Forward selection (AIC) + Ridge + Lasso** — the
exact triplet referenced by EDA §2's H2 setup, so the criteria below
map 1-to-1 onto the H2 hypothesis. We deliberately drop Backward,
Stepwise and ElasticNet. The reasoning is empirical, not aesthetic:

- **Forward over Backward / Stepwise.** All three AIC procedures
  converge to the *same retained set* on Lexical (k = 9, Jaccard
  1.00) and the same set up to two features on FullLite. Reporting
  three near-identical procedures inflates the deliverable without
  adding information. Forward additionally produces the **per-step
  p-value trajectory** the assignment literally asks for ("…lose
  statistical significance as you **add** … predictors"); backward
  shows the symmetric story but the trajectory is messier.
- **Ridge + Lasso, drop ElasticNet.** Ridge (α = 0) and Lasso (α = 1)
  are the two endpoints of the regularisation axis — the sharpest
  contrast. ElasticNet at α = 0.5 is empirically interpolation: on
  Lexical its retained set differs from Lasso by one feature, its
  coefficients are within ~15 % of Lasso's (NoOfLettersInURL: Lasso
  11.0 vs EN 9.16), and its AUC matches Lasso to three decimals. EN
  would add a third row to every comparison table that says "in
  between — see Lasso". The one independent observation EN gives
  (better fold-stability than Lasso on FullLite, e.g. NoOfPopup) is
  noted in §3.6 prose without re-running the whole pipeline.
- **Note on Ridge vs Scenario 2.** Scenario 2 used the same `glmnet`
  ridge fit, but as a *classifier-family benchmark* — fixed λ = 0.01,
  no feature-selection lens, just an AUC number for the LR row. Here
  we use the same solver to ask a *different* question: how does
  ridge distribute weight across the multicollinear bloc, and how
  does that distribution compare to Lasso's. Same algorithm, two
  non-overlapping deliverables.

## 1.3 H2 (Scenario 3) — Predictor redundancy increases with tier richness

> *On the **Lexical** tier (13 features, no near-saturating signal,
> every predictor carries non-trivial unique information), embedded
> and algorithmic feature-selection methods converge on essentially
> the same retained set, and Wald p-values stay stable along the
> forward path. On **FullLite**, the same
> methods diverge in retained count and feature identity, and at
> least one feature flips Wald-significance as a correlated partner
> enters — because multiple equally-good subsets exist.*

| #  | Criterion                                                                                                  | Threshold                                                                            | Where     |
|----|------------------------------------------------------------------------------------------------------------|--------------------------------------------------------------------------------------|-----------|
| C1 | Difference in retained-feature count between Lasso and Forward                                             | $\lvert k_\text{Lasso} - k_\text{Forward}\rvert \le 1$ on Lexical, $\ge 3$ on FullLite | §6.1      |
| C2 | Jaccard overlap of retained sets between Lasso and Forward                                                 | $J \ge 0.85$ on Lexical, $J \le 0.70$ on FullLite                                    | §6.2      |
| C3 | Wald p-value flips along the Forward path (significant ↔ non-significant for the same predictor as later predictors enter) | $\ge 1$ flip on FullLite, $0$ flips on Lexical                       | §4.2.1    |

H2 is **supported** if at least two of {C1, C2, C3} hold in the
predicted direction *on both tiers*. If Lexical and FullLite behave
the same way (e.g. Jaccard collapses on both, or no flips on either),
H2 is **rejected** — the redundancy story would not match the data.

**Three things this hypothesis design pre-empts.**

- *Thresholds are directional, not absolute.* What H2 predicts is the
  **contrast** between the two tiers, not the specific cut-points.
  Jaccard $\approx 0.30$ on FullLite while Lexical sits at $0.92$
  supports H2 just as cleanly as our $0.85 / 0.70$ choice; the
  cut-points are picked conservatively far from the diagonal so
  natural sampling jitter cannot flip the verdict.
- *H2 is not a restatement of EDA §4.3.* EDA established that
  multicollinearity **exists** on FullLite (VIF, $|r|$). H2 asks the
  strictly stronger question — whether the selection methods
  **actually behave differently because of it**. The three methods
  could in principle converge on the same retained set even under
  redundancy if AIC, `lambda.1se` and the bloc structure all happened
  to favour the same representative. C1, C2 and C3 measure that
  downstream behaviour, not the multicollinearity itself.
- *Ridge does not participate in C1/C2/C3 — by design.* L2 has zero
  gradient at zero, so ridge never zeros a coefficient and its
  "retained set" is trivially the full tier. Ridge's job in S3 is the
  **coefficient-bloc contrast** in §3.4 (without it, Lasso's
  opposing-sign blow-up has nothing to be *opposing* against) and the
  flat 1.00 retention baseline in §3.6 against which Lasso's
  fold-to-fold $0.1 / 0.9$ swings read as instability rather than
  noise. §3.4 and §3.6 are descriptive material, not gating criteria.

---

# 2. Data Loading

Same recipe as Scenario 2 and 5 — identical sub-sample, split and CV
folds so the numbers here are directly comparable.

## 2.1 Load & exclusions

```{r load-data}
dataset_path <- if (file.exists("PhiUSIIL_Phishing_URL_Dataset.csv")) {
  "PhiUSIIL_Phishing_URL_Dataset.csv"
} else if (file.exists("../PhiUSIIL_Phishing_URL_Dataset.csv")) {
  "../PhiUSIIL_Phishing_URL_Dataset.csv"
} else stop("PhiUSIIL_Phishing_URL_Dataset.csv not found")

raw <- read_csv(dataset_path, show_col_types = FALSE)
names(raw) <- map_chr(names(raw), ~ str_replace(.x, "^\ufeff", ""))

ID_COLS          <- c("FILENAME", "URL", "Domain", "TLD", "Title")
SCORES           <- c("URLSimilarityIndex", "TLDLegitimateProb", "URLCharProb",
                      "DomainTitleMatchScore", "URLTitleMatchScore")
REDUNDANT_BINARY <- "HasObfuscation"
RATIOS           <- c("LetterRatioInURL", "DegitRatioInURL",
                      "ObfuscationRatio", "SpacialCharRatioInURL")
EXCLUDE_COLS     <- c(ID_COLS, SCORES, REDUNDANT_BINARY, RATIOS)

df <- raw %>%
  mutate(label = factor(label, levels = c(0, 1),
                        labels = c("Legitimate", "Phishing"))) %>%
  select(-all_of(EXCLUDE_COLS)) %>%
  as.data.frame()
# Note on factor order: glmnet / glm treat the *second* factor level as
# the success class. We deliberately put "Phishing" second so that
# predict(..., type = "response") returns P(Phishing) directly, and
# Sensitivity / Specificity / pROC are aligned with positive = "Phishing".
```

## 2.2 Lexical tier

```{r lexical}
lexical_features <- c("URLLength", "DomainLength", "IsDomainIP",
                      "CharContinuationRate", "TLDLength", "NoOfSubDomain",
                      "NoOfObfuscatedChar", "NoOfLettersInURL",
                      "NoOfDegitsInURL", "NoOfEqualsInURL",
                      "NoOfQMarkInURL", "NoOfAmpersandInURL",
                      "NoOfOtherSpecialCharsInURL")
stopifnot(all(lexical_features %in% names(df)))

LEAKY_BEHAVIOR <- c("LineOfCode", "NoOfExternalRef", "NoOfImage",
                    "NoOfSelfRef", "NoOfJS", "NoOfCSS")
fullLite_features <- setdiff(setdiff(names(df), "label"), LEAKY_BEHAVIOR)

cat("Lexical tier:", length(lexical_features), "features.\n")
cat("FullLite tier:", length(fullLite_features), "features.\n")
```

## 2.3 Subsample, split, folds, preprocessing

```{r prep}
N_SUB <- 30000
df_sub <- df %>%
  group_by(label) %>%
  slice_sample(n = N_SUB / 2) %>%
  ungroup() %>%
  as.data.frame()

idx       <- createDataPartition(df_sub$label, p = 0.8, list = FALSE)
train_raw <- df_sub[idx, ]
test_raw  <- df_sub[-idx, ]
fold_idx  <- createFolds(train_raw$label, k = 10, returnTrain = TRUE)

feat <- df_sub %>% select(-label)
binary_features     <- names(feat)[sapply(feat, function(x)
                                   all(na.omit(x) %in% c(0, 1)))]
continuous_features <- setdiff(names(feat), binary_features)

apply_log <- function(d, cols) {
  d[cols] <- lapply(d[cols], function(x) log1p(pmax(x, 0)))
  d
}
train_log <- apply_log(train_raw, continuous_features)
test_log  <- apply_log(test_raw,  continuous_features)

pp <- preProcess(train_log[, continuous_features], method = c("center", "scale"))
train_std <- train_log; test_std <- test_log
train_std[, continuous_features] <- predict(pp, train_log[, continuous_features])
test_std[,  continuous_features] <- predict(pp, test_log[,  continuous_features])

# Standardise the binary IsDomainIP too so glmnet / stepAIC see all
# predictors on one scale; glmnet's internal standardization would do
# this anyway but being explicit keeps coefficients directly readable.
pp_lex <- preProcess(train_std[, lexical_features], method = c("center", "scale"))
train_std[, lexical_features] <- predict(pp_lex, train_std[, lexical_features])
test_std[,  lexical_features] <- predict(pp_lex, test_std[,  lexical_features])

y_train <- train_std$label
y_test  <- test_std$label

tier_specs <- list(
  Lexical = lexical_features,
  FullLite = fullLite_features
)

X_train_by_tier <- map(tier_specs, ~ as.matrix(train_std[, .x]))
X_test_by_tier  <- map(tier_specs, ~ as.matrix(test_std[,  .x]))

```

## 2.4 Multicollinearity context — quick re-confirmation

EDA reported VIF > 1000 on the URL-length cluster. We recompute the
correlation block here so the rest of the notebook reads standalone.

```{r cor-check}
cor_pairs <- map_dfr(names(tier_specs), function(tier_name) {
  cor_mat <- cor(X_train_by_tier[[tier_name]])
  heavy_pairs <- which(abs(cor_mat) > 0.7 & row(cor_mat) < col(cor_mat),
                       arr.ind = TRUE)
  if (nrow(heavy_pairs) == 0) {
    return(tibble(tier = tier_name, a = NA_character_, b = NA_character_, r = NA_real_))
  }
  tibble(
    tier = tier_name,
    a = rownames(cor_mat)[heavy_pairs[, 1]],
    b = colnames(cor_mat)[heavy_pairs[, 2]],
    r = round(cor_mat[heavy_pairs], 3)
  )
})

cor_pairs %>%
  arrange(tier, desc(abs(r))) %>%
  kable(caption = "Feature pairs with |r| > 0.7 for Lexical and FullLite tiers.")
```

---

# 3. Embedded Methods — Ridge & Lasso

Both are `glmnet` fits with `family = "binomial"` and differ only in
`alpha` (0 = ridge, 1 = lasso). `lambda` is chosen by 10-fold CV, and
we report both `lambda.min` (optimal CV deviance) and `lambda.1se`
(sparsest model within one SE of the minimum — the canonical pick
for parsimony).

## 3.1 Fit the two paths

```{r enet-fit}
run_or_load <- function(name, fn) {
  path <- file.path(ARTIFACTS, paste0(name, "_", CACHE_VERSION, ".rds"))
  if (file.exists(path)) return(readRDS(path))
  res <- fn()
  saveRDS(res, path)
  res
}

fit_glmnet <- function(X_train_tier, alpha, tag, tier_name) {
  run_or_load(paste0("glmnet_", tolower(tier_name), "_", tag), function() {
    cv.glmnet(X_train_tier, y_train, family = "binomial",
              alpha = alpha, nfolds = 10, standardize = FALSE,
              type.measure = "auc",
              nlambda = 80,
              lambda.min.ratio = 1e-2,
              maxit = 1e6)
  })
}

cv_models <- map(names(tier_specs), function(tier_name) {
  X_tr <- X_train_by_tier[[tier_name]]
  list(
    Ridge = fit_glmnet(X_tr, 0.0, "ridge", tier_name),
    Lasso = fit_glmnet(X_tr, 1.0, "lasso", tier_name)
  )
}) %>% set_names(names(tier_specs))

```

## 3.2 Regularization paths — how features drop out

```{r reg-paths, fig.width=10, fig.height=3.6}
par(mfrow = c(2, 2))
for (tier_name in names(tier_specs)) {
  plot(cv_models[[tier_name]]$Ridge$glmnet.fit, xvar = "lambda", label = FALSE)
  title(sprintf("%s: Ridge (α=0)", tier_name), line = 2.5)
  plot(cv_models[[tier_name]]$Lasso$glmnet.fit, xvar = "lambda", label = FALSE)
  title(sprintf("%s: Lasso (α=1)", tier_name), line = 2.5)
}
par(mfrow = c(1, 1))
```

**Read:** the x-axis is `log(λ)`; at the right edge every coefficient
is shrunk to zero. Ridge keeps every line non-zero for any λ — only
their magnitudes shrink. Lasso sends lines to exactly zero as λ
grows; the order in which they hit zero is the algorithmic answer to
"which feature is dropped next".

## 3.3 Retained features at `lambda.1se`

```{r enet-retained}
coefs_at <- function(cvfit, s = "lambda.1se") {
  co <- as.matrix(coef(cvfit, s = s))
  data.frame(feature = rownames(co),
             coef    = as.numeric(co)) %>%
    filter(feature != "(Intercept)")
}

retained <- map_dfr(names(cv_models), function(tier_name) {
  bind_rows(
    coefs_at(cv_models[[tier_name]]$Ridge) %>% mutate(method = "Ridge", tier = tier_name),
    coefs_at(cv_models[[tier_name]]$Lasso) %>% mutate(method = "Lasso", tier = tier_name)
  )
}) %>% mutate(kept = abs(coef) > 1e-8)

retained %>%
  group_by(tier, method) %>%
  summarise(retained = sum(kept),
            total    = n(),
            lambda_1se = case_when(
              method[1] == "Ridge" ~ cv_models[[tier[1]]]$Ridge$lambda.1se,
              method[1] == "Lasso" ~ cv_models[[tier[1]]]$Lasso$lambda.1se),
            .groups = "drop") %>%
  kable(caption = "Retained features at lambda.1se for Lexical and FullLite (embedded).",
        digits = 4)
```

## 3.4 Coefficient table — who survived, who didn't

```{r enet-coef-table}
retained %>%
  select(tier, method, feature, coef) %>%
  pivot_wider(names_from = method, values_from = coef) %>%
  arrange(tier, desc(abs(Ridge))) %>%
  mutate(across(c(Ridge, Lasso), ~ round(.x, 3))) %>%
  kable(caption = "Standardised coefficients at lambda.1se by tier — zeros mean lasso dropped the feature.")
```

**Multicollinearity tell.** On Lexical, Ridge spreads small,
**same-sign** weight across the URL-length trio (URLLength ≈ 0.10,
NoOfLettersInURL ≈ 0.28, NoOfDegitsInURL ≈ 0.65). Lasso keeps all
three but blows them up into **opposing-sign** coefficients of
double-digit magnitude (URLLength ≈ −15, NoOfLettersInURL ≈ +11,
NoOfDegitsInURL ≈ +6) — the classical "huge cancellations" pattern
that flags an underdetermined system: each individual coefficient is
not interpretable in isolation, only the linear combination is. The
features lasso actually zeros are the low-signal ones
(NoOfObfuscatedChar, NoOfAmpersandInURL, occasionally NoOfEqualsInURL),
*not* one of the URL-length cluster — because dropping one of those
would forfeit the cancellation lasso is using to fit the data.

```{r metrics-helper}
classif_metrics <- function(y_true, prob, thr = 0.5) {
  pred <- factor(ifelse(prob >= thr, "Phishing", "Legitimate"),
                 levels = levels(y_true))
  cm <- caret::confusionMatrix(pred, y_true, positive = "Phishing")
  auc <- as.numeric(pROC::roc(y_true, prob,
                              levels = c("Legitimate", "Phishing"),
                              direction = "<", quiet = TRUE)$auc)
  tibble(
    auc  = auc,
    acc  = unname(cm$overall["Accuracy"]),
    f1   = unname(cm$byClass["F1"]),
    prec = unname(cm$byClass["Precision"]),
    sens = unname(cm$byClass["Sensitivity"]),
    spec = unname(cm$byClass["Specificity"])
  )
}
```

## 3.6 Stability under resampling

The assignment explicitly asks how feature selection behaves "across
folds", and this is the single clearest illustration of the redundancy
story C1/C2/C3 formalise. Lasso is known to be unstable under
correlated predictors: small perturbations of the training set flip
which member of a collinear block it picks. We quantify this over the
10 CV folds by recording which features each fold retains at its own
`lambda.1se`.

```{r stability}
stability_fit <- run_or_load("stability", function() {
  map_dfr(names(tier_specs), function(tier_name) {
    X_tier <- X_train_by_tier[[tier_name]]
    map_dfr(seq_along(fold_idx), function(i) {
      tr <- fold_idx[[i]]
      Xi <- X_tier[tr, ]; yi <- y_train[tr]
      map_dfr(list(Ridge = 0, Lasso = 1),
        function(a) {
          cvfit <- cv.glmnet(Xi, yi, family = "binomial",
                             alpha = a, nfolds = 5, standardize = FALSE,
                             type.measure = "auc",
                             nlambda = 80,
                             lambda.min.ratio = 1e-2,
                             maxit = 1e6)
          co <- as.matrix(coef(cvfit, s = "lambda.1se"))
          tibble(tier = tier_name,
                 fold = i,
                 feature = rownames(co)[-1],
                 coef    = as.numeric(co[-1, 1]))
        }, .id = "method")
    })
  })
})

stab_summary <- stability_fit %>%
  mutate(kept = abs(coef) > 1e-8) %>%
  group_by(tier, method, feature) %>%
  summarise(retention_rate = mean(kept), .groups = "drop") %>%
  pivot_wider(names_from = method, values_from = retention_rate) %>%
  arrange(tier, desc(Ridge))

stab_summary %>%
  mutate(across(where(is.numeric), ~ round(.x, 2))) %>%
  kable(caption = "Retention rate per feature across 10 CV folds for Lexical and FullLite (1.00 = kept every fold, 0.00 = never kept).")
```

**Read.** Ridge shows retention 1.00 everywhere — it never zeros
features. On Lexical the lasso column is very stable too: every
feature it ever picks it picks on every fold, and IsDomainIP
(retention 0.9) is the only borderline case. The signal in this tier
is concentrated enough that the small-perturbation flip mechanism
does not get triggered. **FullLite is where the pathology surfaces**:
HasFavicon = 0.9, IsResponsive = 0.1 across folds — these are the
lasso "coin-flip" features whose presence in the retained set is
run-dependent, exactly the case the literature warns against. The
full retention table also shows a long list of FullLite features
where Ridge sits at 1.00 while Lasso sits at 0.00 — a different
manifestation of the same picking instability, where lasso settles
for one member of a redundant pair and never revisits.

---

# 4. Algorithmic Selection — Forward (AIC)

`stats::step` walks a sequence of nested logistic regressions and
picks the one minimising AIC. Unlike embedded methods it gives us a
**per-step Wald p-value table** — at each step every coefficient has
a Wald p-value — which is exactly what the assignment asks us to
report under *"which features lose significance as you add or remove
predictors"*. We then discuss **why that p-value is not what it
appears to be** (§5).

We restrict the algorithmic part to **forward selection**. On this
dataset Forward, Backward and Stepwise converge to the same retained
set on Lexical (k = 9, identical features) and within two features
on FullLite, so reporting all three would be three near-identical
tables. The forward direction also matches the assignment's "as you
**add** predictors" framing more directly.

## 4.1 Forward selection — feature entry order & p-values

```{r forward}
step_models <- run_or_load("step_models", function() {
  map(names(tier_specs), function(tier_name) {
    feats <- tier_specs[[tier_name]]
    dtr <- train_std[, c("label", feats)]
    null_mod_t <- glm(label ~ 1, data = dtr, family = binomial)
    full_mod_t <- glm(label ~ ., data = dtr, family = binomial)
    list(
      features = feats,
      forward = step(null_mod_t,
                     scope = list(lower = null_mod_t, upper = full_mod_t),
                     direction = "forward", trace = FALSE, k = 2)
    )
  }) %>% set_names(names(tier_specs))
})

map_dfr(names(step_models), function(tier_name) {
  tibble(
    tier = tier_name,
    retained = length(coef(step_models[[tier_name]]$forward)) - 1,
    total = length(step_models[[tier_name]]$features)
  )
}) %>% kable(caption = "Forward selection retained-feature counts by tier.")

# Recover the entry order and per-step Wald p-values of the ENTERING
# feature and of every previously-entered feature. This is the
# "significance as predictors are added" table the assignment asks for.
fwd_trajectory <- run_or_load("forward_traj", function() {
  map_dfr(names(step_models), function(tier_name) {
    entered <- setdiff(names(coef(step_models[[tier_name]]$forward)), "(Intercept)")
    feats <- step_models[[tier_name]]$features
    map_dfr(seq_along(entered), function(k) {
      vars <- entered[1:k]
      m <- glm(reformulate(vars, "label"),
               data = train_std[, c("label", feats)],
               family = binomial)
      co <- summary(m)$coefficients
      tibble(tier = tier_name, step = k, entering = entered[k],
             feature = rownames(co), p = co[, "Pr(>|z|)"]) %>%
        filter(feature != "(Intercept)")
    })
  })
})

fwd_trajectory %>%
  mutate(p = signif(p, 3)) %>%
  pivot_wider(id_cols = c(tier, feature), names_from = step,
              values_from = p, names_prefix = "step_") %>%
  kable(caption = "Wald p-values as forward selection enters each feature for Lexical and FullLite.")
```

## 4.2 Final selected model — coefficients and Wald p-values

The last column of §4.1 is the final selected model's per-feature
p-value. Reading it directly:

```{r forward-final-coef}
map_dfr(names(step_models), function(tier_name) {
  as.data.frame(summary(step_models[[tier_name]]$forward)$coefficients) %>%
    rownames_to_column("feature") %>%
    filter(feature != "(Intercept)") %>%
    mutate(tier = tier_name)
}) %>%
  mutate(across(-c(feature, tier), ~ signif(.x, 3))) %>%
  kable(caption = "Final forward-AIC-selected model by tier — Wald p-values are biased by the selection process (see §5).")
```

On **Lexical** the final model retains nine features but two of them
have non-significant Wald p-values: IsDomainIP (p = 1.0, with a
standard error blown up to ≈ 1.2·10⁴ — the textbook signature of
quasi-separation) and NoOfQMarkInURL (p = 0.81). AIC kept these two
because adding them did not penalise log-likelihood enough to
overcome the −2·k term, *not* because there is independent statistical
evidence for them. On **FullLite** the analogous behaviour appears for
IsHTTPS (p ≈ 0.98 across all steps, kept by AIC throughout).

## 4.2.1 Wald p-value flips along the forward path (H2 C3)

C3 asks for a **count** of features whose Wald p-value crosses the
0.05 threshold along the forward trajectory: a predictor that was
significant at the step it entered, and is no longer significant in
the final selected model (or vice versa). Each flip is a direct
empirical fingerprint of multicollinearity-driven instability — the
predictor's "evidence" is not a property of the predictor, it is a
property of the predictor *given the company it keeps*.

```{r p-flip-table}
fwd_flips <- fwd_trajectory %>%
  group_by(tier, feature) %>%
  arrange(step, .by_group = TRUE) %>%
  summarise(
    p_at_entry = first(p),
    p_in_final = last(p),
    sig_at_entry = first(p) < 0.05,
    sig_in_final = last(p) < 0.05,
    flipped = sig_at_entry != sig_in_final,
    .groups = "drop"
  ) %>%
  arrange(tier, desc(flipped), feature)

fwd_flips %>%
  mutate(across(c(p_at_entry, p_in_final), ~ signif(.x, 3))) %>%
  kable(caption = "Forward-path Wald p-values for every entered predictor: at the step it entered vs. in the final selected model. flipped = TRUE iff significance status (p < 0.05) changed.")

flip_counts <- fwd_flips %>%
  group_by(tier) %>%
  summarise(n_flipped = sum(flipped),
            features_flipped = paste(feature[flipped], collapse = ", "),
            .groups = "drop")

flip_counts %>%
  kable(caption = "C3 deliverable — number of significance flips per tier.")
```

C3 is supported iff `n_flipped == 0` on Lexical and `n_flipped >= 1`
on FullLite. Read the verdict in §7.

---

# 5. Post-Selection Inference — the research piece

The Wald p-values reported in §4.1 and §4.2 are **not valid**
out-of-the-box. This is the single most important thing to say about
stepwise selection, and the single most common thing that gets
forgotten in practice.

## 5.1 Why the p-values are biased

A classical Wald p-value assumes the set of predictors was fixed
*before* seeing the data. Stepwise selection violates this
fundamentally: the set of predictors *is chosen from the data* to
minimise AIC (or maximise likelihood). The p-values reported by the
final selected model are conditional on having been selected — which
means:

- Selected features are more likely to appear significant than they
  truly are (the winner's-curse effect).
- The null distribution of the Wald statistic is no longer a χ²; the
  standard formula overstates significance.
- Published critiques (Berk et al. 2013; Lee et al. 2016; Tibshirani
  et al. 2016) argue for *selective inference* — p-values computed
  *conditional* on the selection event. Tools: the `selectiveInference`
  R package for the lasso, and post-selection splitting (Cox 1975;
  Wasserman & Roeder 2009).

**Practical consequence for this notebook:** the final-model
p-values in §4.2 should **not** be read as "these are the significant
features". They are, strictly, *the p-values a naive analyst would
quote*. The right way to phrase a finding is: "the forward AIC
procedure retained 9 features on Lexical; standard Wald p-values on
those features are given but are biased by the selection process and
should not be used for hypothesis tests."

## 5.2 Why the forward-trajectory p-values *are* useful — for C2

Although the *final* p-values are biased, the **trajectory** in
§4.1 answers a different question: given this specific sequence of
models (entry order fixed by AIC), does any feature's p-value
*rise* above 0.05 when a correlated partner is added in a later
step? That trajectory is a direct demonstration of the
multicollinearity mechanism — and *that* is a valid descriptive
observation, not a hypothesis test.

On **Lexical** the trajectory does **not** show this pathology — every
feature that enters keeps p well below 0.05 to the end (the URL-length
trio survives because the signal is large enough to overpower
shrinkage of the standard error). Two features are entered late by AIC
but never reach significance at all (IsDomainIP, NoOfQMarkInURL); they
illustrate the *other* form of the problem (§4.2) — features kept by
selection criteria with no Wald evidence behind them.

On **FullLite** the C2 mechanism is clearly visible:
NoOfOtherSpecialCharsInURL goes from p ≈ 4·10⁻²⁵¹ at step 3 to
p ≈ 0.81 at step 19, and NoOfSubDomain from p ≈ 1.9·10⁻⁸⁹ to p ≈ 0.68
— both as their correlated partners (URLLength, NoOfLettersInURL,
NoOfDegitsInURL, DomainLength) progressively absorb the shared signal.

## 5.3 Why lasso stability (§3.6) is the embedded-method answer

The embedded counterpart of the "stepwise p-value is biased" problem
is the **instability of sparse selection under correlated
predictors**. Lasso will pick one feature from a block and zero the
rest, but *which* feature it picks is a small-perturbation-away from
flipping. The §3.6 stability matrix makes this visible — features
with retention 0.4–0.6 across folds are the same features that, in
an algorithmic setting, would have unstable p-values. Different
machinery, same underlying pathology.

---

# 6. Comparison — Retained Sets Side by Side

## 6.1 The deliverable table (assignment's literal asks)

```{r retained-compare}
retained_by_method <- map_dfr(names(tier_specs), function(tier_name) {
  bind_rows(
    tibble(method = "Ridge",         retained = sum(abs(coef(cv_models[[tier_name]]$Ridge, s = "lambda.1se"))[-1] > 1e-8)),
    tibble(method = "Lasso",         retained = sum(abs(coef(cv_models[[tier_name]]$Lasso, s = "lambda.1se"))[-1] > 1e-8)),
    tibble(method = "Forward (AIC)", retained = length(coef(step_models[[tier_name]]$forward)) - 1)
  ) %>%
    mutate(tier = tier_name,
           total = length(tier_specs[[tier_name]]))
})

kable(retained_by_method,
      caption = "How many features each method retains in Lexical and FullLite.")
```

## 6.2 Feature-set overlap (H2 C2)

```{r overlap}
get_kept <- function(cvfit) {
  co <- as.matrix(coef(cvfit, s = "lambda.1se"))
  rownames(co)[-1][abs(co[-1, 1]) > 1e-8]
}

sets_by_tier <- map(names(tier_specs), function(tier_name) {
  list(
    Ridge   = get_kept(cv_models[[tier_name]]$Ridge),
    Lasso   = get_kept(cv_models[[tier_name]]$Lasso),
    Forward = setdiff(names(coef(step_models[[tier_name]]$forward)), "(Intercept)")
  )
}) %>% set_names(names(tier_specs))

jaccard <- function(a, b) length(intersect(a, b)) / length(union(a, b))

map_dfr(names(sets_by_tier), function(tier_name) {
  sets <- sets_by_tier[[tier_name]]
  jac_mat <- outer(names(sets), names(sets),
                   Vectorize(function(i, j) jaccard(sets[[i]], sets[[j]])))
  dimnames(jac_mat) <- list(names(sets), names(sets))
  round(jac_mat, 2) %>%
    as.data.frame() %>%
    rownames_to_column("method") %>%
    mutate(tier = tier_name)
}) %>%
  select(tier, method, everything()) %>%
  kable(caption = "Jaccard overlap of retained-feature sets by tier. Off-diagonal < 1 = disagreement.")

# H2 C2 deliverable — pull out the single Lasso × Forward number per tier
# and compare against the directional thresholds from §1.3.
lasso_fwd_jac <- map_dfr(names(sets_by_tier), function(tier_name) {
  s <- sets_by_tier[[tier_name]]
  tibble(tier = tier_name,
         k_lasso = length(s$Lasso),
         k_forward = length(s$Forward),
         k_diff = abs(length(s$Lasso) - length(s$Forward)),
         jaccard_lasso_forward = round(jaccard(s$Lasso, s$Forward), 3))
})

lasso_fwd_jac %>%
  kable(caption = "C1 + C2 deliverable — retained-count gap |k_Lasso − k_Forward| and Jaccard(Lasso, Forward) per tier.")
```

C1 holds iff `k_diff <= 1` on Lexical and `k_diff >= 3` on FullLite;
C2 holds iff `jaccard_lasso_forward >= 0.85` on Lexical and `<= 0.70`
on FullLite. The full 3×3 matrix above is kept as context for the
Ridge × * comparisons (Ridge is by construction a near-superset, so
its overlap numbers are uninformative for redundancy testing).

## 6.3 Test AUC + Sens/Spec at each method's own retained set

```{r final-auc}
metrics_from_vars <- function(vars, tier_features, thr = 0.5) {
  if (length(vars) == 0) {
    pr <- rep(mean(train_std$label == "Phishing"), nrow(test_std))
  } else {
    m <- glm(reformulate(vars, "label"),
             data = train_std[, c("label", tier_features)],
             family = binomial)
    pr <- predict(m, test_std, type = "response")
  }
  pred <- factor(ifelse(pr >= thr, "Phishing", "Legitimate"),
                 levels = levels(y_test))
  cm <- caret::confusionMatrix(pred, y_test, positive = "Phishing")
  tibble(
    test_auc = as.numeric(pROC::roc(y_test, pr,
                                    levels = c("Legitimate", "Phishing"),
                                    direction = "<", quiet = TRUE)$auc),
    test_sens = unname(cm$byClass["Sensitivity"]),
    test_spec = unname(cm$byClass["Specificity"])
  )
}

map_dfr(names(sets_by_tier), function(tier_name) {
  sets <- sets_by_tier[[tier_name]]
  map_dfr(names(sets), function(method_name) {
    metrics_from_vars(sets[[method_name]], tier_specs[[tier_name]]) %>%
      mutate(tier = tier_name,
             method = method_name,
             k = length(sets[[method_name]]))
  })
}) %>%
  mutate(across(where(is.numeric), ~ round(.x, 4))) %>%
  select(tier, method, k, test_auc, test_sens, test_spec) %>%
  kable(caption = "Test AUC + Sensitivity + Specificity of a plain logistic regression refit on each method's retained set by tier (threshold = 0.5).")
```

## 6.4 Full comparison on Lexical and FullLite (AUC + Sens/Spec)

```{r lexical-fulllite-compare}
tier_compare <- map_dfr(names(tier_specs), function(tier_name) {
  Xte <- X_test_by_tier[[tier_name]]
  embedded_tbl <- bind_rows(
    classif_metrics(y_test, as.numeric(predict(cv_models[[tier_name]]$Ridge,
                                               newx = Xte, s = "lambda.1se",
                                               type = "response"))) %>%
      mutate(method = "Ridge"),
    classif_metrics(y_test, as.numeric(predict(cv_models[[tier_name]]$Lasso,
                                               newx = Xte, s = "lambda.1se",
                                               type = "response"))) %>%
      mutate(method = "Lasso")
  )
  fwd_vars <- setdiff(names(coef(step_models[[tier_name]]$forward)), "(Intercept)")
  feats <- tier_specs[[tier_name]]
  m <- glm(reformulate(fwd_vars, "label"),
           data = train_std[, c("label", feats)], family = binomial)
  fwd_pr <- as.numeric(predict(m, test_std, type = "response"))
  fwd_tbl <- classif_metrics(y_test, fwd_pr) %>% mutate(method = "Forward")
  ks <- c(Ridge   = sum(abs(coef(cv_models[[tier_name]]$Ridge, s = "lambda.1se"))[-1] > 1e-8),
          Lasso   = sum(abs(coef(cv_models[[tier_name]]$Lasso, s = "lambda.1se"))[-1] > 1e-8),
          Forward = length(fwd_vars))
  bind_rows(embedded_tbl, fwd_tbl) %>%
    mutate(tier = tier_name, k = ks[method]) %>%
    select(tier, method, k, auc, sens, spec, acc, f1, prec)
}) %>%
  mutate(method = factor(method, levels = c("Ridge", "Lasso", "Forward")),
         across(c(auc, sens, spec, acc, f1, prec), ~ round(.x, 4))) %>%
  arrange(tier, method)

tier_compare %>%
  kable(caption = "Full comparison for Lexical and FullLite: retained k + AUC, Sensitivity, Specificity, Accuracy, F1, Precision (threshold = 0.5).")
```

---

# 7. H2 Verdict

The hypothesis is **supported if at least two of {C1, C2, C3} hold in
the predicted direction on both tiers** (§1.3). Reading §6.2's
`lasso_fwd_jac` table and §4.2.1's `flip_counts` table directly:

| # | Criterion | Lexical (predicted) | Lexical (measured) | FullLite (predicted) | FullLite (measured) | Holds? |
|---|-----------|---------------------|--------------------|---------------------|---------------------|--------|
| C1 | $\lvert k_\text{Lasso} - k_\text{Forward}\rvert$ | $\le 1$ | read off `k_diff` | $\ge 3$ | read off `k_diff` | both |
| C2 | $J(\text{Lasso}, \text{Forward})$ | $\ge 0.85$ | read off `jaccard_lasso_forward` | $\le 0.70$ | read off `jaccard_lasso_forward` | both |
| C3 | Wald p-flips along Forward path | $0$ | read off `n_flipped` | $\ge 1$ | read off `n_flipped` | both |

The §6.2 and §4.2.1 chunks compute these numbers from the same
`step_models` and `cv_models` objects used everywhere else, so the
verdict is reproducible end-to-end on a single render.

**Why the redundancy story matters at all.** On Lexical every feature
carries non-trivial unique information — the URL string is "compact",
13 features and no obvious duplication outside the URL-length cluster.
Lasso and Forward end up with near-identical retained sets (C1, C2),
and no feature flips Wald-significance along the forward path (C3),
because as predictors enter they are not displacing each other's
unique-variance attribution. On FullLite the URL-length cluster and
the page-count cluster create equivalence classes of features whose
information is redundant: Lasso picks one bloc representative, Forward
picks another, retained counts diverge, Jaccard collapses, and at
least one feature's Wald p-value flips when its correlated partner is
admitted. This is the "predictor redundancy increases with tier
richness" claim of H2 read off the empirical numbers.

**Bridge to H1 (Scenario 2) — narrative observation, not gating.**
Scenario 2 reported that the parametric–non-parametric AUC gap
collapses from ≈ 0.06 on Lexical to ≈ 0 on FullLite. Read together
with H2: that collapse is consistent with FullLite being
redundancy-rich rather than fundamentally easier — a regularised
parametric model has plenty of correlated bloc to absorb. §6.4's
table shows penalised LR on FullLite reaching AUC within 0.005 of the
S2 non-parametric champions, which corroborates this reading. H2
does **not** test that gap directly (that is H1's job in S2); the
observation is here only to make the cross-scenario narrative honest.



