---
title: "Phishing Detection — Scenario 3: Feature Selection"
author: "K-o-m-A"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(caret)
library(glmnet)
library(MASS, exclude = "select")
library(pROC)
library(knitr)

set.seed(2026)

ARTIFACTS <- "scenario_3/artifacts"
CACHE_VERSION <- "v3_phishing_second_level"
dir.create(ARTIFACTS, recursive = TRUE, showWarnings = FALSE)
```

---

# 1. Scenario & Hypothesis

The assignment has two explicit deliverables:

> *"Report how many features you ultimately retain, and which features
> lose statistical significance as you add or remove predictors."*

and a comparative imperative wrapped around them:

> *"Compare one algorithmic (forward, backward, mixed selection) and
> two embedded feature-selection methods (lasso, ridge, elastic net)
> for the same classification task. This task may require some
> research..."*

A naive reading — *"run lasso, report N features"* — misses the point.
Feature selection is interesting precisely where methods **disagree**,
and where "statistical significance" stops being a clean property of
a predictor and starts being a fragile function of which *other*
predictors are in the model. This notebook is built around that
observation.

## 1.1 Why Lexical

Scenario 2's §6.1 showed that **Lexical** (13 URL-string features) is
the only tier where classifier family matters — LR-Ridge sits at
0.861 while SVM-RBF reaches 0.997. On the other tiers (Trust 0.92,
Behavior 0.99, FullLite 1.00) the linear classifier already sits at
or near the ceiling, so every sensible selection method converges to
"keep 3–5 features, AUC = 1.0, done." That is not a finding — it is
saturation.

Lexical also inherits the EDA §4.3 finding of a **severe
multicollinearity cluster**: `URLLength`, `NoOfLettersInURL` and
`NoOfDegitsInURL` have VIF > 1000 and pairwise `|r| > 0.84`. That is
the textbook regime where ridge and lasso are known to behave
differently — ridge keeps the entire block with shared weight, lasso
sparsifies and re-distributes the bloc onto large
opposing-sign coefficients on the survivors. Any honest comparison
of the two should be anchored where that contrast is visible, not
hidden.

We therefore run **all of S3 on Lexical**, and report a parallel
FullLite comparison (§3.3, §4.1, §6.4) to show how method differences
play out when signal is abundant.

## 1.2 Method choice — one algorithmic + two embedded

The assignment requires **one** algorithmic and **two** embedded
methods. We pick **Forward selection (AIC) + Ridge + Lasso** and
deliberately drop Backward, Stepwise and ElasticNet. The reasoning is
empirical, not aesthetic:

- **Forward over Backward / Stepwise.** All three AIC procedures
  converge to the *same retained set* on Lexical (k = 9, Jaccard
  1.00) and the same set up to two features on FullLite. Reporting
  three near-identical procedures inflates the deliverable without
  adding information. Forward additionally produces the **per-step
  p-value trajectory** the assignment literally asks for ("…lose
  statistical significance as you **add** … predictors"); backward
  shows the symmetric story but the trajectory is messier.
- **Ridge + Lasso, drop ElasticNet.** Ridge (α = 0) and Lasso (α = 1)
  are the two endpoints of the regularisation axis — the sharpest
  contrast. ElasticNet at α = 0.5 is empirically interpolation: on
  Lexical its retained set differs from Lasso by one feature, its
  coefficients are within ~15 % of Lasso's (NoOfLettersInURL: Lasso
  11.0 vs EN 9.16), and its AUC matches Lasso to three decimals. EN
  would add a third row to every comparison table that says "in
  between — see Lasso". The one independent observation EN gives
  (better fold-stability than Lasso on FullLite, e.g. NoOfPopup) is
  noted in §3.6 prose without re-running the whole pipeline.
- **Note on Ridge vs Scenario 2.** Scenario 2 used the same `glmnet`
  ridge fit, but as a *classifier-family benchmark* — fixed λ = 0.01,
  no feature-selection lens, just an AUC number for the LR row. Here
  we use the same solver to ask a *different* question: how does
  ridge distribute weight across the multicollinear bloc, and how
  does that distribution compare to Lasso's. Same algorithm, two
  non-overlapping deliverables.

## 1.3 H3 — what we claim

> *Ridge and Lasso distribute weight very differently across the
> URL-length collinearity bloc — ridge keeps the full bloc with small
> distributed coefficients, lasso re-pivots onto large opposing-sign
> coefficients on the survivors. The retained *count* therefore
> differs between methods even on Lexical. The algorithmic AIC
> forward procedure keeps features that are not Wald-significant in
> the final selected model — the textbook demonstration that stepwise
> p-values are conditional on selection and not valid hypothesis
> tests. On the higher-signal FullLite tier, forward selection makes
> some features lose Wald significance as correlated partners are
> added.*

| #  | Criterion                                                                                                                                                                                | Threshold                                                                  | Where     |
|----|------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|----------------------------------------------------------------------------|-----------|
| C1 | Ridge and Lasso disagree on retained count and on the shape of the URL-length coefficient bloc                                                                                           | k(Ridge) > k(Lasso); bloc spread for Ridge, concentrated for Lasso         | §3.3, §3.4|
| C2 | At least one feature in the AIC-selected model has Wald p ≥ 0.05 in the final model — *or* a feature loses significance as a correlated partner enters in the forward trajectory        | ≥ 1 feature                                                                | §4.1, §4.2|
| C3 | Lasso's per-feature retention rate across 10 CV folds is below 1.0 for at least one feature where ridge is at 1.0                                                                        | ≥ 1 feature with rate ∈ (0, 1)                                             | §3.6      |

H3 is **supported** if all three criteria hold. The earlier-version
threshold "Jaccard ≤ 0.5 between lasso and ridge" turned out to be
too aggressive: ridge is by construction a superset of lasso, so
Jaccard sits in the 0.7–0.8 band even when the *coefficient profile*
differs sharply. We therefore anchor C1 on retained count and the
coefficient bloc, not on Jaccard.

---

# 2. Data Loading

Same recipe as Scenario 2 and 5 — identical sub-sample, split and CV
folds so the numbers here are directly comparable.

## 2.1 Load & exclusions

```{r load-data}
dataset_path <- if (file.exists("PhiUSIIL_Phishing_URL_Dataset.csv")) {
  "PhiUSIIL_Phishing_URL_Dataset.csv"
} else if (file.exists("../PhiUSIIL_Phishing_URL_Dataset.csv")) {
  "../PhiUSIIL_Phishing_URL_Dataset.csv"
} else stop("PhiUSIIL_Phishing_URL_Dataset.csv not found")

raw <- read_csv(dataset_path, show_col_types = FALSE)
names(raw) <- map_chr(names(raw), ~ str_replace(.x, "^\ufeff", ""))

ID_COLS          <- c("FILENAME", "URL", "Domain", "TLD", "Title")
SCORES           <- c("URLSimilarityIndex", "TLDLegitimateProb", "URLCharProb",
                      "DomainTitleMatchScore", "URLTitleMatchScore")
REDUNDANT_BINARY <- "HasObfuscation"
RATIOS           <- c("LetterRatioInURL", "DegitRatioInURL",
                      "ObfuscationRatio", "SpacialCharRatioInURL")
EXCLUDE_COLS     <- c(ID_COLS, SCORES, REDUNDANT_BINARY, RATIOS)

df <- raw %>%
  mutate(label = factor(label, levels = c(0, 1),
                        labels = c("Legitimate", "Phishing"))) %>%
  select(-all_of(EXCLUDE_COLS)) %>%
  as.data.frame()
# Note on factor order: glmnet / glm treat the *second* factor level as
# the success class. We deliberately put "Phishing" second so that
# predict(..., type = "response") returns P(Phishing) directly, and
# Sensitivity / Specificity / pROC are aligned with positive = "Phishing".
```

## 2.2 Lexical tier

```{r lexical}
lexical_features <- c("URLLength", "DomainLength", "IsDomainIP",
                      "CharContinuationRate", "TLDLength", "NoOfSubDomain",
                      "NoOfObfuscatedChar", "NoOfLettersInURL",
                      "NoOfDegitsInURL", "NoOfEqualsInURL",
                      "NoOfQMarkInURL", "NoOfAmpersandInURL",
                      "NoOfOtherSpecialCharsInURL")
stopifnot(all(lexical_features %in% names(df)))

LEAKY_BEHAVIOR <- c("LineOfCode", "NoOfExternalRef", "NoOfImage",
                    "NoOfSelfRef", "NoOfJS", "NoOfCSS")
fullLite_features <- setdiff(setdiff(names(df), "label"), LEAKY_BEHAVIOR)

cat("Lexical tier:", length(lexical_features), "features.\n")
cat("FullLite tier:", length(fullLite_features), "features.\n")
```

## 2.3 Subsample, split, folds, preprocessing

```{r prep}
N_SUB <- 30000
df_sub <- df %>%
  group_by(label) %>%
  slice_sample(n = N_SUB / 2) %>%
  ungroup() %>%
  as.data.frame()

idx       <- createDataPartition(df_sub$label, p = 0.8, list = FALSE)
train_raw <- df_sub[idx, ]
test_raw  <- df_sub[-idx, ]
fold_idx  <- createFolds(train_raw$label, k = 10, returnTrain = TRUE)

feat <- df_sub %>% select(-label)
binary_features     <- names(feat)[sapply(feat, function(x)
                                   all(na.omit(x) %in% c(0, 1)))]
continuous_features <- setdiff(names(feat), binary_features)

apply_log <- function(d, cols) {
  d[cols] <- lapply(d[cols], function(x) log1p(pmax(x, 0)))
  d
}
train_log <- apply_log(train_raw, continuous_features)
test_log  <- apply_log(test_raw,  continuous_features)

pp <- preProcess(train_log[, continuous_features], method = c("center", "scale"))
train_std <- train_log; test_std <- test_log
train_std[, continuous_features] <- predict(pp, train_log[, continuous_features])
test_std[,  continuous_features] <- predict(pp, test_log[,  continuous_features])

# Standardise the binary IsDomainIP too so glmnet / stepAIC see all
# predictors on one scale; glmnet's internal standardization would do
# this anyway but being explicit keeps coefficients directly readable.
pp_lex <- preProcess(train_std[, lexical_features], method = c("center", "scale"))
train_std[, lexical_features] <- predict(pp_lex, train_std[, lexical_features])
test_std[,  lexical_features] <- predict(pp_lex, test_std[,  lexical_features])

y_train <- train_std$label
y_test  <- test_std$label

tier_specs <- list(
  Lexical = lexical_features,
  FullLite = fullLite_features
)

X_train_by_tier <- map(tier_specs, ~ as.matrix(train_std[, .x]))
X_test_by_tier  <- map(tier_specs, ~ as.matrix(test_std[,  .x]))

```

## 2.4 Multicollinearity context — quick re-confirmation

EDA reported VIF > 1000 on the URL-length cluster. We recompute the
correlation block here so the rest of the notebook reads standalone.

```{r cor-check}
cor_pairs <- map_dfr(names(tier_specs), function(tier_name) {
  cor_mat <- cor(X_train_by_tier[[tier_name]])
  heavy_pairs <- which(abs(cor_mat) > 0.7 & row(cor_mat) < col(cor_mat),
                       arr.ind = TRUE)
  if (nrow(heavy_pairs) == 0) {
    return(tibble(tier = tier_name, a = NA_character_, b = NA_character_, r = NA_real_))
  }
  tibble(
    tier = tier_name,
    a = rownames(cor_mat)[heavy_pairs[, 1]],
    b = colnames(cor_mat)[heavy_pairs[, 2]],
    r = round(cor_mat[heavy_pairs], 3)
  )
})

cor_pairs %>%
  arrange(tier, desc(abs(r))) %>%
  kable(caption = "Feature pairs with |r| > 0.7 for Lexical and FullLite tiers.")
```

---

# 3. Embedded Methods — Ridge & Lasso

Both are `glmnet` fits with `family = "binomial"` and differ only in
`alpha` (0 = ridge, 1 = lasso). `lambda` is chosen by 10-fold CV, and
we report both `lambda.min` (optimal CV deviance) and `lambda.1se`
(sparsest model within one SE of the minimum — the canonical pick
for parsimony).

## 3.1 Fit the two paths

```{r enet-fit}
run_or_load <- function(name, fn) {
  path <- file.path(ARTIFACTS, paste0(name, "_", CACHE_VERSION, ".rds"))
  if (file.exists(path)) return(readRDS(path))
  res <- fn()
  saveRDS(res, path)
  res
}

fit_glmnet <- function(X_train_tier, alpha, tag, tier_name) {
  run_or_load(paste0("glmnet_", tolower(tier_name), "_", tag), function() {
    cv.glmnet(X_train_tier, y_train, family = "binomial",
              alpha = alpha, nfolds = 10, standardize = FALSE,
              type.measure = "auc",
              nlambda = 80,
              lambda.min.ratio = 1e-2,
              maxit = 1e6)
  })
}

cv_models <- map(names(tier_specs), function(tier_name) {
  X_tr <- X_train_by_tier[[tier_name]]
  list(
    Ridge = fit_glmnet(X_tr, 0.0, "ridge", tier_name),
    Lasso = fit_glmnet(X_tr, 1.0, "lasso", tier_name)
  )
}) %>% set_names(names(tier_specs))

```

## 3.2 Regularization paths — how features drop out

```{r reg-paths, fig.width=10, fig.height=3.6}
par(mfrow = c(2, 2))
for (tier_name in names(tier_specs)) {
  plot(cv_models[[tier_name]]$Ridge$glmnet.fit, xvar = "lambda", label = FALSE)
  title(sprintf("%s: Ridge (α=0)", tier_name), line = 2.5)
  plot(cv_models[[tier_name]]$Lasso$glmnet.fit, xvar = "lambda", label = FALSE)
  title(sprintf("%s: Lasso (α=1)", tier_name), line = 2.5)
}
par(mfrow = c(1, 1))
```

**Read:** the x-axis is `log(λ)`; at the right edge every coefficient
is shrunk to zero. Ridge keeps every line non-zero for any λ — only
their magnitudes shrink. Lasso sends lines to exactly zero as λ
grows; the order in which they hit zero is the algorithmic answer to
"which feature is dropped next".

## 3.3 Retained features at `lambda.1se`

```{r enet-retained}
coefs_at <- function(cvfit, s = "lambda.1se") {
  co <- as.matrix(coef(cvfit, s = s))
  data.frame(feature = rownames(co),
             coef    = as.numeric(co)) %>%
    filter(feature != "(Intercept)")
}

retained <- map_dfr(names(cv_models), function(tier_name) {
  bind_rows(
    coefs_at(cv_models[[tier_name]]$Ridge) %>% mutate(method = "Ridge", tier = tier_name),
    coefs_at(cv_models[[tier_name]]$Lasso) %>% mutate(method = "Lasso", tier = tier_name)
  )
}) %>% mutate(kept = abs(coef) > 1e-8)

retained %>%
  group_by(tier, method) %>%
  summarise(retained = sum(kept),
            total    = n(),
            lambda_1se = case_when(
              method[1] == "Ridge" ~ cv_models[[tier[1]]]$Ridge$lambda.1se,
              method[1] == "Lasso" ~ cv_models[[tier[1]]]$Lasso$lambda.1se),
            .groups = "drop") %>%
  kable(caption = "Retained features at lambda.1se for Lexical and FullLite (embedded).",
        digits = 4)
```

## 3.4 Coefficient table — who survived, who didn't

```{r enet-coef-table}
retained %>%
  select(tier, method, feature, coef) %>%
  pivot_wider(names_from = method, values_from = coef) %>%
  arrange(tier, desc(abs(Ridge))) %>%
  mutate(across(c(Ridge, Lasso), ~ round(.x, 3))) %>%
  kable(caption = "Standardised coefficients at lambda.1se by tier — zeros mean lasso dropped the feature.")
```

**Multicollinearity tell.** On Lexical, Ridge spreads small,
**same-sign** weight across the URL-length trio (URLLength ≈ 0.10,
NoOfLettersInURL ≈ 0.28, NoOfDegitsInURL ≈ 0.65). Lasso keeps all
three but blows them up into **opposing-sign** coefficients of
double-digit magnitude (URLLength ≈ −15, NoOfLettersInURL ≈ +11,
NoOfDegitsInURL ≈ +6) — the classical "huge cancellations" pattern
that flags an underdetermined system: each individual coefficient is
not interpretable in isolation, only the linear combination is. The
features lasso actually zeros are the low-signal ones
(NoOfObfuscatedChar, NoOfAmpersandInURL, occasionally NoOfEqualsInURL),
*not* one of the URL-length cluster — because dropping one of those
would forfeit the cancellation lasso is using to fit the data.

## 3.5 Predictive performance — AUC + Sens/Spec at the same `lambda`

If Ridge and Lasso arrive at similar AUC with very different retained
sets and very different coefficient profiles, that itself is the
finding: *multiple explanations fit the same data equally well*,
which is exactly what should make one cautious about reading the
retained set as causal.

```{r enet-auc}
metrics_at <- function(cvfit, X_test_tier, s = "lambda.1se", thr = 0.5) {
  pr <- as.numeric(predict(cvfit, newx = X_test_tier, s = s, type = "response"))
  pred <- factor(ifelse(pr >= thr, "Phishing", "Legitimate"),
                 levels = levels(y_test))
  cm <- caret::confusionMatrix(pred, y_test, positive = "Phishing")
  tibble(
    auc = as.numeric(pROC::roc(y_test, pr,
                               levels = c("Legitimate", "Phishing"),
                               direction = "<", quiet = TRUE)$auc),
    sens = unname(cm$byClass["Sensitivity"]),
    spec = unname(cm$byClass["Specificity"])
  )
}

map_dfr(names(cv_models), function(tier_name) {
  X_te <- X_test_by_tier[[tier_name]]
  m_1se <- bind_rows(
    metrics_at(cv_models[[tier_name]]$Ridge, X_te) %>% mutate(Method = "Ridge"),
    metrics_at(cv_models[[tier_name]]$Lasso, X_te) %>% mutate(Method = "Lasso")
  )
  m_min <- bind_rows(
    metrics_at(cv_models[[tier_name]]$Ridge, X_te, "lambda.min") %>% mutate(Method = "Ridge"),
    metrics_at(cv_models[[tier_name]]$Lasso, X_te, "lambda.min") %>% mutate(Method = "Lasso")
  )
  left_join(m_1se, m_min, by = "Method", suffix = c("_1se", "_min")) %>%
    transmute(
      Tier = tier_name,
      Method,
      `AUC @ 1se` = auc_1se,
      `Sens @ 1se` = sens_1se,
      `Spec @ 1se` = spec_1se,
      `AUC @ min` = auc_min,
      `Sens @ min` = sens_min,
      `Spec @ min` = spec_min
    )
}) %>%
  mutate(across(where(is.numeric), ~ round(.x, 4))) %>%
  kable(caption = "Test AUC + Sensitivity + Specificity at lambda.1se and lambda.min for Lexical and FullLite (threshold = 0.5).")
```

```{r metrics-helper}
classif_metrics <- function(y_true, prob, thr = 0.5) {
  pred <- factor(ifelse(prob >= thr, "Phishing", "Legitimate"),
                 levels = levels(y_true))
  cm <- caret::confusionMatrix(pred, y_true, positive = "Phishing")
  auc <- as.numeric(pROC::roc(y_true, prob,
                              levels = c("Legitimate", "Phishing"),
                              direction = "<", quiet = TRUE)$auc)
  tibble(
    auc  = auc,
    acc  = unname(cm$overall["Accuracy"]),
    f1   = unname(cm$byClass["F1"]),
    prec = unname(cm$byClass["Precision"]),
    sens = unname(cm$byClass["Sensitivity"]),
    spec = unname(cm$byClass["Specificity"])
  )
}
```

## 3.6 Stability under resampling (C3)

Lasso is known to be unstable under correlated predictors: small
perturbations of the training set flip which member of a collinear
block it picks. We quantify this over the 10 CV folds by recording
which features each fold retains at its own `lambda.1se`.

```{r stability}
stability_fit <- run_or_load("stability", function() {
  map_dfr(names(tier_specs), function(tier_name) {
    X_tier <- X_train_by_tier[[tier_name]]
    map_dfr(seq_along(fold_idx), function(i) {
      tr <- fold_idx[[i]]
      Xi <- X_tier[tr, ]; yi <- y_train[tr]
      map_dfr(list(Ridge = 0, Lasso = 1),
        function(a) {
          cvfit <- cv.glmnet(Xi, yi, family = "binomial",
                             alpha = a, nfolds = 5, standardize = FALSE,
                             type.measure = "auc",
                             nlambda = 80,
                             lambda.min.ratio = 1e-2,
                             maxit = 1e6)
          co <- as.matrix(coef(cvfit, s = "lambda.1se"))
          tibble(tier = tier_name,
                 fold = i,
                 feature = rownames(co)[-1],
                 coef    = as.numeric(co[-1, 1]))
        }, .id = "method")
    })
  })
})

stab_summary <- stability_fit %>%
  mutate(kept = abs(coef) > 1e-8) %>%
  group_by(tier, method, feature) %>%
  summarise(retention_rate = mean(kept), .groups = "drop") %>%
  pivot_wider(names_from = method, values_from = retention_rate) %>%
  arrange(tier, desc(Ridge))

stab_summary %>%
  mutate(across(where(is.numeric), ~ round(.x, 2))) %>%
  kable(caption = "Retention rate per feature across 10 CV folds for Lexical and FullLite (1.00 = kept every fold, 0.00 = never kept).")
```

**Read.** Ridge shows retention 1.00 everywhere — it never zeros
features. On Lexical the lasso column is very stable too: every
feature it ever picks it picks on every fold, and IsDomainIP
(retention 0.9) is the only borderline case. The signal in this tier
is concentrated enough that the small-perturbation flip mechanism
does not get triggered. **FullLite is where the pathology surfaces**:
HasFavicon = 0.9, IsResponsive = 0.1 across folds — these are the
lasso "coin-flip" features whose presence in the retained set is
run-dependent, exactly the case the literature warns against. The
full retention table also shows a long list of FullLite features
where Ridge sits at 1.00 while Lasso sits at 0.00 — a different
manifestation of the same picking instability, where lasso settles
for one member of a redundant pair and never revisits.

---

# 4. Algorithmic Selection — Forward (AIC)

`stats::step` walks a sequence of nested logistic regressions and
picks the one minimising AIC. Unlike embedded methods it gives us a
**per-step Wald p-value table** — at each step every coefficient has
a Wald p-value — which is exactly what the assignment asks us to
report under *"which features lose significance as you add or remove
predictors"*. We then discuss **why that p-value is not what it
appears to be** (§5).

We restrict the algorithmic part to **forward selection**. On this
dataset Forward, Backward and Stepwise converge to the same retained
set on Lexical (k = 9, identical features) and within two features
on FullLite, so reporting all three would be three near-identical
tables. The forward direction also matches the assignment's "as you
**add** predictors" framing more directly.

## 4.1 Forward selection — feature entry order & p-values

```{r forward}
step_models <- run_or_load("step_models", function() {
  map(names(tier_specs), function(tier_name) {
    feats <- tier_specs[[tier_name]]
    dtr <- train_std[, c("label", feats)]
    null_mod_t <- glm(label ~ 1, data = dtr, family = binomial)
    full_mod_t <- glm(label ~ ., data = dtr, family = binomial)
    list(
      features = feats,
      forward = step(null_mod_t,
                     scope = list(lower = null_mod_t, upper = full_mod_t),
                     direction = "forward", trace = FALSE, k = 2)
    )
  }) %>% set_names(names(tier_specs))
})

map_dfr(names(step_models), function(tier_name) {
  tibble(
    tier = tier_name,
    retained = length(coef(step_models[[tier_name]]$forward)) - 1,
    total = length(step_models[[tier_name]]$features)
  )
}) %>% kable(caption = "Forward selection retained-feature counts by tier.")

# Recover the entry order and per-step Wald p-values of the ENTERING
# feature and of every previously-entered feature. This is the
# "significance as predictors are added" table the assignment asks for.
fwd_trajectory <- run_or_load("forward_traj", function() {
  map_dfr(names(step_models), function(tier_name) {
    entered <- setdiff(names(coef(step_models[[tier_name]]$forward)), "(Intercept)")
    feats <- step_models[[tier_name]]$features
    map_dfr(seq_along(entered), function(k) {
      vars <- entered[1:k]
      m <- glm(reformulate(vars, "label"),
               data = train_std[, c("label", feats)],
               family = binomial)
      co <- summary(m)$coefficients
      tibble(tier = tier_name, step = k, entering = entered[k],
             feature = rownames(co), p = co[, "Pr(>|z|)"]) %>%
        filter(feature != "(Intercept)")
    })
  })
})

fwd_trajectory %>%
  mutate(p = signif(p, 3)) %>%
  pivot_wider(id_cols = c(tier, feature), names_from = step,
              values_from = p, names_prefix = "step_") %>%
  kable(caption = "Wald p-values as forward selection enters each feature for Lexical and FullLite.")
```

## 4.2 Final selected model — coefficients and Wald p-values

The last column of §4.1 is the final selected model's per-feature
p-value. Reading it directly:

```{r forward-final-coef}
map_dfr(names(step_models), function(tier_name) {
  as.data.frame(summary(step_models[[tier_name]]$forward)$coefficients) %>%
    rownames_to_column("feature") %>%
    filter(feature != "(Intercept)") %>%
    mutate(tier = tier_name)
}) %>%
  mutate(across(-c(feature, tier), ~ signif(.x, 3))) %>%
  kable(caption = "Final forward-AIC-selected model by tier — Wald p-values are biased by the selection process (see §5).")
```

On **Lexical** the final model retains nine features but two of them
have non-significant Wald p-values: IsDomainIP (p = 1.0, with a
standard error blown up to ≈ 1.2·10⁴ — the textbook signature of
quasi-separation) and NoOfQMarkInURL (p = 0.81). AIC kept these two
because adding them did not penalise log-likelihood enough to
overcome the −2·k term, *not* because there is independent statistical
evidence for them. On **FullLite** the analogous behaviour appears for
IsHTTPS (p ≈ 0.98 across all steps, kept by AIC throughout).

## 4.3 AIC trajectory — at what k did AIC stop improving

```{r aic-traj, fig.width=6.5, fig.height=3.4}
# Forward AIC per step, for the AUC-vs-k-curve that the assignment
# implicitly points at ("how many features do you retain").
fwd_aic <- run_or_load("forward_aic", function() {
  map_dfr(names(step_models), function(tier_name) {
    feats <- step_models[[tier_name]]$features
    entered <- setdiff(names(coef(step_models[[tier_name]]$forward)), "(Intercept)")
    ordered_all <- c(entered, setdiff(feats, entered))
    map_dfr(seq_along(ordered_all), function(k) {
      m <- glm(reformulate(ordered_all[1:k], "label"),
               data = train_std[, c("label", feats)],
               family = binomial)
      tibble(tier = tier_name, k = k, aic = AIC(m),
             test_auc = {
               pr <- predict(m, test_std, type = "response")
               as.numeric(pROC::roc(y_test, pr,
                                    levels = c("Legitimate", "Phishing"),
                                    direction = "<", quiet = TRUE)$auc)
             })
    })
  })
})

ggplot(fwd_aic, aes(k)) +
  geom_line(aes(y = aic, colour = "AIC"), linewidth = 0.8) +
  geom_point(aes(y = aic, colour = "AIC")) +
  geom_line(aes(y = (test_auc - min(test_auc)) /
                  (max(test_auc) - min(test_auc)) *
                  (max(aic) - min(aic)) + min(aic),
                colour = "Test AUC (rescaled)"),
            linewidth = 0.8) +
  facet_wrap(~ tier, scales = "free_x") +
  scale_x_continuous(breaks = scales::pretty_breaks()) +
  scale_colour_manual(values = c("AIC" = "steelblue",
                                 "Test AUC (rescaled)" = "firebrick")) +
  labs(title = "Forward-selection trajectory by tier",
       subtitle = "AIC knee vs. Test-AUC plateau — Lexical and FullLite",
       x = "Number of features entered", y = "AIC / rescaled Test AUC",
       colour = NULL) +
  theme_minimal() + theme(legend.position = "top")
```

**Read:** if AIC plateaus at k* and Test AUC also plateaus at k*, the
two criteria agree. Divergences (AIC still improving while AUC
plateaus, or vice versa) are common at small n / large p but at
n ≈ 24 k they are usually small. The curve also answers the H1
question from Scenario 2 from a new angle — if Test AUC climbs
steadily across all 13 features without a sharp knee, the Lexical
signal is **combinatorial** (many weak features adding up) rather
than concentrated in a handful of dominant features.

---

# 5. Post-Selection Inference — the research piece

The Wald p-values reported in §4.1 and §4.2 are **not valid**
out-of-the-box. This is the single most important thing to say about
stepwise selection, and the single most common thing that gets
forgotten in practice.

## 5.1 Why the p-values are biased

A classical Wald p-value assumes the set of predictors was fixed
*before* seeing the data. Stepwise selection violates this
fundamentally: the set of predictors *is chosen from the data* to
minimise AIC (or maximise likelihood). The p-values reported by the
final selected model are conditional on having been selected — which
means:

- Selected features are more likely to appear significant than they
  truly are (the winner's-curse effect).
- The null distribution of the Wald statistic is no longer a χ²; the
  standard formula overstates significance.
- Published critiques (Berk et al. 2013; Lee et al. 2016; Tibshirani
  et al. 2016) argue for *selective inference* — p-values computed
  *conditional* on the selection event. Tools: the `selectiveInference`
  R package for the lasso, and post-selection splitting (Cox 1975;
  Wasserman & Roeder 2009).

**Practical consequence for this notebook:** the final-model
p-values in §4.2 should **not** be read as "these are the significant
features". They are, strictly, *the p-values a naive analyst would
quote*. The right way to phrase a finding is: "the forward AIC
procedure retained 9 features on Lexical; standard Wald p-values on
those features are given but are biased by the selection process and
should not be used for hypothesis tests."

## 5.2 Why the forward-trajectory p-values *are* useful — for C2

Although the *final* p-values are biased, the **trajectory** in
§4.1 answers a different question: given this specific sequence of
models (entry order fixed by AIC), does any feature's p-value
*rise* above 0.05 when a correlated partner is added in a later
step? That trajectory is a direct demonstration of the
multicollinearity mechanism — and *that* is a valid descriptive
observation, not a hypothesis test.

On **Lexical** the trajectory does **not** show this pathology — every
feature that enters keeps p well below 0.05 to the end (the URL-length
trio survives because the signal is large enough to overpower
shrinkage of the standard error). Two features are entered late by AIC
but never reach significance at all (IsDomainIP, NoOfQMarkInURL); they
illustrate the *other* form of the problem (§4.2) — features kept by
selection criteria with no Wald evidence behind them.

On **FullLite** the C2 mechanism is clearly visible:
NoOfOtherSpecialCharsInURL goes from p ≈ 4·10⁻²⁵¹ at step 3 to
p ≈ 0.81 at step 19, and NoOfSubDomain from p ≈ 1.9·10⁻⁸⁹ to p ≈ 0.68
— both as their correlated partners (URLLength, NoOfLettersInURL,
NoOfDegitsInURL, DomainLength) progressively absorb the shared signal.

## 5.3 Why lasso stability (§3.6) is the embedded-method answer

The embedded counterpart of the "stepwise p-value is biased" problem
is the **instability of sparse selection under correlated
predictors**. Lasso will pick one feature from a block and zero the
rest, but *which* feature it picks is a small-perturbation-away from
flipping. The §3.6 stability matrix makes this visible — features
with retention 0.4–0.6 across folds are the same features that, in
an algorithmic setting, would have unstable p-values. Different
machinery, same underlying pathology.

---

# 6. Comparison — Retained Sets Side by Side

## 6.1 The deliverable table (assignment's literal asks)

```{r retained-compare}
retained_by_method <- map_dfr(names(tier_specs), function(tier_name) {
  bind_rows(
    tibble(method = "Ridge",         retained = sum(abs(coef(cv_models[[tier_name]]$Ridge, s = "lambda.1se"))[-1] > 1e-8)),
    tibble(method = "Lasso",         retained = sum(abs(coef(cv_models[[tier_name]]$Lasso, s = "lambda.1se"))[-1] > 1e-8)),
    tibble(method = "Forward (AIC)", retained = length(coef(step_models[[tier_name]]$forward)) - 1)
  ) %>%
    mutate(tier = tier_name,
           total = length(tier_specs[[tier_name]]))
})

kable(retained_by_method,
      caption = "How many features each method retains in Lexical and FullLite.")
```

## 6.2 Feature-set overlap (C1)

```{r overlap}
get_kept <- function(cvfit) {
  co <- as.matrix(coef(cvfit, s = "lambda.1se"))
  rownames(co)[-1][abs(co[-1, 1]) > 1e-8]
}

sets_by_tier <- map(names(tier_specs), function(tier_name) {
  list(
    Ridge   = get_kept(cv_models[[tier_name]]$Ridge),
    Lasso   = get_kept(cv_models[[tier_name]]$Lasso),
    Forward = setdiff(names(coef(step_models[[tier_name]]$forward)), "(Intercept)")
  )
}) %>% set_names(names(tier_specs))

jaccard <- function(a, b) length(intersect(a, b)) / length(union(a, b))

map_dfr(names(sets_by_tier), function(tier_name) {
  sets <- sets_by_tier[[tier_name]]
  jac_mat <- outer(names(sets), names(sets),
                   Vectorize(function(i, j) jaccard(sets[[i]], sets[[j]])))
  dimnames(jac_mat) <- list(names(sets), names(sets))
  round(jac_mat, 2) %>%
    as.data.frame() %>%
    rownames_to_column("method") %>%
    mutate(tier = tier_name)
}) %>%
  select(tier, method, everything()) %>%
  kable(caption = "Jaccard overlap of retained-feature sets by tier. Off-diagonal < 1 = disagreement.")
```

## 6.3 Test AUC + Sens/Spec at each method's own retained set

```{r final-auc}
metrics_from_vars <- function(vars, tier_features, thr = 0.5) {
  if (length(vars) == 0) {
    pr <- rep(mean(train_std$label == "Phishing"), nrow(test_std))
  } else {
    m <- glm(reformulate(vars, "label"),
             data = train_std[, c("label", tier_features)],
             family = binomial)
    pr <- predict(m, test_std, type = "response")
  }
  pred <- factor(ifelse(pr >= thr, "Phishing", "Legitimate"),
                 levels = levels(y_test))
  cm <- caret::confusionMatrix(pred, y_test, positive = "Phishing")
  tibble(
    test_auc = as.numeric(pROC::roc(y_test, pr,
                                    levels = c("Legitimate", "Phishing"),
                                    direction = "<", quiet = TRUE)$auc),
    test_sens = unname(cm$byClass["Sensitivity"]),
    test_spec = unname(cm$byClass["Specificity"])
  )
}

map_dfr(names(sets_by_tier), function(tier_name) {
  sets <- sets_by_tier[[tier_name]]
  map_dfr(names(sets), function(method_name) {
    metrics_from_vars(sets[[method_name]], tier_specs[[tier_name]]) %>%
      mutate(tier = tier_name,
             method = method_name,
             k = length(sets[[method_name]]))
  })
}) %>%
  mutate(across(where(is.numeric), ~ round(.x, 4))) %>%
  select(tier, method, k, test_auc, test_sens, test_spec) %>%
  kable(caption = "Test AUC + Sensitivity + Specificity of a plain logistic regression refit on each method's retained set by tier (threshold = 0.5).")
```

## 6.4 Full comparison on Lexical and FullLite (AUC + Sens/Spec)

```{r lexical-fulllite-compare}
tier_compare <- map_dfr(names(tier_specs), function(tier_name) {
  Xte <- X_test_by_tier[[tier_name]]
  embedded_tbl <- bind_rows(
    classif_metrics(y_test, as.numeric(predict(cv_models[[tier_name]]$Ridge,
                                               newx = Xte, s = "lambda.1se",
                                               type = "response"))) %>%
      mutate(method = "Ridge"),
    classif_metrics(y_test, as.numeric(predict(cv_models[[tier_name]]$Lasso,
                                               newx = Xte, s = "lambda.1se",
                                               type = "response"))) %>%
      mutate(method = "Lasso")
  )
  fwd_vars <- setdiff(names(coef(step_models[[tier_name]]$forward)), "(Intercept)")
  feats <- tier_specs[[tier_name]]
  m <- glm(reformulate(fwd_vars, "label"),
           data = train_std[, c("label", feats)], family = binomial)
  fwd_pr <- as.numeric(predict(m, test_std, type = "response"))
  fwd_tbl <- classif_metrics(y_test, fwd_pr) %>% mutate(method = "Forward")
  ks <- c(Ridge   = sum(abs(coef(cv_models[[tier_name]]$Ridge, s = "lambda.1se"))[-1] > 1e-8),
          Lasso   = sum(abs(coef(cv_models[[tier_name]]$Lasso, s = "lambda.1se"))[-1] > 1e-8),
          Forward = length(fwd_vars))
  bind_rows(embedded_tbl, fwd_tbl) %>%
    mutate(tier = tier_name, k = ks[method]) %>%
    select(tier, method, k, auc, sens, spec, acc, f1, prec)
}) %>%
  mutate(method = factor(method, levels = c("Ridge", "Lasso", "Forward")),
         across(c(auc, sens, spec, acc, f1, prec), ~ round(.x, 4))) %>%
  arrange(tier, method)

tier_compare %>%
  kable(caption = "Full comparison for Lexical and FullLite: retained k + AUC, Sensitivity, Specificity, Accuracy, F1, Precision (threshold = 0.5).")
```

---

# 7. H3 Verdict

**C1 — Ridge and Lasso disagree on retained count and on the
URL-length bloc.** *Supported.* §3.3 gives Ridge = 13, Lasso = 10 on
Lexical (Ridge = 34, Lasso = 25 on FullLite); §3.4 shows Ridge
spreading the URL-length cluster across small same-sign weights
(URLLength ≈ 0.10, NoOfLettersInURL ≈ 0.28, NoOfDegitsInURL ≈ 0.65)
while Lasso compresses it onto large opposing-sign coefficients
(URLLength ≈ −15.2, NoOfLettersInURL ≈ +11.0, NoOfDegitsInURL ≈ +6.3).
The Jaccard overlap §6.2 is 0.77 (Lex) / 0.74 (FL) for Lasso × Ridge
— high because Ridge is a superset, but *which* features Lasso drops
and *how heavily it re-weights the survivors* is the substantive
disagreement.

**C2 — A retained feature has Wald p ≥ 0.05 in the AIC-selected
model, or loses significance as correlated partners enter.**
*Supported on both tiers, by complementary mechanisms.* On
**Lexical** the forward-AIC final model retains IsDomainIP (p = 1.0,
SE ≈ 1.2·10⁴ — the textbook signature of separation) and
NoOfQMarkInURL (p = 0.81), i.e. AIC kept two features that no Wald
test would have accepted. On **FullLite** the §4.1 forward trajectory
shows NoOfOtherSpecialCharsInURL moving from p ≈ 4·10⁻²⁵¹ at step 3
to p ≈ 0.81 at step 19, and NoOfSubDomain from p ≈ 1.9·10⁻⁸⁹ to
p ≈ 0.68, as their URL-length partners are admitted later — exactly
the multicollinearity-driven significance flip the assignment is
asking us to surface.

**C3 — Lasso has at least one fold-unstable retention where ridge is
flat at 1.0.** *Supported on FullLite, marginal on Lexical.* §3.6:
on Lexical, IsDomainIP has Lasso retention 0.9 (one fold flips), all
other features are stably in or out. On FullLite the picture is much
richer — HasFavicon = 0.9, IsResponsive = 0.1, plus a long list of
features at Lasso retention 0 that Ridge keeps at 1.0. The "ask me
tomorrow and you'll get different features" failure mode shows up
where it matters — on the high-dimensional tier — and is suppressed
where the signal is concentrated (Lexical).

**Net.** All three claims hold. The one piece of the original
hypothesis that did **not** survive was the older-version threshold
"Jaccard(Lasso, Ridge) ≤ 0.5 at matched AUC" — measured Jaccard sits
in the 0.7–0.8 band because Ridge is a superset by construction, so
set overlap is the wrong instrument; coefficient *shape* and
fold-level retention rate are the right ones.


