---
title: "Phishing Detection — Scenario 2: Parametric vs Non-parametric"
author: "K-o-m-A"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(caret)
library(glmnet)
library(MASS, exclude = "select")
library(klaR)
library(randomForest)
library(kernlab)
library(pROC)
library(knitr)

set.seed(2026)

ARTIFACTS <- "scenario_2/artifacts"
dir.create(ARTIFACTS, recursive = TRUE, showWarnings = FALSE)
```

---

# 1. Scenario & Hypothesis

The EDA (`phishing.rmd` §2) formulates **H1** as: *on the URL-string-only
tier, non-parametric models extract significantly more signal than parametric
ones.* Initial modelling in the EDA also showed that on the **Full**
feature set (all 40 retained predictors) every model type already reaches
AUC ≈ 0.99 — the dataset is near-perfectly separable there, so the
parametric-vs-non-parametric contrast cannot be seen on Full alone.

To give H1 a fair test, we compare all six models across **five feature
tiers**:

| Tier | Features | What the model sees |
|------|---------:|---------------------|
| **Lexical** (URL-only) | 13 | URL string only — the first-line scenario: no DNS, no HTML fetch. |
| **Trust** | 7 | Domain / title / HTTPS flags — available after a DNS lookup and light page inspection. |
| **Behavior** | 20 | Full page-level content signals — iframes, forms, redirects, hidden fields. |
| **FullLite** | 34 | Full minus six page-count features that each give univariate AUC ≥ 0.95. Keeps > 30 predictors but avoids near-saturating signal. |
| **Full** | 40 | Union of all three — strongest baseline, near-perfect separability. |

`FullLite` drops `LineOfCode`, `NoOfExternalRef`, `NoOfImage`, `NoOfSelfRef`,
`NoOfJS`, `NoOfCSS`. A univariate ROC on the full dataset shows each of these
six features alone gives AUC between 0.958 and 0.990 — phishing pages are
typically shallow HTML shells with almost no scripts, images or external
links, so these counts are near-leakers even though they are legitimate
page-level signals, not computed scores. Keeping them in the feature set
pins every Full-tier model at AUC ≈ 1.0 and washes out the family contrast
we are trying to measure. `FullLite` therefore meets the "> 30 features"
rule while leaving enough room for models to disagree.

The smaller tiers (Lexical, Trust, Behavior) are deliberate ablations that
isolate H1's claim, which is specifically about the URL-only regime.
Scanning all five tiers lets us sharpen H1 from a single-tier claim into a
**gradient statement**: *the non-parametric advantage is largest when the
feature tier is weak and shrinks as more signal is added.*

---

# 2. Data Loading

## 2.1 Load raw file

The dataset is the same PhiUSIIL CSV used by the EDA. We strip the UTF-8
BOM that the provider attaches to the first column name.

```{r load-data}
dataset_path <- if (file.exists("PhiUSIIL_Phishing_URL_Dataset.csv")) {
  "PhiUSIIL_Phishing_URL_Dataset.csv"
} else if (file.exists("../PhiUSIIL_Phishing_URL_Dataset.csv")) {
  "../PhiUSIIL_Phishing_URL_Dataset.csv"
} else stop("PhiUSIIL_Phishing_URL_Dataset.csv not found")

raw <- read_csv(dataset_path, show_col_types = FALSE)
names(raw) <- map_chr(names(raw), ~ str_replace(.x, "^\ufeff", ""))
cat("Loaded", nrow(raw), "rows,", ncol(raw), "columns.\n")
```

## 2.2 Apply EDA feature exclusions

The EDA (§3.3) excluded three groups of features — computed
scores, a redundant binary flag, and linearly dependent ratios. Repeating
those exclusions here guarantees that Scenario 2 is tested on the **same
40 features** the EDA analysed, so the EDA's VIF, SMD, and skewness
findings are directly applicable.

```{r exclusions}
ID_COLS          <- c("FILENAME", "URL", "Domain", "TLD", "Title")
SCORES           <- c("URLSimilarityIndex", "TLDLegitimateProb", "URLCharProb",
                      "DomainTitleMatchScore", "URLTitleMatchScore")
REDUNDANT_BINARY <- "HasObfuscation"
RATIOS           <- c("LetterRatioInURL", "DegitRatioInURL",
                      "ObfuscationRatio", "SpacialCharRatioInURL")
EXCLUDE_COLS     <- c(ID_COLS, SCORES, REDUNDANT_BINARY, RATIOS)

df <- raw %>%
  mutate(label = factor(label, levels = c(1, 0),
                        labels = c("Phishing", "Legitimate"))) %>%
  select(-all_of(EXCLUDE_COLS)) %>%
  as.data.frame()

cat("Feature matrix after exclusions:", ncol(df) - 1, "predictors.\n")
```

## 2.3 Feature tiers

The family map is identical to EDA §3. Each tier is a subset of column
names, so switching tier means subsetting columns — no re-preprocessing.

```{r family-map}
family_map <- tribble(
  ~feature,                     ~family,
  "URLLength",                  "Lexical",
  "DomainLength",               "Lexical",
  "IsDomainIP",                 "Lexical",
  "CharContinuationRate",       "Lexical",
  "TLDLength",                  "Lexical",
  "NoOfSubDomain",              "Lexical",
  "NoOfObfuscatedChar",         "Lexical",
  "NoOfLettersInURL",           "Lexical",
  "NoOfDegitsInURL",            "Lexical",
  "NoOfEqualsInURL",            "Lexical",
  "NoOfQMarkInURL",             "Lexical",
  "NoOfAmpersandInURL",         "Lexical",
  "NoOfOtherSpecialCharsInURL", "Lexical",
  "IsHTTPS",                    "Trust",
  "HasTitle",                   "Trust",
  "HasFavicon",                 "Trust",
  "Robots",                     "Trust",
  "Bank",                       "Trust",
  "Pay",                        "Trust",
  "Crypto",                     "Trust",
  "LineOfCode",                 "Behavior",
  "LargestLineLength",          "Behavior",
  "IsResponsive",               "Behavior",
  "NoOfURLRedirect",            "Behavior",
  "NoOfSelfRedirect",           "Behavior",
  "HasDescription",             "Behavior",
  "NoOfPopup",                  "Behavior",
  "NoOfiFrame",                 "Behavior",
  "HasExternalFormSubmit",      "Behavior",
  "HasSocialNet",               "Behavior",
  "HasSubmitButton",            "Behavior",
  "HasHiddenFields",            "Behavior",
  "HasPasswordField",           "Behavior",
  "HasCopyrightInfo",           "Behavior",
  "NoOfImage",                  "Behavior",
  "NoOfCSS",                    "Behavior",
  "NoOfJS",                     "Behavior",
  "NoOfSelfRef",                "Behavior",
  "NoOfEmptyRef",               "Behavior",
  "NoOfExternalRef",            "Behavior"
)

LEAKY_BEHAVIOR <- c("LineOfCode", "NoOfExternalRef", "NoOfImage",
                    "NoOfSelfRef", "NoOfJS", "NoOfCSS")

tiers <- list(
  Lexical  = family_map %>% filter(family == "Lexical")  %>% pull(feature),
  Trust    = family_map %>% filter(family == "Trust")    %>% pull(feature),
  Behavior = family_map %>% filter(family == "Behavior") %>% pull(feature),
  FullLite = setdiff(family_map$feature, LEAKY_BEHAVIOR),
  Full     = family_map$feature
)

sapply(tiers, length)
```

---

# 3. Modelling Setup

## 3.1 Stratified sub-sampling

The full dataset is ~235 k rows. SVM with an RBF kernel is
\(O(n^2)\)–\(O(n^3)\) in the training size; the full dataset would take
hours per fold. We take a stratified 30 k sub-sample (15 k phishing /
15 k legitimate). After the 80/20 split this leaves 24 k rows per fold —
the largest size at which all six models finish in reasonable time.
Stratification keeps class balance constant across tiers and folds.

```{r subsample}
N_SUB <- 30000
df_sub <- df %>%
  group_by(label) %>%
  slice_sample(n = N_SUB / 2) %>%
  ungroup() %>%
  as.data.frame()

table(df_sub$label)
```

## 3.2 Shared train/test split and CV folds

Every model × tier combination uses **the same rows** for training, the
**same 80/20 split**, and the **same 5-fold CV indices**. The only thing
that varies between runs is which columns the model sees. This is what
makes the per-fold AUCs paired and lets us run a paired Wilcoxon test
later.

```{r splits}
idx       <- createDataPartition(df_sub$label, p = 0.8, list = FALSE)
train_raw <- df_sub[idx, ]
test_raw  <- df_sub[-idx, ]
fold_idx  <- createFolds(train_raw$label, k = 5, returnTrain = TRUE)

cat("Train:", nrow(train_raw), "Test:", nrow(test_raw),
    "CV folds:", length(fold_idx), "\n")
```

## 3.3 Preprocessing recipes

Two parallel recipes are needed:

- **Parametric / distance recipe** — `log1p()` on every continuous feature
  (the EDA flagged \|skew\| > 2 on 19 of 22 continuous predictors), then
  center-and-scale. Binary features are left untouched. Feeds LR, LDA, NB,
  SVM-RBF, KNN.
- **Tree recipe** — no transform. Trees are invariant to monotone feature
  transforms and to scale. Feeds Random Forest.

`preProcess` is fit on the full training set (no per-fold refitting). At
24 k rows with fixed hyperparameters this leakage is negligible and keeps
the code simple.

```{r recipes}
feat <- df_sub %>% select(-label)
binary_features     <- names(feat)[sapply(feat, function(x)
                                   all(na.omit(x) %in% c(0, 1)))]
continuous_features <- setdiff(names(feat), binary_features)

apply_log <- function(d, cols) {
  d[cols] <- lapply(d[cols], function(x) log1p(pmax(x, 0)))
  d
}
train_log <- apply_log(train_raw, continuous_features)
test_log  <- apply_log(test_raw,  continuous_features)

pp <- preProcess(train_log[, continuous_features], method = c("center", "scale"))
train_std <- train_log; test_std  <- test_log
train_std[, continuous_features] <- predict(pp, train_log[, continuous_features])
test_std[,  continuous_features] <- predict(pp, test_log[,  continuous_features])

cat(length(continuous_features), "continuous,",
    length(binary_features), "binary features.\n")
```

## 3.4 Cache preprocessing for downstream scenarios

`scenario_1.rmd` reuses the identical train/test split, CV fold indices
and feature tiers. Writing them to a single `.rds` file here means
Scenario 1 does not have to re-run any upstream code and is guaranteed
to see the same rows this notebook saw.

```{r cache-preprocessing}
saveRDS(
  list(
    train_std = train_std, test_std = test_std,
    train_raw = train_raw, test_raw = test_raw,
    fold_idx  = fold_idx,
    tiers     = tiers,
    family_map = family_map,
    binary_features     = binary_features,
    continuous_features = continuous_features
  ),
  file.path(ARTIFACTS, "preprocessing.rds")
)
cat("Preprocessing cached to ",
    file.path(ARTIFACTS, "preprocessing.rds"), "\n", sep = "")
```

## 3.5 Per-tier fit utility

Fitting the same model on four tiers is four nearly identical pieces of
code, so we wrap the logic once. The function subsets the data to a tier's
features, runs 5-fold CV using the shared `fold_idx`, and returns per-fold
AUCs plus test-set metrics.

```{r fit-utility}
fit_one_tier <- function(tier_features, data_train, data_test,
                         method, tuneGrid = NULL, ...) {
  tr <- data_train[, c("label", tier_features)]
  te <- data_test[,  c("label", tier_features)]

  ctrl <- trainControl(
    method          = "cv",
    index           = fold_idx,
    classProbs      = TRUE,
    summaryFunction = twoClassSummary,
    savePredictions = "final"
  )
  t0 <- Sys.time()
  fit <- caret::train(label ~ ., data = tr,
                      method = method, trControl = ctrl,
                      metric = "ROC", tuneGrid = tuneGrid, ...)
  elapsed <- as.numeric(difftime(Sys.time(), t0, units = "secs"))

  test_prob <- predict(fit, te, type = "prob")[, "Phishing"]
  test_pred <- predict(fit, te)
  test_auc  <- as.numeric(pROC::roc(te$label, test_prob,
                                    levels = c("Legitimate", "Phishing"),
                                    direction = "<", quiet = TRUE)$auc)
  cm <- caret::confusionMatrix(test_pred, te$label, positive = "Phishing")

  list(cv_per_fold = fit$resample,
       test_auc = test_auc,
       test_acc = unname(cm$overall["Accuracy"]),
       test_f1  = unname(cm$byClass["F1"]),
       test_sens = unname(cm$byClass["Sensitivity"]),
       test_spec = unname(cm$byClass["Specificity"]),
       train_secs = elapsed)
}

run_or_load <- function(name, fn) {
  path <- file.path(ARTIFACTS, paste0("res_", name, ".rds"))
  if (file.exists(path)) {
    cached <- readRDS(path)
    if (setequal(names(cached$by_tier), names(tiers))) return(cached)
    cat(sprintf("  [%s] cache tiers (%s) != expected (%s) — refitting\n",
                name, paste(names(cached$by_tier), collapse = ","),
                paste(names(tiers), collapse = ",")))
  }
  res <- fn(); saveRDS(res, path); res
}

print_tier_summary <- function(res) {
  for (tier in names(res$by_tier)) {
    r <- res$by_tier[[tier]]
    cat(sprintf("  %-15s [%-8s] CV ROC %.4f ± %.4f | test AUC %.4f | train %.1fs\n",
                res$model, tier,
                mean(r$cv_per_fold$ROC), sd(r$cv_per_fold$ROC),
                r$test_auc, r$train_secs))
  }
}
```

`run_or_load` is a small cache wrapper: the first knit fits everything
and writes `scenario_2/artifacts/res_<model>.rds`; subsequent knits load
those files and finish in seconds. When the tier set changes (e.g.
`FullLite` added) the cache is detected as stale and the model is refit.
`print_tier_summary` echoes one line per tier so every fit chunk shows
its CV ROC ± SD, held-out test AUC and wall-clock train time.

---

# 4. Parametric Models

## 4.1 Logistic Regression (ridge)

The EDA §4.3.3 showed a severe collinearity cluster around `URLLength`
(VIF > 1000) on the Lexical tier, and several |r| > 0.7 pairs on Full.
Ordinary (unpenalised) logistic regression becomes numerically unstable
under that collinearity — the coefficient estimates swing wildly between
folds. We therefore use **ridge-regularised** LR (α = 0, λ = 0.01), which
keeps coefficients bounded without doing any feature selection.

```{r fit-lr}
res_lr <- run_or_load("lr", function() {
  out <- list(model = "LogReg-Ridge", family = "parametric", by_tier = list())
  for (tier in names(tiers)) {
    cat(sprintf("  LogReg-Ridge   [%-8s] fitting (%d features)...\n",
                tier, length(tiers[[tier]])))
    out$by_tier[[tier]] <- fit_one_tier(
      tiers[[tier]], train_std, test_std,
      method = "glmnet",
      tuneGrid = data.frame(alpha = 0, lambda = 0.01))
  }
  out
})
print_tier_summary(res_lr)
```

## 4.2 Linear Discriminant Analysis (LDA)

LDA assumes Gaussian class-conditional densities with a shared covariance
matrix. The EDA showed heavy skewness on 19 of 22 continuous features;
`log1p` + center/scale in our preprocessing recipe partially repairs this,
so LDA is applicable in principle. LDA has no tunable hyperparameters,
which makes it a clean "linear baseline".

```{r fit-lda}
res_lda <- run_or_load("lda", function() {
  out <- list(model = "LDA", family = "parametric", by_tier = list())
  for (tier in names(tiers)) {
    cat(sprintf("  LDA            [%-8s] fitting (%d features)...\n",
                tier, length(tiers[[tier]])))
    out$by_tier[[tier]] <- fit_one_tier(
      tiers[[tier]], train_std, test_std, method = "lda")
  }
  out
})
print_tier_summary(res_lda)
```

## 4.3 Naive Bayes

NB assumes conditional **independence** of features given the class. EDA
§4.3 already showed that this assumption is violated: |r| > 0.7 between
URL-length cluster members, and between Behavior ref-count features. We
include NB knowing this, to quantify **how much the independence
assumption costs** on each tier. Kernel density is used so stratified
folds with zero-variance features (e.g. `Robots` on Trust) do not crash
the Gaussian version of NB.

```{r fit-nb, message=FALSE, warning=FALSE}
res_nb <- run_or_load("nb", function() {
  out <- list(model = "NaiveBayes", family = "parametric", by_tier = list())
  for (tier in names(tiers)) {
    cat(sprintf("  NaiveBayes     [%-8s] fitting (%d features)...\n",
                tier, length(tiers[[tier]])))
    out$by_tier[[tier]] <- suppressWarnings(fit_one_tier(
      tiers[[tier]], train_std, test_std,
      method = "nb",
      tuneGrid = data.frame(fL = 1, usekernel = TRUE, adjust = 1)))
  }
  out
})
print_tier_summary(res_nb)
```

---

# 5. Non-parametric Models

## 5.1 Random Forest

Random Forest is the natural non-parametric counterpart to LR: it fits
piecewise-constant decision regions via ensembled trees, and is invariant
to monotone feature transforms. `mtry = floor(sqrt(p))` scales the random
feature subset with the tier size; 500 trees is more than enough for AUC
to stabilise on this sample size (per EDA skewness + class balance).

```{r fit-rf}
res_rf <- run_or_load("rf", function() {
  out <- list(model = "RandomForest", family = "non-parametric", by_tier = list())
  for (tier in names(tiers)) {
    p    <- length(tiers[[tier]])
    mtry <- max(1, floor(sqrt(p)))
    cat(sprintf("  RandomForest   [%-8s] fitting (%d features, mtry=%d)...\n",
                tier, p, mtry))
    out$by_tier[[tier]] <- fit_one_tier(
      tiers[[tier]], train_raw, test_raw,
      method = "rf",
      tuneGrid = data.frame(mtry = mtry),
      ntree = 500)
  }
  out
})
print_tier_summary(res_rf)
```

## 5.2 SVM with RBF kernel

The RBF kernel fits smooth non-linear boundaries — a natural alternative
to tree ensembles and distance-based methods. RBF distances are
scale-sensitive, so SVM uses the parametric recipe (`log1p` + center /
scale). `C = 1, σ = 0.1` are sensible defaults for standardised inputs;
we did not tune further because the main question is the parametric vs
non-parametric contrast, not within-family optimisation.

This is the single slowest chunk in the notebook; first knit takes
roughly 15 minutes on the four tiers combined, re-knit is instant (cached
by `run_or_load`).

```{r fit-svm}
res_svm <- run_or_load("svm", function() {
  out <- list(model = "SVM-RBF", family = "non-parametric", by_tier = list())
  for (tier in names(tiers)) {
    cat(sprintf("  SVM-RBF        [%-8s] fitting (%d features)...\n",
                tier, length(tiers[[tier]])))
    out$by_tier[[tier]] <- fit_one_tier(
      tiers[[tier]], train_std, test_std,
      method = "svmRadial",
      tuneGrid = data.frame(C = 1, sigma = 0.1))
  }
  out
})
print_tier_summary(res_svm)
```

## 5.3 k-Nearest Neighbours

KNN is the simplest non-parametric classifier: it stores the training set
and classifies each test point by majority vote among its `k` closest
neighbours under Euclidean distance. Distance-based → it consumes the
parametric recipe. `k = 25` is a mild smoother that keeps the decision
surface from being jagged.

One twist: when a tier consists **only of binary features** (the Trust
tier has 7 binary indicators), many training points sit at identical
coordinates, and caret's `knn3` refuses to run with "too many ties in
knn". We add a microscopic Gaussian jitter (SD = 10⁻³, less than 0.1 %
of feature scale) to break distance ties without changing decisions.

```{r fit-knn}
jitter_features <- function(d, cols, sd = 1e-3) {
  for (c in cols) d[[c]] <- d[[c]] + rnorm(nrow(d), 0, sd)
  d
}

res_knn <- run_or_load("knn", function() {
  out <- list(model = "KNN", family = "non-parametric", by_tier = list())
  for (tier in names(tiers)) {
    feats <- tiers[[tier]]
    cat(sprintf("  KNN            [%-8s] fitting (%d features)...\n",
                tier, length(feats)))
    tr_j  <- jitter_features(train_std, feats)
    te_j  <- jitter_features(test_std,  feats)
    out$by_tier[[tier]] <- fit_one_tier(
      feats, tr_j, te_j,
      method = "knn",
      tuneGrid = data.frame(k = 25))
  }
  out
})
print_tier_summary(res_knn)
```

---

# 6. Comparison

## 6.1 Per-(model, tier) summary

We first assemble a long table with one row per `(model, tier, fold)` and
then a wide summary of mean ± sd CV AUC plus held-out test metrics.

```{r build-long}
all_results <- list(res_lr, res_lda, res_nb, res_rf, res_svm, res_knn)

auc_long <- map_dfr(all_results, function(r) {
  map_dfr(names(r$by_tier), function(tier)
    r$by_tier[[tier]]$cv_per_fold %>%
      transmute(model = r$model, family = r$family, tier = tier,
                fold = Resample, auc = ROC))
})

test_tbl <- map_dfr(all_results, function(r) {
  map_dfr(names(r$by_tier), function(tier) {
    t <- r$by_tier[[tier]]
    tibble(model = r$model, family = r$family, tier = tier,
           test_auc = t$test_auc, test_acc = t$test_acc,
           test_f1  = t$test_f1,  train_secs = t$train_secs)
  })
})

summary_tbl <- auc_long %>%
  group_by(model, family, tier) %>%
  summarise(cv_auc_mean = mean(auc), cv_auc_sd = sd(auc), .groups = "drop") %>%
  left_join(test_tbl, by = c("model", "family", "tier")) %>%
  mutate(tier = factor(tier, levels = c("Lexical", "Trust",
                                        "Behavior", "FullLite", "Full"))) %>%
  arrange(tier, family, desc(cv_auc_mean))

summary_tbl %>%
  mutate(across(c(cv_auc_mean, cv_auc_sd, test_auc, test_acc, test_f1),
                ~ round(.x, 4)),
         train_secs = round(train_secs, 1)) %>%
  kable()
```

**How to read this table:** within each tier the two families should
cluster differently if H1 holds. On Lexical, the three non-parametric
rows should sit well above the three parametric ones; as we move through
Trust → Behavior → FullLite → Full, the gap should shrink and eventually
vanish. FullLite is the most informative high-dimensional check because
it keeps > 30 features but removes the six near-leaky Behavior counts —
if parametric and non-parametric models still disagree there, the
gradient claim survives at realistic feature sizes.

## 6.2 Per-tier paired Wilcoxon test

For each tier we take the per-fold mean AUC of the three non-parametric
models and of the three parametric models, and compare them with a
one-sided paired Wilcoxon (non-parametric > parametric). Because we share
folds across models, the test is fully paired: each row in the compared
series comes from the same five validation sets.

```{r h1-test}
per_tier_test <- auc_long %>%
  group_by(tier, fold, family) %>%
  summarise(auc = mean(auc), .groups = "drop") %>%
  pivot_wider(names_from = family, values_from = auc) %>%
  group_by(tier) %>%
  summarise(
    mean_np  = mean(`non-parametric`),
    mean_p   = mean(parametric),
    diff     = mean(`non-parametric` - parametric),
    wilcox_V = wilcox.test(`non-parametric`, parametric,
                           paired = TRUE, alternative = "greater")$statistic,
    wilcox_p = wilcox.test(`non-parametric`, parametric,
                           paired = TRUE, alternative = "greater")$p.value,
    .groups = "drop"
  ) %>%
  mutate(tier = factor(tier, levels = c("Lexical", "Trust",
                                        "Behavior", "FullLite", "Full"))) %>%
  arrange(tier)

per_tier_test %>%
  mutate(across(c(mean_np, mean_p, diff), ~ round(.x, 4)),
         wilcox_p = signif(wilcox_p, 3)) %>%
  kable()
```

A caveat on the p-value column: with only **five paired observations**
(one per CV fold) the smallest p-value a one-sided paired Wilcoxon can
ever produce is \(1/32 \approx 0.031\). Three of the four tiers reach
exactly this floor — meaning every non-parametric fold beat the matched
parametric fold. The p-value alone therefore does not distinguish a
large effect (Lexical, diff ≈ 0.10) from a negligible one (Behavior,
diff ≈ 0.0004). **Read the `diff` column, not the p-value**, to judge
effect size.

## 6.3 CV AUC by model × tier

```{r plot-auc, fig.width=11, fig.height=5, message=FALSE}
p <- auc_long %>%
  mutate(tier  = factor(tier, levels = c("Lexical", "Trust",
                                         "Behavior", "FullLite", "Full")),
         model = fct_reorder(model, auc, median)) %>%
  ggplot(aes(model, auc, colour = family)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, fill = NA) +
  geom_jitter(width = 0.08, height = 0, size = 1.6, alpha = 0.6) +
  facet_wrap(~ tier, nrow = 1) +
  scale_colour_manual(values = c(parametric = "steelblue",
                                 `non-parametric` = "firebrick")) +
  labs(
    title    = "Scenario 2 — CV AUC by model across five feature tiers",
    subtitle = "Each dot is one CV fold; shared folds across models",
    x = NULL, y = "ROC AUC (5-fold CV)"
  ) +
  coord_flip() +
  theme_minimal(base_size = 11)

ggsave(file.path(ARTIFACTS, "scenario2_auc_by_tier.png"), p,
       width = 11, height = 5, dpi = 150)
p
```

The Lexical panel (leftmost) shows the widest spread and the clearest
separation between parametric (blue) and non-parametric (red) models.
Moving right, the boxes tighten and the two colour groups converge.
The rightmost two panels (FullLite, Full) pin every model at AUC ≈ 1,
though FullLite — with six near-leaky Behavior counts removed — retains
slightly more spread than Full and is the fairer "many-features" tier
for judging the family contrast.

## 6.4 H1 verdict against the pre-registered criteria

`phishing.rmd` §2 pins three measurable criteria for H1. We evaluate
each one directly from the per-fold AUC table and report pass / fail
rather than relying on narrative inspection.

```{r h1-verdict}
fold_family <- auc_long %>%
  group_by(tier, fold, family) %>%
  summarise(auc = mean(auc), .groups = "drop") %>%
  pivot_wider(names_from = family, values_from = auc) %>%
  mutate(diff = `non-parametric` - parametric)

diff_lexical  <- fold_family %>% filter(tier == "Lexical")  %>% pull(diff)
diff_fulllite <- fold_family %>% filter(tier == "FullLite") %>% pull(diff)

delta_lex       <- mean(diff_lexical)
delta_fulllite  <- mean(diff_fulllite)
fold_unanimous  <- all(diff_lexical > 0)
wilcox_lex_p    <- wilcox.test(diff_lexical, alternative = "greater")$p.value

c1_pass <- delta_lex >= 0.05
c1_weak <- delta_lex >= 0.02 && delta_lex < 0.05
c2_pass <- (delta_lex > delta_fulllite) && (delta_fulllite >= 0)
c3_pass <- fold_unanimous

verdict <- tibble(
  Criterion = c(
    "1. Primary — ΔAUC_Lexical ≥ 0.05",
    "2. Secondary — ΔAUC_Lexical > ΔAUC_FullLite ≥ 0",
    "3. Tertiary — Fold sign unanimous on Lexical (p = 0.031)"
  ),
  Realised = c(
    sprintf("ΔAUC_Lexical = %.4f", delta_lex),
    sprintf("ΔAUC_Lexical = %.4f vs ΔAUC_FullLite = %.4f",
            delta_lex, delta_fulllite),
    sprintf("%d / 5 folds non-param > param (Wilcoxon p = %.3f)",
            sum(diff_lexical > 0), wilcox_lex_p)
  ),
  Status = c(
    if (c1_pass) "PASS" else if (c1_weak) "WEAK" else "FAIL",
    if (c2_pass) "PASS" else "FAIL",
    if (c3_pass) "PASS" else "FAIL"
  )
)
kable(verdict)

falsified <-
  (delta_lex < 0.02) ||
  (delta_fulllite >= delta_lex) ||
  (!fold_unanimous)

supported <- c1_pass && c2_pass && c3_pass

cat(sprintf(
  "\nOverall H1 verdict: %s\n",
  if (supported) "SUPPORTED (all three criteria pass)"
  else if (falsified) "FALSIFIED (see falsification rule in EDA §2)"
  else "PARTIAL (some criteria weak but none triggers falsification)"
))
```

This chunk closes the hypothesis loop: the criteria are set in the EDA,
the numbers come from Scenario 2's shared folds, and the verdict is
machine-produced rather than argued in prose.

---

# 7. Conclusion — H1 gradient

H1 from the EDA was a single-tier claim about URL-only detection. The
five-tier comparison lets us state the stronger, more informative
version:

1. **Non-parametric > parametric when the feature tier is weak.** On
   the **Lexical** tier the non-parametric advantage is **+0.101 AUC**
   (0.986 vs 0.884), and every single CV fold shows the same ordering.
   This is exactly the regime H1 targets: URL counts interact
   non-linearly (digit runs × length × subdomain depth), and linear or
   Gaussian decision boundaries cannot exploit these interactions.
2. **Trust tier also favours non-parametric, but only marginally**
   (diff ≈ +0.003). With only seven features — most of them near-pure
   binary flags — there is little non-linear structure left for
   non-parametric models to find.
3. **Both families converge on rich tiers.** On Behavior every model
   hits AUC ≈ 1 (diff ≈ +0.0004) because six page-count features each
   individually give univariate AUC ≥ 0.95. FullLite removes those six
   near-leakers but still saturates — any remaining Behavior signal is
   enough to separate classes linearly. Full adds them back and simply
   confirms the ceiling.
4. **FullLite is the honest > 30-feature benchmark.** It meets the
   assignment's "several dozen parameters" rule without relying on
   features that single-handedly solve the task. The family contrast
   on FullLite tells us whether non-parametric flexibility still buys
   anything in a realistic high-dimensional regime — read its row in
   Section 6.2 for the decisive diff.

**Practical takeaway.** A URL-only first-line phishing detector benefits
substantially from a non-parametric model (SVM-RBF or KNN gain ~0.10 AUC
over LR / NB on Lexical). Once the system can afford to fetch the page
and compute Behavior-level features, model choice becomes almost
irrelevant and simpler, more explainable models (LR, LDA) are good
enough. That is exactly the rationale for deploying a fast
non-parametric URL-only filter *in front of* a slower page-level
analyser, rather than relying on a single model at either extreme.
