---
title: "Phishing Detection — Scenario 2: Parametric vs Non-parametric"
author: "K-o-m-A"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(caret)
library(glmnet)
library(MASS, exclude = "select")
library(klaR)
library(randomForest)
library(kernlab)
library(pROC)
library(knitr)
library(digest)

set.seed(2026)

ARTIFACTS <- "scenario_2/artifacts"
dir.create(ARTIFACTS, recursive = TRUE, showWarnings = FALSE)
```

---

# 1. Scenario & Hypothesis

The EDA (`phishing.rmd` §2) states **H1** as: *on the URL-only tier,
non-parametric models pick up meaningfully more signal than parametric
ones.* Section 5 of the EDA also points out that on the **Full** feature
set (all 40 remaining predictors) every model type already reaches
AUC ≈ 0.99 — the dataset is near-perfectly separable there, so the
parametric-vs-non-parametric contrast can't be seen on Full alone.

To give H1 a fair test, we compare all six models across **four feature
tiers**:

| Tier | Features | What the model sees |
|------|---------:|---------------------|
| **Lexical** (URL-only) | 13 | URL string only — the first-line case: no DNS, no HTML fetch. |
| **Trust** | 7 | Domain / title / HTTPS flags — available after a DNS lookup and light page inspection. |
| **Behavior** | 14 | Page-level content signals (iframes, forms, redirects, hidden fields), **with the six near-leaky page counts removed** for the same reason we remove them from FullLite. |
| **FullLite** | 34 | All three families combined minus those six near-leakers. The high-dimensional benchmark (> 30 predictors) without single-feature near-saturating signal. |

Six Behavior features — `LineOfCode`, `NoOfExternalRef`, `NoOfImage`,
`NoOfSelfRef`, `NoOfJS`, `NoOfCSS` — each individually give a univariate
AUC between 0.958 and 0.990 on this dataset. Phishing pages are usually
shallow HTML shells with almost no scripts, images or external links, so
these counts effectively solve the task on their own, even though they
are genuine page-level signals rather than computed scores. Leaving them
in any tier pins every model at AUC ≈ 1.0 and flattens the family
contrast we are trying to measure. We therefore exclude them
**consistently everywhere they would otherwise appear** — both from
Behavior and from the high-dimensional `FullLite` tier. A "Full" tier
that puts them back is redundant once `FullLite` already saturates (see
the results tables in §6), and it would only let us defend the same
AUC ≈ 1.0 twice.

The smaller tiers (Lexical, Trust, Behavior) are deliberate ablations that
isolate H1's claim, which is specifically about the URL-only regime. By
scanning all four tiers we can sharpen H1 from a single-tier claim into a
**gradient statement**: *the non-parametric advantage is largest when the
feature tier is weak and shrinks as more signal is added.*

---

# 2. Data Loading

## 2.1 Load raw file

Same PhiUSIIL CSV as in the EDA. We strip the UTF-8 BOM the provider
attaches to the first column name.

```{r load-data}
dataset_path <- if (file.exists("PhiUSIIL_Phishing_URL_Dataset.csv")) {
  "PhiUSIIL_Phishing_URL_Dataset.csv"
} else if (file.exists("../PhiUSIIL_Phishing_URL_Dataset.csv")) {
  "../PhiUSIIL_Phishing_URL_Dataset.csv"
} else stop("PhiUSIIL_Phishing_URL_Dataset.csv not found")

raw <- read_csv(dataset_path, show_col_types = FALSE)
names(raw) <- map_chr(names(raw), ~ str_replace(.x, "^\ufeff", ""))
cat("Loaded", nrow(raw), "rows,", ncol(raw), "columns.\n")
```

## 2.2 Apply EDA feature exclusions

The EDA (§3.3) dropped three groups of features: computed scores, one
redundant binary flag, and linearly dependent ratios. Repeating the same
exclusions here makes sure Scenario 2 is tested on the **same 40 features**
the EDA analysed, so the EDA's VIF, SMD and skewness findings apply here
directly.

```{r exclusions}
ID_COLS          <- c("FILENAME", "URL", "Domain", "TLD", "Title")
SCORES           <- c("URLSimilarityIndex", "TLDLegitimateProb", "URLCharProb",
                      "DomainTitleMatchScore", "URLTitleMatchScore")
REDUNDANT_BINARY <- "HasObfuscation"
RATIOS           <- c("LetterRatioInURL", "DegitRatioInURL",
                      "ObfuscationRatio", "SpacialCharRatioInURL")
EXCLUDE_COLS     <- c(ID_COLS, SCORES, REDUNDANT_BINARY, RATIOS)

df <- raw %>%
  mutate(label = factor(label, levels = c(1, 0),
                        labels = c("Phishing", "Legitimate"))) %>%
  select(-all_of(EXCLUDE_COLS)) %>%
  as.data.frame()

cat("Feature matrix after exclusions:", ncol(df) - 1, "predictors.\n")
```

## 2.3 Feature tiers

The family map is the same as in EDA §3. Each tier is just a subset of
column names, so switching tier means subsetting columns — no
re-preprocessing needed.

```{r family-map}
family_map <- tribble(
  ~feature,                     ~family,
  "URLLength",                  "Lexical",
  "DomainLength",               "Lexical",
  "IsDomainIP",                 "Lexical",
  "CharContinuationRate",       "Lexical",
  "TLDLength",                  "Lexical",
  "NoOfSubDomain",              "Lexical",
  "NoOfObfuscatedChar",         "Lexical",
  "NoOfLettersInURL",           "Lexical",
  "NoOfDegitsInURL",            "Lexical",
  "NoOfEqualsInURL",            "Lexical",
  "NoOfQMarkInURL",             "Lexical",
  "NoOfAmpersandInURL",         "Lexical",
  "NoOfOtherSpecialCharsInURL", "Lexical",
  "IsHTTPS",                    "Trust",
  "HasTitle",                   "Trust",
  "HasFavicon",                 "Trust",
  "Robots",                     "Trust",
  "Bank",                       "Trust",
  "Pay",                        "Trust",
  "Crypto",                     "Trust",
  "LineOfCode",                 "Behavior",
  "LargestLineLength",          "Behavior",
  "IsResponsive",               "Behavior",
  "NoOfURLRedirect",            "Behavior",
  "NoOfSelfRedirect",           "Behavior",
  "HasDescription",             "Behavior",
  "NoOfPopup",                  "Behavior",
  "NoOfiFrame",                 "Behavior",
  "HasExternalFormSubmit",      "Behavior",
  "HasSocialNet",               "Behavior",
  "HasSubmitButton",            "Behavior",
  "HasHiddenFields",            "Behavior",
  "HasPasswordField",           "Behavior",
  "HasCopyrightInfo",           "Behavior",
  "NoOfImage",                  "Behavior",
  "NoOfCSS",                    "Behavior",
  "NoOfJS",                     "Behavior",
  "NoOfSelfRef",                "Behavior",
  "NoOfEmptyRef",               "Behavior",
  "NoOfExternalRef",            "Behavior"
)

LEAKY_BEHAVIOR <- c("LineOfCode", "NoOfExternalRef", "NoOfImage",
                    "NoOfSelfRef", "NoOfJS", "NoOfCSS")

tiers <- list(
  Lexical  = family_map %>% filter(family == "Lexical")  %>% pull(feature),
  Trust    = family_map %>% filter(family == "Trust")    %>% pull(feature),
  Behavior = setdiff(family_map %>% filter(family == "Behavior") %>% pull(feature),
                     LEAKY_BEHAVIOR),
  FullLite = setdiff(family_map$feature, LEAKY_BEHAVIOR)
)

sapply(tiers, length)
```

---

# 3. Modelling Setup

## 3.1 Stratified sub-sampling

The full dataset has ~235 k rows. SVM with an RBF kernel is
\(O(n^2)\)–\(O(n^3)\) in the training size, so the full dataset would
take hours per fold. Instead we take a stratified 30 k sub-sample
(15 k phishing / 15 k legitimate). After the 80/20 split this leaves
24 k rows per fold — the largest size where all six models still finish
in reasonable time. Stratification keeps the class balance constant
across tiers and folds.

```{r subsample}
N_SUB <- 30000
df_sub <- df %>%
  group_by(label) %>%
  slice_sample(n = N_SUB / 2) %>%
  ungroup() %>%
  as.data.frame()

table(df_sub$label)
```

## 3.2 Shared train/test split and CV folds

Every model × tier combination uses **the same rows** for training, the
**same 80/20 split**, and the **same 10-fold CV indices**. The only thing
that changes between runs is which columns the model gets to see. That
is what makes the per-fold AUCs paired, so we can run a paired Wilcoxon
test later.

```{r splits}
idx       <- createDataPartition(df_sub$label, p = 0.8, list = FALSE)
train_raw <- df_sub[idx, ]
test_raw  <- df_sub[-idx, ]
fold_idx  <- createFolds(train_raw$label, k = 10, returnTrain = TRUE)

cat("Train:", nrow(train_raw), "Test:", nrow(test_raw),
    "CV folds:", length(fold_idx), "\n")
```

## 3.3 Preprocessing recipes

We need two parallel recipes:

- **Parametric / distance recipe** — `log1p()` on every continuous
  feature (the EDA flagged \|skew\| > 2 on 19 of 22 continuous
  predictors), then centre and scale. Binary features are left alone.
  Feeds LR, LDA, NB, SVM-RBF, KNN.
- **Tree recipe** — no transform. Trees don't care about monotone
  transforms or scale. Feeds Random Forest.

`preProcess` is fit once on the full training set (no per-fold refit).
At 24 k rows with fixed hyperparameters the leakage is negligible, and
it keeps the code simple.

```{r recipes}
feat <- df_sub %>% select(-label)
binary_features     <- names(feat)[sapply(feat, function(x)
                                   all(na.omit(x) %in% c(0, 1)))]
continuous_features <- setdiff(names(feat), binary_features)

apply_log <- function(d, cols) {
  d[cols] <- lapply(d[cols], function(x) log1p(pmax(x, 0)))
  d
}
train_log <- apply_log(train_raw, continuous_features)
test_log  <- apply_log(test_raw,  continuous_features)

pp <- preProcess(train_log[, continuous_features], method = c("center", "scale"))
train_std <- train_log; test_std  <- test_log
train_std[, continuous_features] <- predict(pp, train_log[, continuous_features])
test_std[,  continuous_features] <- predict(pp, test_log[,  continuous_features])

cat(length(continuous_features), "continuous,",
    length(binary_features), "binary features.\n")
```

## 3.4 Per-tier fit utility

Fitting the same model across four tiers would mean four almost identical
blocks of code, so we wrap the logic in one function. It subsets the
data to a tier's features, runs 10-fold CV using the shared `fold_idx`,
and returns per-fold AUCs plus test-set metrics.

```{r fit-utility}
fit_one_tier <- function(tier_features, data_train, data_test,
                         method, tuneGrid = NULL, ...) {
  tr <- data_train[, c("label", tier_features)]
  te <- data_test[,  c("label", tier_features)]

  ctrl <- trainControl(
    method          = "cv",
    index           = fold_idx,
    classProbs      = TRUE,
    summaryFunction = twoClassSummary,
    savePredictions = "final"
  )
  t0 <- Sys.time()
  fit <- caret::train(label ~ ., data = tr,
                      method = method, trControl = ctrl,
                      metric = "ROC", tuneGrid = tuneGrid, ...)
  elapsed <- as.numeric(difftime(Sys.time(), t0, units = "secs"))

  test_prob <- predict(fit, te, type = "prob")[, "Phishing"]
  test_pred <- predict(fit, te)
  test_auc  <- as.numeric(pROC::roc(te$label, test_prob,
                                    levels = c("Legitimate", "Phishing"),
                                    direction = "<", quiet = TRUE)$auc)
  cm <- caret::confusionMatrix(test_pred, te$label, positive = "Phishing")

  train_prob <- predict(fit, tr, type = "prob")[, "Phishing"]
  train_pred <- predict(fit, tr)
  train_auc  <- as.numeric(pROC::roc(tr$label, train_prob,
                                     levels = c("Legitimate", "Phishing"),
                                     direction = "<", quiet = TRUE)$auc)
  train_cm   <- caret::confusionMatrix(train_pred, tr$label, positive = "Phishing")

  list(cv_per_fold = fit$resample,
       train_auc = train_auc,
       train_acc = unname(train_cm$overall["Accuracy"]),
       test_auc = test_auc,
       test_acc = unname(cm$overall["Accuracy"]),
       test_f1  = unname(cm$byClass["F1"]),
       test_prec = unname(cm$byClass["Precision"]),
       test_sens = unname(cm$byClass["Sensitivity"]),
       test_spec = unname(cm$byClass["Specificity"]),
       train_secs = elapsed)
}

tiers_fingerprint <- function(x)
  digest::digest(lapply(x[sort(names(x))], sort))

run_or_load <- function(name, fn) {
  path <- file.path(ARTIFACTS, paste0("res_", name, ".rds"))
  fp   <- tiers_fingerprint(tiers)
  if (file.exists(path)) {
    cached <- readRDS(path)
    tiers_ok   <- setequal(names(cached$by_tier), names(tiers))
    schema_ok  <- all(c("train_auc", "train_acc", "test_prec") %in%
                      names(cached$by_tier[[1]]))
    fp_ok      <- isTRUE(cached$tiers_fp == fp)
    if (tiers_ok && schema_ok && fp_ok) return(cached)
    cat(sprintf("  [%s] cache stale (tiers_ok=%s, schema_ok=%s, fp_ok=%s) - refitting\n",
                name, tiers_ok, schema_ok, fp_ok))
  }
  res <- fn()
  res$tiers_fp <- fp
  saveRDS(res, path)
  res
}

print_tier_summary <- function(res) {
  tbl <- map_dfr(names(res$by_tier), function(tier) {
    r <- res$by_tier[[tier]]
    gap <- r$train_auc - r$test_auc
    tibble(
      Tier        = tier,
      `Train AUC` = sprintf("%.4f", r$train_auc),
      `CV ROC`    = sprintf("%.4f +/- %.4f",
                            mean(r$cv_per_fold$ROC), sd(r$cv_per_fold$ROC)),
      `Test AUC`  = sprintf("%.4f", r$test_auc),
      `Gap`       = sprintf("%+.4f", gap),
      Accuracy    = sprintf("%.4f", r$test_acc),
      F1          = sprintf("%.4f", r$test_f1),
      Prec        = sprintf("%.4f", r$test_prec),
      Sens        = sprintf("%.4f", r$test_sens),
      Spec        = sprintf("%.4f", r$test_spec),
      `Train (s)` = sprintf("%.1f", r$train_secs)
    )
  })
  knitr::kable(
    tbl,
    align   = "lrrrrrrrrrr",
    caption = sprintf("%s - %s",
                      res$model, res$family)
  )
}
```

`run_or_load` is a small cache wrapper: the first knit fits everything
and writes `scenario_2/artifacts/res_<model>.rds`; later knits just load
those files and finish in seconds. If the tier set on disk doesn't match
the tier set defined in the notebook any more, the cache is flagged as
stale and the model is refit automatically. `print_tier_summary` returns a `kable` table (one row per tier) with
CV ROC ± SD, held-out test AUC, Accuracy, F1, Sensitivity, Specificity,
and wall-clock train time. Each fit chunk therefore produces a small
model-titled table rather than a dense one-liner per tier.

---

# 4. Parametric Models

## 4.1 Logistic Regression (ridge)

EDA §4.3 showed a severe collinearity cluster around `URLLength`
(VIF > 1000) on the Lexical tier, plus several |r| > 0.7 pairs on the
full 40-feature set.
Plain (unpenalised) logistic regression becomes numerically unstable in
that setting — the coefficients swing wildly from fold to fold. So we
use **ridge-regularised** LR (α = 0, λ = 0.01), which keeps the
coefficients bounded without doing any feature selection.

```{r fit-lr}
res_lr <- run_or_load("lr", function() {
  out <- list(model = "LogReg-Ridge", family = "parametric", by_tier = list())
  for (tier in names(tiers)) {
    cat(sprintf("  LogReg-Ridge   [%-8s] fitting (%d features)...\n",
                tier, length(tiers[[tier]])))
    out$by_tier[[tier]] <- fit_one_tier(
      tiers[[tier]], train_std, test_std,
      method = "glmnet",
      tuneGrid = data.frame(alpha = 0, lambda = 0.01))
  }
  out
})
print_tier_summary(res_lr)
```

On **Lexical** LR-Ridge sits at AUC 0.852
with Sens 0.921 / Spec 0.716 — at threshold 0.5 the model blocks
roughly 28 % of legitimate URLs to keep 92 % of phish. For the
corporate proxy this is the "unusable filter" regime: the false-block
rate is an order of magnitude above what a helpdesk can absorb.
From **Trust** onwards the Sens/Spec gap closes (≈ 0.86–0.99 across both
axes).

LDA assumes Gaussian class-conditional densities with a shared covariance
matrix. The EDA flagged heavy skewness on 19 of 22 continuous features;
`log1p` + centre/scale in our preprocessing recipe cleans this up
reasonably well, so LDA is fine in principle. LDA has no tunable
hyperparameters, which makes it a clean "linear baseline".

```{r fit-lda}
res_lda <- run_or_load("lda", function() {
  out <- list(model = "LDA", family = "parametric", by_tier = list())
  for (tier in names(tiers)) {
    cat(sprintf("  LDA            [%-8s] fitting (%d features)...\n",
                tier, length(tiers[[tier]])))
    out$by_tier[[tier]] <- fit_one_tier(
      tiers[[tier]], train_std, test_std, method = "lda")
  }
  out
})
print_tier_summary(res_lda)
```

On **Lexical** LDA posts AUC 0.930 —
visibly better than LR — yet the threshold-0.5 operating point is
*worse*: Sens 0.971 / Spec 0.688, i.e. ~31 % of legitimate URLs
blocked. The Gaussian class-conditional assumption pushes the decision
surface toward the legit side of the score distribution, so a good
rank metric ends up with a badly miscalibrated cut. On **Trust**
(0.860 / 0.861) and **FullLite** (0.989 / 0.994) the Sens/Spec values
collapse to a tight ±0.005 band; **Behavior** (0.905 / 0.981) keeps a
residual ~8 pp skew toward Spec, but the asymmetry is a fraction of
the Lexical gap. This is the cleanest
case in the report where AUC and the proxy-relevant operating point
disagree: LDA beats LR on AUC but is *more* unusable as a Lexical
block/allow filter without threshold recalibration.

NB assumes features are conditionally **independent** given the class.
EDA §4.3 already showed that this is not true here: |r| > 0.7 inside
the URL-length cluster and between Behavior ref-count features. We run
NB anyway, to measure **how much the independence assumption actually
costs** on each tier. We use a kernel density estimate so that folds
with a zero-variance feature don't crash the Gaussian version of NB.

```{r fit-nb, message=FALSE, warning=FALSE}
res_nb <- run_or_load("nb", function() {
  out <- list(model = "NaiveBayes", family = "parametric", by_tier = list())
  for (tier in names(tiers)) {
    cat(sprintf("  NaiveBayes     [%-8s] fitting (%d features)...\n",
                tier, length(tiers[[tier]])))
    out$by_tier[[tier]] <- suppressWarnings(fit_one_tier(
      tiers[[tier]], train_std, test_std,
      method = "nb",
      tuneGrid = data.frame(fL = 1, usekernel = TRUE, adjust = 1)))
  }
  out
})
print_tier_summary(res_nb)
```

NB is the textbook case of AUC hiding
operating-point pathology. On **Lexical** AUC is 0.849 — not terrible
— but Sens 0.999 / Spec **0.383** means the model calls phishing for
nearly two thirds of legitimate URLs at threshold 0.5. This is the
direct empirical fingerprint of the broken independence assumption
flagged in EDA §4.3: the URL-length cluster (|r| > 0.7) is counted as
independent evidence six or seven times over, so phish-consistent
scores collapse toward 1. On **Behavior** the same violation *flips
direction* — Sens 0.732 / Spec 0.996 — because the correlated block
this time is the page-count ref group, and its repeated "evidence"
pulls probabilities the other way. AUC on Behavior is a respectable
0.990, yet at the default threshold NB misses 27 % of phish. In proxy
terms: NB on either tier needs recalibration or a tuned threshold
before it is deployable at all.

## 5.1 Random Forest

Random Forest is the natural non-parametric counterpart to LR: it builds
piecewise-constant decision regions as an ensemble of trees and is
invariant to monotone feature transforms. `mtry = floor(sqrt(p))` scales
the random feature subset with the tier size. We use **`ntree = 300`**:
OOB AUC on this dataset converges at ~200–300 trees, and going higher
spends extra compute without measurable AUC gain.

```{r fit-rf}
res_rf <- run_or_load("rf", function() {
  out <- list(model = "RandomForest", family = "non-parametric", by_tier = list())
  for (tier in names(tiers)) {
    p    <- length(tiers[[tier]])
    mtry <- max(1, floor(sqrt(p)))
    cat(sprintf("  RandomForest   [%-8s] fitting (%d features, mtry=%d)...\n",
                tier, p, mtry))
    out$by_tier[[tier]] <- fit_one_tier(
      tiers[[tier]], train_raw, test_raw,
      method = "rf",
      tuneGrid = data.frame(mtry = mtry),
      ntree = 300)
  }
  out
})
print_tier_summary(res_rf)
```

On **Lexical** RF posts AUC 0.969 but the
threshold-0.5 split is Sens 0.967 / Spec 0.745 — Spec is the soft
spot, driven by the bagging vote under deep unpruned trees (leaves
that see few legits vote "phish"). From **Trust** upward the model
becomes symmetric: Behavior Sens 0.972 / Spec 0.978, FullLite 0.997 /
0.996. RF is therefore the model that would most benefit from a
simple **threshold shift** on Lexical — the AUC is already there, the
0.5 cut just isn't the cost-optimal one. Unlike NB the problem is not
a structural model assumption, it's a calibration artefact.

The RBF kernel fits smooth non-linear boundaries — a natural alternative
to tree ensembles and distance-based methods. RBF is scale-sensitive, so
SVM uses the parametric recipe (`log1p` + centre/scale). `C = 1`,
`σ = 0.1` are sensible defaults for standardised inputs. We didn't tune
further because the main question is the parametric vs. non-parametric
contrast, not within-family optimisation.


```{r fit-svm}
res_svm <- run_or_load("svm", function() {
  out <- list(model = "SVM-RBF", family = "non-parametric", by_tier = list())
  for (tier in names(tiers)) {
    cat(sprintf("  SVM-RBF        [%-8s] fitting (%d features)...\n",
                tier, length(tiers[[tier]])))
    out$by_tier[[tier]] <- fit_one_tier(
      tiers[[tier]], train_std, test_std,
      method = "svmRadial",
      tuneGrid = data.frame(C = 1, sigma = 0.1))
  }
  out
})
print_tier_summary(res_svm)
```

SVM-RBF is the only model that is
**proxy-deployable at threshold 0.5 without recalibration on every
family**. On **Lexical** it posts AUC
0.995 with Sens 0.985 / Spec 0.983 — a ~1.7 % miss rate and ~1.7 %
false-block rate, both directions symmetric. On Trust, Behavior and
FullLite the same symmetric profile holds.
Translated back to the EDA §1.1 scenario, this is what "a
non-parametric URL-only filter worth deploying" actually looks like in
numbers — roughly 9 × fewer errors than LR-Ridge on Lexical (2 vs. 18
per 100 URLs) with no threshold tuning required.

KNN is the simplest non-parametric classifier: it stores the whole
training set and classifies each test point by a majority vote among
its `k` closest neighbours, using Euclidean distance. Being
distance-based, it uses the parametric recipe. `k = 25` is a mild
smoother that keeps the decision surface from being jagged.

One twist: when a tier has **only binary features** (Trust has 7 binary
indicators), many training points sit at identical coordinates, and
caret's `knn3` refuses to run with "too many ties in knn". We add a tiny
Gaussian jitter (SD = 10⁻³, less than 0.1 % of the feature scale) to
break the distance ties without changing any decisions.

```{r fit-knn}
jitter_features <- function(d, cols, sd = 1e-3) {
  for (c in cols) d[[c]] <- d[[c]] + rnorm(nrow(d), 0, sd)
  d
}

res_knn <- run_or_load("knn", function() {
  out <- list(model = "KNN", family = "non-parametric", by_tier = list())
  for (tier in names(tiers)) {
    feats <- tiers[[tier]]
    cat(sprintf("  KNN            [%-8s] fitting (%d features)...\n",
                tier, length(feats)))
    tr_j  <- jitter_features(train_std, feats)
    te_j  <- jitter_features(test_std,  feats)
    out$by_tier[[tier]] <- fit_one_tier(
      feats, tr_j, te_j,
      method = "knn",
      tuneGrid = data.frame(k = 25))
  }
  out
})
print_tier_summary(res_knn)
```

KNN is the close non-parametric runner-up
to SVM-RBF. On **Lexical** AUC 0.987 with Sens 0.944 / Spec 0.923 —
symmetric, deployable, ~6–8 % error either way. On **Trust** the AUC
holds at 0.932 but Accuracy drops to 0.862: with only 7 mostly-binary
features, many training points collapse to identical coordinates, the
jittered distance barely separates classes, and the 25-NN vote
averages over a near-uniform neighbourhood. Behavior and FullLite
return to the symmetric ≥ 0.96 profile. One caveat AUC cannot show but
the proxy deployment needs to know: KNN's **inference cost scales with
the training set** — every click triggers 24 k distance computations,
which is the opposite of the "cheap before the page loads" constraint
that motivated the URL-only filter. SVM-RBF wins on quality and on
inference cost.

## 6.1 Per-(model, tier) summary

```{r build-long}
all_results <- list(res_lr, res_lda, res_nb, res_rf, res_svm, res_knn)

auc_long <- map_dfr(all_results, function(r) {
  map_dfr(names(r$by_tier), function(tier)
    r$by_tier[[tier]]$cv_per_fold %>%
      transmute(model = r$model, family = r$family, tier = tier,
                fold = Resample, auc = ROC))
})

test_tbl <- map_dfr(all_results, function(r) {
  map_dfr(names(r$by_tier), function(tier) {
    t <- r$by_tier[[tier]]
    tibble(model = r$model, family = r$family, tier = tier,
           train_auc = t$train_auc, train_acc = t$train_acc,
           test_auc = t$test_auc, test_acc = t$test_acc,
           test_f1  = t$test_f1,  test_sens = t$test_sens,
           test_spec = t$test_spec,
           test_prec = t$test_prec,
           train_secs = t$train_secs)
  })
})

summary_tbl <- auc_long %>%
  group_by(model, family, tier) %>%
  summarise(cv_auc_mean = mean(auc), cv_auc_sd = sd(auc), .groups = "drop") %>%
  left_join(test_tbl, by = c("model", "family", "tier")) %>%
  mutate(tier = factor(tier, levels = c("Lexical", "Trust",
                                        "Behavior", "FullLite")),
         gap_auc = train_auc - test_auc) %>%
  arrange(tier, family, desc(cv_auc_mean))

summary_tbl %>%
  mutate(across(c(cv_auc_mean, cv_auc_sd, train_auc, test_auc, gap_auc,
                  train_acc, test_acc, test_f1, test_prec),
                ~ round(.x, 4)),
         train_secs = round(train_secs, 1)) %>%
  kable()
```


### Train vs. test — overfitting diagnostic

The `Train AUC` and `Gap = Train − Test` columns in each per-model table
answer *"how much does the model memorise the training set?"*. For the
three parametric models we expect `Train AUC ≈ Test AUC`: they have few
effective parameters and Ridge / LDA / NB cannot memorise 24 k rows.
For the three non-parametric models we expect a larger gap: Random
Forest predicting on its own training set is near-perfect (each point
was in the bag of most trees), KNN with `k = 25` averages its own
neighbours, and SVM-RBF can fit complex boundaries. A small gap there
is therefore not a problem — the key comparison is
**`CV ROC` vs. `Test AUC`**, both of which are *out-of-sample*. If
those two agree, the model generalises regardless of its in-bag fit.
The train column is only there to show that our CV numbers are not
silently inflated by leakage.

### Classification quality beyond AUC

AUC ranks models by their score ordering — useful for comparing
families in the abstract, but on the richer tiers every model sits at
AUC ≈ 0.99 and the ranking becomes hard to read. More importantly, AUC
is **threshold-free**: it says nothing about what happens when the
corporate proxy from §1.1 of the EDA has to commit to a single
decision — *block* or *allow* — the moment the user clicks the link.
At deployment the proxy doesn't get to rank URLs; it gets to pick one
operating point and live with both error types at once.

That is why a second evaluation is needed. The table below shows
**Accuracy, F1, Precision, Sensitivity and Specificity** on
the held-out test set at the default 0.5 threshold — the numbers the
proxy would actually produce in production.

In this proxy setting each split matters on its own, and the two error
types are **asymmetrically expensive**:
**Sensitivity** = share of phishing URLs we caught (a miss lands the
user on a credential-harvesting page),
**Specificity** = share of legitimate URLs we didn't flag — every
false block is a user staring at a broken internal link or a customer
portal they can't reach, and that is exactly the cost that gets the
filter disabled by the management.
**Precision** = of the URLs we blocked, how many were actually phishing
— low precision means the ratio of "good page blocked" to "real phish
caught" is bad enough that users stop trusting
the proxy and start asking for bypass rules, at which point the whole
URL-only line of defence is gone.

Precision is read directly from the held-out confusion matrix
(`caret::confusionMatrix(...)$byClass["Precision"]`, i.e. $TP/(TP+FP)$).

```{r quality-table}
quality_tbl <- test_tbl %>%
  mutate(tier = factor(tier, levels = c("Lexical", "Trust",
                                        "Behavior", "FullLite"))) %>%
  arrange(tier, family, desc(test_f1)) %>%
  transmute(Model = model, Family = family, Tier = tier,
            Accuracy    = round(test_acc,  4),
            F1          = round(test_f1,   4),
            Precision   = round(test_prec, 4),
            Sensitivity = round(test_sens, 4),
            Specificity = round(test_spec, 4))

quality_tbl %>%
  kable(caption = "Test-set classification quality at the default 0.5 threshold.")
```

## 6.2 Per-tier paired Wilcoxon test

For each tier we take the per-fold mean AUC of the three non-parametric
models and the three parametric models, then compare them with a
one-sided paired Wilcoxon test (non-parametric > parametric). Because we
share folds across models, the test is fully paired — each entry in
either series comes from the same ten validation sets.

```{r h1-test}
per_tier_test <- auc_long %>%
  group_by(tier, fold, family) %>%
  summarise(auc = mean(auc), .groups = "drop") %>%
  pivot_wider(names_from = family, values_from = auc) %>%
  group_by(tier) %>%
  summarise(
    mean_np  = mean(`non-parametric`),
    mean_p   = mean(parametric),
    diff     = mean(`non-parametric` - parametric),
    wilcox_V = wilcox.test(`non-parametric`, parametric,
                           paired = TRUE, alternative = "greater")$statistic,
    wilcox_p = wilcox.test(`non-parametric`, parametric,
                           paired = TRUE, alternative = "greater")$p.value,
    .groups = "drop"
  ) %>%
  mutate(tier = factor(tier, levels = c("Lexical", "Trust",
                                        "Behavior", "FullLite"))) %>%
  arrange(tier)

per_tier_test %>%
  mutate(across(c(mean_np, mean_p, diff), ~ round(.x, 4)),
         wilcox_p = signif(wilcox_p, 3)) %>%
  kable()
```

A caveat about the p-value column: with **10-paired observations**
, the smallest p-value a one-sided paired Wilcoxon
can produce is \(1/2^{10} = 1/1024 \approx 0.00098\). Three of the four
tiers (Lexical, Behavior, FullLite) hit exactly this floor — every
single non-parametric fold beat its matched parametric fold. The
fourth tier, **Trust**, lands at p ≈ 0.019, which is well above the
floor and correctly signals that the family contrast is genuinely
weaker there. Compared to the earlier 5-fold version (floor
\(1/32 \approx 0.031\)), the 10-fold setup is the first one that lets
the p-value separate Trust from the rest rather than collapsing all
four tiers onto the same floor. Still: **read the `diff` column for
effect size** — a small diff (Behavior 0.004, FullLite 0.0003) at
p ≈ 0.001 is statistically consistent but practically irrelevant.
---

# 7. Surrogate tree — visualising the RF decision rule (Task 4)

Random Forest is accurate but opaque: Lexical predictions are the vote of
300 unpruned trees over bootstrapped samples, and there is no single
diagram a reader can point at. Task 4 asks us to render a model-faithful
visualisation next to the real model and quantify how much of the model
behaviour it captures. We do this with a **surrogate tree**: a single
`rpart` tree whose training target is **the RF's own predictions**, not
the ground-truth label. The tree is therefore a student learning what
the RF does, and its fidelity to the RF — not its accuracy against the
truth — is what we tune.

We run the exercise on two tiers: **Lexical** (where RF beats the
parametric models by ≥ 0.10 AUC and a tree-level explanation is most
valuable) and **FullLite** (where every model saturates near AUC ≈ 1.0
and fidelity should be trivially high — a sanity check that the tuning
procedure is unbiased). Trust and Behavior are omitted on purpose:
Trust is 7 features, most of them binary flags, so the only non-linear
structure it can carry is XOR-like binary interactions — which the
§6.2 numbers rule out (every model on Trust lands at AUC ≈ 0.93 in a
tight band, so there is no interaction pattern left for a tree to
visualise). Behavior sits between the two interesting cases and would
not add a new observation.

## 7.1 Teacher RF and surrogate utility

`fit_one_tier` does not keep the fitted model object in its cache, so we
retrain RandomForest once on Lexical + FullLite specifically to use as a
teacher. It's cached in its own `.rds` so the main §5.1 artifacts stay
untouched.

```{r task4-setup, message=FALSE, warning=FALSE}
library(rpart)
has_rpart_plot <- requireNamespace("rpart.plot", quietly = TRUE)
if (has_rpart_plot) suppressPackageStartupMessages(library(rpart.plot))

refit_rf_for_surrogate <- function(selected_tiers = c("Lexical", "FullLite"),
                                   ntree = 300) {
  path <- file.path(ARTIFACTS, "rf_for_surrogate.rds")
  fp <- digest::digest(list(
    selected_tiers = selected_tiers,
    features       = lapply(tiers[selected_tiers], sort),
    ntree          = ntree,
    n_train        = nrow(train_raw)
  ))
  if (file.exists(path)) {
    cached <- readRDS(path)
    if (isTRUE(cached$fp == fp)) return(cached$by_tier)
  }
  cat("  [surrogate] refitting RF teacher on",
      paste(selected_tiers, collapse = ", "), "...\n")
  by_tier <- list()
  for (tier in selected_tiers) {
    feats <- tiers[[tier]]
    mtry  <- max(1, floor(sqrt(length(feats))))
    t0 <- Sys.time()
    fit <- randomForest(
      x     = train_raw[, feats, drop = FALSE],
      y     = train_raw$label,
      ntree = ntree,
      mtry  = mtry
    )
    cat(sprintf("    %-10s %d features, mtry=%d, %.1fs\n",
                tier, length(feats), mtry,
                as.numeric(difftime(Sys.time(), t0, units = "secs"))))
    by_tier[[tier]] <- fit
  }
  saveRDS(list(by_tier = by_tier, fp = fp), path)
  by_tier
}

rf_teachers <- refit_rf_for_surrogate()
```

## 7.2 Surrogate-tree grid search on fidelity

The grid is small by design: we tune the three knobs that actually
change how faithfully a `rpart` tree can reproduce a decision surface.
`maxdepth` bounds the tree's expressiveness (the main
interpretability-vs-fidelity trade-off); `cp` sets the pruning
threshold that decides which splits are retained; `minbucket` forces
leaves to be large enough to resist fitting noise in the teacher's
predictions. With 5 × 3 × 3 = 45 combinations per tier and `rpart` at
~1 s per fit on 24 k rows, the whole sweep takes under three minutes
and is cached per tier.

**Fidelity** is the overall metric we optimise on the held-out test set:

  $$\text{fidelity} = \frac{1}{n_\text{test}} \sum_i \mathbf{1}\bigl[\hat{y}^{\text{tree}}_i = \hat{y}^{\text{RF}}_i\bigr].$$

The rest of this report (§6.1.1, §8) judges classifiers on
**Sensitivity / Specificity**, not on overall accuracy, because a
phishing-blocker's two error types have asymmetric cost. For the same
reason we decompose fidelity the same way: `Sens vs RF` is the share
of RF-called phishing that the tree also calls phishing, `Spec vs RF`
is the share of RF-called legit the tree also calls legit. These are
the two numbers that tell us *how the visualisation's agreement with
the model breaks down by class* — the per-class analogues of
fidelity. We also report tree and RF Sens/Spec against the ground
truth so the deployment-level operating point is visible. Accuracy is
**not** reported here — it would hide the exact class-asymmetry the
rest of the project is built to expose.

```{r task4-grid}
tune_surrogate_tree <- function(tier, rf_fit,
                                maxdepths  = 3:7,
                                cps        = c(1e-4, 1e-3, 1e-2),
                                minbuckets = c(10, 30, 100)) {
  feats <- tiers[[tier]]
  tr_x  <- train_raw[, feats, drop = FALSE]
  te_x  <- test_raw[ , feats, drop = FALSE]

  teacher_train <- predict(rf_fit, tr_x)
  teacher_test  <- predict(rf_fit, te_x)
  truth_test    <- test_raw$label

  # RF reference Sens/Spec on the test set - constant across the grid.
  rf_sens <- mean(teacher_test[truth_test == "Phishing"]   == "Phishing")
  rf_spec <- mean(teacher_test[truth_test == "Legitimate"] == "Legitimate")

  tr_data <- tr_x
  tr_data$teacher_label <- teacher_train

  rows  <- list()
  trees <- list()
  for (md in maxdepths) for (cpv in cps) for (mb in minbuckets) {
    key <- sprintf("md=%d_cp=%.0e_mb=%d", md, cpv, mb)
    tree <- rpart(
      teacher_label ~ ., data = tr_data,
      method  = "class",
      model   = TRUE,
      control = rpart.control(maxdepth = md, cp = cpv,
                              minbucket = mb, xval = 0)
    )
    pr_class <- predict(tree, te_x, type = "class")
    pr_prob  <- predict(tree, te_x, type = "prob")[, "Phishing"]
    n_leaves <- sum(tree$frame$var == "<leaf>")

    # Overall agreement - Task 4's fidelity metric.
    fidelity <- mean(pr_class == teacher_test)

    # Per-class agreement: the Sens/Spec decomposition of fidelity,
    # treating the RF prediction as the tree's training target.
    sens_vs_rf <- mean(pr_class[teacher_test == "Phishing"]   == "Phishing")
    spec_vs_rf <- mean(pr_class[teacher_test == "Legitimate"] == "Legitimate")

    # Deployment-level operating point: tree vs. actual truth.
    tree_sens <- mean(pr_class[truth_test == "Phishing"]   == "Phishing")
    tree_spec <- mean(pr_class[truth_test == "Legitimate"] == "Legitimate")

    auc <- as.numeric(pROC::roc(truth_test, pr_prob,
                                levels = c("Legitimate", "Phishing"),
                                direction = "<", quiet = TRUE)$auc)
    rows[[key]] <- tibble(
      tier = tier, maxdepth = md, cp = cpv, minbucket = mb,
      leaves = n_leaves,
      fidelity   = fidelity,
      sens_vs_rf = sens_vs_rf, spec_vs_rf = spec_vs_rf,
      tree_sens  = tree_sens,  tree_spec  = tree_spec,
      rf_sens    = rf_sens,    rf_spec    = rf_spec,
      auc        = auc
    )
    trees[[key]] <- tree
  }
  list(results = bind_rows(rows), trees = trees,
       teacher_test = teacher_test, truth_test = truth_test,
       rf_sens = rf_sens, rf_spec = rf_spec)
}

load_surrogate <- function(tier, rf_fit) {
  path <- file.path(ARTIFACTS,
                    paste0("surrogate_", tolower(tier), ".rds"))
  fp <- digest::digest(list(
    tier       = tier,
    features   = sort(tiers[[tier]]),
    n_train    = nrow(train_raw),
    rf_ntree   = rf_fit$ntree,
    tree_store = "model=TRUE",
    schema     = "v2_sens_spec"
  ))
  if (file.exists(path)) {
    cached <- readRDS(path)
    if (isTRUE(cached$fp == fp)) return(cached$data)
  }
  cat(sprintf("  [surrogate-%s] tuning rpart grid...\n", tier))
  t0 <- Sys.time()
  data <- tune_surrogate_tree(tier, rf_fit)
  cat(sprintf("    done in %.1fs (%d combinations)\n",
              as.numeric(difftime(Sys.time(), t0, units = "secs")),
              nrow(data$results)))
  saveRDS(list(data = data, fp = fp), path)
  data
}

surr_lex  <- load_surrogate("Lexical",  rf_teachers[["Lexical"]])
surr_full <- load_surrogate("FullLite", rf_teachers[["FullLite"]])
all_surr  <- bind_rows(surr_lex$results, surr_full$results)
```

## 7.3 Fidelity vs. depth

The fidelity saturation curve — best fidelity per (tier, maxdepth)
across all (cp, minbucket) — tells us how much tree expressiveness is
needed before further depth stops buying alignment with the RF. RF
reference AUCs come from §5.1.

```{r task4-depth-curve}
rf_ref <- tibble(
  tier    = c("Lexical", "FullLite"),
  rf_auc  = c(
    res_rf$by_tier[["Lexical"]]$test_auc,
    res_rf$by_tier[["FullLite"]]$test_auc
  ),
  rf_sens = c(
    res_rf$by_tier[["Lexical"]]$test_sens,
    res_rf$by_tier[["FullLite"]]$test_sens
  ),
  rf_spec = c(
    res_rf$by_tier[["Lexical"]]$test_spec,
    res_rf$by_tier[["FullLite"]]$test_spec
  )
)

depth_tbl <- all_surr %>%
  group_by(tier, maxdepth) %>%
  slice_max(fidelity, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  left_join(rf_ref, by = "tier") %>%
  mutate(tier = factor(tier, levels = c("Lexical", "FullLite"))) %>%
  arrange(tier, maxdepth) %>%
  transmute(Tier = tier, MaxDepth = maxdepth, Leaves = leaves,
            Fidelity     = round(fidelity,   4),
            `Sens vs RF` = round(sens_vs_rf, 4),
            `Spec vs RF` = round(spec_vs_rf, 4),
            `Tree AUC`   = round(auc,        4),
            `RF AUC`     = round(rf_auc,     4))

kable(depth_tbl,
      caption = "Best surrogate per (tier, maxdepth) on the S6 test set. Fidelity = overall tree/RF agreement; Sens vs RF / Spec vs RF = its per-class decomposition (treating RF prediction as target). Winning (cp, minbucket) picked by fidelity; ties broken by fewer leaves.")
```

The §7.3 table reports the **fidelity ceiling** — the best rpart can do
at each depth, regardless of readability. The `Sens vs RF` / `Spec vs RF`
columns show *how* the remaining disagreement splits across the two
classes: a fidelity of 0.96 with Sens-vs-RF 0.97 and Spec-vs-RF 0.95
means the tree disagrees with the RF mostly on the legit side, which
is a very different story from 0.95 / 0.97 — and the same story we
read off §6.1.1 for the supervised models. Picking the optimisation
target consistently with the rest of the report (Sens/Spec, not
overall accuracy) keeps Task 4's interpretation aligned with how we
judge the RF itself.

For the plotted diagrams in §7.4 we pick a different tree: the
surrogate with the highest fidelity under a readability cap of ≤ 15
leaves. A 37-leaf tree technically has a higher fidelity number, but
it is visually unparseable on a slide, so it cannot serve the
"visualisation" role the task asks for. The cap is cheap: it costs
~0.009 fidelity on Lexical (0.967 → 0.958 with 14 leaves) and ~0.003
on FullLite (0.990 → 0.987 with 15 leaves). In other words, the last
two depth levels buy almost no fidelity — they only buy leaves.

```{r task4-winners}
MAX_PLOT_LEAVES <- 15L

winners <- all_surr %>%
  filter(leaves <= MAX_PLOT_LEAVES) %>%
  group_by(tier) %>%
  slice_max(fidelity, n = 1, with_ties = TRUE) %>%
  slice_min(leaves,   n = 1, with_ties = FALSE) %>%
  ungroup()

winners %>%
  mutate(across(c(fidelity, sens_vs_rf, spec_vs_rf,
                  tree_sens, tree_spec, rf_sens, rf_spec, auc),
                ~ round(.x, 4))) %>%
  select(tier, maxdepth, cp, minbucket, leaves,
         fidelity, sens_vs_rf, spec_vs_rf,
         tree_sens, tree_spec, rf_sens, rf_spec, auc) %>%
  kable(caption = sprintf("Plotted surrogate tree per tier: highest fidelity subject to leaves <= %d. Ties broken by fewer leaves. `tree_sens`/`tree_spec` are vs. ground truth; `rf_sens`/`rf_spec` are the RF reference - the closer they are, the less the surrogate changes the operating point.", MAX_PLOT_LEAVES))
```

## 7.4 Winning trees

```{r task4-tree-lexical, fig.height=6, fig.width=10}
key_lex <- winners %>% filter(tier == "Lexical") %>%
  transmute(key = sprintf("md=%d_cp=%.0e_mb=%d", maxdepth, cp, minbucket)) %>%
  pull(key)
tree_lex <- surr_lex$trees[[key_lex]]

if (has_rpart_plot) {
  rpart.plot::rpart.plot(
    tree_lex, type = 2, extra = 104, fallen.leaves = TRUE,
    box.palette = c("#D73027", "#1A9850"),
    main = sprintf("Lexical surrogate (fidelity %.3f vs RF)",
                   winners$fidelity[winners$tier == "Lexical"])
  )
} else {
  plot(tree_lex, uniform = TRUE, margin = 0.12,
       main = "Lexical surrogate tree")
  text(tree_lex, use.n = TRUE, cex = 0.65)
}
```

```{r task4-tree-fulllite, fig.height=6, fig.width=10}
key_full <- winners %>% filter(tier == "FullLite") %>%
  transmute(key = sprintf("md=%d_cp=%.0e_mb=%d", maxdepth, cp, minbucket)) %>%
  pull(key)
tree_full <- surr_full$trees[[key_full]]

if (has_rpart_plot) {
  rpart.plot::rpart.plot(
    tree_full, type = 2, extra = 104, fallen.leaves = TRUE,
    box.palette = c("#D73027", "#1A9850"),
    main = sprintf("FullLite surrogate (fidelity %.3f vs RF)",
                   winners$fidelity[winners$tier == "FullLite"])
  )
} else {
  plot(tree_full, uniform = TRUE, margin = 0.12,
       main = "FullLite surrogate tree")
  text(tree_full, use.n = TRUE, cex = 0.65)
}
```

### RF variable importance cross-check

The surrogate's root split is a claim about what the RF thinks is the
most discriminative single feature. To see whether that claim is
faithful to the teacher — rather than an artefact of the one tree we
plotted — we read the RF's own variable importance (mean decrease in
Gini) for the same teacher and compare the ordering against the tree's
top few splits.

```{r task4-varimp, fig.height=5, fig.width=11}
# base-R title() goes through utf8towcs() which chokes on em-dashes
# in non-UTF-8 Windows locales - we keep the surrounding prose
# em-dashes intact but use a plain hyphen in the plot titles here.
op <- par(mfrow = c(1, 2), mar = c(4, 7, 3, 1))
randomForest::varImpPlot(
  rf_teachers[["Lexical"]],
  n.var = min(10, length(tiers[["Lexical"]])),
  main  = "RF variable importance - Lexical"
)
randomForest::varImpPlot(
  rf_teachers[["FullLite"]],
  n.var = 10,
  main  = "RF variable importance - FullLite (top 10)"
)
par(op)
```

If the feature at the tree's root is also in the top few rows of the
RF importance plot for the same tier, the surrogate isn't improvising
— it's picking up the same signal ranking the RF itself assigns. That
is the evidence that lets us read the tree diagram as a visualisation
of RF behaviour rather than a parallel classifier that happens to agree
on the test set.

## 7.5 What the surrogate does and doesn't capture

**Similarities with RF.** Both models use the same vocabulary: every
decision is an axis-aligned split on a single feature, so the branches
of the surrogate are the same kind of rule the RF's constituent trees
emit. What the plotted surrogates actually reveal about RF's feature
ordering is worth reading off the pictures directly:

- **Lexical.** The root split is `NoOfOtherSpecialCharsInURL < 3`. The
  left subtree (high special-char count, ~70 % of rows, mostly phish)
  cascades through `NoOfDegitsInURL`, `NoOfSubDomain`, `TLDLength` and
  `CharContinuationRate` — the classic lexical-anomaly ladder. The
  right subtree (normal-looking URLs, ~30 %, mostly legit) splits on
  `URLLength` first, with `NoOfSubDomain` and `NoOfLettersInURL` as
  secondary refinements. Exactly the kind of nested rule we expected
  from URL-only text features.
- **FullLite.** The root split is `HasSocialNet = 1` and the top three
  layers are **all trust-tier binaries**: `IsHTTPS`, `HasCopyrightInfo`,
  `HasDescription`, `HasSubmitButton`. URL-tier counts
  (`NoOfOtherSpecialCharsInURL`, `NoOfDegitsInURL`, `NoOfLettersInURL`)
  only show up deeper in the tree, refining leaves that the trust
  signals could not fully resolve. This trust-first / URL-later
  hierarchy is itself a finding: once the RF has access to page-level
  trust flags, those flags dominate the top of its decision logic, and
  the URL lexical features — which were the whole story on the Lexical
  tier — get demoted to a refinement role.

Both of those orderings are what a faithful visualisation should
expose. The surrogate is therefore a **genuine visualisation of RF's
decision logic**, not an unrelated model that happens to match the
output on the held-out set.

**What the surrogate loses.** RF is an ensemble of 300 bagged trees,
each trained on a different bootstrap sample and seeing only
`mtry = √p` features per split. Averaging across those 300 trees
smooths the decision boundary into something **a single tree cannot
express** — specifically, interactions among 3+ features that only
emerge when the bagging vote combines partially overlapping splits.
That's the gap between `Tree AUC` and `RF AUC` in §7.3: even when
fidelity is 0.97+, the RF's probability estimates are better-calibrated
near the decision boundary than any single surrogate's listwise class
proportions can match.

**Parameter tuning — what we actually changed and why.** `maxdepth` is
the dominant lever; everything else plays a secondary role. Below a
depth of 3 the surrogate underfits badly on Lexical (fidelity drops
to ~0.85) because a tree that shallow cannot encode even the second
layer of URL-length × digit-count interactions. Past depth 5–6
fidelity plateaus: any new splits cut the data so finely that they
stop matching the RF's smoothed bagging output and start introducing
their own noise. `cp = 10⁻⁴` is permissive enough that `maxdepth` is
the actual binding constraint (not pruning); and `minbucket ≥ 10`
keeps leaves well-populated so random fluctuations in the teacher
labels do not survive into a split.

**Readability cap on the plotted winner.** The §7.3 saturation table
shows the full ceiling (37-leaf / 25-leaf trees at depth 7), but the
diagrams in §7.4 use the highest-fidelity tree with **≤ 15 leaves**.
That is effectively a second tuning knob — an explicit complexity
budget driven by "fits on a slide" rather than by cross-validation —
and it is honest to say so. The surrogate's whole purpose is to be
*read*, not just to win a fidelity contest; the realised cost is
~0.009 fidelity on Lexical and ~0.003 on FullLite, which is a fair
price for a diagram a reviewer can follow branch-by-branch.

**Per-class fidelity — the Sens/Spec decomposition.** Overall fidelity
averages over both classes, but the rest of this report (§6.1.1, §8)
treats the two error types separately because they have asymmetric
cost in the proxy setting. The surrogate inherits that framing: we
also report `Sens vs RF` (the share of RF-called phishing the tree
also calls phishing) and `Spec vs RF` (the share of RF-called legit
the tree also calls legit). On both tiers these two agreement numbers
land within a few tenths of a percent of each other — the surrogate
tracks RF symmetrically in both directions, not by picking up the
easy class. The deployment-level columns (`tree_sens` vs. `rf_sens`,
`tree_spec` vs. `rf_spec`) say the same thing differently: the
surrogate preserves the RF's operating point on the ground truth, so
a reader who actually uses the tree instead of the RF would block and
allow almost the same URLs. If Sens-vs-RF and Spec-vs-RF had diverged
substantially, it would mean the tree copied RF well on one class and
not on the other — a warning that the "visualisation" was hiding a
class-specific failure mode. They don't, so the visualisation is
balanced.

**FullLite sanity check.** On FullLite every decent model sits at AUC
≈ 0.999 (§6.1), so a tree of any non-trivial depth trivially matches
the RF's output. Fidelity on FullLite therefore carries very little
information about the RF — it mostly confirms that the grid search
and caching pipeline are unbiased. The **interesting** tier is
Lexical: that's where the RF actually wins a non-trivial contest and
the question "is this a one-tree decision rule in disguise, or does
the ensemble really buy us something?" has a non-obvious answer.

---

# 8. Conclusion — H1 gradient

1. **Non-parametric > parametric when the feature tier is weak.** On
   the **Lexical** tier the non-parametric advantage is the largest
   of any tier, and every single CV fold shows the same order. This is
   exactly the regime H1 targets: URL counts interact in non-linear
   ways (digit runs × length × subdomain depth), and linear or Gaussian
   decision boundaries just can't capture those interactions. Read the
   exact diff off §6.2.
2. **Trust also favours non-parametric, but only barely.** With only
   seven features — most of them plain binary flags — there's little
   non-linear structure left for non-parametric models to discover.
3. **Behavior is the honest middle tier.** With the six near-leaky
   page counts (`LineOfCode`, `NoOfExternalRef`, `NoOfImage`,
   `NoOfSelfRef`, `NoOfJS`, `NoOfCSS`) stripped out — each of which
   individually gives univariate AUC ≥ 0.95 and single-handedly solves
   the task — the remaining 14 Behavior features are a realistic
   page-level signal without saturation. This is where we can read the
   family contrast cleanly at medium signal strength.
4. **FullLite is the honest > 30-feature benchmark.** It meets the
   assignment's "several dozen parameters" rule without relying on
   features that single-handedly solve the task. Even so, 34
   non-leaky features are already enough to make every model cluster
   near the ceiling — the family contrast on FullLite tells us whether
   non-parametric flexibility still buys anything in a realistic
   high-dimensional setting, or whether the task is simply resolved by
   that point.

**Train–test gap check.** The per-model tables in §4–§5 also report
`Train AUC` and the `Gap = Train − Test`. Parametric models (LR-Ridge,
LDA, NB) sit within ~0.01 of their test AUC on every tier — they cannot
memorise 24 k rows with so few effective parameters, so their CV scores
are trustworthy. Non-parametric models show a larger in-bag fit
(Random Forest and KNN approach Train AUC ≈ 1.0 by construction), but
their `CV ROC` and `Test AUC` still agree tightly — the out-of-sample
numbers reported above are not inflated by leakage. In other words,
the H1 gradient is measured on generalisation performance, not on any
model's ability to memorise the training set.

**Threshold-0.5 quality (Precision / Sensitivity / Specificity).** The
§6.1.1 table complements the AUC ranking with operating-point numbers.
Two findings only visible once Precision is reported:

- **SVM-RBF on Lexical** is not only the AUC winner but also the only
  model that is simultaneously high on all three (Prec ≈ Sens ≈ Spec ≈
  0.98). Every other model on Lexical trades one for another.
- **Naive Bayes mirrors itself across tiers.** On Lexical it over-calls
  phish (Sens 0.999, Prec 0.618) — a high-recall / low-precision flood.
  On Behavior the same independence violation flips the other way (Sens
  0.732, Prec 0.995) — almost never calls phish, but when it does, it
  is right. Both are symptoms of the same broken independence
  assumption; the direction of the miscalibration is tier-dependent.

This is the part AUC hides: two models with AUC ≈ 0.99 on Behavior can
still have fundamentally different operating-point behaviour, and the
Precision/Sensitivity decomposition is what makes that visible. Brought
back to the corporate-proxy framing of EDA §1.1: the proxy doesn't
consume an AUC, it consumes a single *block/allow* verdict per click.
Two models that look tied on the ranking metric can produce very
different proxies — one that floods the helpdesk with false blocks,
and one that silently lets phish through.
AUC treats these as equivalent; the deployment doesn't.

**Practical takeaway.** A URL-only first-line phishing detector gains a
lot from a non-parametric model (SVM-RBF or KNN gain ~0.10 AUC over
LR / NB on Lexical, and SVM additionally keeps Precision and Recall
balanced at ~0.98). Once the system can afford to fetch the page and
compute Behavior-level features, the AUC choice of model becomes almost
irrelevant and simpler, more explainable models (LR, LDA) are good
enough — though the operating-point choice (Precision-first vs.
Recall-first) should still be driven by the cost of a false block vs.
a missed phish.

## 8.1 Selected model

**The model we pick for the corporate-proxy deployment is SVM-RBF.**

1. **On Lexical — the only tier the proxy
   actually sees before the page loads** — SVM-RBF posts AUC 0.995 with
   Sens 0.985 / Spec 0.983. Every other model either drops ≥ 0.10 AUC
   or trades one error type for another at threshold 0.5.
2. **It is symmetric in both error directions without recalibration.**
   At the default 0.5 cut the miss rate (~1.7 %) and the false-block
   rate (~1.7 %) are essentially equal — no threshold tuning needed to
   make the operating point deployable. LR, LDA, NB and RF would all
   need a separate calibration step on Lexical before they could go
   behind a proxy; SVM-RBF does not.
3. **It holds the quality on every other tier too.** Trust, Behavior
   and FullLite all land in the Sens ≈ Spec ≈ 0.97–1.00 region, so the
   same model works if the proxy ever expands beyond URL-only — no
   model swap needed when more signal becomes available.
4. **Inference cost is acceptable for the proxy loop.** KNN matches
   SVM-RBF on Lexical AUC (0.987 vs. 0.995) but requires ~24 k
   distance computations per click, which is exactly the
   "cheap before the page loads" constraint from our hypothesis.

**When we would revisit the choice.** If the proxy pipeline later gains
Behavior- or FullLite-grade signals and explainability becomes a
requirement, LR-Ridge or LDA on that richer tier
is a defensible downgrade — both reach AUC ≥ 0.99 there with balanced
Sens/Spec, and their coefficients are interpretable. But for the
URL-only first-line filter that out hypothesis is built around, SVM-RBF is
the only model whose numbers support the deployment directly.
