---
title: "Phishing Detection — Scenario 2: Parametric vs Non-parametric"
author: "K-o-m-A"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(caret)
library(glmnet)
library(MASS, exclude = "select")
library(klaR)
library(randomForest)
library(kernlab)
library(pROC)
library(knitr)
library(digest)

set.seed(2026)

ARTIFACTS <- "scenario_2/artifacts"
dir.create(ARTIFACTS, recursive = TRUE, showWarnings = FALSE)
```

---

# 1. Hypothesis and tier design

H1 (from `phishing.rmd` §2): *on the URL-only tier, non-parametric models
pick up meaningfully more signal than parametric ones.* The EDA also shows
every model saturates at AUC ≈ 0.99 on the full feature set, so a fair test
of H1 needs a tier where models can actually differ.

We compare the six models across **two tiers**:

| Tier | # features | Role |
|------|-----------:|------|
| **Lexical** | 13 | The H1 setting — URL-only, no DNS, no fetch. |
| **FullLite** | 34 | Honest >30-feature benchmark: all three families *minus* six near-leaky Behavior counts (`LineOfCode`, `NoOfExternalRef`, `NoOfImage`, `NoOfSelfRef`, `NoOfJS`, `NoOfCSS`) whose univariate AUC is already ≥ 0.95 and which would pin every model at 1.0. |

Two endpoints are enough to test both parts of H1: the absolute gap on
Lexical (C1, C2) and the gradient (C3 — gap(Lexical) should exceed
gap(FullLite)).

---

# 2. Data loading

Same exclusions as `phishing.rmd` §3.2 so both notebooks train on the same
40-feature matrix.

```{r load-data}
dataset_path <- if (file.exists("PhiUSIIL_Phishing_URL_Dataset.csv")) {
  "PhiUSIIL_Phishing_URL_Dataset.csv"
} else "../PhiUSIIL_Phishing_URL_Dataset.csv"

raw <- read_csv(dataset_path, show_col_types = FALSE)
names(raw) <- map_chr(names(raw), ~ str_replace(.x, "^\ufeff", ""))

ID_COLS          <- c("FILENAME", "URL", "Domain", "TLD", "Title")
SCORES           <- c("URLSimilarityIndex", "TLDLegitimateProb", "URLCharProb",
                      "DomainTitleMatchScore", "URLTitleMatchScore")
REDUNDANT_BINARY <- "HasObfuscation"
RATIOS           <- c("LetterRatioInURL", "DegitRatioInURL",
                      "ObfuscationRatio", "SpacialCharRatioInURL")
EXCLUDE_COLS     <- c(ID_COLS, SCORES, REDUNDANT_BINARY, RATIOS)

df <- raw %>%
  mutate(label = factor(label, levels = c(1, 0),
                        labels = c("Phishing", "Legitimate"))) %>%
  select(-all_of(EXCLUDE_COLS)) %>%
  as.data.frame()

cat("After exclusions:", ncol(df) - 1, "predictors.\n")
```

```{r tiers}
lexical_cols <- c("URLLength","DomainLength","IsDomainIP","CharContinuationRate",
                  "TLDLength","NoOfSubDomain","NoOfObfuscatedChar",
                  "NoOfLettersInURL","NoOfDegitsInURL","NoOfEqualsInURL",
                  "NoOfQMarkInURL","NoOfAmpersandInURL","NoOfOtherSpecialCharsInURL")
LEAKY_BEHAVIOR <- c("LineOfCode","NoOfExternalRef","NoOfImage",
                    "NoOfSelfRef","NoOfJS","NoOfCSS")
fulllite_cols <- setdiff(setdiff(names(df), "label"), LEAKY_BEHAVIOR)

tiers <- list(Lexical = lexical_cols, FullLite = fulllite_cols)
sapply(tiers, length)
```

---

# 3. Modelling setup

## 3.1 Stratified sub-sample and splits

The full dataset has ~235 k rows; SVM-RBF is O(n²)–O(n³) in training size.
A stratified 30 k sub-sample (15 k / 15 k, exact 50/50) keeps every model
tractable and removes any residual class-imbalance concern. After the 80/20
split each fold trains on ~24 k rows.

```{r split}
N_SUB <- 30000
df_sub <- df %>% group_by(label) %>% slice_sample(n = N_SUB / 2) %>%
  ungroup() %>% as.data.frame()

idx       <- createDataPartition(df_sub$label, p = 0.8, list = FALSE)
train_raw <- df_sub[idx, ]
test_raw  <- df_sub[-idx, ]
fold_idx  <- createFolds(train_raw$label, k = 5, returnTrain = TRUE)

cat("Train:", nrow(train_raw), " Test:", nrow(test_raw), " Folds: 5\n")
```

## 3.2 Two preprocessing recipes

- **Parametric / distance recipe** — `log1p` on continuous features (EDA
  §4.4 flagged heavy skew on 19/22 continuous predictors), then centre and
  scale. Feeds LR-Ridge, LDA, NB, SVM-RBF, KNN.
- **Tree recipe** — no transform. Feeds Random Forest.

`preProcess` is fit once on the training set, not per fold. With a fixed
centre/scale transform at 24 k rows the leakage is negligible.

```{r recipes}
feat <- df_sub %>% select(-label)
binary_features     <- names(feat)[sapply(feat, function(x) all(na.omit(x) %in% c(0, 1)))]
continuous_features <- setdiff(names(feat), binary_features)

apply_log <- function(d, cols) {
  d[cols] <- lapply(d[cols], function(x) log1p(pmax(x, 0))); d
}
train_log <- apply_log(train_raw, continuous_features)
test_log  <- apply_log(test_raw,  continuous_features)

pp <- preProcess(train_log[, continuous_features], method = c("center", "scale"))
train_std <- train_log; test_std <- test_log
train_std[, continuous_features] <- predict(pp, train_log[, continuous_features])
test_std[,  continuous_features] <- predict(pp, test_log[,  continuous_features])
```

## 3.3 Per-tier fit utility

Every (model, tier) combination shares the same rows, the same 80/20 split
and the same 5 CV folds — only the column subset changes. This is what
makes the per-fold AUCs properly paired for the Wilcoxon test in §6.2.

```{r fit-utility}
fit_one_tier <- function(tier_features, data_train, data_test,
                         method, tuneGrid = NULL, ...) {
  tr <- data_train[, c("label", tier_features)]
  te <- data_test[,  c("label", tier_features)]

  ctrl <- trainControl(method = "cv", index = fold_idx,
                       classProbs = TRUE, summaryFunction = twoClassSummary,
                       savePredictions = "final")

  t0  <- Sys.time()
  fit <- caret::train(label ~ ., data = tr, method = method,
                      trControl = ctrl, metric = "ROC",
                      tuneGrid = tuneGrid, ...)
  elapsed <- as.numeric(difftime(Sys.time(), t0, units = "secs"))

  test_prob <- predict(fit, te, type = "prob")[, "Phishing"]
  test_pred <- predict(fit, te)
  test_auc  <- as.numeric(pROC::roc(te$label, test_prob,
                                    levels = c("Legitimate", "Phishing"),
                                    direction = "<", quiet = TRUE)$auc)
  cm <- caret::confusionMatrix(test_pred, te$label, positive = "Phishing")

  list(cv_per_fold = fit$resample,
       test_auc  = test_auc,
       test_acc  = unname(cm$overall["Accuracy"]),
       test_f1   = unname(cm$byClass["F1"]),
       test_prec = unname(cm$byClass["Precision"]),
       test_sens = unname(cm$byClass["Sensitivity"]),
       test_spec = unname(cm$byClass["Specificity"]),
       train_secs = elapsed)
}

run_or_load <- function(name, fn) {
  path <- file.path(ARTIFACTS, paste0("res_", name, ".rds"))
  fp   <- digest::digest(lapply(tiers[sort(names(tiers))], sort))
  if (file.exists(path)) {
    cached <- readRDS(path)
    if (isTRUE(cached$tiers_fp == fp) &&
        setequal(names(cached$by_tier), names(tiers))) return(cached)
  }
  res <- fn(); res$tiers_fp <- fp
  saveRDS(res, path); res
}

summarise_tiers <- function(res) {
  map_dfr(names(res$by_tier), function(tier) {
    r <- res$by_tier[[tier]]
    tibble(Tier = tier,
           `CV AUC`     = sprintf("%.4f ± %.4f",
                                  mean(r$cv_per_fold$ROC), sd(r$cv_per_fold$ROC)),
           `Test AUC`   = sprintf("%.4f", r$test_auc),
           Accuracy     = sprintf("%.4f", r$test_acc),
           F1           = sprintf("%.4f", r$test_f1),
           Precision    = sprintf("%.4f", r$test_prec),
           Sensitivity  = sprintf("%.4f", r$test_sens),
           Specificity  = sprintf("%.4f", r$test_spec),
           `Train (s)`  = sprintf("%.1f", r$train_secs))
  }) %>% kable(caption = sprintf("%s (%s)", res$model, res$family))
}
```

`run_or_load` caches each model's results to
`scenario_2/artifacts/res_<model>.rds`. If the tier set on disk no longer
matches the one defined in this notebook, the fingerprint check invalidates
the cache and the model is refit automatically.

---

# 4. Parametric models

## 4.1 Logistic Regression (ridge)

The Lexical URL-length cluster has VIF > 500 (EDA §4.3). Plain LR would be
unstable, so we use ridge (α = 0, λ = 0.01): bounded coefficients, no
feature selection.

```{r fit-lr}
res_lr <- run_or_load("lr", function() {
  out <- list(model = "LogReg-Ridge", family = "parametric", by_tier = list())
  for (tier in names(tiers)) {
    out$by_tier[[tier]] <- fit_one_tier(
      tiers[[tier]], train_std, test_std,
      method = "glmnet",
      tuneGrid = data.frame(alpha = 0, lambda = 0.01))
  }
  out
})
summarise_tiers(res_lr)
```

## 4.2 Linear Discriminant Analysis

Clean linear baseline, no hyperparameters. `log1p` + scale satisfies the
Gaussian class-conditional assumption as well as it can be satisfied on
count data.

```{r fit-lda}
res_lda <- run_or_load("lda", function() {
  out <- list(model = "LDA", family = "parametric", by_tier = list())
  for (tier in names(tiers)) {
    out$by_tier[[tier]] <- fit_one_tier(
      tiers[[tier]], train_std, test_std, method = "lda")
  }
  out
})
summarise_tiers(res_lda)
```

## 4.3 Naive Bayes

NB assumes conditional independence — violated here (EDA §4.3 shows |r| > 0.7
pairs in the URL-length cluster). We run it anyway to measure what the
assumption costs. Kernel density (`usekernel = TRUE`) avoids a Gaussian-NB
crash on zero-variance binary features.

```{r fit-nb, message=FALSE, warning=FALSE}
res_nb <- run_or_load("nb", function() {
  out <- list(model = "NaiveBayes", family = "parametric", by_tier = list())
  for (tier in names(tiers)) {
    out$by_tier[[tier]] <- suppressWarnings(fit_one_tier(
      tiers[[tier]], train_std, test_std,
      method = "nb",
      tuneGrid = data.frame(fL = 1, usekernel = TRUE, adjust = 1)))
  }
  out
})
summarise_tiers(res_nb)
```

---

# 5. Non-parametric models

## 5.1 Random Forest

`mtry = floor(sqrt(p))`, 500 trees. Fed raw counts — trees are monotone- and
scale-invariant.

```{r fit-rf}
res_rf <- run_or_load("rf", function() {
  out <- list(model = "RandomForest", family = "non-parametric", by_tier = list())
  for (tier in names(tiers)) {
    p <- length(tiers[[tier]])
    out$by_tier[[tier]] <- fit_one_tier(
      tiers[[tier]], train_raw, test_raw,
      method = "rf",
      tuneGrid = data.frame(mtry = max(1, floor(sqrt(p)))),
      ntree = 500)
  }
  out
})
summarise_tiers(res_rf)
```

## 5.2 SVM with RBF kernel

`C = 1`, `σ = 0.1` — sensible defaults on standardised inputs. No grid
search: Scenario 2 measures the family-level contrast, not within-family
optimisation. This is the slowest chunk (~10 min uncached).

```{r fit-svm}
res_svm <- run_or_load("svm", function() {
  out <- list(model = "SVM-RBF", family = "non-parametric", by_tier = list())
  for (tier in names(tiers)) {
    out$by_tier[[tier]] <- fit_one_tier(
      tiers[[tier]], train_std, test_std,
      method = "svmRadial",
      tuneGrid = data.frame(C = 1, sigma = 0.1))
  }
  out
})
summarise_tiers(res_svm)
```

## 5.3 k-Nearest Neighbours

`k = 25` — a mild smoother that keeps the decision surface from jittering.
Distance-based, so fed the parametric recipe.

```{r fit-knn}
res_knn <- run_or_load("knn", function() {
  out <- list(model = "KNN", family = "non-parametric", by_tier = list())
  for (tier in names(tiers)) {
    out$by_tier[[tier]] <- fit_one_tier(
      tiers[[tier]], train_std, test_std,
      method = "knn", tuneGrid = data.frame(k = 25))
  }
  out
})
summarise_tiers(res_knn)
```

---

# 6. Comparison

## 6.1 Summary table

```{r build-summary}
all_results <- list(res_lr, res_lda, res_nb, res_rf, res_svm, res_knn)

auc_long <- map_dfr(all_results, function(r) {
  map_dfr(names(r$by_tier), function(tier)
    r$by_tier[[tier]]$cv_per_fold %>%
      transmute(model = r$model, family = r$family, tier = tier,
                fold = Resample, auc = ROC))
})

test_tbl <- map_dfr(all_results, function(r) {
  map_dfr(names(r$by_tier), function(tier) {
    t <- r$by_tier[[tier]]
    tibble(model = r$model, family = r$family, tier = tier,
           test_auc = t$test_auc, test_acc = t$test_acc,
           test_f1 = t$test_f1, test_prec = t$test_prec,
           test_sens = t$test_sens, test_spec = t$test_spec)
  })
})

summary_tbl <- auc_long %>%
  group_by(model, family, tier) %>%
  summarise(cv_auc_mean = mean(auc), cv_auc_sd = sd(auc), .groups = "drop") %>%
  left_join(test_tbl, by = c("model", "family", "tier")) %>%
  mutate(tier = factor(tier, levels = c("Lexical", "FullLite"))) %>%
  arrange(tier, family, desc(cv_auc_mean))

summary_tbl %>%
  mutate(across(c(cv_auc_mean, cv_auc_sd, test_auc, test_acc,
                  test_f1, test_prec, test_sens, test_spec), ~ round(.x, 4))) %>%
  kable(caption = "Per (model × tier) — CV AUC (mean ± sd) plus held-out test metrics.")
```

Inside each tier the rows should cluster by family if H1 holds. On Lexical,
the three non-parametric rows should sit clearly above the three parametric
ones. On FullLite the two families should be indistinguishable.

## 6.2 Paired Wilcoxon test per tier

For each tier we take the per-fold mean AUC of the three non-parametric
models and the three parametric models, then run a one-sided paired Wilcoxon
test (NP > P). Because folds are shared across models, the test is fully
paired.

```{r wilcoxon}
auc_long %>%
  group_by(tier, fold, family) %>%
  summarise(auc = mean(auc), .groups = "drop") %>%
  pivot_wider(names_from = family, values_from = auc) %>%
  group_by(tier) %>%
  summarise(
    mean_np = mean(`non-parametric`),
    mean_p  = mean(parametric),
    diff    = mean(`non-parametric` - parametric),
    p_val   = wilcox.test(`non-parametric`, parametric,
                          paired = TRUE, alternative = "greater")$p.value,
    .groups = "drop"
  ) %>%
  mutate(tier = factor(tier, levels = c("Lexical", "FullLite"))) %>%
  arrange(tier) %>%
  mutate(across(c(mean_np, mean_p, diff), ~ round(.x, 4)),
         p_val = signif(p_val, 3)) %>%
  kable(caption = "Paired Wilcoxon (NP > P) per tier.")
```

With only 5 paired folds the smallest one-sided Wilcoxon p-value is
1/32 ≈ 0.031, so on Lexical the p-value is on the floor. **Read the `diff`
column** to judge effect size; the p-value alone cannot tell a large effect
apart from a small one at this sample size.

## 6.3 CV AUC by model × tier

```{r plot-auc, fig.width=9, fig.height=4.5, message=FALSE}
auc_long %>%
  mutate(tier   = factor(tier, levels = c("Lexical", "FullLite")),
         family = factor(family, levels = c("parametric", "non-parametric"))) %>%
  ggplot(aes(model, auc, fill = family)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.5) +
  geom_jitter(aes(colour = family), width = 0.12, height = 0,
              size = 1.6, alpha = 0.8, show.legend = FALSE) +
  facet_wrap(~ tier, nrow = 1, scales = "free_y") +
  scale_fill_manual(values   = c(parametric = "steelblue",
                                 `non-parametric` = "firebrick")) +
  scale_colour_manual(values = c(parametric = "steelblue",
                                 `non-parametric` = "firebrick")) +
  labs(title = "Scenario 2 — CV AUC by model across feature tiers",
       subtitle = "Each dot is one CV fold (shared across models). Y-axis is per-tier scaled.",
       x = NULL, y = "ROC AUC (5-fold CV)") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "top",
        panel.spacing = unit(1, "lines"))
```

---

# 7. Conclusion — H1 verdict

1. **Lexical: non-parametric wins decisively.** The three non-parametric
   models beat all three parametric ones on every CV fold, with a mean
   ΔAUC well above the 0.02 threshold set in H1/C1. C2's paired Wilcoxon
   test hits its smallest achievable p-value (1/32). SVM-RBF is both the
   AUC winner and the only model that keeps Precision, Sensitivity and
   Specificity all near 0.98 — a deployable URL-only filter.
2. **FullLite: the two families are indistinguishable.** With 34 non-leaky
   features every model lands at AUC ≈ 0.99 and the family diff collapses
   to ~0.0003. C3 (gradient) holds: the non-parametric advantage is real
   only when the feature pool is weak.
3. **Practical takeaway.** A URL-only first-line filter benefits a lot
   from a non-parametric model (≈0.10 AUC over LR/NB on Lexical). Once the
   system can afford to fetch the page and compute page-level features,
   the choice of model class becomes almost irrelevant and simpler,
   more explainable models (LR, LDA) are fine. This is the argument for
   putting a fast non-parametric URL-only filter *in front of* a slower
   page-level analyser rather than relying on one single model at either
   extreme.
