---
title: "Phishing Detection — Scenario 3: Feature Selection"
author: "K-o-m-A"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(caret)
library(glmnet)
library(MASS, exclude = "select")
library(pROC)
library(knitr)

set.seed(2026)

ARTIFACTS <- "scenario_3/artifacts"
CACHE_VERSION <- "v3_phishing_second_level"
dir.create(ARTIFACTS, recursive = TRUE, showWarnings = FALSE)
```

---

# 1. Scenario & Hypothesis
## 1.1 Why Lexical

In Scenario 2 §6.1, **Lexical** (13 URL-string features) was the only
tier where the choice of classifier family actually mattered: LR-Ridge
stayed at 0.861 while SVM-RBF reached 0.997. On every other tier the
linear classifier was already at or near the ceiling, so any sensible
selection method just lands at "keep 3–5 features, AUC ≈ 1.0, done".
That tells us nothing about feature selection, only that those tiers
are too easy to differentiate the methods.

Lexical also carries the multicollinearity cluster flagged in EDA §4.3:
`URLLength`, `NoOfLettersInURL` and `NoOfDegitsInURL` have VIF above
1000 and pairwise `|r| > 0.84`. This is the regime where ridge and
lasso are known to diverge: ridge spreads the weight across the bloc,
lasso keeps the bloc but redistributes it into large opposing-sign
coefficients on the survivors. We need to run on a tier where that
contrast actually shows up.

So we run S3 on **Lexical and FullLite**, treating them as two
redundancy regimes. Lexical is a *small isolated cluster* tier: one
3-feature URL-length bloc inside 13 features whose other pairwise
correlations are mild. FullLite is a *rich multi-cluster* tier: that
same URL-length cluster, a page-count cluster, and several binary
near-duplicates, all in 34 features. H2 below predicts that the gap
between Lasso and Forward grows from the first regime to the second.
Trust has only seven features, so there is no room for bloc-level
redundancy to manifest; Behavior with the six near-leakers stripped
sits dimensionally between Lexical and FullLite. Adding either would
just blur the comparison.

## 1.2 Method choice — one algorithmic + two embedded

The assignment requires one algorithmic and two embedded methods. We
use **Forward selection (AIC) + Ridge + Lasso**, the same triplet EDA
§2 already set up the H2 hypothesis around, so the criteria below map
onto H2 directly. Backward, Stepwise and ElasticNet are dropped. The
reasoning is empirical:

- **Forward over Backward / Stepwise.** The three AIC procedures
  is used as the single algorithmic method because the assignment
  asks for one algorithmic method and Forward directly gives the
  "as you **add** predictors" trajectory the assignment asks for.
- **Ridge + Lasso, no ElasticNet.** Ridge (α = 0) and Lasso (α = 1)
  sit at the two ends of the regularisation axis, which is the
  sharpest contrast we can run. ElasticNet at α = 0.5 ends up
  interpolating between them: on Lexical its retained set differs
  from Lasso by one feature, its coefficients land within ~15% of
  Lasso's (NoOfLettersInURL: Lasso 11.0 vs EN 9.16), and its AUC
  matches Lasso to three decimals. Adding EN would mean a third row
  in every table that just reads "somewhere between Ridge and Lasso".
  The one independent thing EN does show (slightly better
  fold-stability than Lasso on FullLite, e.g. NoOfPopup) we mention
  in §3.5 instead of re-running the whole pipeline.
- **Ridge vs Scenario 2 — same fit, different question.** Scenario 2
  used glmnet ridge as a classifier-family benchmark with a fixed
  λ = 0.01: one AUC number for the LR row, no feature-selection lens.
  Here we use the same solver to ask how ridge distributes weight
  across the multicollinear bloc, and how that distribution compares
  to Lasso's. Same algorithm, separate deliverables.

## 1.3 H2 (Scenario 3) — Predictor redundancy increases with tier richness

> *The divergence between the Lasso and Forward retained sets grows
> as the tier becomes richer in correlated-feature blocs. On
> **Lexical** (small isolated URL-length cluster), the two methods
> agree more on which features matter; on **FullLite** (multiple
> overlapping clusters), they agree less.*

| # | Criterion | Threshold | Where |
|---|---|---|---|
| C1 | Difference in retained-feature count between Lasso and Forward | $\lvert k_\text{Lasso} - k_\text{Forward}\rvert \le 1$ on Lexical, $\ge 3$ on FullLite | §6.2 |
| C2 | Jaccard overlap of retained sets between Lasso and Forward | $J \ge 0.85$ on Lexical, $J \le 0.70$ on FullLite | §6.2 |
| C3 | Wald p-value flips along the Forward path | $\ge 1$ flip on FullLite, $0$ flips on Lexical | §4.2.1 |

H2 is supported iff **all three** criteria (C1, C2, C3) hold.

**Two things worth being upfront about.**

- *H2 is not a restatement of EDA §4.3.* EDA showed that
  multicollinearity exists on both tiers. H2 asks the stronger
  question: whether selection methods actually behave more differently
  as the bloc structure gets richer. In principle two methods could
  agree on the same retained set even under redundancy, if AIC,
  `lambda.1se` and the bloc structure all happened to favour the same
  representative.
- *Ridge does not participate in C1/C2/C3, and that is the point.* L2 has
  zero gradient at zero, so ridge never zeros a coefficient and its
  retained set is always the full tier. Ridge earns its place in §3.4
  (the coefficient-bloc contrast — without it, Lasso's opposing-sign
  behaviour has nothing to be opposing against) and in §3.5 (a flat
  1.00 retention baseline that lets Lasso's fold-to-fold swings read
  as instability rather than noise).

---

# 2. Data Loading

Same recipe as Scenario 2 and 5 — identical sub-sample, split and CV
folds so the numbers here are directly comparable.

## 2.1 Load & exclusions

```{r load-data}
dataset_path <- if (file.exists("PhiUSIIL_Phishing_URL_Dataset.csv")) {
  "PhiUSIIL_Phishing_URL_Dataset.csv"
} else if (file.exists("../PhiUSIIL_Phishing_URL_Dataset.csv")) {
  "../PhiUSIIL_Phishing_URL_Dataset.csv"
} else stop("PhiUSIIL_Phishing_URL_Dataset.csv not found")

raw <- read_csv(dataset_path, show_col_types = FALSE)
names(raw) <- map_chr(names(raw), ~ str_replace(.x, "^\ufeff", ""))

ID_COLS          <- c("FILENAME", "URL", "Domain", "TLD", "Title")
SCORES           <- c("URLSimilarityIndex", "TLDLegitimateProb", "URLCharProb",
                      "DomainTitleMatchScore", "URLTitleMatchScore")
REDUNDANT_BINARY <- "HasObfuscation"
RATIOS           <- c("LetterRatioInURL", "DegitRatioInURL",
                      "ObfuscationRatio", "SpacialCharRatioInURL")
EXCLUDE_COLS     <- c(ID_COLS, SCORES, REDUNDANT_BINARY, RATIOS)

df <- raw %>%
  mutate(label = factor(label, levels = c(0, 1),
                        labels = c("Legitimate", "Phishing"))) %>%
  select(-all_of(EXCLUDE_COLS)) %>%
  as.data.frame()
# Note on factor order: glmnet / glm treat the *second* factor level as
# the success class. We deliberately put "Phishing" second so that
# predict(..., type = "response") returns P(Phishing) directly, and
# Sensitivity / Specificity / pROC are aligned with positive = "Phishing".
```

## 2.2 Lexical tier

```{r lexical}
lexical_features <- c("URLLength", "DomainLength", "IsDomainIP",
                      "CharContinuationRate", "TLDLength", "NoOfSubDomain",
                      "NoOfObfuscatedChar", "NoOfLettersInURL",
                      "NoOfDegitsInURL", "NoOfEqualsInURL",
                      "NoOfQMarkInURL", "NoOfAmpersandInURL",
                      "NoOfOtherSpecialCharsInURL")
stopifnot(all(lexical_features %in% names(df)))

LEAKY_BEHAVIOR <- c("LineOfCode", "NoOfExternalRef", "NoOfImage",
                    "NoOfSelfRef", "NoOfJS", "NoOfCSS")
fullLite_features <- setdiff(setdiff(names(df), "label"), LEAKY_BEHAVIOR)

cat("Lexical tier:", length(lexical_features), "features.\n")
cat("FullLite tier:", length(fullLite_features), "features.\n")
```

## 2.3 Subsample, split, folds, preprocessing

```{r prep}
N_SUB <- 30000
df_sub <- df %>%
  group_by(label) %>%
  slice_sample(n = N_SUB / 2) %>%
  ungroup() %>%
  as.data.frame()

idx       <- createDataPartition(df_sub$label, p = 0.8, list = FALSE)
train_raw <- df_sub[idx, ]
test_raw  <- df_sub[-idx, ]
fold_idx  <- createFolds(train_raw$label, k = 10, returnTrain = TRUE)

feat <- df_sub %>% select(-label)
binary_features     <- names(feat)[sapply(feat, function(x)
                                   all(na.omit(x) %in% c(0, 1)))]
continuous_features <- setdiff(names(feat), binary_features)

apply_log <- function(d, cols) {
  d[cols] <- lapply(d[cols], function(x) log1p(pmax(x, 0)))
  d
}
train_log <- apply_log(train_raw, continuous_features)
test_log  <- apply_log(test_raw,  continuous_features)

pp <- preProcess(train_log[, continuous_features], method = c("center", "scale"))
train_std <- train_log; test_std <- test_log
train_std[, continuous_features] <- predict(pp, train_log[, continuous_features])
test_std[,  continuous_features] <- predict(pp, test_log[,  continuous_features])

# Standardise the binary IsDomainIP too so glmnet / stepAIC see all
# predictors on one scale; glmnet's internal standardization would do
# this anyway but being explicit keeps coefficients directly readable.
pp_lex <- preProcess(train_std[, lexical_features], method = c("center", "scale"))
train_std[, lexical_features] <- predict(pp_lex, train_std[, lexical_features])
test_std[,  lexical_features] <- predict(pp_lex, test_std[,  lexical_features])

y_train <- train_std$label
y_test  <- test_std$label

tier_specs <- list(
  Lexical = lexical_features,
  FullLite = fullLite_features
)

X_train_by_tier <- map(tier_specs, ~ as.matrix(train_std[, .x]))
X_test_by_tier  <- map(tier_specs, ~ as.matrix(test_std[,  .x]))

```

## 2.4 Multicollinearity context — quick re-confirmation

EDA reported VIF > 1000 on the URL-length cluster. We recompute the
correlation block here so the rest of the notebook reads standalone.

```{r cor-check}
cor_pairs <- map_dfr(names(tier_specs), function(tier_name) {
  cor_mat <- cor(X_train_by_tier[[tier_name]])
  heavy_pairs <- which(abs(cor_mat) > 0.7 & row(cor_mat) < col(cor_mat),
                       arr.ind = TRUE)
  if (nrow(heavy_pairs) == 0) {
    return(tibble(tier = tier_name, a = NA_character_, b = NA_character_, r = NA_real_))
  }
  tibble(
    tier = tier_name,
    a = rownames(cor_mat)[heavy_pairs[, 1]],
    b = colnames(cor_mat)[heavy_pairs[, 2]],
    r = round(cor_mat[heavy_pairs], 3)
  )
})

cor_pairs %>%
  arrange(tier, desc(abs(r))) %>%
  kable(caption = "Feature pairs with |r| > 0.7 for Lexical and FullLite tiers.")
```

---

# 3. Embedded Methods — Ridge & Lasso

Both are `glmnet` fits with `family = "binomial"` and differ only in
`alpha` (0 = ridge, 1 = lasso). `lambda` is chosen by 10-fold CV, and
we report both `lambda.min` (optimal CV deviance) and `lambda.1se`
(sparsest model within one SE of the minimum — the canonical pick
for parsimony).

## 3.1 Fit the two paths

```{r enet-fit}
run_or_load <- function(name, fn) {
  path <- file.path(ARTIFACTS, paste0(name, "_", CACHE_VERSION, ".rds"))
  if (file.exists(path)) return(readRDS(path))
  res <- fn()
  saveRDS(res, path)
  res
}

fit_glmnet <- function(X_train_tier, alpha, tag, tier_name) {
  run_or_load(paste0("glmnet_", tolower(tier_name), "_", tag), function() {
    cv.glmnet(X_train_tier, y_train, family = "binomial",
              alpha = alpha, nfolds = 10, standardize = FALSE,
              type.measure = "auc",
              nlambda = 80,
              lambda.min.ratio = 1e-2,
              maxit = 1e6)
  })
}

cv_models <- map(names(tier_specs), function(tier_name) {
  X_tr <- X_train_by_tier[[tier_name]]
  list(
    Ridge = fit_glmnet(X_tr, 0.0, "ridge", tier_name),
    Lasso = fit_glmnet(X_tr, 1.0, "lasso", tier_name)
  )
}) %>% set_names(names(tier_specs))

```

## 3.2 Regularization paths — how features drop out

```{r reg-paths, fig.width=10, fig.height=3.6}
par(mfrow = c(2, 2))
for (tier_name in names(tier_specs)) {
  plot(cv_models[[tier_name]]$Ridge$glmnet.fit, xvar = "lambda", label = FALSE)
  title(sprintf("%s: Ridge (α=0)", tier_name), line = 2.5)
  plot(cv_models[[tier_name]]$Lasso$glmnet.fit, xvar = "lambda", label = FALSE)
  title(sprintf("%s: Lasso (α=1)", tier_name), line = 2.5)
}
par(mfrow = c(1, 1))
```

The x-axis is `log(λ)` and the right edge is where every coefficient
has been shrunk to zero. Ridge keeps every line non-zero for any λ;
only the magnitudes shrink. Lasso sends lines to zero as λ grows, and
the order in which they hit zero is the embedded answer to "which
feature is dropped next".

## 3.3 Retained features at `lambda.1se`

```{r enet-retained}
coefs_at <- function(cvfit, s = "lambda.1se") {
  co <- as.matrix(coef(cvfit, s = s))
  data.frame(feature = rownames(co),
             coef    = as.numeric(co)) %>%
    filter(feature != "(Intercept)")
}

retained <- map_dfr(names(cv_models), function(tier_name) {
  bind_rows(
    coefs_at(cv_models[[tier_name]]$Ridge) %>% mutate(method = "Ridge", tier = tier_name),
    coefs_at(cv_models[[tier_name]]$Lasso) %>% mutate(method = "Lasso", tier = tier_name)
  )
}) %>% mutate(kept = abs(coef) > 1e-8)

retained %>%
  group_by(tier, method) %>%
  summarise(retained = sum(kept),
            total    = n(),
            lambda_1se = case_when(
              method[1] == "Ridge" ~ cv_models[[tier[1]]]$Ridge$lambda.1se,
              method[1] == "Lasso" ~ cv_models[[tier[1]]]$Lasso$lambda.1se),
            .groups = "drop") %>%
  kable(caption = "Retained features at lambda.1se for Lexical and FullLite (embedded).",
        digits = 4)
```

## 3.4 Coefficient table — who survived, who didn't

```{r enet-coef-table}
retained %>%
  select(tier, method, feature, coef) %>%
  pivot_wider(names_from = method, values_from = coef) %>%
  arrange(tier, desc(abs(Ridge))) %>%
  mutate(across(c(Ridge, Lasso), ~ round(.x, 3))) %>%
  kable(caption = "Standardised coefficients at lambda.1se by tier — zeros mean lasso dropped the feature.")
```

On Lexical, Ridge spreads small and fairly uniform weight across the
URL-length trio (URLLength, NoOfLettersInURL, NoOfDegitsInURL — see
the table). Lasso keeps the trio non-zero too, but with opposing
signs and substantially larger magnitudes. This is the cancellations
pattern that comes up under an underdetermined design: no individual
coefficient in the trio is meaningful on its own, only the linear
combination is. The features Lasso zeros sit outside the trio, not
inside it — dropping a member of the trio would break the
cancellation Lasso relies on to fit the data.

```{r metrics-helper}
classif_metrics <- function(y_true, prob, thr = 0.5) {
  pred <- factor(ifelse(prob >= thr, "Phishing", "Legitimate"),
                 levels = levels(y_true))
  cm <- caret::confusionMatrix(pred, y_true, positive = "Phishing")
  auc <- as.numeric(pROC::roc(y_true, prob,
                              levels = c("Legitimate", "Phishing"),
                              direction = "<", quiet = TRUE)$auc)
  tibble(
    auc  = auc,
    acc  = unname(cm$overall["Accuracy"]),
    f1   = unname(cm$byClass["F1"]),
    prec = unname(cm$byClass["Precision"]),
    sens = unname(cm$byClass["Sensitivity"]),
    spec = unname(cm$byClass["Specificity"])
  )
}
```

## 3.5 Stability under resampling

The assignment explicitly asks how feature selection behaves "across
folds", and this is the single clearest illustration of the
redundancy story H2 formalises. Lasso is known to be unstable under
correlated predictors: small perturbations of the training set flip
which member of a collinear block it picks. We quantify this over the
10 CV folds by recording which features each fold retains at its own
`lambda.1se`.

```{r stability}
stability_fit <- run_or_load("stability", function() {
  map_dfr(names(tier_specs), function(tier_name) {
    X_tier <- X_train_by_tier[[tier_name]]
    map_dfr(seq_along(fold_idx), function(i) {
      tr <- fold_idx[[i]]
      Xi <- X_tier[tr, ]; yi <- y_train[tr]
      map_dfr(list(Ridge = 0, Lasso = 1),
        function(a) {
          cvfit <- cv.glmnet(Xi, yi, family = "binomial",
                             alpha = a, nfolds = 5, standardize = FALSE,
                             type.measure = "auc",
                             nlambda = 80,
                             lambda.min.ratio = 1e-2,
                             maxit = 1e6)
          co <- as.matrix(coef(cvfit, s = "lambda.1se"))
          tibble(tier = tier_name,
                 fold = i,
                 feature = rownames(co)[-1],
                 coef    = as.numeric(co[-1, 1]))
        }, .id = "method")
    })
  })
})

stab_summary <- stability_fit %>%
  mutate(kept = abs(coef) > 1e-8) %>%
  group_by(tier, method, feature) %>%
  summarise(retention_rate = mean(kept), .groups = "drop") %>%
  pivot_wider(names_from = method, values_from = retention_rate) %>%
  arrange(tier, desc(Ridge))

stab_summary %>%
  mutate(across(where(is.numeric), ~ round(.x, 2))) %>%
  kable(caption = "Retention rate per feature across 10 CV folds for Lexical and FullLite (1.00 = kept every fold, 0.00 = never kept).")
```

Ridge sits at retention 1.00 everywhere; that's just L2 never zeroing
a coefficient, regardless of fold. On Lexical the Lasso column is
almost binary: features are either picked in every fold or never
picked, with maybe one or two borderline cases. FullLite is where the
instability shows. The Lasso column contains a band of intermediate
retentions — features picked in some folds and not in others — which
is the coin-flip behaviour the literature describes for lasso under
correlated predictors. The same table also has a long stretch of
FullLite features where Ridge stays at 1.00 and Lasso sits at 0.00.
That is the same picking instability seen from the other side: lasso
commits to one member of a redundant bloc and never revisits the
others.

---

# 4. Algorithmic Selection — Forward (AIC)

`stats::step` walks a sequence of nested logistic regressions and
picks the one with the lowest AIC. Compared to the embedded methods
it has one extra output we care about: at every step each coefficient
has a Wald p-value, which is what the assignment refers to when it
asks for *"which features lose significance as you add or remove
predictors"*. The catch is that those p-values are not quite what
they look like, which is what §5 is about.

We use forward selection as the algorithmic representative because it
matches the assignment's "as you add predictors" framing directly.

## 4.1 Forward selection — feature entry order & p-values

```{r forward}
step_models <- run_or_load("step_models", function() {
  map(names(tier_specs), function(tier_name) {
    feats <- tier_specs[[tier_name]]
    dtr <- train_std[, c("label", feats)]
    null_mod_t <- glm(label ~ 1, data = dtr, family = binomial)
    full_mod_t <- glm(label ~ ., data = dtr, family = binomial)
    list(
      features = feats,
      forward = step(null_mod_t,
                     scope = list(lower = null_mod_t, upper = full_mod_t),
                     direction = "forward", trace = FALSE, k = 2)
    )
  }) %>% set_names(names(tier_specs))
})

map_dfr(names(step_models), function(tier_name) {
  tibble(
    tier = tier_name,
    retained = length(coef(step_models[[tier_name]]$forward)) - 1,
    total = length(step_models[[tier_name]]$features)
  )
}) %>% kable(caption = "Forward selection retained-feature counts by tier.")

# Recover the entry order and per-step Wald p-values of the ENTERING
# feature and of every previously-entered feature. This is the
# "significance as predictors are added" table the assignment asks for.
fwd_trajectory <- run_or_load("forward_traj", function() {
  map_dfr(names(step_models), function(tier_name) {
    entered <- setdiff(names(coef(step_models[[tier_name]]$forward)), "(Intercept)")
    feats <- step_models[[tier_name]]$features
    map_dfr(seq_along(entered), function(k) {
      vars <- entered[1:k]
      m <- glm(reformulate(vars, "label"),
               data = train_std[, c("label", feats)],
               family = binomial)
      co <- summary(m)$coefficients
      tibble(tier = tier_name, step = k, entering = entered[k],
             feature = rownames(co), p = co[, "Pr(>|z|)"]) %>%
        filter(feature != "(Intercept)")
    })
  })
})

fwd_trajectory %>%
  mutate(p = signif(p, 3)) %>%
  pivot_wider(id_cols = c(tier, feature), names_from = step,
              values_from = p, names_prefix = "step_") %>%
  kable(caption = "Wald p-values as forward selection enters each feature for Lexical and FullLite.")
```

## 4.2 Final selected model — coefficients and Wald p-values

The last column of §4.1 is the final selected model's per-feature
p-value. Reading it directly:

```{r forward-final-coef}
map_dfr(names(step_models), function(tier_name) {
  as.data.frame(summary(step_models[[tier_name]]$forward)$coefficients) %>%
    rownames_to_column("feature") %>%
    filter(feature != "(Intercept)") %>%
    mutate(tier = tier_name)
}) %>%
  mutate(across(-c(feature, tier), ~ signif(.x, 3))) %>%
  kable(caption = "Final forward-AIC-selected model by tier — Wald p-values are biased by the selection process (see §5).")
```

On both tiers the final selected model contains a handful of
features whose Wald p-value sits well above 0.05. The most striking
are IsDomainIP on Lexical (p ≈ 1, standard error blown up by orders
of magnitude — that is quasi-separation) and IsHTTPS on FullLite
(p ≈ 1, same story). AIC kept these because adding them did not cost
enough log-likelihood to outweigh the −2·k term, not because there is
independent statistical evidence that they matter. The other
non-significant retained features for each tier are visible in the
table above.

## 4.2.1 Wald p-value flips along the forward path

A *flip* is a feature whose Wald p-value crosses 0.05 somewhere along
the forward trajectory: significant at the step it entered and
non-significant in the final model, or vice versa. Each flip is the
multicollinearity story showing up at the level of an individual
p-value: the "evidence" attached to a predictor is not really a
property of the predictor, it is a property of the predictor together
with whatever else made it into the model. This is descriptive
support for H2: we expect more flips on the multi-cluster tier than
on the small-cluster one, in the same direction as the Jaccard gating
criterion in §6.2.

```{r p-flip-table}
fwd_flips <- fwd_trajectory %>%
  group_by(tier, feature) %>%
  arrange(step, .by_group = TRUE) %>%
  summarise(
    p_at_entry = first(p),
    p_in_final = last(p),
    sig_at_entry = first(p) < 0.05,
    sig_in_final = last(p) < 0.05,
    flipped = sig_at_entry != sig_in_final,
    .groups = "drop"
  ) %>%
  arrange(tier, desc(flipped), feature)

fwd_flips %>%
  mutate(across(c(p_at_entry, p_in_final), ~ signif(.x, 3))) %>%
  kable(caption = "Forward-path Wald p-values for every entered predictor: at the step it entered vs. in the final selected model. flipped = TRUE iff significance status (p < 0.05) changed.")

flip_counts <- fwd_flips %>%
  group_by(tier) %>%
  summarise(n_flipped = sum(flipped),
            features_flipped = paste(feature[flipped], collapse = ", "),
            .groups = "drop")

flip_counts %>%
  kable(caption = "C3 deliverable — number of significance flips per tier.")
```

This table maps directly to C3 in §1.3: for C3 to pass we need at
least one flip on FullLite and zero flips on Lexical.

---

# 5. Post-Selection Inference — the research piece

The Wald p-values reported in §4.1 and §4.2 are not valid out of the
box. This is the most important caveat about stepwise selection, and
also the one that most easily gets glossed over in practice.

## 5.1 Why the p-values are biased

A classical Wald p-value assumes the predictor set was fixed *before*
the data was seen. Stepwise selection breaks that assumption: the
predictor set is chosen *from* the data to minimise AIC (or maximise
likelihood). The p-values in the final selected model are therefore
conditional on having survived the selection, which has three
consequences:

- Selected features look more significant than they really are (the
  winner's-curse effect).
- The null distribution of the Wald statistic is no longer a χ², so
  the standard formula overstates significance.
- The post-selection inference literature (Berk et al. 2013; Lee et
  al. 2016; Tibshirani et al. 2016) argues for *selective inference* —
  p-values computed conditional on the selection event itself.
  Practical tools: the `selectiveInference` R package for lasso, and
  data splitting (Cox 1975; Wasserman & Roeder 2009).

**What this means for the notebook.** The final-model p-values in
§4.2 should not be read as "these are the significant features". They
are the numbers a naive reader would quote, no more. The honest
phrasing for a finding here is: "the forward AIC procedure retained a
subset of predictors; standard Wald p-values on those predictors are
reported but are biased by the selection procedure and should not be
used as hypothesis tests."

## 5.2 Why the forward-trajectory p-values *are* useful

The final p-values are biased, but the trajectory in §4.1 is
answering a different question. Given the entry order AIC produced,
does any feature's p-value climb above 0.05 once a correlated partner
enters at a later step? The trajectory shows the multicollinearity
mechanism directly, and that is a valid descriptive observation
rather than a hypothesis test.

The flip count and entry-vs-final p-value table in §4.2.1 condenses
the trajectory into one number per feature. Some features enter
significantly and then lose significance once a correlated partner
absorbs their signal; others enter late on AIC grounds and never
become significant at all (the quasi-separation cases from §4.2).
Both are the same underlying instability: the Wald p-value of a
predictor depends on which other predictors share the model with it.

## 5.3 Why lasso stability (§3.5) is the embedded-method answer

The embedded counterpart of "stepwise p-values are biased" is the
instability of sparse selection under correlated predictors. Lasso
picks one feature out of a bloc and zeros the rest, but which one it
picks can flip on a small perturbation of the training set. The §3.5
stability matrix shows this directly: features with retention rates
around 0.4–0.6 across folds are the same features that would have
unstable Wald p-values in an algorithmic setting. Different
mechanism, same underlying problem.

---

# 6. Comparison — Retained Sets Side by Side

## 6.1 The deliverable table (assignment's literal asks)

```{r retained-compare}
retained_by_method <- map_dfr(names(tier_specs), function(tier_name) {
  bind_rows(
    tibble(method = "Ridge",         retained = sum(abs(coef(cv_models[[tier_name]]$Ridge, s = "lambda.1se"))[-1] > 1e-8)),
    tibble(method = "Lasso",         retained = sum(abs(coef(cv_models[[tier_name]]$Lasso, s = "lambda.1se"))[-1] > 1e-8)),
    tibble(method = "Forward (AIC)", retained = length(coef(step_models[[tier_name]]$forward)) - 1)
  ) %>%
    mutate(tier = tier_name,
           total = length(tier_specs[[tier_name]]))
})

kable(retained_by_method,
      caption = "How many features each method retains in Lexical and FullLite.")
```

## 6.2 Feature-set overlap (H2 gating)

```{r overlap}
get_kept <- function(cvfit) {
  co <- as.matrix(coef(cvfit, s = "lambda.1se"))
  rownames(co)[-1][abs(co[-1, 1]) > 1e-8]
}

sets_by_tier <- map(names(tier_specs), function(tier_name) {
  list(
    Ridge   = get_kept(cv_models[[tier_name]]$Ridge),
    Lasso   = get_kept(cv_models[[tier_name]]$Lasso),
    Forward = setdiff(names(coef(step_models[[tier_name]]$forward)), "(Intercept)")
  )
}) %>% set_names(names(tier_specs))

jaccard <- function(a, b) length(intersect(a, b)) / length(union(a, b))

map_dfr(names(sets_by_tier), function(tier_name) {
  sets <- sets_by_tier[[tier_name]]
  jac_mat <- outer(names(sets), names(sets),
                   Vectorize(function(i, j) jaccard(sets[[i]], sets[[j]])))
  dimnames(jac_mat) <- list(names(sets), names(sets))
  round(jac_mat, 2) %>%
    as.data.frame() %>%
    rownames_to_column("method") %>%
    mutate(tier = tier_name)
}) %>%
  select(tier, method, everything()) %>%
  kable(caption = "Jaccard overlap of retained-feature sets by tier. Off-diagonal < 1 = disagreement.")

# C1 + C2 deliverable — retained-count gap and Jaccard with explicit
# threshold checks from §1.3.
lasso_fwd_jac <- map_dfr(names(sets_by_tier), function(tier_name) {
  s <- sets_by_tier[[tier_name]]
  tibble(tier = tier_name,
         k_lasso = length(s$Lasso),
         k_forward = length(s$Forward),
         k_diff = abs(length(s$Lasso) - length(s$Forward)),
         jaccard_lasso_forward = round(jaccard(s$Lasso, s$Forward), 3))
})

lasso_fwd_jac %>%
  mutate(
    C1_pass = case_when(
      tier == "Lexical" ~ k_diff <= 1,
      tier == "FullLite" ~ k_diff >= 3,
      TRUE ~ NA
    ),
    C2_pass = case_when(
      tier == "Lexical" ~ jaccard_lasso_forward >= 0.85,
      tier == "FullLite" ~ jaccard_lasso_forward <= 0.70,
      TRUE ~ NA
    )
  ) %>%
  kable(caption = "C1 + C2 deliverable — retained-count gap, Jaccard(Lasso, Forward), and per-tier pass/fail.")
```

The table above provides the direct C1/C2 check against fixed EDA
thresholds. The full 3×3 matrix is left in for context on Ridge × *
comparisons only; Ridge is always a near-superset, so those overlaps
do not enter the C1/C2 criteria.

## 6.3 Predictive performance at each method's retained set

```{r lexical-fulllite-compare}
tier_compare <- map_dfr(names(tier_specs), function(tier_name) {
  Xte <- X_test_by_tier[[tier_name]]
  embedded_tbl <- bind_rows(
    classif_metrics(y_test, as.numeric(predict(cv_models[[tier_name]]$Ridge,
                                               newx = Xte, s = "lambda.1se",
                                               type = "response"))) %>%
      mutate(method = "Ridge"),
    classif_metrics(y_test, as.numeric(predict(cv_models[[tier_name]]$Lasso,
                                               newx = Xte, s = "lambda.1se",
                                               type = "response"))) %>%
      mutate(method = "Lasso")
  )
  fwd_vars <- setdiff(names(coef(step_models[[tier_name]]$forward)), "(Intercept)")
  feats <- tier_specs[[tier_name]]
  m <- glm(reformulate(fwd_vars, "label"),
           data = train_std[, c("label", feats)], family = binomial)
  fwd_pr <- as.numeric(predict(m, test_std, type = "response"))
  fwd_tbl <- classif_metrics(y_test, fwd_pr) %>% mutate(method = "Forward")
  ks <- c(Ridge   = sum(abs(coef(cv_models[[tier_name]]$Ridge, s = "lambda.1se"))[-1] > 1e-8),
          Lasso   = sum(abs(coef(cv_models[[tier_name]]$Lasso, s = "lambda.1se"))[-1] > 1e-8),
          Forward = length(fwd_vars))
  bind_rows(embedded_tbl, fwd_tbl) %>%
    mutate(tier = tier_name, k = ks[method]) %>%
    select(tier, method, k, auc, sens, spec)
}) %>%
  mutate(method = factor(method, levels = c("Ridge", "Lasso", "Forward")),
         across(c(auc, sens, spec), ~ round(.x, 4))) %>%
  arrange(tier, method)

tier_compare %>%
  kable(caption = "Test AUC + Sensitivity + Specificity per method by tier (threshold = 0.5). For Forward, predictions come from a plain logistic regression refit on the retained set.")
```

---

# 7. H2 Verdict

H2 is **supported iff all three criteria C1, C2 and C3 hold**
(§1.3).

```{r h2-verdict}
verdict <- lasso_fwd_jac %>%
  select(tier, jaccard_lasso_forward, k_diff) %>%
  pivot_wider(names_from = tier, values_from = c(jaccard_lasso_forward, k_diff))

flip_compact <- flip_counts %>%
  select(tier, n_flipped) %>%
  pivot_wider(names_from = tier, values_from = n_flipped, names_prefix = "flips_")

bind_cols(verdict, flip_compact) %>%
  mutate(
    C1_pass = (k_diff_Lexical <= 1) & (k_diff_FullLite >= 3),
    C2_pass = (jaccard_lasso_forward_Lexical >= 0.85) & (jaccard_lasso_forward_FullLite <= 0.70),
    C3_pass = (flips_FullLite >= 1) & (flips_Lexical == 0),
    H2_supported = C1_pass & C2_pass & C3_pass
  ) %>%
  kable(caption = "H2 verdict — strict EDA criteria: C1, C2, C3 must all be TRUE.")
```

**Reading the verdict.** H2 is supported only when `C1_pass`,
`C2_pass`, and `C3_pass` are all `TRUE`. If any one is `FALSE`, H2 is
not supported under the strict EDA criteria.

**Why the redundancy story matters here.** Lexical has one isolated
correlation cluster, the URL-length trio, and the rest of its
features carry non-overlapping information. Lasso and Forward
therefore agree more often: both pick the trio plus the obvious
unique signals. FullLite combines that same URL-length cluster with a
page-count cluster and several binary near-duplicates, which gives
several equivalence classes of redundant predictors at once. Lasso
picks one bloc representative and Forward picks another, so the
retained sets drift apart. That is what §6.2's gating comparison is
measuring.
