---
title: "Phishing Detection — Scenario 2: Parametric vs Non-parametric"
author: "K-o-m-A"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(caret)
library(glmnet)
library(MASS, exclude = "select")
library(klaR)
library(randomForest)
library(kernlab)
library(pROC)
library(knitr)
library(digest)

set.seed(2026)

ARTIFACTS <- "scenario_2/artifacts"
dir.create(ARTIFACTS, recursive = TRUE, showWarnings = FALSE)
```

---

# 1. Scenario & Hypothesis

The EDA (`phishing.rmd` §2) states **H1** as: *on the URL-only tier,
non-parametric models pick up meaningfully more signal than parametric
ones.* Section 5 of the EDA also points out that on the **Full** feature
set (all 40 remaining predictors) every model type already reaches
AUC ≈ 0.99 — the dataset is near-perfectly separable there, so the
parametric-vs-non-parametric contrast can't be seen on Full alone.

To give H1 a fair test, we compare all six models across **four feature
tiers**:

| Tier | Features | What the model sees |
|------|---------:|---------------------|
| **Lexical** (URL-only) | 13 | URL string only — the first-line case: no DNS, no HTML fetch. |
| **Trust** | 7 | Domain / title / HTTPS flags — available after a DNS lookup and light page inspection. |
| **Behavior** | 14 | Page-level content signals (iframes, forms, redirects, hidden fields), **with the six near-leaky page counts removed** for the same reason we remove them from FullLite. |
| **FullLite** | 34 | All three families combined minus those six near-leakers. The high-dimensional benchmark (> 30 predictors) without single-feature near-saturating signal. |

Six Behavior features — `LineOfCode`, `NoOfExternalRef`, `NoOfImage`,
`NoOfSelfRef`, `NoOfJS`, `NoOfCSS` — each individually give a univariate
AUC between 0.958 and 0.990 on this dataset. Phishing pages are usually
shallow HTML shells with almost no scripts, images or external links, so
these counts effectively solve the task on their own, even though they
are genuine page-level signals rather than computed scores. Leaving them
in any tier pins every model at AUC ≈ 1.0 and flattens the family
contrast we are trying to measure. We therefore exclude them
**consistently everywhere they would otherwise appear** — both from
Behavior and from the high-dimensional `FullLite` tier. A "Full" tier
that puts them back is redundant once `FullLite` already saturates (see
the results tables in §6), and it would only let us defend the same
AUC ≈ 1.0 twice.

The smaller tiers (Lexical, Trust, Behavior) are deliberate ablations that
isolate H1's claim, which is specifically about the URL-only regime. By
scanning all four tiers we can sharpen H1 from a single-tier claim into a
**gradient statement**: *the non-parametric advantage is largest when the
feature tier is weak and shrinks as more signal is added.*

---

# 2. Data Loading

## 2.1 Load raw file

Same PhiUSIIL CSV as in the EDA. We strip the UTF-8 BOM the provider
attaches to the first column name.

```{r load-data}
dataset_path <- if (file.exists("PhiUSIIL_Phishing_URL_Dataset.csv")) {
  "PhiUSIIL_Phishing_URL_Dataset.csv"
} else if (file.exists("../PhiUSIIL_Phishing_URL_Dataset.csv")) {
  "../PhiUSIIL_Phishing_URL_Dataset.csv"
} else stop("PhiUSIIL_Phishing_URL_Dataset.csv not found")

raw <- read_csv(dataset_path, show_col_types = FALSE)
names(raw) <- map_chr(names(raw), ~ str_replace(.x, "^\ufeff", ""))
cat("Loaded", nrow(raw), "rows,", ncol(raw), "columns.\n")
```

## 2.2 Apply EDA feature exclusions

The EDA (§3.3) dropped three groups of features: computed scores, one
redundant binary flag, and linearly dependent ratios. Repeating the same
exclusions here makes sure Scenario 2 is tested on the **same 40 features**
the EDA analysed, so the EDA's VIF, SMD and skewness findings apply here
directly.

```{r exclusions}
ID_COLS          <- c("FILENAME", "URL", "Domain", "TLD", "Title")
SCORES           <- c("URLSimilarityIndex", "TLDLegitimateProb", "URLCharProb",
                      "DomainTitleMatchScore", "URLTitleMatchScore")
REDUNDANT_BINARY <- "HasObfuscation"
RATIOS           <- c("LetterRatioInURL", "DegitRatioInURL",
                      "ObfuscationRatio", "SpacialCharRatioInURL")
EXCLUDE_COLS     <- c(ID_COLS, SCORES, REDUNDANT_BINARY, RATIOS)

df <- raw %>%
  mutate(label = factor(label, levels = c(1, 0),
                        labels = c("Phishing", "Legitimate"))) %>%
  select(-all_of(EXCLUDE_COLS)) %>%
  as.data.frame()

cat("Feature matrix after exclusions:", ncol(df) - 1, "predictors.\n")
```

## 2.3 Feature tiers

The family map is the same as in EDA §3. Each tier is just a subset of
column names, so switching tier means subsetting columns — no
re-preprocessing needed.

```{r family-map}
family_map <- tribble(
  ~feature,                     ~family,
  "URLLength",                  "Lexical",
  "DomainLength",               "Lexical",
  "IsDomainIP",                 "Lexical",
  "CharContinuationRate",       "Lexical",
  "TLDLength",                  "Lexical",
  "NoOfSubDomain",              "Lexical",
  "NoOfObfuscatedChar",         "Lexical",
  "NoOfLettersInURL",           "Lexical",
  "NoOfDegitsInURL",            "Lexical",
  "NoOfEqualsInURL",            "Lexical",
  "NoOfQMarkInURL",             "Lexical",
  "NoOfAmpersandInURL",         "Lexical",
  "NoOfOtherSpecialCharsInURL", "Lexical",
  "IsHTTPS",                    "Trust",
  "HasTitle",                   "Trust",
  "HasFavicon",                 "Trust",
  "Robots",                     "Trust",
  "Bank",                       "Trust",
  "Pay",                        "Trust",
  "Crypto",                     "Trust",
  "LineOfCode",                 "Behavior",
  "LargestLineLength",          "Behavior",
  "IsResponsive",               "Behavior",
  "NoOfURLRedirect",            "Behavior",
  "NoOfSelfRedirect",           "Behavior",
  "HasDescription",             "Behavior",
  "NoOfPopup",                  "Behavior",
  "NoOfiFrame",                 "Behavior",
  "HasExternalFormSubmit",      "Behavior",
  "HasSocialNet",               "Behavior",
  "HasSubmitButton",            "Behavior",
  "HasHiddenFields",            "Behavior",
  "HasPasswordField",           "Behavior",
  "HasCopyrightInfo",           "Behavior",
  "NoOfImage",                  "Behavior",
  "NoOfCSS",                    "Behavior",
  "NoOfJS",                     "Behavior",
  "NoOfSelfRef",                "Behavior",
  "NoOfEmptyRef",               "Behavior",
  "NoOfExternalRef",            "Behavior"
)

LEAKY_BEHAVIOR <- c("LineOfCode", "NoOfExternalRef", "NoOfImage",
                    "NoOfSelfRef", "NoOfJS", "NoOfCSS")

tiers <- list(
  Lexical  = family_map %>% filter(family == "Lexical")  %>% pull(feature),
  Trust    = family_map %>% filter(family == "Trust")    %>% pull(feature),
  Behavior = setdiff(family_map %>% filter(family == "Behavior") %>% pull(feature),
                     LEAKY_BEHAVIOR),
  FullLite = setdiff(family_map$feature, LEAKY_BEHAVIOR)
)

sapply(tiers, length)
```

---

# 3. Modelling Setup

## 3.1 Stratified sub-sampling

The full dataset has ~235 k rows. SVM with an RBF kernel is
\(O(n^2)\)–\(O(n^3)\) in the training size, so the full dataset would
take hours per fold. Instead we take a stratified 30 k sub-sample
(15 k phishing / 15 k legitimate). After the 80/20 split this leaves
24 k rows per fold — the largest size where all six models still finish
in reasonable time. Stratification keeps the class balance constant
across tiers and folds.

```{r subsample}
N_SUB <- 30000
df_sub <- df %>%
  group_by(label) %>%
  slice_sample(n = N_SUB / 2) %>%
  ungroup() %>%
  as.data.frame()

table(df_sub$label)
```

## 3.2 Shared train/test split and CV folds

Every model × tier combination uses **the same rows** for training, the
**same 80/20 split**, and the **same 5-fold CV indices**. The only thing
that changes between runs is which columns the model gets to see. That
is what makes the per-fold AUCs paired, so we can run a paired Wilcoxon
test later.

```{r splits}
idx       <- createDataPartition(df_sub$label, p = 0.8, list = FALSE)
train_raw <- df_sub[idx, ]
test_raw  <- df_sub[-idx, ]
fold_idx  <- createFolds(train_raw$label, k = 5, returnTrain = TRUE)

cat("Train:", nrow(train_raw), "Test:", nrow(test_raw),
    "CV folds:", length(fold_idx), "\n")
```

## 3.3 Preprocessing recipes

We need two parallel recipes:

- **Parametric / distance recipe** — `log1p()` on every continuous
  feature (the EDA flagged \|skew\| > 2 on 19 of 22 continuous
  predictors), then centre and scale. Binary features are left alone.
  Feeds LR, LDA, NB, SVM-RBF, KNN.
- **Tree recipe** — no transform. Trees don't care about monotone
  transforms or scale. Feeds Random Forest.

`preProcess` is fit once on the full training set (no per-fold refit).
At 24 k rows with fixed hyperparameters the leakage is negligible, and
it keeps the code simple.

```{r recipes}
feat <- df_sub %>% select(-label)
binary_features     <- names(feat)[sapply(feat, function(x)
                                   all(na.omit(x) %in% c(0, 1)))]
continuous_features <- setdiff(names(feat), binary_features)

apply_log <- function(d, cols) {
  d[cols] <- lapply(d[cols], function(x) log1p(pmax(x, 0)))
  d
}
train_log <- apply_log(train_raw, continuous_features)
test_log  <- apply_log(test_raw,  continuous_features)

pp <- preProcess(train_log[, continuous_features], method = c("center", "scale"))
train_std <- train_log; test_std  <- test_log
train_std[, continuous_features] <- predict(pp, train_log[, continuous_features])
test_std[,  continuous_features] <- predict(pp, test_log[,  continuous_features])

cat(length(continuous_features), "continuous,",
    length(binary_features), "binary features.\n")
```

## 3.4 Per-tier fit utility

Fitting the same model across four tiers would mean four almost identical
blocks of code, so we wrap the logic in one function. It subsets the
data to a tier's features, runs 5-fold CV using the shared `fold_idx`,
and returns per-fold AUCs plus test-set metrics.

```{r fit-utility}
fit_one_tier <- function(tier_features, data_train, data_test,
                         method, tuneGrid = NULL, ...) {
  tr <- data_train[, c("label", tier_features)]
  te <- data_test[,  c("label", tier_features)]

  ctrl <- trainControl(
    method          = "cv",
    index           = fold_idx,
    classProbs      = TRUE,
    summaryFunction = twoClassSummary,
    savePredictions = "final"
  )
  t0 <- Sys.time()
  fit <- caret::train(label ~ ., data = tr,
                      method = method, trControl = ctrl,
                      metric = "ROC", tuneGrid = tuneGrid, ...)
  elapsed <- as.numeric(difftime(Sys.time(), t0, units = "secs"))

  test_prob <- predict(fit, te, type = "prob")[, "Phishing"]
  test_pred <- predict(fit, te)
  test_auc  <- as.numeric(pROC::roc(te$label, test_prob,
                                    levels = c("Legitimate", "Phishing"),
                                    direction = "<", quiet = TRUE)$auc)
  cm <- caret::confusionMatrix(test_pred, te$label, positive = "Phishing")

  train_prob <- predict(fit, tr, type = "prob")[, "Phishing"]
  train_pred <- predict(fit, tr)
  train_auc  <- as.numeric(pROC::roc(tr$label, train_prob,
                                     levels = c("Legitimate", "Phishing"),
                                     direction = "<", quiet = TRUE)$auc)
  train_cm   <- caret::confusionMatrix(train_pred, tr$label, positive = "Phishing")

  list(cv_per_fold = fit$resample,
       train_auc = train_auc,
       train_acc = unname(train_cm$overall["Accuracy"]),
       test_auc = test_auc,
       test_acc = unname(cm$overall["Accuracy"]),
       test_f1  = unname(cm$byClass["F1"]),
       test_sens = unname(cm$byClass["Sensitivity"]),
       test_spec = unname(cm$byClass["Specificity"]),
       train_secs = elapsed)
}

tiers_fingerprint <- function(x)
  digest::digest(lapply(x[sort(names(x))], sort))

run_or_load <- function(name, fn) {
  path <- file.path(ARTIFACTS, paste0("res_", name, ".rds"))
  fp   <- tiers_fingerprint(tiers)
  if (file.exists(path)) {
    cached <- readRDS(path)
    tiers_ok   <- setequal(names(cached$by_tier), names(tiers))
    schema_ok  <- all(c("train_auc", "train_acc") %in%
                      names(cached$by_tier[[1]]))
    fp_ok      <- isTRUE(cached$tiers_fp == fp)
    if (tiers_ok && schema_ok && fp_ok) return(cached)
    cat(sprintf("  [%s] cache stale (tiers_ok=%s, schema_ok=%s, fp_ok=%s) — refitting\n",
                name, tiers_ok, schema_ok, fp_ok))
  }
  res <- fn()
  res$tiers_fp <- fp
  saveRDS(res, path)
  res
}

derive_precision <- function(sens, spec)
  sens / (sens + (1 - spec))

print_tier_summary <- function(res) {
  tbl <- map_dfr(names(res$by_tier), function(tier) {
    r <- res$by_tier[[tier]]
    gap <- r$train_auc - r$test_auc
    prec <- derive_precision(r$test_sens, r$test_spec)
    tibble(
      Tier        = tier,
      `Train AUC` = sprintf("%.4f", r$train_auc),
      `CV ROC`    = sprintf("%.4f ± %.4f",
                            mean(r$cv_per_fold$ROC), sd(r$cv_per_fold$ROC)),
      `Test AUC`  = sprintf("%.4f", r$test_auc),
      `Gap`       = sprintf("%+.4f", gap),
      Accuracy    = sprintf("%.4f", r$test_acc),
      F1          = sprintf("%.4f", r$test_f1),
      Prec        = sprintf("%.4f", prec),
      Sens        = sprintf("%.4f", r$test_sens),
      Spec        = sprintf("%.4f", r$test_spec),
      `Train (s)` = sprintf("%.1f", r$train_secs)
    )
  })
  knitr::kable(
    tbl,
    align   = "lrrrrrrrrrr",
    caption = sprintf("%s — %s  (Gap = Train AUC − Test AUC; > 0.01 hints at overfit; Prec derived from Sens/Spec on balanced test set)",
                      res$model, res$family)
  )
}
```

`run_or_load` is a small cache wrapper: the first knit fits everything
and writes `scenario_2/artifacts/res_<model>.rds`; later knits just load
those files and finish in seconds. If the tier set on disk doesn't match
the tier set defined in the notebook any more, the cache is flagged as
stale and the model is refit automatically. `print_tier_summary` returns a `kable` table (one row per tier) with
CV ROC ± SD, held-out test AUC, Accuracy, F1, Sensitivity, Specificity,
and wall-clock train time. Each fit chunk therefore produces a small
model-titled table rather than a dense one-liner per tier.

---

# 4. Parametric Models

## 4.1 Logistic Regression (ridge)

EDA §4.3 showed a severe collinearity cluster around `URLLength`
(VIF > 1000) on the Lexical tier, plus several |r| > 0.7 pairs on the
full 40-feature set.
Plain (unpenalised) logistic regression becomes numerically unstable in
that setting — the coefficients swing wildly from fold to fold. So we
use **ridge-regularised** LR (α = 0, λ = 0.01), which keeps the
coefficients bounded without doing any feature selection.

```{r fit-lr}
res_lr <- run_or_load("lr", function() {
  out <- list(model = "LogReg-Ridge", family = "parametric", by_tier = list())
  for (tier in names(tiers)) {
    cat(sprintf("  LogReg-Ridge   [%-8s] fitting (%d features)...\n",
                tier, length(tiers[[tier]])))
    out$by_tier[[tier]] <- fit_one_tier(
      tiers[[tier]], train_std, test_std,
      method = "glmnet",
      tuneGrid = data.frame(alpha = 0, lambda = 0.01))
  }
  out
})
print_tier_summary(res_lr)
```

## 4.2 Linear Discriminant Analysis (LDA)

LDA assumes Gaussian class-conditional densities with a shared covariance
matrix. The EDA flagged heavy skewness on 19 of 22 continuous features;
`log1p` + centre/scale in our preprocessing recipe cleans this up
reasonably well, so LDA is fine in principle. LDA has no tunable
hyperparameters, which makes it a clean "linear baseline".

```{r fit-lda}
res_lda <- run_or_load("lda", function() {
  out <- list(model = "LDA", family = "parametric", by_tier = list())
  for (tier in names(tiers)) {
    cat(sprintf("  LDA            [%-8s] fitting (%d features)...\n",
                tier, length(tiers[[tier]])))
    out$by_tier[[tier]] <- fit_one_tier(
      tiers[[tier]], train_std, test_std, method = "lda")
  }
  out
})
print_tier_summary(res_lda)
```

## 4.3 Naive Bayes

NB assumes features are conditionally **independent** given the class.
EDA §4.3 already showed that this is not true here: |r| > 0.7 inside
the URL-length cluster and between Behavior ref-count features. We run
NB anyway, to measure **how much the independence assumption actually
costs** on each tier. We use a kernel density estimate so that folds
with a zero-variance feature (e.g. `Robots` on Trust) don't crash the
Gaussian version of NB.

```{r fit-nb, message=FALSE, warning=FALSE}
res_nb <- run_or_load("nb", function() {
  out <- list(model = "NaiveBayes", family = "parametric", by_tier = list())
  for (tier in names(tiers)) {
    cat(sprintf("  NaiveBayes     [%-8s] fitting (%d features)...\n",
                tier, length(tiers[[tier]])))
    out$by_tier[[tier]] <- suppressWarnings(fit_one_tier(
      tiers[[tier]], train_std, test_std,
      method = "nb",
      tuneGrid = data.frame(fL = 1, usekernel = TRUE, adjust = 1)))
  }
  out
})
print_tier_summary(res_nb)
```

---

# 5. Non-parametric Models

## 5.1 Random Forest

Random Forest is the natural non-parametric counterpart to LR: it builds
piecewise-constant decision regions as an ensemble of trees and is
invariant to monotone feature transforms. `mtry = floor(sqrt(p))` scales
the random feature subset with the tier size; 500 trees is plenty for
the AUC to stabilise at this sample size.

```{r fit-rf}
res_rf <- run_or_load("rf", function() {
  out <- list(model = "RandomForest", family = "non-parametric", by_tier = list())
  for (tier in names(tiers)) {
    p    <- length(tiers[[tier]])
    mtry <- max(1, floor(sqrt(p)))
    cat(sprintf("  RandomForest   [%-8s] fitting (%d features, mtry=%d)...\n",
                tier, p, mtry))
    out$by_tier[[tier]] <- fit_one_tier(
      tiers[[tier]], train_raw, test_raw,
      method = "rf",
      tuneGrid = data.frame(mtry = mtry),
      ntree = 500)
  }
  out
})
print_tier_summary(res_rf)
```

## 5.2 SVM with RBF kernel

The RBF kernel fits smooth non-linear boundaries — a natural alternative
to tree ensembles and distance-based methods. RBF is scale-sensitive, so
SVM uses the parametric recipe (`log1p` + centre/scale). `C = 1`,
`σ = 0.1` are sensible defaults for standardised inputs. We didn't tune
further because the main question is the parametric vs. non-parametric
contrast, not within-family optimisation.

This is the single slowest chunk in the notebook. The first knit takes
about 15 minutes across all tiers combined; re-knits are instant thanks
to the `run_or_load` cache.

```{r fit-svm}
res_svm <- run_or_load("svm", function() {
  out <- list(model = "SVM-RBF", family = "non-parametric", by_tier = list())
  for (tier in names(tiers)) {
    cat(sprintf("  SVM-RBF        [%-8s] fitting (%d features)...\n",
                tier, length(tiers[[tier]])))
    out$by_tier[[tier]] <- fit_one_tier(
      tiers[[tier]], train_std, test_std,
      method = "svmRadial",
      tuneGrid = data.frame(C = 1, sigma = 0.1))
  }
  out
})
print_tier_summary(res_svm)
```

## 5.3 k-Nearest Neighbours

KNN is the simplest non-parametric classifier: it stores the whole
training set and classifies each test point by a majority vote among
its `k` closest neighbours, using Euclidean distance. Being
distance-based, it uses the parametric recipe. `k = 25` is a mild
smoother that keeps the decision surface from being jagged.

One twist: when a tier has **only binary features** (Trust has 7 binary
indicators), many training points sit at identical coordinates, and
caret's `knn3` refuses to run with "too many ties in knn". We add a tiny
Gaussian jitter (SD = 10⁻³, less than 0.1 % of the feature scale) to
break the distance ties without changing any decisions.

```{r fit-knn}
jitter_features <- function(d, cols, sd = 1e-3) {
  for (c in cols) d[[c]] <- d[[c]] + rnorm(nrow(d), 0, sd)
  d
}

res_knn <- run_or_load("knn", function() {
  out <- list(model = "KNN", family = "non-parametric", by_tier = list())
  for (tier in names(tiers)) {
    feats <- tiers[[tier]]
    cat(sprintf("  KNN            [%-8s] fitting (%d features)...\n",
                tier, length(feats)))
    tr_j  <- jitter_features(train_std, feats)
    te_j  <- jitter_features(test_std,  feats)
    out$by_tier[[tier]] <- fit_one_tier(
      feats, tr_j, te_j,
      method = "knn",
      tuneGrid = data.frame(k = 25))
  }
  out
})
print_tier_summary(res_knn)
```

---

# 6. Comparison

## 6.1 Per-(model, tier) summary

First we build a long table with one row per `(model, tier, fold)`, then
a wide summary with mean ± SD CV AUC and held-out test metrics.

```{r build-long}
all_results <- list(res_lr, res_lda, res_nb, res_rf, res_svm, res_knn)

auc_long <- map_dfr(all_results, function(r) {
  map_dfr(names(r$by_tier), function(tier)
    r$by_tier[[tier]]$cv_per_fold %>%
      transmute(model = r$model, family = r$family, tier = tier,
                fold = Resample, auc = ROC))
})

test_tbl <- map_dfr(all_results, function(r) {
  map_dfr(names(r$by_tier), function(tier) {
    t <- r$by_tier[[tier]]
    tibble(model = r$model, family = r$family, tier = tier,
           train_auc = t$train_auc, train_acc = t$train_acc,
           test_auc = t$test_auc, test_acc = t$test_acc,
           test_f1  = t$test_f1,  test_sens = t$test_sens,
           test_spec = t$test_spec,
           test_prec = derive_precision(t$test_sens, t$test_spec),
           train_secs = t$train_secs)
  })
})

summary_tbl <- auc_long %>%
  group_by(model, family, tier) %>%
  summarise(cv_auc_mean = mean(auc), cv_auc_sd = sd(auc), .groups = "drop") %>%
  left_join(test_tbl, by = c("model", "family", "tier")) %>%
  mutate(tier = factor(tier, levels = c("Lexical", "Trust",
                                        "Behavior", "FullLite")),
         gap_auc = train_auc - test_auc) %>%
  arrange(tier, family, desc(cv_auc_mean))

summary_tbl %>%
  mutate(across(c(cv_auc_mean, cv_auc_sd, train_auc, test_auc, gap_auc,
                  train_acc, test_acc, test_f1, test_prec),
                ~ round(.x, 4)),
         train_secs = round(train_secs, 1)) %>%
  kable()
```

**How to read this table:** inside each tier the two families should
cluster differently if H1 holds. On Lexical, the three non-parametric
rows should sit clearly above the three parametric ones. Moving through
Trust → Behavior → FullLite, the gap should shrink and eventually
vanish. FullLite is the honest high-dimensional check: it keeps > 30
features but removes the six near-leaky Behavior counts. If parametric
and non-parametric models still disagree there, the gradient claim
holds up at realistic feature sizes too.

### Train vs. test — overfitting diagnostic

The `Train AUC` and `Gap = Train − Test` columns in each per-model table
answer *"how much does the model memorise the training set?"*. For the
three parametric models we expect `Train AUC ≈ Test AUC`: they have few
effective parameters and Ridge / LDA / NB cannot memorise 24 k rows.
For the three non-parametric models we expect a larger gap: Random
Forest predicting on its own training set is near-perfect (each point
was in the bag of most trees), KNN with `k = 25` averages its own
neighbours, and SVM-RBF can fit complex boundaries. A small gap there
is therefore not a problem — the key comparison is
**`CV ROC` vs. `Test AUC`**, both of which are *out-of-sample*. If
those two agree, the model generalises regardless of its in-bag fit.
The train column is only there to show that our CV numbers are not
silently inflated by leakage.

### Classification quality beyond AUC

AUC ranks models by their score ordering — useful, but on the richer
tiers every model sits at AUC ≈ 0.99 and the ranking becomes hard to
read. The table below shows **Accuracy, F1, Precision, Sensitivity
(= Recall) and Specificity** on the held-out test set at the default
0.5 threshold.

In our security setting each split matters on its own:
**Sensitivity (Recall)** = share of phishing URLs we caught (missing
one is the costly error), **Specificity** = share of legitimate URLs
we didn't flag (over-flagging hurts user trust), **Precision** = of
the URLs we blocked, how many were actually phishing (low precision
means users see many false alarms and disable the filter).

Precision is derived analytically from Sensitivity and Specificity
using the test-set class balance (≈ 50/50 after stratified splitting):
$\text{Prec} = \frac{\text{Sens}}{\text{Sens} + (1 - \text{Spec})}$. This
is exact for a 1:1 prevalence and within 10⁻⁴ of the
`caret::confusionMatrix` value for our near-balanced split, so no
re-fit is needed.

```{r quality-table}
quality_tbl <- test_tbl %>%
  mutate(tier = factor(tier, levels = c("Lexical", "Trust",
                                        "Behavior", "FullLite"))) %>%
  arrange(tier, family, desc(test_f1)) %>%
  transmute(Model = model, Family = family, Tier = tier,
            Accuracy    = round(test_acc,  4),
            F1          = round(test_f1,   4),
            Precision   = round(test_prec, 4),
            Sensitivity = round(test_sens, 4),
            Specificity = round(test_spec, 4))

quality_tbl %>%
  kable(caption = "Test-set classification quality at the default 0.5 threshold.")
```

**How to read this table:**

- **Lexical** translates the AUC story into practical terms: SVM-RBF
  reaches Accuracy ≈ 0.98 and F1 ≈ 0.98, while LR-Ridge and NaiveBayes
  sit around 0.69–0.82. The gap that showed up as +0.10 AUC is the
  difference between a deployable filter and an unusable one.
- **Behavior / FullLite** is where Accuracy and F1 earn their keep.
  Once the six near-leaky page counts are stripped out, Behavior no
  longer saturates — AUCs sit in a visibly non-trivial range and the
  Accuracy / F1 spread across models becomes a more honest signal of
  which classifier would actually deploy well. FullLite is the reverse
  case: AUC is already near-saturated, so Accuracy and F1 are what
  separate "great" from "essentially perfect".
- **Sensitivity vs. Specificity** shows *how* each model errs. Models
  that lean heavily on one (high Sens, lower Spec) will over-block
  legitimate sites; the opposite pattern lets phishing through. In
  production the right threshold depends on the cost ratio, but at
  0.5 the table tells us which models are balanced by default.

## 6.2 Per-tier paired Wilcoxon test

For each tier we take the per-fold mean AUC of the three non-parametric
models and the three parametric models, then compare them with a
one-sided paired Wilcoxon test (non-parametric > parametric). Because we
share folds across models, the test is fully paired — each entry in
either series comes from the same five validation sets.

```{r h1-test}
per_tier_test <- auc_long %>%
  group_by(tier, fold, family) %>%
  summarise(auc = mean(auc), .groups = "drop") %>%
  pivot_wider(names_from = family, values_from = auc) %>%
  group_by(tier) %>%
  summarise(
    mean_np  = mean(`non-parametric`),
    mean_p   = mean(parametric),
    diff     = mean(`non-parametric` - parametric),
    wilcox_V = wilcox.test(`non-parametric`, parametric,
                           paired = TRUE, alternative = "greater")$statistic,
    wilcox_p = wilcox.test(`non-parametric`, parametric,
                           paired = TRUE, alternative = "greater")$p.value,
    .groups = "drop"
  ) %>%
  mutate(tier = factor(tier, levels = c("Lexical", "Trust",
                                        "Behavior", "FullLite"))) %>%
  arrange(tier)

per_tier_test %>%
  mutate(across(c(mean_np, mean_p, diff), ~ round(.x, 4)),
         wilcox_p = signif(wilcox_p, 3)) %>%
  kable()
```

A caveat about the p-value column: with only **five paired
observations** (one per CV fold), the smallest p-value a one-sided
paired Wilcoxon can possibly produce is \(1/32 \approx 0.031\). Three
of the four tiers hit exactly this floor — meaning every non-parametric
fold beat its matched parametric fold. So the p-value alone can't tell
apart a large effect (Lexical, diff ≈ 0.10) from a negligible one
(Behavior, diff ≈ 0.0004). **Read the `diff` column, not the p-value**,
if you want to judge effect size.

## 6.3 CV AUC by model × tier

```{r plot-auc, fig.width=10, fig.height=14, message=FALSE}
plot_list <- map(c("Lexical", "Trust", "Behavior", "FullLite"), function(t) {
  auc_long %>%
    filter(tier == t) %>%
    mutate(family = factor(family, levels = c("parametric", "non-parametric")),
           model  = fct_reorder(model, auc, median)) %>%
    ggplot(aes(model, auc, fill = family)) +
    geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.5) +
    geom_jitter(aes(colour = family),
                width = 0.12, height = 0, size = 1.6, alpha = 0.8,
                show.legend = FALSE) +
    coord_flip() +
    scale_fill_manual(values   = c(parametric = "steelblue",
                                   `non-parametric` = "firebrick")) +
    scale_colour_manual(values = c(parametric = "steelblue",
                                   `non-parametric` = "firebrick")) +
    labs(title = t, x = NULL, y = "ROC AUC (5-fold CV)") +
    theme_minimal(base_size = 11) +
    theme(legend.position = if (t == "Lexical") "top" else "none")
})

library(patchwork)
wrap_plots(plot_list, ncol = 1) +
  plot_annotation(
    title    = "Scenario 2 -- CV AUC by model across feature tiers",
    subtitle = "Each dot is one CV fold (shared across models)"
  )
```

The Lexical panel shows the widest spread and the clearest separation
between parametric (blue) and non-parametric (red) models. As we go
down the tiers, the boxes tighten and the two colour groups merge.
The FullLite panel pins every model at AUC ≈ 1 — the dataset is
near-perfectly separable on a rich page-level feature set even after
the six near-leaky counts are removed, so the family contrast is
essentially resolved by then.

---

# 7. Conclusion — H1 gradient

H1 in the EDA was a single-tier claim about URL-only detection. The
four-tier comparison lets us make the stronger, more informative
version:

1. **Non-parametric > parametric when the feature tier is weak.** On
   the **Lexical** tier the non-parametric advantage is the largest
   of any tier, and every single CV fold shows the same order. This is
   exactly the regime H1 targets: URL counts interact in non-linear
   ways (digit runs × length × subdomain depth), and linear or Gaussian
   decision boundaries just can't capture those interactions. Read the
   exact diff off §6.2.
2. **Trust also favours non-parametric, but only barely.** With only
   seven features — most of them plain binary flags — there's little
   non-linear structure left for non-parametric models to discover.
3. **Behavior is the honest middle tier.** With the six near-leaky
   page counts (`LineOfCode`, `NoOfExternalRef`, `NoOfImage`,
   `NoOfSelfRef`, `NoOfJS`, `NoOfCSS`) stripped out — each of which
   individually gives univariate AUC ≥ 0.95 and single-handedly solves
   the task — the remaining 14 Behavior features are a realistic
   page-level signal without saturation. This is where we can read the
   family contrast cleanly at medium signal strength.
4. **FullLite is the honest > 30-feature benchmark.** It meets the
   assignment's "several dozen parameters" rule without relying on
   features that single-handedly solve the task. Even so, 34
   non-leaky features are already enough to make every model cluster
   near the ceiling — the family contrast on FullLite tells us whether
   non-parametric flexibility still buys anything in a realistic
   high-dimensional setting, or whether the task is simply resolved by
   that point.

**Train–test gap check.** The per-model tables in §4–§5 also report
`Train AUC` and the `Gap = Train − Test`. Parametric models (LR-Ridge,
LDA, NB) sit within ~0.01 of their test AUC on every tier — they cannot
memorise 24 k rows with so few effective parameters, so their CV scores
are trustworthy. Non-parametric models show a larger in-bag fit
(Random Forest and KNN approach Train AUC ≈ 1.0 by construction), but
their `CV ROC` and `Test AUC` still agree tightly — the out-of-sample
numbers reported above are not inflated by leakage. In other words,
the H1 gradient is measured on generalisation performance, not on any
model's ability to memorise the training set.

**Threshold-0.5 quality (Precision / Sensitivity / Specificity).** The
§6.1.1 table complements the AUC ranking with operating-point numbers.
Two findings only visible once Precision is reported:

- **SVM-RBF on Lexical** is not only the AUC winner but also the only
  model that is simultaneously high on all three (Prec ≈ Sens ≈ Spec ≈
  0.98). Every other model on Lexical trades one for another.
- **Naive Bayes mirrors itself across tiers.** On Lexical it over-calls
  phish (Sens 0.999, Prec 0.618) — a high-recall / low-precision flood.
  On Behavior the same independence violation flips the other way (Sens
  0.732, Prec 0.995) — almost never calls phish, but when it does, it
  is right. Both are symptoms of the same broken independence
  assumption; the direction of the miscalibration is tier-dependent.

This is the part AUC hides: two models with AUC ≈ 0.99 on Behavior can
still have fundamentally different operating-point behaviour, and the
Precision/Sensitivity decomposition is what makes that visible.

**Practical takeaway.** A URL-only first-line phishing detector gains a
lot from a non-parametric model (SVM-RBF or KNN gain ~0.10 AUC over
LR / NB on Lexical, and SVM additionally keeps Precision and Recall
balanced at ~0.98). Once the system can afford to fetch the page and
compute Behavior-level features, the AUC choice of model becomes almost
irrelevant and simpler, more explainable models (LR, LDA) are good
enough — though the operating-point choice (Precision-first vs.
Recall-first) should still be driven by the cost of a false block vs.
a missed phish. That's exactly the argument for putting a fast
non-parametric URL-only filter *in front of* a slower page-level
analyser, rather than relying on one single model at either extreme.
