---
title: "Phishing Detection — Scenario 1: Cross-family Feature Informativeness"
author: "K-o-m-A"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(caret)
library(glmnet)
library(MASS, exclude = "select")
library(klaR)
library(randomForest)
library(kernlab)
library(pROC)
library(knitr)
library(ggrepel)

set.seed(2026)

S2_ARTIFACTS <- "scenario_2/artifacts"   # imported — preprocessing + shared state
S1_ARTIFACTS <- "scenario_1/artifacts"
dir.create(S1_ARTIFACTS, recursive = TRUE, showWarnings = FALSE)
```

---

# 1. Scenario & Hypothesis

Scenario 1 from the assignment asks us to compare **three statistical or
machine-learning methods** across **two feature-space partitioning
approaches**, identify which features remain informative across **both
method families**, and discuss the implications for explainability and
the benefits of each approach.

We reuse the model line-up from Scenario 2 — three parametric methods
(LR-ridge, LDA, Naive Bayes; hyperplane-partitioning) and three
non-parametric methods (Random Forest, SVM-RBF, KNN;
axis-aligned / kernel partitioning). The two feature-space partitioning
approaches are the **Lexical** (13 features, URL-string only) and
**FullLite** (34 features, Full minus six near-leaky page-count
signals) tiers defined in `scenario_2.rmd` §1.

**H2 (Scenario 1):** A compact "core set" of features appears in the
top-k permutation importance of **both** method families on both tiers.
Features ranked highly by only one family reveal that family's structural
bias: non-parametric-only features exploit non-linear thresholds or
interactions that a linear decision boundary cannot capture, while
parametric-only features contribute linearly and additively — invisible
to axis-aligned splits unless explicitly thresholded. The core set is
the method-agnostic feature footprint a deployment-grade scanner should
rely on.

Why two tiers rather than one: Lexical forces every model to work from
weak, collinear URL counts; FullLite is the realistic "many-features"
regime. If the core set is *stable* between the two tiers, method-agnostic
informativeness is a property of the feature itself, not the feature pool.

---

# 2. Setup — Load cached preprocessing from Scenario 2

`scenario_2.rmd` §3.4 writes a `preprocessing.rds` artifact containing the
same stratified sub-sample, train/test split, CV fold indices and tier
definitions that Scenario 2 uses. Loading it here guarantees the two
notebooks operate on identical rows and columns.

```{r load-preprocessing}
pp_path <- file.path(S2_ARTIFACTS, "preprocessing.rds")
if (!file.exists(pp_path)) {
  stop("Preprocessing artifacts not found at ", pp_path,
       "\nRun scenario_2.rmd first to generate them.")
}
pp <- readRDS(pp_path)
train_std <- pp$train_std; test_std <- pp$test_std
train_raw <- pp$train_raw; test_raw <- pp$test_raw
fold_idx  <- pp$fold_idx
tiers     <- pp$tiers
family_map <- pp$family_map
binary_features     <- pp$binary_features
continuous_features <- pp$continuous_features
rm(pp)

cat("Loaded train:", nrow(train_std), "rows, test:", nrow(test_std), "rows\n")
cat("Tiers available: ", paste(names(tiers), collapse = ", "), "\n")
```

---

# 3. Selected feature-space partitioning approaches

```{r select-tiers}
S1_TIERS <- c("Lexical", "FullLite")
for (t in S1_TIERS) {
  cat(sprintf("  %-10s — %d features\n", t, length(tiers[[t]])))
}
```

`Lexical` — URL-only features (no DNS lookup, no page fetch). Weak
per-feature signal (lowest median SMD in EDA §4.2.2), dense collinearity
(URL-length cluster with VIF > 500).

`FullLite` — Full minus `LineOfCode`, `NoOfExternalRef`, `NoOfImage`,
`NoOfSelfRef`, `NoOfJS`, `NoOfCSS` (each with univariate AUC ≥ 0.95).
Keeps > 30 features but avoids one-feature-solves-the-task saturation.

We deliberately omit `Trust`, `Behavior` and `Full` from Scenario 1: Full
is AUC-saturated, and Trust / Behavior alone add little to the
cross-family comparison beyond what FullLite captures.

---

# 4. Model fitting (six models × two tiers)

Same six methods and hyperparameters as Scenario 2 — LR-ridge (glmnet,
α = 0, λ = 0.01), LDA (no tuning), NB (Gaussian+kernel),
RF (500 trees, mtry = √p), SVM-RBF (C = 1, σ = 0.1), KNN (k = 25, with
micro-jitter on the Lexical tier to break distance ties). Fits are
cached to `scenario_1/artifacts/fit_<model>_<tier>.rds`.

```{r model-spec}
model_specs <- list(
  LR_Ridge = list(
    family = "parametric", method = "glmnet",
    recipe = "std", jitter = FALSE,
    tuneGrid = data.frame(alpha = 0, lambda = 0.01)),
  LDA = list(
    family = "parametric", method = "lda",
    recipe = "std", jitter = FALSE,
    tuneGrid = NULL),
  NaiveBayes = list(
    family = "parametric", method = "nb",
    recipe = "std", jitter = FALSE,
    tuneGrid = data.frame(fL = 1, usekernel = TRUE, adjust = 1)),
  RandomForest = list(
    family = "non-parametric", method = "rf",
    recipe = "raw", jitter = FALSE,
    tuneGrid = NULL),
  SVM_RBF = list(
    family = "non-parametric", method = "svmRadial",
    recipe = "std", jitter = FALSE,
    tuneGrid = data.frame(C = 1, sigma = 0.1)),
  KNN = list(
    family = "non-parametric", method = "knn",
    recipe = "std", jitter = TRUE,
    tuneGrid = data.frame(k = 25))
)
```

```{r fit-utility-s1}
jitter_features <- function(d, cols, sd = 1e-3) {
  for (c in cols) d[[c]] <- d[[c]] + rnorm(nrow(d), 0, sd)
  d
}

fit_model_tier <- function(model_name, spec, tier_name) {
  path <- file.path(S1_ARTIFACTS,
                    sprintf("fit_%s_%s.rds", model_name, tier_name))
  if (file.exists(path)) return(readRDS(path))

  feats <- tiers[[tier_name]]
  if (spec$recipe == "std") {
    tr <- train_std; te <- test_std
  } else {
    tr <- train_raw; te <- test_raw
  }
  if (spec$jitter) {
    tr <- jitter_features(tr, feats)
    te <- jitter_features(te, feats)
  }
  tr <- tr[, c("label", feats)]
  te <- te[, c("label", feats)]

  ctrl <- trainControl(method = "cv", index = fold_idx,
                       classProbs = TRUE,
                       summaryFunction = twoClassSummary,
                       savePredictions = "final")

  tuneGrid   <- spec$tuneGrid
  extra_args <- list()
  if (spec$method == "rf") {
    mtry       <- max(1, floor(sqrt(length(feats))))
    tuneGrid   <- data.frame(mtry = mtry)
    extra_args$ntree <- 500
  }

  call_args <- c(list(label ~ ., data = tr,
                      method = spec$method, trControl = ctrl,
                      metric = "ROC", tuneGrid = tuneGrid),
                 extra_args)

  t0  <- Sys.time()
  fit <- if (spec$method == "nb")
           suppressWarnings(do.call(caret::train, call_args))
         else do.call(caret::train, call_args)
  elapsed <- as.numeric(difftime(Sys.time(), t0, units = "secs"))

  res <- list(fit = fit,
              test_X = te[, feats, drop = FALSE],
              test_y = te$label,
              features = feats,
              train_secs = elapsed,
              family = spec$family,
              model = model_name,
              tier = tier_name)
  saveRDS(res, path)
  res
}
```

```{r fit-all, message=FALSE, warning=FALSE}
fits <- list()
for (model_name in names(model_specs)) {
  for (tier_name in S1_TIERS) {
    key <- paste(model_name, tier_name, sep = "__")
    t0  <- Sys.time()
    fits[[key]] <- fit_model_tier(model_name,
                                  model_specs[[model_name]],
                                  tier_name)
    cat(sprintf("  %-15s [%-9s]  %.1fs\n",
                model_name, tier_name,
                as.numeric(difftime(Sys.time(), t0, units = "secs"))))
  }
}
```

---

# 5. Permutation Importance

Native importance metrics across the six models are *not* mutually
comparable — LR gives signed coefficients, RF gives Gini drops, SVM-RBF
has no direct per-feature importance, and NB/KNN have none at all. We
therefore use **permutation importance on the held-out test set** as the
single unified metric: for each feature `f`, importance is the mean AUC
drop when `f`'s values are randomly shuffled across the test set,
averaged over `n_perm = 5` permutations.

```{r perm-importance-fn}
compute_perm_imp <- function(res, n_perm = 5) {
  fit   <- res$fit
  X     <- res$test_X
  y     <- res$test_y
  feats <- res$features

  auc_of <- function(X_use) {
    prob <- suppressWarnings(
      predict(fit, X_use, type = "prob")[, "Phishing"])
    as.numeric(pROC::roc(y, prob,
                         levels = c("Legitimate", "Phishing"),
                         direction = "<", quiet = TRUE)$auc)
  }

  baseline <- auc_of(X)
  imp <- sapply(feats, function(f) {
    drops <- replicate(n_perm, {
      Xp <- X
      Xp[[f]] <- sample(Xp[[f]])
      baseline - auc_of(Xp)
    })
    mean(drops)
  })
  list(baseline = baseline, importance = imp)
}
```

```{r perm-importance-run, message=FALSE, warning=FALSE}
imp_path <- file.path(S1_ARTIFACTS, "permutation_importance.rds")
if (file.exists(imp_path)) {
  perm_results <- readRDS(imp_path)
  cat("Loaded cached permutation importance.\n")
} else {
  perm_results <- list()
  for (key in names(fits)) {
    t0 <- Sys.time()
    perm_results[[key]] <- compute_perm_imp(fits[[key]], n_perm = 5)
    cat(sprintf("  perm-imp %-30s %.1fs\n", key,
                as.numeric(difftime(Sys.time(), t0, units = "secs"))))
  }
  saveRDS(perm_results, imp_path)
}
```

Long-format importance table:

```{r imp-long}
imp_long <- map_dfr(names(perm_results), function(key) {
  parts <- strsplit(key, "__")[[1]]
  model_name <- parts[1]; tier_name <- parts[2]
  spec <- model_specs[[model_name]]
  tibble(
    model      = model_name,
    family     = spec$family,
    tier       = tier_name,
    feature    = names(perm_results[[key]]$importance),
    importance = perm_results[[key]]$importance
  )
})

imp_long %>%
  group_by(model, tier) %>%
  summarise(max_imp = max(importance),
            mean_imp = mean(importance), .groups = "drop") %>%
  kable(digits = 4,
        caption = "Per-model sanity check: maximum and mean importance per tier")
```

---

# 6. Cross-family Feature Informativeness

## 6.1 Per-feature family-averaged importance

For each `(tier, feature)` pair we compute the mean permutation
importance across the three models within each family. The resulting
`(imp_param, imp_nonparam)` tuple places the feature in a 2D plane where
the diagonal represents features informative in both families.

```{r cross-family-mean}
imp_family <- imp_long %>%
  group_by(tier, feature, family) %>%
  summarise(mean_imp = mean(importance), .groups = "drop") %>%
  pivot_wider(names_from = family, values_from = mean_imp) %>%
  rename(imp_param    = parametric,
         imp_nonparam = `non-parametric`)

imp_family %>%
  arrange(tier, desc(imp_nonparam + imp_param)) %>%
  head(12) %>%
  kable(digits = 4,
        caption = "Top features by total cross-family importance")
```

## 6.2 Cross-family importance scatter

```{r scatter-plot, fig.width=11, fig.height=5}
p <- imp_family %>%
  ggplot(aes(imp_param, imp_nonparam, label = feature)) +
  geom_abline(slope = 1, intercept = 0,
              colour = "grey65", linetype = "dashed") +
  geom_hline(yintercept = 0, colour = "grey85") +
  geom_vline(xintercept = 0, colour = "grey85") +
  geom_point(size = 2.4, alpha = 0.75, colour = "steelblue") +
  ggrepel::geom_text_repel(size = 3, max.overlaps = 25,
                           min.segment.length = 0.2) +
  facet_wrap(~ tier, scales = "free") +
  labs(
    title = "Cross-family permutation importance",
    subtitle = "On-diagonal: informative in both families (core set). Off-diagonal: family-specific.",
    x = "Mean importance — parametric family (LR-ridge, LDA, NB)",
    y = "Mean importance — non-parametric family (RF, SVM-RBF, KNN)"
  ) +
  theme_minimal(base_size = 11)

ggsave(file.path(S1_ARTIFACTS, "cross_family_scatter.png"), p,
       width = 11, height = 5, dpi = 150)
p
```

## 6.3 Jaccard top-k overlap

```{r jaccard}
jaccard_topk <- function(df_tier, k) {
  top_p  <- df_tier %>% arrange(desc(imp_param))    %>%
            slice_head(n = k) %>% pull(feature)
  top_np <- df_tier %>% arrange(desc(imp_nonparam)) %>%
            slice_head(n = k) %>% pull(feature)
  inter  <- intersect(top_p, top_np)
  unio   <- union(top_p, top_np)
  list(k = k,
       jaccard = length(inter) / length(unio),
       core = inter,
       only_param    = setdiff(top_p,  top_np),
       only_nonparam = setdiff(top_np, top_p))
}

jaccard_tbl <- map_dfr(S1_TIERS, function(tier_name) {
  df_tier <- imp_family %>% filter(tier == tier_name)
  ks <- c(3, 5, 10)
  ks <- ks[ks <= nrow(df_tier)]
  map_dfr(ks, function(k) {
    j <- jaccard_topk(df_tier, k)
    tibble(tier = tier_name, k = k,
           jaccard = round(j$jaccard, 3),
           core_size = length(j$core),
           param_only_size = length(j$only_param),
           nonparam_only_size = length(j$only_nonparam))
  })
})

jaccard_tbl %>% kable(caption = "Jaccard top-k overlap between families")
```

## 6.4 Concrete core / family-specific partition

```{r core-specific}
show_partition <- function(tier_name, k) {
  df_tier <- imp_family %>% filter(tier == tier_name)
  k <- min(k, nrow(df_tier))
  j <- jaccard_topk(df_tier, k)
  cat(sprintf("\n── %s tier (top-%d per family) ──\n", tier_name, k))
  cat("  Core (both families):     ",
      if (length(j$core) == 0) "<none>" else paste(j$core, collapse = ", "),
      "\n")
  cat("  Parametric-only top-k:    ",
      if (length(j$only_param) == 0) "<none>"
      else paste(j$only_param, collapse = ", "),
      "\n")
  cat("  Non-parametric-only top-k:",
      if (length(j$only_nonparam) == 0) "<none>"
      else paste(j$only_nonparam, collapse = ", "),
      "\n")
}
show_partition("Lexical",  5)
show_partition("FullLite", 10)
```

## 6.5 Tier-stability of the core set

A feature that is in the core on *both* Lexical and FullLite is
method-agnostic **and** pool-agnostic — the strongest possible evidence
for H2.

```{r tier-stability}
core_lex  <- jaccard_topk(imp_family %>% filter(tier == "Lexical"),  5)$core
core_full <- jaccard_topk(imp_family %>% filter(tier == "FullLite"), 10)$core

stable_core <- intersect(core_lex, core_full)
cat("Core features present in BOTH Lexical (top-5) and FullLite (top-10):\n")
cat("  ", if (length(stable_core) == 0) "<none>"
       else paste(stable_core, collapse = ", "), "\n")
```

---

# 7. Explainability & Benefits of Each Approach

| Aspect | Hyperplane (parametric: LR, LDA, NB) | Axis-aligned / kernel (non-parametric: RF, SVM-RBF, KNN) |
|--------|--------------------------------------|----------------------------------------------------------|
| Coefficient / importance semantics | **Signed magnitude.** LR coefficients give direction (raises or lowers phishing probability) and can be standardised for direct feature ranking. LDA weights are interpretable as discriminant loadings. | **Magnitude-only** from native importance (Gini, distance). Permutation importance unifies magnitude across methods but still carries no direction. SHAP recovers direction at high compute cost. |
| Audit trail for a blocked URL | Additive scorecard: *"blocked because URLLength=87 contributes +0.42, NoOfEqualsInURL=4 contributes +0.31, …"* | A single CART tree offers a readable decision path; RF / SVM-RBF ensembles are only interpretable in aggregate via importances. |
| Inference cost | One dot product per prediction (sub-millisecond on 40 features). | RF: 500 tree traversals. SVM-RBF: kernel evaluation against all support vectors. KNN: distance to every training point. 1–3 orders of magnitude slower. |
| Statistical inference | Wald standard errors, likelihood-ratio tests, confidence intervals available in closed form. | Bootstrap or permutation-based only. No closed-form CIs. |
| Non-linear interactions | Require manual feature engineering (polynomial, interaction terms). | Captured automatically via tree splits or RBF kernel warping. |
| Scale / skew sensitivity | Requires `log1p` + center-scale (EDA flagged 19/22 skewed features). | Trees are invariant to monotone transforms; SVM-RBF and KNN still need scale-normalisation. |
| Extrapolation beyond training range | Linear extrapolation (can be wrong but predictable). | RF returns flat prediction; KNN falls back to the closest training point. |

**Practical benefit profile:**

- **Parametric models** fit first when inference cost, regulatory
  auditability, or small-sample confidence intervals are binding. A
  URL-only filter at the SMTP gateway that must decide in a few
  milliseconds cannot afford a 500-tree traversal per message.
- **Non-parametric models** fit first when non-linear structure is the
  dominant signal and latency / auditability are not binding. Scenario 2
  evidence: the Lexical tier non-parametric advantage is ≈ +0.10 AUC,
  precisely because URL counts interact non-linearly.
- **Core features** (informative in both families) are the safe
  deployment subset — whichever classifier ends up in production, they
  will carry weight. **Family-specific features** expose the modelling
  assumption, not a universal signal.

---

# 8. Conclusion — H2 (Scenario 1) Evidence

Concrete findings are produced by the code in Sections 6.3–6.5. The
notebook pattern is:

1. **Core set on Lexical.** The top-5 features ranked by both families
   (reported in §6.4) are the robust URL-string signals a first-line
   scanner should rely on regardless of model family. Small overlap
   (Jaccard ≈ 0.3–0.5 at k = 5) is expected here because Lexical
   features are weak and collinear; the families disagree about which
   representative of the URL-length cluster to prefer.
2. **Family-specific features on Lexical.** Non-parametric-only entries
   (see §6.4) are the features that depend on thresholds / interactions
   — invisible to linear additivity. Parametric-only entries are the
   features with a clean linear-scale signal that tree splits fail to
   isolate on weak-signal data.
3. **FullLite expansion.** Jaccard top-10 overlap is expected to rise
   on FullLite because the Trust tier's binary flags (`IsHTTPS`, `Bank`,
   `Pay`, `Crypto`) carry monotonic signal that both families pick up
   directly. This tier-level shift — from low agreement (Lexical) to
   high agreement (FullLite) — is itself part of H2's claim: core-set
   size scales with signal strength.
4. **Stable core (§6.5).** Features present in the core of **both**
   tiers form the hard-core method-agnostic subset. These are the
   features a production scanner can commit to regardless of which
   classifier or feature pool it ends up using.

**Practical deployment takeaway:** the Lexical-tier core set is the
minimum-viable feature footprint for a fast first-line URL-only filter;
the FullLite core set is the safe, explainability-first deployment
subset once the scanner can afford page-level features. Features
informative in only one family should not be the *only* rationale for a
decision, because their relevance is a property of the method, not the
URL.
