---
title: "Scenario 2 — Parametric vs. Non-parametric Classification"
author: "K-o-m-A"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r libraries, message=FALSE, warning=FALSE}
library(MASS)
library(tidyverse)
library(caret)
library(glmnet)
library(class)
library(e1071)
library(randomForest)
library(pROC)
library(car)
library(klaR)
```

---

# 1. Data Loading & Preparation

We reload the dataset and apply the same cleaning as in the EDA notebook.

```{r load-data}
raw <- read_csv(
  "PhiUSIIL_Phishing_URL_Dataset.csv",
  show_col_types = FALSE
)
names(raw) <- map_chr(names(raw), ~ str_replace(.x, "^\ufeff", ""))

ID_COLS <- c("FILENAME", "URL", "Domain", "TLD", "Title")

df <- raw %>%
  mutate(label_bin = factor(label, levels = c(0, 1),
                            labels = c("Legitimate", "Phishing"))) %>%
  select(-all_of(ID_COLS))

feature_names <- setdiff(names(df), c("label", "label_bin"))
cat("Observations:", nrow(df), "| Predictors:", length(feature_names), "\n")

# Feature-family mapping (from EDA) — used to test hypotheses via model outputs
family_map <- tribble(
  ~feature,                    ~family,
  "URLLength",                 "Lexical",
  "DomainLength",              "Lexical",
  "IsDomainIP",                "Lexical",
  "URLSimilarityIndex",        "Trust",
  "CharContinuationRate",      "Lexical",
  "TLDLegitimateProb",         "Trust",
  "URLCharProb",               "Trust",
  "TLDLength",                 "Lexical",
  "NoOfSubDomain",             "Lexical",
  "HasObfuscation",            "Lexical",
  "NoOfObfuscatedChar",        "Lexical",
  "ObfuscationRatio",          "Lexical",
  "NoOfLettersInURL",          "Lexical",
  "LetterRatioInURL",          "Lexical",
  "NoOfDegitsInURL",           "Lexical",
  "DegitRatioInURL",           "Lexical",
  "NoOfEqualsInURL",           "Lexical",
  "NoOfQMarkInURL",            "Lexical",
  "NoOfAmpersandInURL",        "Lexical",
  "NoOfOtherSpecialCharsInURL","Lexical",
  "SpacialCharRatioInURL",     "Lexical",
  "IsHTTPS",                   "Trust",
  "LineOfCode",                "Behavior",
  "LargestLineLength",         "Behavior",
  "HasTitle",                  "Trust",
  "DomainTitleMatchScore",     "Trust",
  "URLTitleMatchScore",        "Trust",
  "HasFavicon",                "Trust",
  "Robots",                    "Trust",
  "IsResponsive",              "Behavior",
  "NoOfURLRedirect",           "Behavior",
  "NoOfSelfRedirect",          "Behavior",
  "HasDescription",            "Behavior",
  "NoOfPopup",                 "Behavior",
  "NoOfiFrame",                "Behavior",
  "HasExternalFormSubmit",     "Behavior",
  "HasSocialNet",              "Behavior",
  "HasSubmitButton",           "Behavior",
  "HasHiddenFields",           "Behavior",
  "HasPasswordField",          "Behavior",
  "Bank",                      "Trust",
  "Pay",                       "Trust",
  "Crypto",                    "Trust",
  "HasCopyrightInfo",          "Behavior",
  "NoOfImage",                 "Behavior",
  "NoOfCSS",                   "Behavior",
  "NoOfJS",                    "Behavior",
  "NoOfSelfRef",               "Behavior",
  "NoOfEmptyRef",              "Behavior",
  "NoOfExternalRef",           "Behavior"
)
```

## 1.1 Subsampling & Train/Test Split

The full dataset (~235k obs) is too easy for all models — even simple linear
classifiers achieve AUC ≈ 1.0, making parametric vs. non-parametric comparison
meaningless. The features were engineered for production-grade phishing detection
and have very high discriminative power (EDA median SMD: Trust 1.005, Behavior
0.595). We stratified-subsample to 2 000 observations so that models have to
work with limited data and genuine differences in flexibility become visible.

```{r train-test-split}
set.seed(42)
SUBSAMPLE_N <- 10000

sub_idx <- createDataPartition(df$label, p = SUBSAMPLE_N / nrow(df), list = FALSE)
df_sub  <- df[sub_idx, ]
cat("Subsampled to", nrow(df_sub), "obs (from", nrow(df), ")\n")
cat("Subsample class balance:\n")
prop.table(table(df_sub$label_bin))

train_idx <- createDataPartition(df_sub$label, p = 0.8, list = FALSE)
train_df  <- df_sub[train_idx, ]
test_df   <- df_sub[-train_idx, ]

cat("\nTraining set:", nrow(train_df), "obs\n")
cat("Test set:    ", nrow(test_df), "obs\n")
```

## 1.2 Preprocessing

EDA showed heavy skewness in count features and severe multicollinearity.
We apply `log(1+x)` to skewed count features (|skewness| > 2), then centre
and scale all predictors. VIF-based filtering of collinear features follows
in Section 1.3.

```{r preprocessing}
# Compute skewness to identify features needing log-transform
skewness_fn <- function(x) {
  x <- x[!is.na(x)]; n <- length(x); m <- mean(x); s <- sd(x)
  (n / ((n-1)*(n-2))) * sum(((x - m) / s)^3)
}

skew_vals <- map_dbl(feature_names, ~ skewness_fn(train_df[[.x]]))
skewed_feats <- feature_names[abs(skew_vals) > 2]
cat("Log-transforming", length(skewed_feats), "skewed features:\n")
cat(str_wrap(paste(skewed_feats, collapse = ", "), width = 80), "\n")

# Apply log1p
log_transform <- function(data, feats) {
  data %>% mutate(across(all_of(feats), log1p))
}
train_df <- log_transform(train_df, skewed_feats)
test_df  <- log_transform(test_df, skewed_feats)

# Centre and scale using training set parameters
preProcValues <- preProcess(train_df[, feature_names], method = c("center", "scale"))
train_scaled <- as.data.frame(predict(preProcValues, train_df))
test_scaled  <- as.data.frame(predict(preProcValues, test_df))

# Prepare matrices
X_train <- train_scaled[, feature_names]
y_train <- train_scaled$label_bin
X_test  <- test_scaled[, feature_names]
y_test  <- test_scaled$label_bin
```

## 1.3 VIF-Based Feature Filtering

EDA revealed extreme multicollinearity among Lexical features (e.g. `URLLength`
VIF > 2000, `NoOfLettersInURL` > 1000). High VIF inflates coefficient variance in
parametric models (LR, LDA), making them unreliable. We iteratively remove the
feature with the highest VIF until all remaining features fall below the threshold
of 5.

```{r vif-filtering, message=FALSE, warning=FALSE}
VIF_THRESHOLD <- 5

# Work on training data only to avoid data leakage
vif_features <- feature_names

repeat {
  lm_vif <- lm(
    label ~ .,
    data = train_scaled[, c("label", vif_features)]
  )
  vif_vals <- vif(lm_vif)
  max_vif  <- max(vif_vals)

  if (max_vif <= VIF_THRESHOLD) break

  worst <- names(which.max(vif_vals))
  cat("Dropping", worst, "(VIF =", round(max_vif, 1), ")\n")
  vif_features <- setdiff(vif_features, worst)
}

cat("\nRemaining features after VIF filtering:", length(vif_features),
    "out of", length(feature_names), "\n")
cat("Dropped features:\n")
cat(paste(setdiff(feature_names, vif_features), collapse = ", "), "\n")
```

## 1.4 Feature Sets by Family

To test H1 we train every model on four feature sets: each family in isolation
plus all families combined. Comparing per-family AUC directly shows which family
carries the most predictive power.

```{r family-feature-sets}
y_train <- train_scaled$label_bin
y_test  <- test_scaled$label_bin

# Split VIF-filtered features by family
trust_feats    <- family_map %>% filter(family == "Trust")    %>% pull(feature)
lexical_feats  <- family_map %>% filter(family == "Lexical")  %>% pull(feature)
behavior_feats <- family_map %>% filter(family == "Behavior") %>% pull(feature)

trust_feats    <- intersect(trust_feats,    vif_features)
lexical_feats  <- intersect(lexical_feats,  vif_features)
behavior_feats <- intersect(behavior_feats, vif_features)

feature_sets <- list(
  All      = vif_features,
  Trust    = trust_feats,
  Lexical  = lexical_feats,
  Behavior = behavior_feats
)

for (nm in names(feature_sets)) {
  cat(nm, ":", length(feature_sets[[nm]]), "features\n")
}
```

---

# 2. Model Comparison

We compare three parametric models (Logistic Regression, LDA, QDA) against three
non-parametric models (Random Forest, KNN, SVM) using 10-fold stratified CV.

To test **H1**, every model is trained on four feature sets: All, Trust-only,
Lexical-only, and Behavior-only. If Trust-only models consistently outperform the
other single-family models, H1 is confirmed at the modelling level.

## 2.1 Cross-Validation Setup

```{r cv-setup}
ctrl <- trainControl(
  method          = "cv",
  number          = 10,
  classProbs      = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)
```

## 2.2 Preparation

```{r fit-prep}
ridge_grid <- expand.grid(alpha = 0, lambda = 10^seq(-4, 1, length = 20))
knn_grid   <- expand.grid(k = c(3, 5, 7, 11, 15))

# Prepare train/test matrices per feature set (as data.frame to avoid
# "Setting row names on a tibble is deprecated" warnings from caret)
X <- list(); X_te <- list()
for (nm in names(feature_sets)) {
  X[[nm]]    <- as.data.frame(train_scaled[, feature_sets[[nm]]])
  X_te[[nm]] <- as.data.frame(test_scaled[, feature_sets[[nm]]])
}

# QDA (via RDA) handles regularisation internally — no extra preprocessing needed

# Storage for all fits
fits <- list(All = list(), Trust = list(), Lexical = list(), Behavior = list())
```

## 2.3 Logistic Regression (Ridge)

Ridge penalisation (alpha = 0) stabilises coefficients while keeping the model
fully parametric and interpretable.

```{r model-lr, message=FALSE, warning=FALSE}
for (nm in names(feature_sets)) {
  set.seed(42)
  fits[[nm]][["LR"]] <- train(
    x = X[[nm]], y = y_train, method = "glmnet", family = "binomial",
    metric = "ROC", trControl = ctrl, tuneGrid = ridge_grid
  )
  cat("LR ", nm, "CV AUC:", round(max(fits[[nm]][["LR"]]$results$ROC), 4), "\n")
}
```

## 2.4 Linear Discriminant Analysis (LDA)

LDA assumes multivariate normal class distributions with shared covariance.

```{r model-lda, message=FALSE, warning=FALSE}
for (nm in names(feature_sets)) {
  set.seed(42)
  fits[[nm]][["LDA"]] <- train(
    x = X[[nm]], y = y_train, method = "lda",
    metric = "ROC", trControl = ctrl
  )
  cat("LDA", nm, "CV AUC:", round(max(fits[[nm]][["LDA"]]$results$ROC), 4), "\n")
}
```

## 2.5 Quadratic Discriminant Analysis (QDA)

QDA allows per-class covariance matrices. Many features in this dataset are
binary (0/1), which causes singular within-class covariance matrices. We use
Regularised Discriminant Analysis (`method = "rda"` from `klaR`), which
interpolates between LDA (shared covariance) and QDA (per-class covariance)
via tuning parameters `gamma` (shrinkage toward identity) and `lambda`
(shrinkage toward shared covariance). This avoids rank-deficiency errors
while preserving QDA's flexibility where the data supports it.

```{r model-qda, message=FALSE, warning=FALSE}
rda_grid <- expand.grid(gamma = c(0, 0.1, 0.5), lambda = c(0, 0.1, 0.5, 1))
for (nm in names(feature_sets)) {
  set.seed(42)
  fits[[nm]][["QDA"]] <- train(
    x = X[[nm]], y = y_train, method = "rda",
    metric = "ROC", trControl = ctrl,
    tuneGrid = rda_grid
  )
  cat("QDA(rda)", nm, "CV AUC:", round(max(fits[[nm]][["QDA"]]$results$ROC), 4),
      " [", ncol(X[[nm]]), "feats ]\n")
}
```

## 2.6 Random Forest

```{r model-rf, message=FALSE, warning=FALSE}
for (nm in names(feature_sets)) {
  max_mtry <- ncol(X[[nm]])
  rf_grid  <- expand.grid(mtry = unique(pmin(c(3, 5, 7, 10), max_mtry)))
  set.seed(42)
  fits[[nm]][["RF"]] <- train(
    x = X[[nm]], y = y_train, method = "rf", metric = "ROC",
    trControl = ctrl, tuneGrid = rf_grid, ntree = 500
  )
  cat("RF ", nm, "CV AUC:", round(max(fits[[nm]][["RF"]]$results$ROC), 4), "\n")
}
```

## 2.7 K-Nearest Neighbours (KNN)

```{r model-knn, message=FALSE, warning=FALSE}
for (nm in names(feature_sets)) {
  set.seed(42)
  fits[[nm]][["KNN"]] <- train(
    x = X[[nm]], y = y_train, method = "knn", metric = "ROC",
    trControl = ctrl, tuneGrid = knn_grid
  )
  cat("KNN", nm, "CV AUC:", round(max(fits[[nm]][["KNN"]]$results$ROC), 4), "\n")
}
```

## 2.8 Support Vector Machine (RBF)

```{r model-svm, message=FALSE, warning=FALSE}
for (nm in names(feature_sets)) {
  set.seed(42)
  fits[[nm]][["SVM"]] <- train(
    x = X[[nm]], y = y_train, method = "svmRadial", metric = "ROC",
    trControl = ctrl, tuneLength = 5
  )
  cat("SVM", nm, "CV AUC:", round(max(fits[[nm]][["SVM"]]$results$ROC), 4), "\n")
}
```

---

# 3. Cross-Validation Results

```{r cv-summary-table}
# Collect CV AUC from all 24 fits into a single table
cv_summary <- map_dfr(names(fits), function(fs) {
  map_dfr(names(fits[[fs]]), function(mn) {
    tibble(
      FeatureSet = fs,
      Model      = mn,
      Type       = ifelse(mn %in% c("LR","LDA","QDA"), "Parametric", "Non-parametric"),
      CV_AUC     = round(max(fits[[fs]][[mn]]$results$ROC), 4)
    )
  })
})

cv_summary %>%
  pivot_wider(names_from = FeatureSet, values_from = CV_AUC) %>%
  arrange(Type, Model) %>%
  print()
```

```{r cv-grouped-barplot, fig.height=6, fig.width=10}
ggplot(cv_summary, aes(x = Model, y = CV_AUC, fill = FeatureSet)) +
  geom_col(position = "dodge") +
  facet_wrap(~ Type, scales = "free_x") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "CV AUC by model and feature family",
       subtitle = "H1: Trust-only should outperform Lexical-only and Behavior-only",
       x = NULL, y = "10-fold CV AUC", fill = "Feature set") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```

```{r cv-boxplots, fig.height=8, fig.width=10}
# Per-family CV boxplots
par(mfrow = c(2, 2))
for (fs in names(fits)) {
  rs <- resamples(fits[[fs]])
  bwplot(rs, metric = "ROC", main = paste("CV AUC —", fs))
}
```

**Interpretation:** If Trust-only models consistently achieve higher AUC than
Lexical-only and Behavior-only across all six models, H1 is confirmed: Trust
features carry the most discriminative signal. The gap between families also
reveals how much each model type benefits from feature flexibility.

---

# 4. Test Set Evaluation

```{r test-evaluation}
test_results <- map_dfr(names(fits), function(fs) {
  fs_feats <- feature_sets[[fs]]
  X_te     <- test_scaled[, fs_feats]

  map_dfr(names(fits[[fs]]), function(mn) {
    mod  <- fits[[fs]][[mn]]
    x_te <- X_te  # PCA for QDA is handled internally by caret's predict
    pred <- predict(mod, x_te)
    prob <- predict(mod, x_te, type = "prob")
    cm   <- confusionMatrix(pred, y_test, positive = "Phishing")
    roc_obj <- roc(y_test, prob$Phishing,
                   levels = c("Legitimate", "Phishing"), quiet = TRUE)
    tibble(
      Model      = mn,
      FeatureSet = fs,
      Type       = ifelse(mn %in% c("LR","LDA","QDA"), "Parametric", "Non-parametric"),
      Accuracy   = round(cm$overall["Accuracy"], 4),
      Sensitivity = round(cm$byClass["Sensitivity"], 4),
      Specificity = round(cm$byClass["Specificity"], 4),
      F1         = round(cm$byClass["F1"], 4),
      AUC        = round(auc(roc_obj), 4)
    )
  })
})

test_results %>%
  select(Model, Type, FeatureSet, AUC, Accuracy, F1) %>%
  pivot_wider(names_from = FeatureSet, values_from = c(AUC, Accuracy, F1),
              names_glue = "{FeatureSet}_{.value}") %>%
  select(Model, Type, starts_with("All"), starts_with("Trust"),
         starts_with("Lexical"), starts_with("Behavior")) %>%
  print(n = 12)
```

```{r test-family-comparison, fig.height=6, fig.width=10}
ggplot(test_results, aes(x = Model, y = AUC, fill = FeatureSet)) +
  geom_col(position = "dodge") +
  facet_wrap(~ Type, scales = "free_x") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Test AUC by model and feature family",
       subtitle = "H1: Trust-only > Lexical-only, Behavior-only across all models",
       x = NULL, y = "Test AUC", fill = "Feature set") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```

```{r test-roc-per-family, fig.height=8, fig.width=10}
model_cols <- c(LR = "#1f77b4", LDA = "#4a90d9", QDA = "#7ec8e3",
                RF = "#d62728", KNN = "#e57373", SVM = "#ff9896")

par(mfrow = c(2, 2))
for (fs in names(fits)) {
  fs_feats <- feature_sets[[fs]]
  X_te     <- test_scaled[, fs_feats]

  rocs <- map(names(fits[[fs]]), function(mn) {
    roc(y_test, predict(fits[[fs]][[mn]], X_te, type = "prob")$Phishing,
        levels = c("Legitimate", "Phishing"), quiet = TRUE)
  })
  names(rocs) <- names(fits[[fs]])

  plot(rocs[[1]], col = model_cols[1], lwd = 2, main = paste("ROC —", fs))
  for (i in 2:length(rocs)) lines(rocs[[i]], col = model_cols[i], lwd = 2)
  legend("bottomright",
         legend = paste0(names(rocs), " (", sapply(rocs, function(r) round(auc(r), 3)), ")"),
         col = model_cols, lwd = 2, cex = 0.6)
}
```
---

# 5. Hypothesis Verification — H1

EDA formulated **H1 — Trust signals are the primary discriminators of phishing
URLs**. We test this with two complementary approaches:

1. **Family AUC comparison** — does Trust-only outperform Lexical-only and
   Behavior-only across all models?
2. **Feature importance** — do Trust features rank highest in LR and RF when all
   features are used together?

## 5.1 Family AUC Summary

```{r h1-family-auc, fig.height=5, fig.width=8}
family_auc <- test_results %>%
  filter(FeatureSet != "All") %>%
  group_by(FeatureSet, Type) %>%
  summarise(mean_AUC = mean(AUC), .groups = "drop")

ggplot(family_auc, aes(x = FeatureSet, y = mean_AUC, fill = Type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Parametric" = "steelblue", "Non-parametric" = "firebrick")) +
  labs(title = "Mean test AUC per feature family",
       subtitle = "H1: Trust > Lexical, Behavior for both model types",
       x = NULL, y = "Mean test AUC") +
  theme_minimal(base_size = 12)

family_auc %>% mutate(mean_AUC = round(mean_AUC, 4)) %>% print()
```

## 5.2 Feature Importance: LR vs. RF (All features)

```{r feature-importance, fig.height=8, fig.width=10}
# LR (ridge): absolute standardised coefficients at best lambda
fit_lr_all  <- fits[["All"]][["LR"]]
lr_coef_mat <- coef(fit_lr_all$finalModel, s = fit_lr_all$bestTune$lambda)
lr_coefs    <- lr_coef_mat[-1, 1]
lr_imp <- tibble(
  feature    = names(lr_coefs),
  importance = abs(lr_coefs)
) %>%
  mutate(importance = importance / max(importance)) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  mutate(model = "LR (|coef|, normalised)")

# RF: variable importance (MeanDecreaseGini)
fit_rf_all <- fits[["All"]][["RF"]]
rf_imp_raw <- varImp(fit_rf_all)$importance
rf_imp <- tibble(
  feature    = rownames(rf_imp_raw),
  importance = rf_imp_raw$Overall / max(rf_imp_raw$Overall)
) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  mutate(model = "RF (Gini, normalised)")

bind_rows(lr_imp, rf_imp) %>%
  ggplot(aes(x = reorder(feature, importance), y = importance, fill = family)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ model, scales = "free_x") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Feature importance — parametric (LR) vs. non-parametric (RF)",
       subtitle = "H1: Trust features should dominate in both model types",
       x = NULL, y = "Normalised importance", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")
```

## 5.3 H1 Conclusion

```{r h1-conclusion}
cat("=== Top 10 features — LR ===\n")
lr_imp %>% arrange(desc(importance)) %>% slice_head(n = 10) %>%
  select(feature, family, importance) %>%
  mutate(importance = round(importance, 3)) %>% print()

cat("\n=== Top 10 features — RF ===\n")
rf_imp %>% arrange(desc(importance)) %>% slice_head(n = 10) %>%
  select(feature, family, importance) %>%
  mutate(importance = round(importance, 3)) %>% print()

# Family-level AUC ranking
cat("\n=== H1 — Mean test AUC per family (across all 6 models) ===\n")
test_results %>%
  filter(FeatureSet != "All") %>%
  group_by(FeatureSet) %>%
  summarise(mean_AUC = round(mean(AUC), 4), .groups = "drop") %>%
  arrange(desc(mean_AUC)) %>%
  print()
```

**H1 — Trust signals are the primary discriminators of phishing URLs:**

Two lines of evidence:

- **Family AUC comparison:** If Trust-only models consistently achieve higher AUC
  than Lexical-only and Behavior-only across both parametric and non-parametric
  methods, Trust features carry the most predictive power per-family.
- **Feature importance:** If Trust features rank highest in both LR coefficients
  and RF Gini importance when all families are used together, they dominate even
  in the presence of other signal sources.

The question of whether Lexical features lose marginal value when combined with
Trust features (EDA hypothesis H2) is deferred to Scenario 3 (feature selection).

---

# 6. Explainability & Reproducibility (Scenario 2 requirement)

```{r explainability-table}
explain_df <- tribble(
  ~Model, ~Type, ~Explainability, ~Reproducibility,
  "LR (ridge)",  "Parametric",
    "High — ridge coefficients quantify feature effects on log-odds; sign and relative magnitude are interpretable, though shrunk toward zero",
    "Perfect — convex optimisation with unique solution for fixed lambda",
  "LDA", "Parametric",
    "High — discriminant coefficients show which features drive class separation; linear boundary is easy to visualise",
    "Perfect — closed-form solution, fully deterministic",
  "QDA (RDA)", "Parametric",
    "Medium — regularised discriminant analysis interpolates between LDA and QDA; gamma/lambda parameters control shrinkage, making interpretation harder than LDA but more stable",
    "Perfect — closed-form solution, fully deterministic for fixed hyperparameters",
  "RF",  "Non-parametric",
    "Medium — variable importance ranks features but individual predictions require methods like SHAP; ensemble of 500 trees is not directly readable",
    "Seed-dependent — bootstrap sampling and random feature selection introduce stochasticity; fixed seed ensures reproducibility",
  "KNN", "Non-parametric",
    "Low — predictions depend on local neighbourhood; no global model or feature ranking; hard to explain why a specific URL was flagged",
    "Perfect — deterministic given the same data and k; no randomness in the algorithm",
  "SVM", "Non-parametric",
    "Low — RBF kernel maps features to high-dimensional space; support vectors identify the boundary but individual feature contributions are opaque",
    "Seed-dependent — caret's tuning uses resampling; the SVM optimisation itself is deterministic for fixed hyperparameters"
)

explain_df %>%
  select(Model, Type, Explainability, Reproducibility) %>%
  knitr::kable(format = "pipe",
               caption = "Explainability and reproducibility comparison")
```

**Discussion:**

Parametric models offer a clear advantage in **explainability**: LR coefficients and
LDA discriminant loadings directly map features to predictions, which matters for
phishing detection where domain analysts need to understand *why* a URL was flagged.
All parametric models are also **perfectly reproducible** — they use deterministic,
closed-form or convex optimisation.

Non-parametric models trade interpretability for flexibility. RF captures non-linear
effects and interactions, and its variable importance provides some global insight,
but individual predictions are harder to explain. KNN offers no global model at all.
SVM with RBF kernel is similarly opaque. RF's reproducibility depends on the random
seed; KNN and SVM (with fixed hyperparameters) are deterministic.

The choice depends on the use case: if the system must justify its decisions
(e.g. to a security team), parametric models are preferable. If raw predictive
accuracy is the priority, non-parametric models may be worth the interpretability cost.

---

# 7. Summary

```{r scenario2-summary}
cat("=== Best model per type (All features) ===\n")
test_results %>%
  filter(FeatureSet == "All") %>%
  group_by(Type) %>%
  slice_max(AUC, n = 1) %>%
  select(Model, Type, AUC, F1) %>%
  print()

cat("\n=== H1: Family ranking by mean test AUC ===\n")
test_results %>%
  filter(FeatureSet != "All") %>%
  group_by(FeatureSet) %>%
  summarise(mean_AUC = round(mean(AUC), 4), .groups = "drop") %>%
  arrange(desc(mean_AUC)) %>%
  print()

cat("\n=== Parametric vs. Non-parametric gap by family ===\n")
test_results %>%
  group_by(FeatureSet, Type) %>%
  summarise(mean_AUC = round(mean(AUC), 4), .groups = "drop") %>%
  pivot_wider(names_from = Type, values_from = mean_AUC) %>%
  mutate(Gap = `Non-parametric` - Parametric) %>%
  print()
```
