---
title: "Scenario 2 — Parametric vs. Non-parametric Classification"
author: "K-o-m-A"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r libraries, message=FALSE, warning=FALSE}
library(MASS)
library(tidyverse)
library(caret)
library(glmnet)
library(class)
library(e1071)
library(randomForest)
library(pROC)
```

---

# 1. Data Loading & Preparation

We reload the dataset and apply the same cleaning as in the EDA notebook.

```{r load-data}
raw <- read_csv(
  "PhiUSIIL_Phishing_URL_Dataset.csv",
  show_col_types = FALSE
)
names(raw) <- map_chr(names(raw), ~ str_replace(.x, "^\ufeff", ""))

ID_COLS <- c("FILENAME", "URL", "Domain", "TLD", "Title")

df <- raw %>%
  mutate(label_bin = factor(label, levels = c(0, 1),
                            labels = c("Legitimate", "Phishing"))) %>%
  select(-all_of(ID_COLS))

feature_names <- setdiff(names(df), c("label", "label_bin"))
cat("Observations:", nrow(df), "| Predictors:", length(feature_names), "\n")

# Feature-family mapping (from EDA) — used to test hypotheses via model outputs
family_map <- tribble(
  ~feature,                    ~family,
  "URLLength",                 "Lexical",
  "DomainLength",              "Lexical",
  "IsDomainIP",                "Lexical",
  "URLSimilarityIndex",        "Trust",
  "CharContinuationRate",      "Lexical",
  "TLDLegitimateProb",         "Trust",
  "URLCharProb",               "Trust",
  "TLDLength",                 "Lexical",
  "NoOfSubDomain",             "Lexical",
  "HasObfuscation",            "Lexical",
  "NoOfObfuscatedChar",        "Lexical",
  "ObfuscationRatio",          "Lexical",
  "NoOfLettersInURL",          "Lexical",
  "LetterRatioInURL",          "Lexical",
  "NoOfDegitsInURL",           "Lexical",
  "DegitRatioInURL",           "Lexical",
  "NoOfEqualsInURL",           "Lexical",
  "NoOfQMarkInURL",            "Lexical",
  "NoOfAmpersandInURL",        "Lexical",
  "NoOfOtherSpecialCharsInURL","Lexical",
  "SpacialCharRatioInURL",     "Lexical",
  "IsHTTPS",                   "Trust",
  "LineOfCode",                "Behavior",
  "LargestLineLength",         "Behavior",
  "HasTitle",                  "Trust",
  "DomainTitleMatchScore",     "Trust",
  "URLTitleMatchScore",        "Trust",
  "HasFavicon",                "Trust",
  "Robots",                    "Trust",
  "IsResponsive",              "Behavior",
  "NoOfURLRedirect",           "Behavior",
  "NoOfSelfRedirect",          "Behavior",
  "HasDescription",            "Behavior",
  "NoOfPopup",                 "Behavior",
  "NoOfiFrame",                "Behavior",
  "HasExternalFormSubmit",     "Behavior",
  "HasSocialNet",              "Behavior",
  "HasSubmitButton",           "Behavior",
  "HasHiddenFields",           "Behavior",
  "HasPasswordField",          "Behavior",
  "Bank",                      "Trust",
  "Pay",                       "Trust",
  "Crypto",                    "Trust",
  "HasCopyrightInfo",          "Behavior",
  "NoOfImage",                 "Behavior",
  "NoOfCSS",                   "Behavior",
  "NoOfJS",                    "Behavior",
  "NoOfSelfRef",               "Behavior",
  "NoOfEmptyRef",              "Behavior",
  "NoOfExternalRef",           "Behavior"
)
```

## 1.1 Train/Test Split

Stratified 80/20 split. The test set is held out until final evaluation — all
model tuning uses cross-validation on the training set only.

```{r train-test-split}
set.seed(42)
train_idx <- createDataPartition(df$label, p = 0.8, list = FALSE)
train_df  <- df[train_idx, ]
test_df   <- df[-train_idx, ]

cat("Training set:", nrow(train_df), "obs\n")
cat("Test set:    ", nrow(test_df), "obs\n")
cat("Train class balance:\n")
prop.table(table(train_df$label_bin))
```

## 1.2 Preprocessing

EDA showed heavy skewness in count features and severe multicollinearity.
We apply `log(1+x)` to skewed count features (|skewness| > 2), then centre
and scale all predictors.

```{r preprocessing}
# Compute skewness to identify features needing log-transform
skewness_fn <- function(x) {
  x <- x[!is.na(x)]; n <- length(x); m <- mean(x); s <- sd(x)
  (n / ((n-1)*(n-2))) * sum(((x - m) / s)^3)
}

skew_vals <- map_dbl(feature_names, ~ skewness_fn(train_df[[.x]]))
skewed_feats <- feature_names[abs(skew_vals) > 2]
cat("Log-transforming", length(skewed_feats), "skewed features:\n")
cat(str_wrap(paste(skewed_feats, collapse = ", "), width = 80), "\n")

# Apply log1p
log_transform <- function(data, feats) {
  data %>% mutate(across(all_of(feats), log1p))
}
train_df <- log_transform(train_df, skewed_feats)
test_df  <- log_transform(test_df, skewed_feats)

# Centre and scale using training set parameters
preProcValues <- preProcess(train_df[, feature_names], method = c("center", "scale"))
train_scaled <- predict(preProcValues, train_df)
test_scaled  <- predict(preProcValues, test_df)

# Prepare matrices
X_train <- train_scaled[, feature_names]
y_train <- train_scaled$label_bin
X_test  <- test_scaled[, feature_names]
y_test  <- test_scaled$label_bin
```

---

# 2. Model Comparison

We compare three parametric models (Logistic Regression, LDA, QDA) against three
non-parametric models (Random Forest, KNN, SVM) using 10-fold stratified
cross-validation, optimising for AUC.

## 2.1 Cross-Validation Setup

```{r cv-setup}
ctrl <- trainControl(
  method          = "cv",
  number          = 10,
  classProbs      = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)
```

## 2.2 Parametric Models

### Logistic Regression

Logistic regression models the log-odds of phishing as a linear function of the
features. It is the baseline parametric classifier — fully interpretable through
coefficients, but assumes a linear decision boundary and is sensitive to
multicollinearity (which EDA flagged, especially in Lexical features).

```{r model-lr, message=FALSE, warning=FALSE}
set.seed(42)
fit_lr <- train(
  x = X_train, y = y_train,
  method    = "glm",
  family    = "binomial",
  metric    = "ROC",
  trControl = ctrl
)
cat("LR CV AUC:", round(max(fit_lr$results$ROC), 4), "\n")
```

### Linear Discriminant Analysis (LDA)

LDA assumes each class has a multivariate normal distribution with a shared covariance
matrix. It finds the linear combination of features that maximises class separation.
Compared to LR, LDA explicitly models the class-conditional distributions, which can
be an advantage when the normality assumption approximately holds but a disadvantage
with the skewed features we observed.

```{r model-lda, message=FALSE, warning=FALSE}
set.seed(42)
fit_lda <- train(
  x = X_train, y = y_train,
  method    = "lda",
  metric    = "ROC",
  trControl = ctrl
)
cat("LDA CV AUC:", round(max(fit_lda$results$ROC), 4), "\n")
```

### Quadratic Discriminant Analysis (QDA)

QDA relaxes LDA's shared-covariance assumption — each class gets its own covariance
matrix, allowing quadratic (curved) decision boundaries. This is relevant here because
EDA showed unequal class spreads in several features. The cost is more parameters to
estimate, which increases variance.

QDA requires invertible per-class covariance matrices. With severe multicollinearity
in our features (VIF > 2000 for some Lexical pairs), we remove highly correlated
features (|r| > 0.7) before fitting to ensure numerical stability.

```{r model-qda, message=FALSE, warning=FALSE}
cor_train <- cor(X_train)
high_cor  <- findCorrelation(cor_train, cutoff = 0.7)
cat("Removing", length(high_cor), "correlated features for QDA:\n")
cat(names(X_train)[high_cor], sep = ", ")
cat("\n")

X_train_qda <- X_train[, -high_cor]
X_test_qda  <- X_test[, -high_cor]
cat("QDA will use", ncol(X_train_qda), "features\n")

set.seed(42)
fit_qda <- train(
  x = X_train_qda, y = y_train,
  method    = "qda",
  metric    = "ROC",
  trControl = ctrl
)
cat("QDA CV AUC:", round(max(fit_qda$results$ROC), 4), "\n")
```

## 2.3 Non-parametric Models

### Random Forest

Random Forest builds an ensemble of decision trees on bootstrap samples with random
feature subsets at each split. It handles non-linear relationships and interactions
natively, is robust to outliers and multicollinearity, and provides built-in variable
importance. The main tuning parameter is `mtry` — the number of features considered
at each split.

```{r model-rf, message=FALSE, warning=FALSE}
set.seed(42)
fit_rf <- train(
  x = X_train, y = y_train,
  method    = "rf",
  metric    = "ROC",
  trControl = ctrl,
  tuneGrid  = expand.grid(mtry = c(5, 7, 10, 15)),
  ntree     = 500
)
cat("RF best mtry:", fit_rf$bestTune$mtry, "\n")
cat("RF CV AUC:   ", round(max(fit_rf$results$ROC), 4), "\n")
```

### K-Nearest Neighbours (KNN)

KNN classifies each observation by majority vote of its k nearest neighbours in
feature space. It makes no distributional assumptions but is sensitive to the
curse of dimensionality and feature scaling (handled in preprocessing). We tune
k over a range of values.

```{r model-knn, message=FALSE, warning=FALSE}
set.seed(42)
fit_knn <- train(
  x = X_train, y = y_train,
  method    = "knn",
  metric    = "ROC",
  trControl = ctrl,
  tuneGrid  = expand.grid(k = c(3, 5, 7, 11, 15))
)
cat("KNN best k:", fit_knn$bestTune$k, "\n")
cat("KNN CV AUC:", round(max(fit_knn$results$ROC), 4), "\n")
```

### Support Vector Machine (SVM with RBF kernel)

SVM finds the maximum-margin hyperplane separating classes. With the radial basis
function (RBF) kernel, it can capture non-linear boundaries. The key tuning parameters
are `C` (regularisation strength) and `sigma` (kernel width). SVM is effective in
high-dimensional spaces but less interpretable than linear models.

```{r model-svm, message=FALSE, warning=FALSE}
set.seed(42)
fit_svm <- train(
  x = X_train, y = y_train,
  method    = "svmRadial",
  metric    = "ROC",
  trControl = ctrl,
  tuneLength = 5
)
cat("SVM best C:    ", fit_svm$bestTune$C, "\n")
cat("SVM best sigma:", round(fit_svm$bestTune$sigma, 5), "\n")
cat("SVM CV AUC:    ", round(max(fit_svm$results$ROC), 4), "\n")
```

---

# 3. Cross-Validation Results

```{r cv-comparison, fig.height=5, fig.width=8}
cv_results <- resamples(list(
  LR  = fit_lr,
  LDA = fit_lda,
  QDA = fit_qda,
  RF  = fit_rf,
  KNN = fit_knn,
  SVM = fit_svm
))

summary(cv_results)
```

```{r cv-boxplot, fig.height=5, fig.width=8}
bwplot(cv_results, metric = "ROC",
       main = "10-fold CV AUC comparison — Parametric vs. Non-parametric")
```

```{r cv-dotplot, fig.height=5, fig.width=8}
dotplot(cv_results, metric = "ROC",
        main = "Mean CV AUC with 95% CI")
```

**Interpretation:** Models with higher median and tighter spread are preferred — they
are both more accurate and more stable. The parametric vs. non-parametric grouping
reveals whether the flexibility of non-parametric methods pays off on this dataset.

---

# 4. Test Set Evaluation

We evaluate all models on the held-out test set to check for overfitting and get
unbiased performance estimates.

```{r test-evaluation}
models <- list(
  LR = fit_lr, LDA = fit_lda, QDA = fit_qda,
  RF = fit_rf, KNN = fit_knn, SVM = fit_svm
)

test_results <- map_dfr(names(models), function(name) {
  mod  <- models[[name]]
  x_te <- if (name == "QDA") X_test_qda else X_test
  pred <- predict(mod, x_te)
  prob <- predict(mod, x_te, type = "prob")

  cm <- confusionMatrix(pred, y_test, positive = "Phishing")
  roc_obj <- roc(y_test, prob$Phishing, levels = c("Legitimate", "Phishing"),
                 quiet = TRUE)

  tibble(
    Model       = name,
    Accuracy    = round(cm$overall["Accuracy"], 4),
    Sensitivity = round(cm$byClass["Sensitivity"], 4),
    Specificity = round(cm$byClass["Specificity"], 4),
    F1          = round(cm$byClass["F1"], 4),
    AUC         = round(auc(roc_obj), 4),
    Type        = ifelse(name %in% c("LR", "LDA", "QDA"),
                         "Parametric", "Non-parametric")
  )
})

test_results %>%
  arrange(desc(AUC)) %>%
  print()
```

```{r test-roc, fig.height=6, fig.width=7}
roc_list <- map(names(models), function(name) {
  x_te <- if (name == "QDA") X_test_qda else X_test
  prob <- predict(models[[name]], x_te, type = "prob")
  roc(y_test, prob$Phishing, levels = c("Legitimate", "Phishing"), quiet = TRUE)
})
names(roc_list) <- names(models)

model_cols <- c(
  LR = "#1f77b4", LDA = "#4a90d9", QDA = "#7ec8e3",
  RF = "#d62728", KNN = "#e57373", SVM = "#ff9896"
)

plot(roc_list[[1]], col = model_cols[1], lwd = 2,
     main = "ROC Curves — Test Set")
for (i in 2:length(roc_list)) {
  lines(roc_list[[i]], col = model_cols[i], lwd = 2)
}
legend("bottomright",
       legend = paste0(names(roc_list), " (AUC=",
                       sapply(roc_list, function(r) round(auc(r), 3)), ")"),
       col = model_cols, lwd = 2, cex = 0.8)
```

---

# 5. Hypothesis Verification Through Models

EDA formulated three hypotheses about which feature families drive phishing detection.
Here we test whether these findings hold *inside the models* — not just in raw
descriptive statistics — by extracting feature importance from LR (parametric)
and RF (non-parametric) and comparing them by family.

## 5.1 Feature Importance: LR Coefficients vs. RF Importance

```{r feature-importance, fig.height=8, fig.width=10}
# LR: absolute standardised coefficients
lr_coefs <- coef(fit_lr$finalModel)[-1]  # drop intercept
lr_imp <- tibble(
  feature    = names(lr_coefs),
  importance = abs(as.numeric(lr_coefs))
) %>%
  mutate(importance = importance / max(importance)) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  mutate(model = "LR (|coef|, normalised)")

# RF: variable importance (MeanDecreaseGini)
rf_imp_raw <- varImp(fit_rf)$importance
rf_imp <- tibble(
  feature    = rownames(rf_imp_raw),
  importance = rf_imp_raw$Overall / max(rf_imp_raw$Overall)
) %>%
  left_join(family_map, by = "feature") %>%
  replace_na(list(family = "Other")) %>%
  mutate(model = "RF (Gini, normalised)")

# Combined plot
bind_rows(lr_imp, rf_imp) %>%
  ggplot(aes(x = reorder(feature, importance), y = importance, fill = family)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ model, scales = "free_x") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Feature importance — parametric (LR) vs. non-parametric (RF)",
       subtitle = "Coloured by family to test H1 (Trust dominates) and H3 (Lexical is weakest)",
       x = NULL, y = "Normalised importance", fill = "Family") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")
```

## 5.2 Family-Level Importance Comparison

```{r family-importance, fig.height=4, fig.width=8}
family_imp <- bind_rows(lr_imp, rf_imp) %>%
  group_by(model, family) %>%
  summarise(
    median_imp = median(importance),
    mean_imp   = mean(importance),
    max_imp    = max(importance),
    n          = n(),
    .groups    = "drop"
  )

ggplot(family_imp, aes(family, median_imp, fill = family)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ model) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Median feature importance by family",
       subtitle = "H1: Trust > others | H3: Lexical is weakest",
       x = NULL, y = "Median normalised importance") +
  theme_minimal(base_size = 12)

family_imp %>%
  select(model, family, n, median_imp, mean_imp, max_imp) %>%
  mutate(across(where(is.numeric) & !n, ~ round(.x, 3))) %>%
  arrange(model, desc(median_imp)) %>%
  print()
```

## 5.3 Hypothesis Conclusions from Models

```{r hypothesis-conclusions}
# Top 10 features per model
cat("=== Top 10 features — LR ===\n")
lr_imp %>% arrange(desc(importance)) %>% slice_head(n = 10) %>%
  select(feature, family, importance) %>%
  mutate(importance = round(importance, 3)) %>% print()

cat("\n=== Top 10 features — RF ===\n")
rf_imp %>% arrange(desc(importance)) %>% slice_head(n = 10) %>%
  select(feature, family, importance) %>%
  mutate(importance = round(importance, 3)) %>% print()
```

**H1 — Trust signals are primary discriminators:**
We verify whether Trust features consistently rank highest in importance across both
the parametric (LR) and non-parametric (RF) model. If `URLSimilarityIndex` and
`DomainTitleMatchScore` appear in the top features of both models, H1 is confirmed
not just descriptively (EDA) but also *inside the fitted models*.

**H2 — Behavioral flag profiles:**
We check whether legitimacy-mimicry flags (`HasSocialNet`, `HasCopyrightInfo`,
`HasDescription`) rank higher than deception flags (`HasExternalFormSubmit`,
`HasHiddenFields`) in both LR and RF — confirming that the two-mechanism structure
identified in EDA translates into actual model signal.

**H3 — Lexical features are secondary:**
If the Lexical family has the lowest median importance in both models, H3 is confirmed:
lexical features contribute but are not the primary drivers once Trust and Behavioral
features are present.

---

# 6. Explainability & Reproducibility (Scenario 2 requirement)

```{r explainability-table}
explain_df <- tribble(
  ~Model, ~Type, ~Explainability, ~Reproducibility,
  "LR",  "Parametric",
    "High — coefficients directly quantify feature effects on log-odds; sign and magnitude are interpretable",
    "Perfect — deterministic optimisation, same result every run",
  "LDA", "Parametric",
    "High — discriminant coefficients show which features drive class separation; linear boundary is easy to visualise",
    "Perfect — closed-form solution, fully deterministic",
  "QDA", "Parametric",
    "Medium — class-specific covariances make interpretation harder than LDA; quadratic boundary is less intuitive",
    "Perfect — closed-form solution, fully deterministic",
  "RF",  "Non-parametric",
    "Medium — variable importance ranks features but individual predictions require methods like SHAP; ensemble of 500 trees is not directly readable",
    "Seed-dependent — bootstrap sampling and random feature selection introduce stochasticity; fixed seed ensures reproducibility",
  "KNN", "Non-parametric",
    "Low — predictions depend on local neighbourhood; no global model or feature ranking; hard to explain why a specific URL was flagged",
    "Perfect — deterministic given the same data and k; no randomness in the algorithm",
  "SVM", "Non-parametric",
    "Low — RBF kernel maps features to high-dimensional space; support vectors identify the boundary but individual feature contributions are opaque",
    "Seed-dependent — caret's tuning uses resampling; the SVM optimisation itself is deterministic for fixed hyperparameters"
)

explain_df %>%
  select(Model, Type, Explainability, Reproducibility) %>%
  knitr::kable(format = "pipe",
               caption = "Explainability and reproducibility comparison")
```

**Discussion:**

Parametric models offer a clear advantage in **explainability**: LR coefficients and
LDA discriminant loadings directly map features to predictions, which matters for
phishing detection where domain analysts need to understand *why* a URL was flagged.
All parametric models are also **perfectly reproducible** — they use deterministic,
closed-form or convex optimisation.

Non-parametric models trade interpretability for flexibility. RF captures non-linear
effects and interactions, and its variable importance provides some global insight,
but individual predictions are harder to explain. KNN offers no global model at all.
SVM with RBF kernel is similarly opaque. RF's reproducibility depends on the random
seed; KNN and SVM (with fixed hyperparameters) are deterministic.

The choice depends on the use case: if the system must justify its decisions
(e.g. to a security team), parametric models are preferable. If raw predictive
accuracy is the priority, non-parametric models may be worth the interpretability cost.

---

# 7. Summary

```{r scenario2-summary}
best_param <- test_results %>%
  filter(Type == "Parametric") %>%
  slice_max(AUC, n = 1)

best_nonparam <- test_results %>%
  filter(Type == "Non-parametric") %>%
  slice_max(AUC, n = 1)

cat("Best parametric:     ", best_param$Model,
    "(AUC =", best_param$AUC, ")\n")
cat("Best non-parametric: ", best_nonparam$Model,
    "(AUC =", best_nonparam$AUC, ")\n")
cat("AUC difference:      ",
    round(best_nonparam$AUC - best_param$AUC, 4), "\n")
```
